### 2023-03-21 manuscript/appendix/appendixB.Rmd
add appendix files

---
output: 
  stevetemplates::article:
    fig_caption: true
#bibliography: master.bib
title: "Estimating process-based model parameters from species distribution data"
subtitle: "Supplementary Appendix B"
author:
- name: Victor Van der Meersch
  affiliation: CEFE, CNRS
- name: Isabelle Chuine
  affiliation: CEFE, CNRS
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontfamily: libertine
fontsize: 11pt
# spacing: double
appendix: yes
appendixletter: B
endnote: no
header-includes:
  - \usepackage{pdflscape}
  - \usepackage{xcolor}
  - \usepackage{graphicx}
  - \usepackage{float}
  - \floatplacement{figure}{H}
  - \usepackage[width=.9\textwidth, textfont=it, font=small]{caption}
  - \usepackage{libertine}

---

```{r setup_appendixB, include=FALSE}

knitr::opts_chunk$set(cache=TRUE,
                      message=FALSE, warning=FALSE,
                      echo = FALSE,
                      fig.path='figs/',
                      cache.path = '_cache/',
                      fig.process = function(x) {
                      x2 = sub('-\\d+([.][a-z]+)$', '\\1', x)
                      if (file.rename(x, x2)) x2 else x
                      })

library(dplyr)
library(ggplot2)
library(ggthemes)
library(ggbeeswarm)
library(showtext)
library(kableExtra)
options(kableExtra.latex.load_packages = FALSE)
font_add("Linux Libertine G", regular = "LinLibertine_R_G.ttf")

```

\hfill \break

```{r, results='asis'}
table <- data.frame(
  species = c('Fagus sylvatica', 'Quercus ilex', 'Abies alba'),
  number = c("718.898", "78.979", "119.891"),
  link = c('https://doi.org/10.15468/dl.e9wasa', 'https://doi.org/10.15468/dl.2a4haw', ' https://doi.org/10.15468/dl.my6c9t ')
)
kable(table, booktabs = TRUE, escape = FALSE, align="ccc",
      col.names = c("Species", "Number of occurrences", "Download link"),
      caption = "GBIF download links") %>%
  kable_classic(full_width = F, html_font = "Linux Libertine G") %>%
  kable_styling(position = "center", latex_options = "hold_position")

```

\hfill \break

```{=latex}
\begin{figure}[htbp]

{\centering \includegraphics{figs/files/presence_clustering}

}

\caption{Stratified random sampling of beech presence records based on climate clusters.}\label{fig:pres_clustering}
\end{figure}

```

```{=latex}
\begin{landscape}
\begin{figure}[htbp]

{\centering \includegraphics{figs/files/occurrence_processing}

}

\caption{Processing of holm oak occurrence records. GBIF: Global Biodiversity Information Facility, AFE: Atlas Flora Europeae, EVM: EuroVegMap.}\label{fig:occ_processing}
\end{figure}
\end{landscape}
```

\newpage

### 2023-03-21 manuscript/appendix/appendixC.Rmd
add appendix files

---
output: 
  stevetemplates::article:
    fig_caption: true
#bibliography: master.bib
title: "Estimating process-based model parameters from species distribution data"
subtitle: "Supplementary Appendix C"
author:
- name: Victor Van der Meersch
  affiliation: CEFE, CNRS
- name: Isabelle Chuine
  affiliation: CEFE, CNRS
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontfamily: libertine
fontsize: 11pt
# spacing: double
appendix: yes
appendixletter: C
endnote: no
header-includes:
  - \usepackage{pdflscape}
  - \usepackage{xcolor}
  - \usepackage{graphicx}
  - \usepackage{float}
  - \floatplacement{figure}{H}
  - \usepackage[width=.9\textwidth, textfont=it, font=small]{caption}
  - \usepackage{libertine}

---

```{r setup_appendixC, include=FALSE}

knitr::opts_chunk$set(cache=TRUE,
                      message=FALSE, warning=FALSE,
                      echo = FALSE,
                      fig.path='figs/',
                      cache.path = '_cache/',
                      fig.process = function(x) {
                      x2 = sub('-\\d+([.][a-z]+)$', '\\1', x)
                      if (file.rename(x, x2)) x2 else x
                      })

library(dplyr)
library(ggplot2)
library(ggthemes)
library(ggbeeswarm)
library(showtext)
library(kableExtra)
options(kableExtra.latex.load_packages = FALSE)
font_add("Linux Libertine G", regular = "LinLibertine_R_G.ttf")

```

```{r distributionmaps, fig.cap= "Species distributions of \\textbf{(A)} beech, \\textbf{(B)} holm oak and \\textbf{(C)} silver fir. Green cells are $0.1^\\circ$ cells where species is present, orange cells where species is supposed to be absent.", fig.height = 7.8, fig.width = 4, fig.pos="H", fig.showtext=TRUE, eval = T}

fagus_map <- ggplot() + 
  geom_raster(data = fsyl_presabs[fsyl_presabs$pres==1,], aes(x = lon, y = lat), fill = "#0C8346") +
  geom_raster(data = fsyl_presabs[fsyl_presabs$pres==0,], aes(x = lon, y = lat), fill = "#D57A66") +
  geom_sf(data = eu_map_cropped, fill = 'grey', alpha=0.1, color='grey', size = 0.05) +
  theme_void() +
  ylab("") +
  xlab("") +
  theme(legend.title.align = 0.5, legend.position = "bottom", 
        legend.key.height = unit(0.2, 'cm'), legend.key.width = unit(0.5, 'cm'),
        legend.title = element_text(family = "Linux Libertine G", size = 9),
        legend.text = element_text(family = "Linux Libertine G"))

quercus_map <- ggplot() + 
  geom_raster(data = qile_presabs[qile_presabs$pres==1,], aes(x = lon, y = lat), fill = "#0C8346") +
  geom_raster(data = qile_presabs[qile_presabs$pres==0,], aes(x = lon, y = lat), fill = "#D57A66") +
  geom_sf(data = eu_map_cropped, fill = 'grey', alpha=0.1, color='grey', size = 0.05) +
  theme_void() +
  ylab("") +
  xlab("") +
  theme(legend.title.align = 0.5, legend.position = "bottom", 
        legend.key.height = unit(0.2, 'cm'), legend.key.width = unit(0.5, 'cm'),
        legend.title = element_text(family = "Linux Libertine G", size = 9),
        legend.text = element_text(family = "Linux Libertine G"))

abies_map <- ggplot() + 
  geom_raster(data = aalb_presabs[aalb_presabs$pres==1,], aes(x = lon, y = lat), fill = "#0C8346") +
  geom_raster(data = aalb_presabs[aalb_presabs$pres==0,], aes(x = lon, y = lat), fill = "#D57A66") +
  geom_sf(data = eu_map_cropped, fill = 'grey', alpha=0.1, color='grey', size = 0.05) +
  theme_void() +
  ylab("") +
  xlab("") +
  theme(legend.title.align = 0.5, legend.position = "bottom", 
        legend.key.height = unit(0.2, 'cm'), legend.key.width = unit(0.5, 'cm'),
        legend.title = element_text(family = "Linux Libertine G", size = 9),
        legend.text = element_text(family = "Linux Libertine G"))

p <- plot_grid(fagus_map, quercus_map, abies_map,
               labels=c("(A)", "(B)", "(C)"), ncol = 1,
               align = "h", label_size = 12, label_colour = "#495057",
               hjust = 0)

p

```

### 2023-03-21 manuscript/appendix/appendixD.Rmd
add appendix files

---
output: 
  stevetemplates::article:
    fig_caption: true
#bibliography: master.bib
title: "Estimating process-based model parameters from species distribution data"
subtitle: "Supplementary Appendix D"
author:
- name: Victor Van der Meersch
  affiliation: CEFE, CNRS
- name: Isabelle Chuine
  affiliation: CEFE, CNRS
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontfamily: libertine
fontsize: 11pt
# spacing: double
appendix: yes
appendixletter: D
endnote: no
include-after: true
header-includes:
  - \usepackage{pdflscape}
  - \usepackage{xcolor}
  - \usepackage{graphicx}
  - \usepackage{float}
  - \floatplacement{figure}{H}
  - \usepackage[width=.9\textwidth, textfont=it, font=small]{caption}
  - \usepackage{libertine}
  - \definecolor{darkgreen}{RGB}{112,173,71}

---

```{r setup_appendixD, include=FALSE}

knitr::opts_chunk$set(cache=TRUE,
                      message=FALSE, warning=FALSE,
                      echo = FALSE,
                      fig.path='figs/',
                      cache.path = '_cache/',
                      fig.process = function(x) {
                      x2 = sub('-\\d+([.][a-z]+)$', '\\1', x)
                      if (file.rename(x, x2)) x2 else x
                      })

library(dplyr)
library(ggplot2)
library(ggthemes)
library(ggbeeswarm)
library(showtext)
library(kableExtra)
options(kableExtra.latex.load_packages = FALSE)
font_add("Linux Libertine G", regular = "LinLibertine_R_G.ttf")

```

\hfill \break

## The AUC as an objective function

The AUC is the area under the receiver operating characteristic curve, which plots sensitivity (correctly predicted positive fraction) as a function of commission error (falsely predicted positive fraction), as the probability threshold discriminating presence/absence varies. It is a discrimination metric, which has been widely used in the species distribution modelling literature.

## Box constraint handling

With this constraint handling - implemented by default in the R package *cmaes* [@Trautmann2011] - each evaluated solution is guaranteed to lie within the feasible space.
Let's say we have a parameter vector $x$.
For each parameter $x_i$, we have a lower bound $lb_i$ and an upper bound $ub_i$.
If a parameter $x_i$ violates one of this bound, we set $x_i$ to a new value $x_i^{repaired}$ equal to the closest boundary value ($lb_i$ or $ub_i$). We thus obtained a new parameter set $x^{repaired}$, with a minimal $\|x-x^{repaired}\|$ value. This new feasible solution $x^{repaired}$ is used for the evaluation of the objective function $AUC_{model}(x^{repaired})$, and to compute a penalty term $pen=\sum\limits_{i}(x_i-x_i^{repaired})^2=\|x^{repaired}-x\|^2$. Then $x^{repaired}$ is discarded, and the algorithm computes the penalized objective function of $x^{repaired}$ as follows: $AUC_{model}(x)=AUC_{model}(x^{repaired})+pen$. This boundary handling could be improved with adaptive weights [see @Hansen2009].

## Ecological infeasibility constraint

We added a simple way to handle ecological constraint (e.g. unfolding before flowering in beech mixed bud) with a death penalty. When a parameter vector $x$ violates a constraint, it is rejected and generated again. The main drawback of this approach is that CMA-ES does not use information from unfeasible points. An other approach could be to set $AUC_{model}(x)=0$. However, as our feasible space was large, the death penalty constraint worked well in our case.

We applied an inequality constraint on both \textit{Fagus sylvatica} and \textit{Quercus ilex}, which have mixed buds (leaves and flowers within the same bud): unfolding must happen before flowering. On the contrary, we did not apply any inequality constraint on \textit{Abies alba} simple bud phenology parameters.

### 2023-03-21 manuscript/appendix/appendixE.Rmd
add appendix files

---
output: 
  stevetemplates::article:
    fig_caption: true
#bibliography: master.bib
title: "Estimating process-based model parameters from species distribution data"
subtitle: "Supplementary Appendix E"
author:
- name: Victor Van der Meersch
  affiliation: CEFE, CNRS
- name: Isabelle Chuine
  affiliation: CEFE, CNRS
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontfamily: libertine
fontsize: 11pt
# spacing: double
appendix: yes
appendixletter: E
endnote: no
header-includes:
  - \usepackage{pdflscape}
  - \usepackage{xcolor}
  - \usepackage{graphicx}
  - \usepackage{float}
  - \floatplacement{figure}{H}
  - \usepackage[width=.9\textwidth, textfont=it, font=small]{caption}
  - \usepackage{libertine}

---

```{r setup_appendixE, include=FALSE}

knitr::opts_chunk$set(cache=TRUE,
                      message=FALSE, warning=FALSE,
                      echo = FALSE,
                      fig.path='figs/',
                      cache.path = '_cache/',
                      fig.process = function(x) {
                      x2 = sub('-\\d+([.][a-z]+)$', '\\1', x)
                      if (file.rename(x, x2)) x2 else x
                      })

library(dplyr)
library(ggplot2)
library(ggthemes)
library(ggbeeswarm)
library(showtext)
library(kableExtra)
options(kableExtra.latex.load_packages = FALSE)
font_add("Linux Libertine G", regular = "LinLibertine_R_G.ttf", bold = "LinLibertine_RB_G.ttf", italic = "LinLibertine_RI_G.ttf")

```

\hfill \break

```{r ABC_quercus_data}

# Calibrations
cal_folder <- 'D:/calibrations/phenofit/quercus_ilex/1000pres_1000abs/paper_data'
sim_folder <- 'D:/simulations/phenofit/backward/quercus_ilex/paper_data'
abc_files <- mixedsort(list.files(path = file.path(cal_folder, "ABC"), pattern = "\\.rds$", full.names = T, recursive = T))

auc_data_ABC_cal <- data.frame()
for(i in 1:length(abc_files)){
  auc_data_ABC_cal[i, 'subset'] <- paste0(strsplit(abc_files[i], "_")[[1]][7])

  # Load AUC on calibration_points
  abc_fit <- readRDS(abc_files[i])
  auc_cal <- 1-round(min(abc_fit$stats), 6)

  auc_data_ABC_cal[i, 'cal'] <- auc_cal

}

auc_data_ABC_cal$method <- "ABC"
quercus_auc_data_cal$method <- "CMAES"

quercus_auc_data_cal_comparison <- rbind(quercus_auc_data_cal[quercus_auc_data_cal$subset %in% c("subset1"), c("subset", "cal", "method")], 
                                 auc_data_ABC_cal[auc_data_ABC_cal$subset %in% c("subset1"),],
                                 quercus_auc_data_cal[quercus_auc_data_cal$subset %in% c("subset2"), c("subset", "cal", "method")], 
                                 auc_data_ABC_cal[auc_data_ABC_cal$subset %in% c("subset2"),])

if(recompute_quercus_AUC_ABC){
  simulation_files <- mixedsort(list.files(path = file.path(sim_folder, "ABC"), pattern = "Fitness", full.names = T, recursive = T))

  auc_data_ABC_tot <- data.frame()
  fitness_ABC_data <- as.data.frame(cbind(alt$lat, alt$lon))
  for(i in 1:(length(simulation_files))){
    auc_data_ABC_tot[i, 'subset'] <- strsplit(strsplit(simulation_files[i], "/")[[1]][8], "_")[[1]][1]

    # Compute AUC on every pres/abs points
    fitness_presabs <- read_mean_outputvalue_phenofit(simulation_files[i], points = qile_presabs)
    auc_tot <- round(auc(roc(fitness_presabs, as.factor(qile_presabs$pres))),6)
    auc_data_ABC_tot[i, 'tot'] <- auc_tot

    # Load fitness
    # fitness <- read_mean_outputvalue_phenofit(simulation_files[i])
    # fitness_ABC_data[,2+i] <- t(fitness)

  }

  # Consensus 
  # names(fitness_ABC_data)[1:2] <- c("lat", "lon")
  # fitness_ABC_data$mean <- rowMeans(fitness_ABC_data[,3:ncol(fitness_ABC_data)])
  # fitness_ABC_data$mad <- rowMads(as.matrix(fitness_ABC_data[,3:ncol(fitness_ABC_data)]))
  # fitness_ABC_data$sd <- rowSds(as.matrix(fitness_ABC_data[,3:ncol(fitness_ABC_data)]))

  saveRDS(auc_data_ABC_tot, file.path(wd, "cache", "quercus_auc_data_ABC_tot.rds"))
  # saveRDS(fitness_ABC_data, file.path(wd, "cache", "quercus_fitness_ABC_data.rds"))
}else{
  auc_data_ABC_tot <- readRDS(file.path(wd, "cache", "quercus_auc_data_ABC_tot.rds"))
  # fitness_ABC_data <- readRDS(file.path(wd, "cache", "quercus_fitness_ABC_data.rds"))
}

auc_data_ABC_tot$method <- "ABC"
quercus_auc_data_tot$method <- "CMAES"

quercus_auc_data_tot_comparison <- rbind(quercus_auc_data_tot[quercus_auc_data_tot$subset %in% c("subset1"), c("subset", "tot", "method")], 
                                 auc_data_ABC_tot[auc_data_ABC_tot$subset %in% c("subset1"),],
                                 quercus_auc_data_tot[quercus_auc_data_tot$subset %in% c("subset2"), c("subset", "tot", "method")], 
                                 auc_data_ABC_tot[auc_data_ABC_tot$subset %in% c("subset2"),])

```

```{r comparisonquercusABCCMAES, include=TRUE, fig.height = 2.7, fig.width = 4, fig.align = "center", fig.cap = "Comparison of CMA-ES and ABC-rejection methods, with two holm oak occurrence subsets, on \\textbf{(A)} calibration AUC (only calibration points) and \\textbf{(B)} total AUC (every presence/absence points). Each point is a calibration run. The black horizontal bars represent the pairwise Mann–Whitney tests between the two methods on the same subset. CMA-ES is red and ABC green.", fig.pos="H", fig.showtext=TRUE, eval = F}

quercus_auc_data_cal_comparison$sub_met <- paste0(quercus_auc_data_cal_comparison$subset, quercus_auc_data_cal_comparison$method)

positions_cal <- c(rep(1,5), rep(1.05,5), rep(1.15,5), rep(1.20,5))
quercus_auc_data_cal_comparison$positions_cal <- positions_cal

wilcoxon_test_cal <- compare_means(cal ~ positions_cal,  data = quercus_auc_data_cal_comparison)

p_auc_cal <- quercus_auc_data_cal_comparison %>% ggplot(aes(y = cal, x = factor(method, levels = c("CMAES", "ABC")), 
                                                      color = factor(method, levels = c("CMAES", "ABC")), 
                                                      fill = factor(method, levels = c("CMAES", "ABC")))) +
  facet_wrap(~subset) +
  geom_boxplot(width = 0.7, alpha = 0.2, outlier.shape = NA, fatten = 1, lwd=0.3) +
  geom_beeswarm(alpha = 0.8, cex = 2.5, size = 0.5, dodge.width = 0.5) + 
  scale_fill_manual(values = c("#DB9D85", "#6DBC86")) +
  scale_color_manual(values = c("#DB9D85", "#6DBC86")) +
  stat_compare_means(comparisons = list(c("ABC", "CMAES")), symnum.args = symnum.args,
                     tip.length = 0.005, label.y = 0.963,
                     size=3, vjust = 0.6, bracket.size = 0.2) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        axis.text.y = element_text(family = "Linux Libertine G"),
        axis.title.y = element_text(family = "Linux Libertine G"),
        axis.ticks.x=element_blank(),
        panel.grid.major = element_line(size = 0.35),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        strip.text.x = element_blank(),
        panel.spacing.x = unit(1.3, "lines"),
        legend.position="none") +
  coord_cartesian(ylim = c(0.915, 0.965)) +
  labs(y = "")

quercus_auc_data_tot_comparison$sub_met <- paste0(quercus_auc_data_tot_comparison$subset, quercus_auc_data_tot_comparison$method)

positions_tot <- c(rep(1,5), rep(1.05,5), rep(1.15,5), rep(1.20,5))
quercus_auc_data_tot_comparison$positions_tot <- positions_tot

p_auc_tot <- na.omit(quercus_auc_data_tot_comparison) %>% ggplot(aes(y = tot, x = factor(method, levels = c("CMAES", "ABC")), 
                                                      color = factor(method, levels = c("CMAES", "ABC")), 
                                                      fill = factor(method, levels = c("CMAES", "ABC")))) +
  facet_wrap(~subset) +
  geom_boxplot(width = 0.7, alpha = 0.2, outlier.shape = NA, fatten = 1, lwd=0.3) +
  geom_beeswarm(alpha = 0.8, cex = 2.5, size = 0.5, dodge.width = 0.5) + 
  scale_fill_manual(values = c("#DB9D85", "#6DBC86")) +
  scale_color_manual(values = c("#DB9D85", "#6DBC86")) +
  stat_compare_means(comparisons = list(c("ABC", "CMAES")), symnum.args = symnum.args,
                     tip.length = 0.005, label.y = 0.963,
                     size=3, vjust = 0.6, bracket.size = 0.2) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        axis.text.y = element_text(family = "Linux Libertine G"),
        axis.title.y = element_text(family = "Linux Libertine G"),
        axis.ticks.x=element_blank(),
        panel.grid.major = element_line(size = 0.35),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        strip.text.x = element_blank(),
        panel.spacing.x = unit(1.3, "lines"),
        legend.position="none") +
  coord_cartesian(ylim = c(0.915, 0.965)) +
  labs(y = "")

plot_grid(p_auc_cal, p_auc_tot, labels=c("(A)", "(B)"), ncol = 2, nrow = 1,
          hjust = 0, vjust = 1, align = "h", label_size = 12, label_colour = "#495057")

```

```{r calibCMAESquercus, include=TRUE, fig.height = 2.7, fig.width = 3, fig.align = "center", fig.cap = "CMA-ES calibration using the PHENOFIT model and holm oak: \\textbf{(A)} calibration AUC (only calibration cells) and \\textbf{(B)} total AUC (every presence/absence cells). Each color is a different sub-sampling of occurrence data, each point is a calibration run. The black horizontal bars represent the pairwise Mann–Whitney tests between the two subsets.", fig.pos="H", fig.showtext=TRUE, results='hide'}

positions <- c(rep(1,5), rep(1.1,5))
quercus_auc_data_cal$positions <- positions

# wilcoxon_test_cal <- compare_means(cal ~ positions,  data = auc_data_cal) 
# wilcoxon_test_s7 <- wilcoxon_test_cal %>% filter(p.signif != "ns") %>% filter(group1 == "1.7" | group2 == "1.7")
# list_comparisons <- lapply(1:nrow(wilcoxon_test_s7), function(i){
#   return(c(wilcoxon_test_s7$group1[i], wilcoxon_test_s7$group2[i]))})
# list_positions <- c(0.875, 0.9225, 0.9245, 0.917, 0.9145, 0.8785, 0.882, 0.894, 0.888)

symnum.args <- list(cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 1), symbols = c("****", "***", "**", "*", "NS."))

p_auc_cal <- quercus_auc_data_cal %>% ggplot(aes(y = cal, x = as.character(positions), group = subset, 
                    color = factor(subset, levels = paste0("subset", 1:10)), 
                    fill = factor(subset, levels = paste0("subset", 1:10)))) +
  geom_boxplot(width = 0.5, alpha = 0.3, outlier.shape = NA, fatten = 1, lwd=0.3) +
  geom_beeswarm(alpha = 0.8, cex = 2.5, size = 0.5) +
  # stat_compare_means(comparisons = list_comparisons, symnum.args = symnum.args, 
  #                    tip.length = c(-0.005, rep(0.005,4), rep(-0.005,4)),
  #                    label.y = list_positions, size=2, vjust = 0.6, bracket.size = 0.2) + 
  stat_compare_means(comparisons = list(c("1", "1.1")), symnum.args = symnum.args,
                     tip.length = 0.005, label.y = 0.965,
                     size=2, vjust = 0.1, bracket.size = 0.2) +
  scale_fill_manual(values = hcl.colors(10, palette = "Dynamic")) +
  scale_color_manual(values = hcl.colors(10, palette = "Dynamic")) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        axis.text.y = element_text(family = "Linux Libertine G"),
        axis.title.y = element_text(family = "Linux Libertine G", margin = margin(t = 0, r = 10, b = 0, l = 0)),
        axis.ticks.x=element_blank(),
        panel.grid.major = element_line(size = 0.2),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position="none") +
  coord_cartesian(ylim = c(0.94, 0.967)) +
  labs(y = "AUC")

quercus_auc_data_tot$positions <- positions

# wilcoxon_test_tot <- compare_means(tot ~ positions,  data = auc_data_tot) 
# wilcoxon_test_s7 <- wilcoxon_test_tot %>% filter(p.signif != "ns") %>% filter(group1 == "1.7" | group2 == "1.7")
# list_comparisons <- lapply(1:nrow(wilcoxon_test_s7), function(i){
#   return(c(wilcoxon_test_s7$group1[i], wilcoxon_test_s7$group2[i]))})

p_auc_tot <- na.omit(quercus_auc_data_tot) %>% ggplot(aes(y = tot, as.character(positions), group = subset, 
                    color = factor(subset, levels = paste0("subset", 1:10)), fill = factor(subset, levels = paste0("subset", 1:10)))) +
  geom_boxplot(width = 0.5, alpha = 0.2, outlier.shape = NA, fatten = 1, lwd=0.3) +
  geom_beeswarm(alpha = 0.8, cex = 2.5, size = 0.5) +
  stat_compare_means(comparisons = list(c("1", "1.1")), symnum.args = symnum.args,
                     tip.length = 0.005, label.y = 0.965,
                     size=2, vjust = 0.1, bracket.size = 0.2) +
  scale_fill_manual(values = hcl.colors(10, palette = "Dynamic")) +
  scale_color_manual(values = hcl.colors(10, palette = "Dynamic")) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        axis.text.y = element_text(family = "Linux Libertine G"),
        axis.title.y = element_text(family = "Linux Libertine G"),
        axis.ticks.x=element_blank(),
        panel.grid.major = element_line(size = 0.2),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position="none") +
  coord_cartesian(ylim = c(0.94, 0.967)) +
  labs(y = "")

plot_grid(p_auc_cal, p_auc_tot, labels=c("(A)", "(B)"), ncol = 2, nrow = 1,
          hjust = 0, vjust = 1.2, align = "h", label_size = 12, label_colour = "#495057")

```

```{r calibCMAESabies, include=TRUE, fig.height = 2.7, fig.width = 3, fig.align = "center", fig.cap = "CMA-ES calibration using the PHENOFIT model and silver fir: \\textbf{(A)} calibration AUC (only calibration cells) and \\textbf{(B)} total AUC (every presence/absence cells). Each color is a different sub-sampling of occurrence data, each point is a calibration run. The black horizontal bars represent the pairwise Mann–Whitney tests between the two subsets.", fig.pos="H", fig.showtext=TRUE, results='hide'}

positions <- c(rep(1,5), rep(1.1,5))
abies_auc_data_cal$positions <- positions

# wilcoxon_test_cal <- compare_means(cal ~ positions,  data = auc_data_cal) 
# wilcoxon_test_s7 <- wilcoxon_test_cal %>% filter(p.signif != "ns") %>% filter(group1 == "1.7" | group2 == "1.7")
# list_comparisons <- lapply(1:nrow(wilcoxon_test_s7), function(i){
#   return(c(wilcoxon_test_s7$group1[i], wilcoxon_test_s7$group2[i]))})
# list_positions <- c(0.875, 0.9225, 0.9245, 0.917, 0.9145, 0.8785, 0.882, 0.894, 0.888)

p_auc_cal <- abies_auc_data_cal %>% ggplot(aes(y = cal, x = as.character(positions), group = subset, 
                    color = factor(subset, levels = paste0("subset", 1:10)), 
                    fill = factor(subset, levels = paste0("subset", 1:10)))) +
  geom_boxplot(width = 0.5, alpha = 0.3, outlier.shape = NA, fatten = 1, lwd=0.3) +
  geom_beeswarm(alpha = 0.8, cex = 2.5, size = 0.5) +
  # stat_compare_means(comparisons = list_comparisons, symnum.args = symnum.args, 
  #                    tip.length = c(-0.005, rep(0.005,4), rep(-0.005,4)),
  #                    label.y = list_positions, size=2, vjust = 0.6, bracket.size = 0.2) + 
  stat_compare_means(comparisons = list(c("1", "1.1")), symnum.args = symnum.args,
                     tip.length = 0.005, label.y = 0.92,
                     size=2, vjust = 0.1, bracket.size = 0.2) +
  scale_fill_manual(values = hcl.colors(10, palette = "Dynamic")) +
  scale_color_manual(values = hcl.colors(10, palette = "Dynamic")) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        axis.text.y = element_text(family = "Linux Libertine G"),
        axis.title.y = element_text(family = "Linux Libertine G", margin = margin(t = 0, r = 10, b = 0, l = 0)),
        axis.ticks.x=element_blank(),
        panel.grid.major = element_line(size = 0.2),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position="none") +
  coord_cartesian(ylim = c(0.87, 0.923)) +
  labs(y = "AUC")

abies_auc_data_tot$positions <- positions

# wilcoxon_test_tot <- compare_means(tot ~ positions,  data = auc_data_tot) 
# wilcoxon_test_s7 <- wilcoxon_test_tot %>% filter(p.signif != "ns") %>% filter(group1 == "1.7" | group2 == "1.7")
# list_comparisons <- lapply(1:nrow(wilcoxon_test_s7), function(i){
#   return(c(wilcoxon_test_s7$group1[i], wilcoxon_test_s7$group2[i]))})

p_auc_tot <- na.omit(abies_auc_data_tot) %>% ggplot(aes(y = tot, as.character(positions), group = subset, 
                    color = factor(subset, levels = paste0("subset", 1:10)), fill = factor(subset, levels = paste0("subset", 1:10)))) +
  geom_boxplot(width = 0.5, alpha = 0.2, outlier.shape = NA, fatten = 1, lwd=0.3) +
  geom_beeswarm(alpha = 0.8, cex = 2.5, size = 0.5) +
  stat_compare_means(comparisons = list(c("1", "1.1")), symnum.args = symnum.args,
                     tip.length = 0.005, label.y = 0.92,
                     size=2, vjust = 0.1, bracket.size = 0.2) +
  scale_fill_manual(values = hcl.colors(10, palette = "Dynamic")) +
  scale_color_manual(values = hcl.colors(10, palette = "Dynamic")) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        axis.text.y = element_text(family = "Linux Libertine G"),
        axis.title.y = element_text(family = "Linux Libertine G"),
        axis.ticks.x=element_blank(),
        panel.grid.major = element_line(size = 0.2),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position="none") +
  coord_cartesian(ylim = c(0.87, 0.923)) +
  labs(y = "")

plot_grid(p_auc_cal, p_auc_tot, labels=c("(A)", "(B)"), ncol = 2, nrow = 1,
          hjust = 0, vjust = 1.2, align = "h", label_size = 12, label_colour = "#495057")

```

### 2023-03-21 manuscript/appendix/appendixF.Rmd
add appendix files

---
output: 
  stevetemplates::article:
    fig_caption: true
#bibliography: master.bib
title: "Estimating process-based model parameters from species distribution data"
subtitle: "Supplementary Appendix F"
author:
- name: Victor Van der Meersch
  affiliation: CEFE, CNRS
- name: Isabelle Chuine
  affiliation: CEFE, CNRS
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: "left=0.7in,right=0.7in,top=0.3in,bottom=0.7in"
fontfamily: libertine
fontsize: 11pt
# spacing: double
appendix: yes
appendixletter: F
endnote: no
header-includes:
  - \usepackage{pdflscape}
  - \usepackage{xcolor}
  - \usepackage{graphicx}
  - \usepackage{float}
  - \floatplacement{figure}{H}
  - \usepackage[width=.9\textwidth, textfont=it, font=small]{caption}
  - \usepackage{libertine}

---

```{r setup_appendixF, include=FALSE}

knitr::opts_chunk$set(cache=TRUE,
                      message=FALSE, warning=FALSE,
                      echo = FALSE,
                      fig.path='figs/',
                      cache.path = '_cache/',
                      fig.process = function(x) {
                      x2 = sub('-\\d+([.][a-z]+)$', '\\1', x)
                      if (file.rename(x, x2)) x2 else x
                      })

library(dplyr)
library(ggplot2)
library(ggthemes)
library(ggbeeswarm)
library(showtext)
library(kableExtra)
options(kableExtra.latex.load_packages = FALSE)
font_add("Linux Libertine G", regular = "LinLibertine_R_G.ttf", bold = "LinLibertine_RB_G.ttf", italic = "LinLibertine_RI_G.ttf")

cal_folder <- 'D:/calibrations/phenofit/fagus_sylvatica/1000pres_1000abs/paper_data'

sim_folder <- 'D:/simulations/phenofit/backward/fagus_sylvatica/paper_data'

```

\hfill \break

```{r phenofit_cmaes_output_maps}

phenofit_fagus_map <- ggplot(data=phenofit_fagus_best_fit, aes(x = lon, y = lat)) + 
  geom_raster(aes(fill = fitness), color = NA) +
  theme_void() +
  labs(fill = "Fitness") +
  ylab("") +
  xlab("") +
  scale_fill_gradient2(low = "#EBEBD3", mid = "#B4E0AA", high = "#488B49", limits = c(0,1), breaks = c(0,0.5,1), midpoint = 0.5) +
  guides(fill = guide_colorbar(title.position = "top", direction = "horizontal", frame.colour = "black", frame.linewidth = 0.3,
                               ticks = FALSE)) +
  theme(legend.title.align = 0.5, legend.position = "bottom", 
        legend.key.height = unit(0.2, 'cm'), legend.key.width = unit(0.5, 'cm'),
        legend.title = element_text(family = "Linux Libertine G", size = 9),
        legend.text = element_text(family = "Linux Libertine G")) +
  annotate("text", x = -4, y = 65, label = paste("AUC =", round(max(fagus_auc_data_tot$tot), 2)), colour = 'darkgrey', size = 3.6, 
           family = "Linux Libertine G", fontface = "bold")

phenofit_quercus_map <- ggplot(data=phenofit_quercus_best_fit, aes(x = lon, y = lat)) + 
  geom_raster(aes(fill = fitness), color = NA) +
  theme_void() +
  labs(fill = "Fitness") +
  ylab("") +
  xlab("") +
  scale_fill_gradient2(low = "#EBEBD3", mid = "#B4E0AA", high = "#488B49", limits = c(0,1), breaks = c(0,0.5,1), midpoint = 0.5) +
  guides(fill = guide_colorbar(title.position = "top", direction = "horizontal", frame.colour = "black", frame.linewidth = 0.3,
                               ticks = FALSE)) +
  theme(legend.title.align = 0.5, legend.position = "bottom", 
        legend.key.height = unit(0.2, 'cm'), legend.key.width = unit(0.5, 'cm'),
        legend.title = element_text(family = "Linux Libertine G", size = 9),
        legend.text = element_text(family = "Linux Libertine G")) +
  annotate("text", x = -4, y = 65, label = paste("AUC =", round(max(quercus_auc_data_tot$tot), 2)), colour = 'darkgrey', size = 3.6, 
           family = "Linux Libertine G", fontface = "bold")

phenofit_abies_map <- ggplot(data=phenofit_abies_best_fit, aes(x = lon, y = lat)) + 
  geom_raster(aes(fill = fitness), color = NA) +
  theme_void() +
  labs(fill = "Fitness") +
  ylab("") +
  xlab("") +
  scale_fill_gradient2(low = "#EBEBD3", mid = "#B4E0AA", high = "#488B49", limits = c(0,1), breaks = c(0,0.5,1), midpoint = 0.5) +
  guides(fill = guide_colorbar(title.position = "top", direction = "horizontal", frame.colour = "black", frame.linewidth = 0.3,
                               ticks = FALSE)) +
  theme(legend.title.align = 0.5, legend.position = "bottom", 
        legend.key.height = unit(0.2, 'cm'), legend.key.width = unit(0.5, 'cm'),
        legend.title = element_text(family = "Linux Libertine G", size = 9),
        legend.text = element_text(family = "Linux Libertine G")) +
  annotate("text", x = -4, y = 65, label = paste("AUC =", round(max(abies_auc_data_tot$tot), 2)), colour = 'darkgrey', size = 3.6, 
           family = "Linux Libertine G", fontface = "bold")

```

```{r castanea_cmaes_output_maps}

castanea_fagus_map <- ggplot(data=castanea_fagus_best_fit, aes(x = lon, y = lat)) + 
  geom_raster(aes(fill = biomass_reserves), color = NA) +
  theme_void() +
  labs(fill = "Carbon reserves") +
  ylab("") +
  xlab("") +
  scale_fill_gradient2(low = "#EBEBD3", mid = "#B4E0AA", high = "#488B49", limits = c(0,22), breaks = c(0, 10, 20), midpoint = 10,
                       na.value = "#EAEAEA") +
  guides(fill = guide_colorbar(title.position = "top", direction = "horizontal", frame.colour = "black", frame.linewidth = 0.3,
                               ticks = FALSE)) +
  theme(legend.title.align = 0.5, legend.position = "bottom", 
        legend.key.height = unit(0.15, 'cm'), legend.key.width = unit(0.5, 'cm'),
        legend.title = element_text(family = "Linux Libertine G", size = 8),
        legend.text = element_text(family = "Linux Libertine G", size = 8)) +
  annotate("text", x = -4, y = 65, label = paste("AUC =", round(castanea_fagus_best_auc, 2)), colour = 'darkgrey', size = 3.6, 
           family = "Linux Libertine G", fontface = "bold")

castanea_quercus_map <- ggplot(data=castanea_quercus_best_fit, aes(x = lon, y = lat)) + 
  geom_raster(aes(fill = biomass_reserves), color = NA) +
  theme_void() +
  labs(fill = "Carbon reserves") +
  ylab("") +
  xlab("") +
  scale_fill_gradient2(low = "#EBEBD3", mid = "#B4E0AA", high = "#488B49", limits = c(0,436), breaks = c(0, 200, 400), midpoint = 200,
                       na.value = "#EAEAEA") +
  guides(fill = guide_colorbar(title.position = "top", direction = "horizontal", frame.colour = "black", frame.linewidth = 0.3,
                               ticks = FALSE)) +
  theme(legend.title.align = 0.5, legend.position = "bottom", 
        legend.key.height = unit(0.15, 'cm'), legend.key.width = unit(0.5, 'cm'),
        legend.title = element_text(family = "Linux Libertine G", size = 8),
        legend.text = element_text(family = "Linux Libertine G", size = 8)) +
  annotate("text", x = -4, y = 65, label = paste("AUC =", round(castanea_quercus_best_auc, 2)), colour = 'darkgrey', size = 3.6, 
           family = "Linux Libertine G", fontface = "bold")

castanea_abies_map <- ggplot(data=castanea_abies_best_fit, aes(x = lon, y = lat)) + 
  geom_raster(aes(fill = biomass_reserves), color = NA) +
  theme_void() +
  labs(fill = "Carbon reserves") +
  ylab("") +
  xlab("") +
  scale_fill_gradient2(low = "#EBEBD3", mid = "#B4E0AA", high = "#488B49", limits = c(0,5457), breaks = c(0, 2500, 5000), midpoint = 2500,
                       na.value = "#EAEAEA") +
  guides(fill = guide_colorbar(title.position = "top", direction = "horizontal", frame.colour = "black", frame.linewidth = 0.3,
                               ticks = FALSE)) +
  theme(legend.title.align = 0.5, legend.position = "bottom", 
        legend.key.height = unit(0.15, 'cm'), legend.key.width = unit(0.5, 'cm'),
        legend.title = element_text(family = "Linux Libertine G", size = 8),
        legend.text = element_text(family = "Linux Libertine G", size = 8)) +
  annotate("text", x = -4, y = 65, label = paste("AUC =", round(castanea_abies_best_auc, 2)), colour = 'darkgrey', size = 3.6, 
           family = "Linux Libertine G", fontface = "bold")

```

```{r phenofit_forward_output_maps}

phenofit_forward_fagus_map <- ggplot(data=phenofit_fagus_forward_fit, aes(x = lon, y = lat)) + 
  geom_raster(aes(fill = fitness), color = NA) +
  theme_void() +
  labs(fill = "Fitness") +
  ylab("") +
  xlab("") +
  scale_fill_gradient2(low = "#EBEBD3", mid = "#B4E0AA", high = "#488B49", limits = c(0,1), breaks = c(0,0.5,1), midpoint = 0.5) +
  guides(fill = guide_colorbar(title.position = "top", direction = "horizontal", frame.colour = "black", frame.linewidth = 0.3,
                               ticks = FALSE)) +
  theme(legend.title.align = 0.5, legend.position = "bottom", 
        legend.key.height = unit(0.2, 'cm'), legend.key.width = unit(0.5, 'cm'),
        legend.title = element_text(family = "Linux Libertine G", size = 9),
        legend.text = element_text(family = "Linux Libertine G")) +
  annotate("text", x = -4, y = 65, label = paste("AUC =", round(phenofit_fagus_forward_auc, 2)), colour = 'darkgrey', size = 3.6, 
           family = "Linux Libertine G", fontface = "bold")

phenofit_forward_quercus_map <- ggplot(data=phenofit_quercus_forward_fit, aes(x = lon, y = lat)) + 
  geom_raster(aes(fill = fitness), color = NA) +
  theme_void() +
  labs(fill = "Fitness") +
  ylab("") +
  xlab("") +
  scale_fill_gradient2(low = "#EBEBD3", mid = "#B4E0AA", high = "#488B49", limits = c(0,1), breaks = c(0,0.5,1), midpoint = 0.5) +
  guides(fill = guide_colorbar(title.position = "top", direction = "horizontal", frame.colour = "black", frame.linewidth = 0.3,
                               ticks = FALSE)) +
  theme(legend.title.align = 0.5, legend.position = "bottom", 
        legend.key.height = unit(0.2, 'cm'), legend.key.width = unit(0.5, 'cm'),
        legend.title = element_text(family = "Linux Libertine G", size = 9),
        legend.text = element_text(family = "Linux Libertine G")) +
  annotate("text", x = -4, y = 65, label = paste("AUC =", round(phenofit_quercus_forward_auc, 2)), colour = 'darkgrey', size = 3.6, 
           family = "Linux Libertine G", fontface = "bold")

phenofit_forward_abies_map <- ggplot(data=phenofit_abies_forward_fit, aes(x = lon, y = lat)) + 
  geom_raster(aes(fill = fitness), color = NA) +
  theme_void() +
  labs(fill = "Fitness") +
  ylab("") +
  xlab("") +
  scale_fill_gradient2(low = "#EBEBD3", mid = "#B4E0AA", high = "#488B49", limits = c(0,1), breaks = c(0,0.5,1), midpoint = 0.5) +
  guides(fill = guide_colorbar(title.position = "top", direction = "horizontal", frame.colour = "black", frame.linewidth = 0.3,
                               ticks = FALSE)) +
  theme(legend.title.align = 0.5, legend.position = "bottom", 
        legend.key.height = unit(0.15, 'cm'), legend.key.width = unit(0.5, 'cm'),
        legend.title = element_text(family = "Linux Libertine G", size = 8),
        legend.text = element_text(family = "Linux Libertine G", size = 8)) +
  annotate("text", x = -4, y = 65, label = paste("AUC =", round(phenofit_abies_forward_auc, 2)), colour = 'darkgrey', size = 3.6, 
           family = "Linux Libertine G", fontface = "bold")

```

```{r castanea_forward_output_maps}

castanea_forward_fagus_map <- ggplot(data=castanea_fagus_forward_fit, aes(x = lon, y = lat)) + 
  geom_raster(aes(fill = biomass_reserves), color = NA) +
  theme_void() +
  labs(fill = "Carbon reserves") +
  ylab("") +
  xlab("") +
  scale_fill_gradient2(low = "#EBEBD3", mid = "#B4E0AA", high = "#488B49", limits = c(0,204), breaks = c(0, 100, 200), midpoint = 100,
                       na.value = "#EAEAEA") +
  guides(fill = guide_colorbar(title.position = "top", direction = "horizontal", frame.colour = "black", frame.linewidth = 0.3,
                               ticks = FALSE)) +
  theme(legend.title.align = 0.5, legend.position = "bottom", 
        legend.key.height = unit(0.15, 'cm'), legend.key.width = unit(0.5, 'cm'),
        legend.title = element_text(family = "Linux Libertine G", size = 8),
        legend.text = element_text(family = "Linux Libertine G", size = 8)) +
  annotate("text", x = -4, y = 65, label = paste("AUC =", round(castanea_fagus_forward_auc, 2)), colour = 'darkgrey', size = 3.6, 
           family = "Linux Libertine G", fontface = "bold")

castanea_forward_quercus_map <- ggplot(data=castanea_quercus_forward_fit, aes(x = lon, y = lat)) + 
  geom_raster(aes(fill = biomass_reserves), color = NA) +
  theme_void() +
  labs(fill = "Carbon reserves") +
  ylab("") +
  xlab("") +
  scale_fill_gradient2(low = "#EBEBD3", mid = "#B4E0AA", high = "#488B49", limits = c(0,3800), breaks = c(0, 1000, 2000, 3000), midpoint = 1900,
                       na.value = "#EAEAEA") +
  guides(fill = guide_colorbar(title.position = "top", direction = "horizontal", frame.colour = "black", frame.linewidth = 0.3,
                               ticks = FALSE)) +
  theme(legend.title.align = 0.5, legend.position = "bottom", 
        legend.key.height = unit(0.15, 'cm'), legend.key.width = unit(0.5, 'cm'),
        legend.title = element_text(family = "Linux Libertine G", size = 8),
        legend.text = element_text(family = "Linux Libertine G", size = 8)) +
  annotate("text", x = -4, y = 65, label = paste("AUC =", round(castanea_quercus_forward_auc, 2)), colour = 'darkgrey', size = 3.6, 
           family = "Linux Libertine G", fontface = "bold")

castanea_forward_abies_map <- ggplot(data=castanea_abies_forward_fit, aes(x = lon, y = lat)) + 
  geom_tile(aes(fill = biomass_reserves), color = NA) +
  theme_void() +
  labs(fill = "Carbon reserves") +
  ylab("") +
  xlab("") +
  scale_fill_gradient2(low = "#EBEBD3", mid = "#B4E0AA", high = "#488B49", limits = c(0,117), breaks = c(0, 50, 100), midpoint = 55,
                       na.value = "#EAEAEA") +
  guides(fill = guide_colorbar(title.position = "top", direction = "horizontal", frame.colour = "black", frame.linewidth = 0.3,
                               ticks = FALSE)) +
  theme(legend.title.align = 0.5, legend.position = "bottom", 
        legend.key.height = unit(0.15, 'cm'), legend.key.width = unit(0.5, 'cm'),
        legend.title = element_text(family = "Linux Libertine G", size = 8),
        legend.text = element_text(family = "Linux Libertine G", size = 8)) +
  annotate("text", x = -4, y = 65, label = paste("AUC =", round(castanea_abies_forward_auc, 2)), colour = 'darkgrey', size = 3.6, 
           family = "Linux Libertine G", fontface = "bold")

```

```{r phenofit_output_maps, include=TRUE, fig.height = 7, fig.width = 4.5, fig.align = "center", fig.cap = "Fitness index predicted by PHENOFIT with the expert and the inverse calibrations.", fig.pos="H", fig.showtext=TRUE, eval = T}

p <- plot_grid(
  "",
  plot_grid(
    "",
    phenofit_forward_fagus_map + theme(legend.position = "none"), 
    phenofit_fagus_map + theme(legend.position = "none"),
    "",
    phenofit_forward_quercus_map + theme(legend.position = "none"), 
    phenofit_quercus_map + theme(legend.position = "none"),
    "",
    phenofit_forward_abies_map + theme(legend.position = "none"), 
    phenofit_abies_map + theme(legend.position = "none"),
    ncol=3, rel_widths = c(1, 5, 5),
    rel_heights = c(5, 5, 5)),
  get_legend(phenofit_fagus_map),
  ncol = 1,
  rel_heights = c(2, 20, 2))

ggdraw(p) +
  draw_label("Expert calibration", color = "#495057", size = 10, angle = 0, y = 0.92, x = 0.29,
             fontfamily = "Linux Libertine G", fontface = "italic") +
  draw_label("Inverse calibration", color = "#495057", size = 10, angle = 0, y = 0.92, x = 0.75,
             fontfamily = "Linux Libertine G", fontface = "italic") +
  draw_label("Fagus sylvatica", color = "#495057", size = 10, angle = 90, y = 0.75, x = 0.04,
             fontfamily = "Linux Libertine G", fontface = "italic") +
  draw_label("Quercus ilex", color = "#495057", size = 10, angle = 90, y = 0.46, x = 0.04,
             fontfamily = "Linux Libertine G", fontface = "italic") +
  draw_label("Abies alba", color = "#495057", size = 10, angle = 90, y = 0.18, x = 0.04,
             fontfamily = "Linux Libertine G", fontface = "italic")

```

```{r best_forward_maps, include=TRUE, fig.height = 8, fig.width = 4.5, fig.align = "center", fig.cap = "Carbon reserves predicted by CASTANEA with the expert and the inverse calibrations. Note that CASTANEA cannot be used in high-latitude regions (grey area).", fig.pos="H", fig.showtext=TRUE, eval = T}

p <- plot_grid(
  "",
  plot_grid(
    "",
    castanea_forward_fagus_map + theme(legend.position = "none"), 
    castanea_fagus_map + theme(legend.position = "none"),
    "", get_legend(castanea_forward_fagus_map), get_legend(castanea_fagus_map),
    "",
    castanea_forward_quercus_map + theme(legend.position = "none"), 
    castanea_quercus_map + theme(legend.position = "none"),
    "", get_legend(castanea_forward_quercus_map), get_legend(castanea_quercus_map),
    "",
    castanea_forward_abies_map + theme(legend.position = "none"), 
    castanea_abies_map + theme(legend.position = "none"),
    "",
    get_legend(castanea_forward_abies_map),
    get_legend(castanea_abies_map),
    ncol=3, rel_widths = c(1, 5, 5),
    rel_heights = c(5, 1, 5, 1, 5, 1)),
  ncol = 1,
  rel_heights = c(2,20))

ggdraw(p) +
  draw_label("Expert calibration", color = "#495057", size = 10, angle = 0, y = 0.92, x = 0.28,
             fontfamily = "Linux Libertine G", fontface = "italic") +
  draw_label("Inverse calibration", color = "#495057", size = 10, angle = 0, y = 0.92, x = 0.75,
             fontfamily = "Linux Libertine G", fontface = "italic") +
  draw_label("Fagus sylvatica", color = "#495057", size = 10, angle = 90, y = 0.75, x = 0.04,
             fontfamily = "Linux Libertine G", fontface = "italic") +
  draw_label("Quercus ilex", color = "#495057", size = 10, angle = 90, y = 0.46, x = 0.04,
             fontfamily = "Linux Libertine G", fontface = "italic") +
  draw_label("Abies alba", color = "#495057", size = 10, angle = 90, y = 0.18, x = 0.04,
             fontfamily = "Linux Libertine G", fontface = "italic")

```

### 2023-03-21 manuscript/appendix/appendixG.Rmd
add appendix files

---
output: 
  stevetemplates::article:
    fig_caption: true
#bibliography: master.bib
title: "Estimating process-based model parameters from species distribution data"
subtitle: "Supplementary Appendix G"
author:
- name: Victor Van der Meersch
  affiliation: CEFE, CNRS
- name: Isabelle Chuine
  affiliation: CEFE, CNRS
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontfamily: libertine
fontsize: 11pt
# spacing: double
appendix: yes
appendixletter: G
endnote: no
header-includes:
  - \usepackage{pdflscape}
  - \usepackage{xcolor}
  - \usepackage{graphicx}
  - \usepackage{float}
  - \floatplacement{figure}{H}
  - \usepackage[width=.9\textwidth, textfont=it, font=small]{caption}
  - \usepackage{libertine}

---

```{r setup_appendixG, include=FALSE}

knitr::opts_chunk$set(cache=TRUE,
                      message=FALSE, warning=FALSE,
                      echo = FALSE,
                      fig.path='figs/',
                      cache.path = '_cache/',
                      fig.process = function(x) {
                      x2 = sub('-\\d+([.][a-z]+)$', '\\1', x)
                      if (file.rename(x, x2)) x2 else x
                      })

library(dplyr)
library(ggplot2)
library(ggthemes)
library(ggbeeswarm)
library(showtext)
library(kableExtra)
options(kableExtra.latex.load_packages = FALSE)
font_add("Linux Libertine G", regular = "LinLibertine_R_G.ttf")

cal_folder <- 'D:/calibrations/phenofit/fagus_sylvatica/1000pres_1000abs/paper_data'

sim_folder <- 'D:/simulations/phenofit/backward/fagus_sylvatica/paper_data'

```

\hfill \break

## Fagus sylvatica leaf unfolding submodel

This model, called UniChill [@Chuine2000], is a sequential two-phase model (endodormancy and ecodormancy phases).  
The endodormancy phase begins at day $t_0$. 
The daily rate of chilling $R_c$ is defined as a threshold function of the daily mean temperature $T_d$:
$$ R_c(T_d) = \left\{
\begin{array}{ll}
      0 & T_d \geq T_b \\
      1 & T_d < T_b \\
\end{array} 
\right. $$
where $T_b$ is the threshold temperature below which the bud  accumulates chilling units.  
The endodormancy releases at day $t_c$ when the accumulated rate of chilling has reached the level $C_{crit}$:
$$\sum\limits_{t_0}^{t_c} R_c(T_d) \geq C_{crit} $$
Then, the ecodormancy phase begins. The daily rate of forcing $R_f$ is defined as a sigmoid function of the daily mean temperature $T_d$:
$$R_f(T_d) = \frac{1}{1 + e^{-d_T(T_d-T_{50})}} $$
where $d_T$ is the slope and $T_{50}$ the mid-response temperature.
Bud break occurs at day $t_f$ when the accumulated rate of forcing has reached the level $F_{crit}$:
$$\sum\limits_{t_c}^{t_f} R_f(T_d) \geq F_{crit} $$
Thus, the UniChill model has 6 parameters: $t_0$, $T_b$ and $C_{crit}$ for the first phase, $d_T$, $T_{50}$ and $F_{crit}$ for the second phase.

```{r leafpardensity, include=TRUE, fig.height = 4, fig.width = 4, fig.align = "center", fig.cap = "Beech leaf unfolding model parameter density. Y-axis and X-axis limits are lower and upper bounds used during calibration.", fig.pos="H", fig.showtext=TRUE, eval=T}

par_names <- unique(data_leaf_par$var) 
names(par_names) <- c("t_0", "T_b", "d_T", "T_{50}", "C_{crit}", "F_{crit}")

parameter_name1 <- "T_b"
parameter_name2 <- "C_{crit}"
plot_name <- "plot_Vb_Ccrit"
breaks1 <- c(0, 5)
breaks2 <- c(50, 100)
source(file.path(wd, "scripts", "generate_density_plot.R"))

parameter_name1 <- "d_T"
parameter_name2 <- "F_{crit}"
plot_name <- "plot_d_Fcrit"
breaks1 <- c(-1.5, -0.5)
breaks2 <- c(40, 80)
source(file.path(wd, "scripts", "generate_density_plot.R"))

parameter_name1 <- "T_{50}"
parameter_name2 <- "F_{crit}"
plot_name <- "plot_e_Fcrit"
breaks1 <- c(10, 15)
source(file.path(wd, "scripts", "generate_density_plot.R"))

parameter_name1 <- "F_{crit}"
parameter_name2 <- "C_{crit}"
plot_name <- "plot_Fcrit_Ccrit"
breaks1 <- c(40, 80)
breaks2 <- c(50, 100)
source(file.path(wd, "scripts", "generate_density_plot.R"))

p <- plot_grid(plot_Vb_Ccrit + theme(legend.position = "none"), plot_Fcrit_Ccrit + theme(legend.position = "none"),
               plot_d_Fcrit + theme(legend.position = "none"), plot_e_Fcrit + theme(legend.position = "none"), 
          labels=c("(A)", "(B)", "(C)", "(D)"), ncol = 2, nrow = 2,
          hjust = 0, vjust = 1, align = "h", scale = 0.95)

legend <- get_legend(plot_Vb_Ccrit)

plot_grid(p, legend, ncol = 1, nrow = 2, rel_heights= c(0.9, 0.25))

```

```{r parametersplots, include=TRUE, fig.height = 4, fig.width = 6, fig.align = "center", fig.cap = "Impacts of replications of CMA-ES calibration, with PHENOFIT model and beech.", fig.pos="H", fig.showtext=TRUE, eval= F}

data_fruit_par <- species_parameters %>%
  dplyr::filter(grepl("fruit", species_parameters$var))
data_fruit_par$var_name <- factor(data_fruit_par$var, labels = c('fruitmat1' = parse(text=TeX('$aa$')),
                                                               'fruitmat2' = parse(text=TeX('$bb$')),
                                                               'fruitmat3' = parse(text=TeX('$F_{crit}$')),
                                                               'fruitmat4' = parse(text=TeX('$T_{opt}$')),
                                                               'fruitmat5' = parse(text=TeX('$Mat_{moy}$')),
                                                               'fruitmat6' = parse(text=TeX('$\\sigma$')),
                                                               'fruitmat7' = parse(text=TeX('$PFE_{50}$'))))

data_fruit_par$var_name <- factor(data_fruit_par$var, labels = c('fruitmat1' = parse(text=TeX('$aa$')),
                                                               'fruitmat2' = parse(text=TeX('$bb$')),
                                                               'fruitmat3' = parse(text=TeX('$cc$')),
                                                               'fruitmat4' = parse(text=TeX('$dd$')),
                                                               'fruitmat5' = parse(text=TeX('$ee$')),
                                                               'fruitmat6' = parse(text=TeX('$sig$')),
                                                               'fruitmat7' = parse(text=TeX('$gg$'))))

fruit_parameter_plot <- data_fruit_par %>%
  ggplot(aes(y = value, x = factor(1))) +
  geom_violin(fill = 'lightgrey', alpha= 0.2) +
  geom_hline(aes(yintercept = init), linetype = 'dashed', col = 'darkred') +
  geom_beeswarm(col = 'darkred', 
                alpha = 0.5, cex = 2.5, size = 0.4) +
  facet_wrap("var_name", scales="free",
             labeller = label_parsed) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        axis.text.y = element_text(family = "Linux Libertine G"),
        axis.title.y = element_text(family = "Linux Libertine G"),
        strip.text.x = element_text(family = "Linux Libertine G", face = "bold", size = 11),
        axis.ticks.x=element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position="none") +
  geom_blank(aes(y = lb)) +
  geom_blank(aes(y = ub)) +
  labs(y = "")

data_frost_par <- species_parameters %>%
  dplyr::filter(grepl("frost", species_parameters$var))
data_frost_par$var_name <- factor(data_frost_par$var, labels = c('frost1' = parse(text=TeX('$FH_{max1}^{fr}$')),
                                                               'frost2' = parse(text=TeX('$FH_{max2}^{fr}$')),
                                                               'frost3' = parse(text=TeX('$FH_{min}^{fe}$')),
                                                               'frost4' = parse(text=TeX('$FH_{min}^{f1}$')),
                                                               'frost5' = parse(text=TeX('$Te_1$')),
                                                               'frost6' = parse(text=TeX('$Te_2$')),
                                                               'frost7' = parse(text=TeX('$FH_{max}^{tfe}$')),
                                                               'frost8' = parse(text=TeX('$FH_{max}^{tf1}$')),
                                                               'frost9' = parse(text=TeX('$FH_{max}^{pfe}$')),
                                                               'frost10' = parse(text=TeX('$FH_{max}^{pf1}$')),
                                                               'frost11' = parse(text=TeX('$NL_1$')),
                                                               'frost12' = parse(text=TeX('$NL_2$'))))
frost_parameter_plot <- data_frost_par %>%
  ggplot(aes(y = value, x = factor(1))) +
  geom_violin(fill = 'lightgrey', alpha= 0.2) +
  geom_hline(aes(yintercept = init), linetype = 'dashed', col = 'darkred') +
  geom_beeswarm(col = 'darkred', 
                alpha = 0.5, cex = 2.5, size = 0.4) +
  facet_wrap("var_name", scales="free",
             labeller = label_parsed) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        axis.text.y = element_text(family = "Linux Libertine G"),
        axis.title.y = element_text(family = "Linux Libertine G"),
        strip.text.x = element_text(family = "Linux Libertine G", face = "bold", size = 11),
        axis.ticks.x=element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position="none") +
  geom_blank(aes(y = lb)) +
  geom_blank(aes(y = ub)) +
  labs(y = "")

data_senes_par <- species_parameters %>%
  dplyr::filter(grepl("senes", species_parameters$var))
data_senes_par$var_name <- factor(data_senes_par$var, labels = c('senes1' = parse(text=TeX('$P_b$')),
                                                               'senes2' = parse(text=TeX('$T_b$')),
                                                               'senes3' = parse(text=TeX('$\\alpha$')),
                                                               'senes4' = parse(text=TeX('$\\beta$')),
                                                               'senes5' = parse(text=TeX('$S_{crit}$')),
                                                               'senes6' = parse(text=TeX('$\\sigma$'))))
senes_parameter_plot <- data_senes_par %>%
  ggplot(aes(y = value, x = factor(1))) +
  geom_violin(fill = 'lightgrey', alpha= 0.2) +
  geom_hline(aes(yintercept = init), linetype = 'dashed', col = 'darkred') +
  geom_beeswarm(col = 'darkred', 
                alpha = 0.5, cex = 2.5, size = 0.4) +
  facet_wrap("var_name", scales="free",
             labeller = label_parsed) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        axis.text.y = element_text(family = "Linux Libertine G"),
        axis.title.y = element_text(family = "Linux Libertine G"),
        strip.text.x = element_text(family = "Linux Libertine G", face = "bold", size = 11),
        axis.ticks.x=element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position="none") +
  geom_blank(aes(y = lb)) +
  geom_blank(aes(y = ub)) +
  labs(y = "")

plot_grid(fruit_parameter_plot, fruit_parameter_plot, labels=c("(A)", "(B)"), ncol = 2, nrow = 2,
          hjust = 0, vjust = 1, align = "h", label_size = 12, 
          label_colour = "#495057")

plot_grid(fruit_parameter_plot, mad_map, labels=c("(A)", "(B)"), ncol = 2, nrow = 1,
          hjust = 0, vjust = 1, align = "h", label_size = 12, rel_widths = c(1, 0.9),
          label_colour = "#495057")

```

```{r consensusfitnessmap, include=TRUE, fig.height = 4, fig.width = 7, fig.align = "center", fig.cap = "Median absolute deviation of beech \\textbf{(A)} leaf unfolding date and \\textbf{(B)} fitness, predicted with 100 calibrated parameter sets of PHENOFIT.", fig.pos="H", fig.showtext=TRUE}

mad_unfolding <- ggplot(data=unfold_date_data, aes(x = lon, y = unfold_date_data$lat)) + 
  geom_tile(aes(fill = mad), color = NA) +
  theme_void() +
  scale_fill_gradient2(name = stringr::str_wrap("Median absolute deviation", width = 30), 
                       low = "#0A9396", mid ="#e9d8a6", high = "#bb3e03", 
                       limits = c(0,max(unfold_date_data$mad)), 
                       midpoint = 60,
                       guide = guide_colourbar(title.position = "top", frame.colour = "black", frame.linewidth = 0.5, ticks = FALSE)) +
  theme(legend.title.align = 0.5, legend.direction = "horizontal", legend.position = "bottom", 
        legend.key.height = unit(0.2, 'cm'), legend.key.width = unit(1, 'cm'),
        legend.title = element_text(family = "Linux Libertine G", size = 9),
        legend.text = element_text(family = "Linux Libertine G"))

mad_fitness <- ggplot(data=fitness_data, aes(x = lon, y = lat)) + 
  geom_tile(aes(fill = mad), color = NA) +
  theme_void() +
  ylab("") +
  xlab("") +
  scale_fill_gradient2(name = stringr::str_wrap("Median absolute deviation", width = 30), 
                       low = "#0A9396", mid ="#e9d8a6", high = "#bb3e03", 
                       limits = c(0,max(fitness_data$mad)), 
                       midpoint = 0.25,
                       guide = guide_colourbar(title.position = "top", frame.colour = "black", frame.linewidth = 0.5, ticks = FALSE)) +
  theme(legend.title.align = 0.5, legend.position = "bottom", legend.direction = "horizontal",
        legend.key.height = unit(0.2, 'cm'), legend.key.width = unit(1, 'cm'),
        legend.title = element_text(family = "Linux Libertine G", size = 9),
        legend.text = element_text(family = "Linux Libertine G"))

plot_grid(mad_unfolding, mad_fitness, labels=c("(A)", "(B)"), ncol = 2, nrow = 1,
          hjust = 0, vjust = 1, align = "h", label_size = 12,
          label_colour = "#495057")

fitness_data_fsyl <- left_join(fsyl_presabs, fitness_data)

```

The median standard deviation of unfolding date across Europe was about `r round(median(unfold_date_data$sd),1)` days. On beech presence points, it was about `r round(median(unfold_date_data_fsyl[unfold_date_data_fsyl$pres ==1, 'sd']),1)` days. Nearly `r round(nrow(unfold_date_data[unfold_date_data$mad < 30,])/nrow(unfold_date_data)*100,1)`% of cells had a median absolute deviation lower than 30 days (\hyperref[fig:consensusfitnessmap]{Figure F.2.A.}).
The median standard deviation of fitness across Europe was about `r round(median(fitness_data$sd),3)`. On beech presence points, it was about `r round(median(fitness_data_fsyl[fitness_data_fsyl$pres ==1, 'sd']),3)`. Nearly `r round(nrow(fitness_data[fitness_data$mad < 0.1,])/nrow(fitness_data)*100,1)`% and `r round(nrow(fitness_data[fitness_data$mad < 0.2,])/nrow(fitness_data)*100,1)`% of total cells had a median absolute deviation lower than 0.1 and 0.2 respectively (\hyperref[fig:consensusfitnessmap]{Figure F.2.B.}).

```{r unfoldingdatesbackforw, include=TRUE, fig.height = 4, fig.width = 7, fig.align = "center", fig.cap = "Mean leaf unfolding day of beech with \\textbf{(A)} best CMA-ES calibrated parameters and \\textbf{(B)} expert parameters. Values above June solstice day (167) are in grey. Note that PHENOFIT model assign a value of 365 when unfolding has not happened at all due to climate conditions.", fig.pos="H", fig.showtext=TRUE}

best_unfoldingdate <- read_mean_outputvalue_phenofit(file.path(file.path(sim_folder, "CMAES"), phenofit_fagus_best_rep, "LeafUnfoldingDate.txt"))
backward_unfoldingdate <- data.frame(lat= alt$lat, lon =alt$lon, date = t(best_unfoldingdate))
backward_unfoldingdate[backward_unfoldingdate$date > 172, "date"] <- NA # 21 june

backward_unfolding_map <- ggplot(data=backward_unfoldingdate, aes(x = lon, y = lat)) + 
  geom_tile(aes(fill = date), color = NA) +
  theme_void() +
  scale_fill_gradient2(name = stringr::str_wrap("Leaf unfolding day", width = 30), 
                       low = "#0A9396", mid ="#e9d8a6", high = "#bb3e03", 
                       limits = c(100,max(backward_unfoldingdate$date)), 
                       midpoint = 140,
                       guide = guide_colourbar(title.position = "top", frame.colour = "black", frame.linewidth = 0.5, ticks = FALSE),
                       na.value = "grey50") +
  theme(legend.title.align = 0.5, legend.direction = "horizontal", legend.position = "bottom", 
        legend.key.height = unit(0.2, 'cm'), legend.key.width = unit(1, 'cm'),
        legend.title = element_text(family = "Linux Libertine G", size = 9),
        legend.text = element_text(family = "Linux Libertine G"))

forward_sim_folder <- "D:/simulations/phenofit/forward"
unfoldingdate <- read_mean_outputvalue_phenofit(file.path(forward_sim_folder, "fagus_sylvatica/EvolLett2019", "LeafUnfoldingDate.txt"))
forward_unfoldingdate <- data.frame(lat= alt$lat, lon =alt$lon, date = t(unfoldingdate))
forward_unfoldingdate[forward_unfoldingdate$date > 172, "date"] <- NA

forward_unfolding_map <- ggplot(data=forward_unfoldingdate, aes(x = lon, y = lat)) + 
  geom_tile(aes(fill = date), color = NA) +
  theme_void() +
  scale_fill_gradient2(name = stringr::str_wrap("Leaf unfolding day", width = 30), 
                       low = "#0A9396", mid ="#e9d8a6", high = "#bb3e03", 
                       limits = c(100,max(forward_unfoldingdate$date)), 
                       midpoint = 140,
                       guide = guide_colourbar(title.position = "top", frame.colour = "black", frame.linewidth = 0.5, ticks = FALSE),
                       na.value = "grey50") +
  theme(legend.title.align = 0.5, legend.direction = "horizontal", legend.position = "bottom", 
        legend.key.height = unit(0.2, 'cm'), legend.key.width = unit(1, 'cm'),
        legend.title = element_text(family = "Linux Libertine G", size = 9),
        legend.text = element_text(family = "Linux Libertine G"))

plot_grid(backward_unfolding_map , forward_unfolding_map, labels=c("(A)", "(B)"), ncol = 2, nrow = 1,
          hjust = 0, vjust = 1, align = "h", label_size = 12,
          label_colour = "#495057")

fitness_data_fsyl <- left_join(fsyl_presabs, fitness_data)

```

### 2023-03-21 manuscript/paper.Rmd
add manuscript Rmarkdown/Latex files

---
output: 
  stevetemplates::article:
    keep_tex: true
    fig_caption: true
    number_sections: yes
bibliography: bib/paper.bib
csl: bib/mee.csl
title: "A novel approach to inverse calibration of process-based models using evolutionary algorithm and species distribution data"
author:
- name: Victor Van der Meersch \footnote{Corresponding author - victor.vandermeersch@cefe.cnrs.fr}, Isabelle Chuine
  affiliation: CEFE, Université de Montpellier, CNRS, EPHE, IRD, Montpellier, France
abstract: "1. Two main types of species distribution models are used to project species range shifts in future climatic conditions: correlative and process-based models. Although there is some continuity between these two types of models, they are fundamentally different in their hypotheses (statistical relationships *vs* mechanistic relationships) and their calibration methods (dependent *vs* generally independent of the species observed distributions). \n\n 2. One of the limitations to the use of process-based models is the difficulty to parameterize them for a large number of species, while one of the advantages of correlative SDMs is their easier calibration using solely species occurrences}. We investigated the feasibility of using an evolutionary algorithm (called covariance matrix adaptation evolution strategy, CMA-ES) to calibrate process-based models using species distribution data. This method is well established in some fields (robotics, aerospace research, ...), but has never been used, to our knowledge, in ecology, despite its ability to deal with very large space dimensions. Using tree species occurrence data across Europe, we adapted the CMA-ES algorithm to find appropriate values of model parameters. We estimated simultaneously 27 to 77 parameters of two process-based models simulating forest tree's ecophysiology for three species with varying range sizes and geographical distributions. \n\n 3. CMA-ES provided parameter estimates leading to better prediction of species distribution than parameter estimates based on experts knowledge. Our results also revealed that some model parameters and processes were strongly dependent, and different parameter combinations could therefore lead to high model accuracy. \n\n 4. \\textcolor{darkgreen}{We conclude that} CMA-ES is an efficient state-of-the-art method to calibrate process-based models with a large number of parameters using species occurrence data. Inverse modelling using CMA-ES is a powerful method to calibrate process-based parameters which can hardly be measured. However, the method does not warranty that parameter estimates are correct because of several sources of bias, similarly to correlative models, and expert knowledge is required to validate results. \n\n"
keywords: "calibration, evolutionary algorithm, cma-es, species distribution model, process-based model, trees"
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontfamily: libertine
fontsize: 11pt
spacing: double
endnote: no
biblatex: true
link-citations: true

header-includes:
  - \usepackage{graphicx}
  - \usepackage{float}
  - \floatplacement{figure}{H}
  - \usepackage[width=.9\textwidth, font=small]{caption}
  - \usepackage{libertine}
  - \usepackage[left]{lineno}
  - \linenumbers
  - \renewcommand*{\thefootnote}{\fnsymbol{footnote}}
  - \usepackage{xcolor}
  - \definecolor{darkgreen}{RGB}{112,173,71}

---

```{r setup, include=FALSE}

wd <- 'C:/Users/vandermeersch/Documents/CEFE/phd/present_calibration/calibration_paper/second_revision'

knitr::opts_chunk$set(message=FALSE, warning=FALSE,
                      echo = FALSE, cache.lazy = FALSE,
                      fig.path='figs/',
                      cache.path = 'cache/',
                      fig.process = function(x) {
                      x2 = sub('-\\d+([.][a-z]+)$', '\\1', x)
                      if (file.rename(x, x2)) x2 else x
                      })

library(dplyr)
library(ggplot2)
library(ggthemes)
library(ggbeeswarm)
library(showtext)
library(gtools)
library(kableExtra)
options(kableExtra.latex.load_packages = FALSE)
library(AUC)
library(cowplot)
library(ggpubr)
library(knitr)
library(matrixStats) #rowMedians
library(latex2exp)
library(future.apply)
library(extrafont)

font_add("Linux Libertine G", regular = "LinLibertine_R_G.ttf", bold = "LinLibertine_RB_G.ttf", italic = "LinLibertine_RI_G.ttf")
font_import(prompt = F, pattern = "LinLibertine")
loadfonts(device = "all", quiet = TRUE)

source(file.path(wd, "functions", "read_species_file.R"))
source(file.path(wd, "functions", "read_mean_outputvalue_phenofit.R"))
source(file.path(wd, "functions", "read_mean_outputvalue_castanea.R"))

theme_set(theme_cowplot(font_family = "Linux Libertine G"))

# triggers
## fagus and phenofit
recompute_AUC_CMAES <- F
recompute_AUC_ABC <- F
recompute_unfold_date <- F

## quercus and phenofit
recompute_quercus_AUC_CMAES <- F
recompute_quercus_AUC_ABC <- F

## abies and phenofit
recompute_abies_AUC_CMAES <- F

## auc castanea
load_castanea_forward_values <- F
load_castanea_backward_values <- F

# Way to get lat/lon
climate_folder <- "D:/climate/ERA5-Land/phenofit_format/transformed"
alt_file <- paste0(climate_folder, "/ERA5LAND_", "Altitude.fit")
alt <- fread(alt_file, showProgress=F)
colnames(alt) <- c("lat", "lon", "alt")
alt$lat <- round(alt$lat, 1)
alt$lon <- round(alt$lon, 1)
alt$points <- as.numeric(rownames(alt))

```

```{r data, include=FALSE}

sp_folder <- "D:/species/processed"
qile_presabs <- readRDS(file.path(sp_folder, "quercus_ilex/quercus_ilex_presabs_woUkraine.rds"))
fsyl_presabs <- readRDS(file.path(sp_folder, "fagus_sylvatica/fagus_sylvatica_presabs_woUkraine.rds"))
aalb_presabs <- readRDS(file.path(sp_folder, "abies_alba/abies_alba_presabs_woUkraine.rds"))

```

```{r mapmonde, include=FALSE}

library(rnaturalearth)
library(sf)
library(tidyverse)

# Europe map
countries_EUForest <- c("Austria", "Belarus", "Belgium", "Bulgaria", "Croatia", "Cyprus", "Czech Republic", "Denmark", "Estonia", "Finland", "France",
                        "Germany", "Greece", "Hungary", "Ireland", "Italy", "Latvia", "Lithuania", "Moldova", "Netherlands", "Norway", "Poland", "Portugal",
                        "Romania", "Slovakia", "Slovenia", "Spain", "Sweden", "Switzerland", "United Kingdom")
other_countries <- c("Bosnia and Herzegovina", "Republic of Serbia", "Macedonia", "Greece",
                     "Kosovo", "Albania", "Montenegro", "Malta", "Liechtenstein", "Luxembourg")
countries <- c(countries_EUForest, other_countries)
world_map <- ne_countries(scale="medium",returnclass = 'sf')
eu_map <- world_map %>% filter(sovereignt %in% countries)
eu_map_cropped <- eu_map %>% 
  st_crop(st_bbox(c(xmin = -12, xmax = 45, ymax = 71, ymin = 32), crs = st_crs(4326)))

ERA5_points <- st_as_sf(alt, coords = c("lon", "lat"), crs = st_crs(4326))
ERA5_points <- ERA5_points[st_intersects(ERA5_points, st_buffer(eu_map_cropped, 10000)) %>% lengths > 0,] 
ERA5_points <- ERA5_points %>%
  mutate(lat = round(unlist(map(ERA5_points$geometry,2)),1),
         lon = round(unlist(map(ERA5_points$geometry,1)),1)) %>%
  st_drop_geometry() %>%
  dplyr::select(-c("points"))
names(ERA5_points)[1] <- "pres"

ERA5_grid_points <- left_join(ERA5_points, alt,  by = c('lat', 'lon'))
ERA5_grid_points <- ERA5_grid_points[, c("lat", "lon", "points")]

```

# Introduction

The speed and magnitude of projected climate changes are profoundly affecting species distributions, ecological communities and ecosystem processes, and numerous ecological systems are now approaching tipping points [@Lenton2008; @Barnosky2012; @Steffen2018; but see @Brook2013]. Large uncertainties on the persistence and the resilience of ecosystems exist. Ecological forecasting has now become a critical tool for managers and decision-makers [@Urban2015], and robust predictive approaches are necessary to provide reliable projections of species geographic range shifts and ecosystem functioning [@Mouquet2015]. Forecasting the dynamics of ecological systems for the upcoming decades and centuries is very difficult, because ecological systems are extremely complex, influenced by a lot of factors and processes, and climatic conditions with no analogues in the recent past are forecasted to become common [@Williams2007; @Radeloff2015; @Fitzpatrick2018]. Ecological models have thus increased in complexity over the last 50 years, incorporating more and more processes described with various degrees of complexity depending on their objectives.

Nowadays, two main types of species distribution models (SDM) are used to project species range shifts in future climatic conditions: correlative and process-based models [@Dormann2012]. The vast majority of currently used SDMs are correlative: they seek to find statistical relationships between various environmental descriptors and species presence and absence. They assume there is an equilibrium between species distribution and environment [equilibrium postulate, @Guisan2005], that there is no adaptive responses within a generation [no trait plasticity, @Berzaghi2020], and that species niche is stable over macroevolutionary time [niche conservatism, @Pearman2008]. Most of them include a fairly large number of predictors (particularly in machine-learning approaches), and consider flexible transformations (linear, quadratic...) and interactions between them [@Merow2014]. Even though some authors advocate for "putting more biology into SDMs" [@Higgins2012], parameters have no a priori defined ecological meaning [@Dormann2012] and shape of response curves to environmental variables is generally not constrained based on biological considerations. Although these models are not always used correctly [@Araujo2019; @Santini2021], their flexibility makes them an important tool in predictive ecology [@Mouquet2015]. They have been widely used especially to generate species range projections under current and future climates [e.g. @Guisan2005]. Nevertheless, their ability to accurately describe the effects of climate on species distributions has recently been questioned [e.g. @Fourcade2018; @Journe2020; @Warren2021].  
For all these reasons, another kind of models has been developed. Process-based models aim to translate into mathematical equations our knowledge about the physiological and ecological processes involved in an organism’s life, such as growth, reproduction, survival, movement, and interactions with other livebeings. Process-based models take more time to develop and are more challenging to use, but they might provide a greater comprehension of the complexity of ecosystem dynamics and more robust projections in novel conditions [@Evans2012; @Zurell2016; @Urban2016; @Singer2016]. A wide variety of process-based models exists, from quite simple models [e.g. @Kleidon2000] to much more complex ones [e.g. @Dufrene2005]. They all rely on an explicit representation of processes at stake, with a direct biological interpretation - at least in principle [@Connolly2017]. Process-based models may also include some phenomenological relationships on lower-level processes, but never on the pattern itself. The choices about the specific processes to include into the model are made based on theory, empirical observations and the objectives of the research, and modeler subjectivity may play an important role. One of the challenges is to build a model with the appropriate amount of complexity: a too simplistic model might be unrealistic whereas a very complex model could be far beyond our ability to understand it (because of interconnected mechanisms) and calibrate it. Each model relies on different hypotheses with its own balance of complexity, accuracy and parsimony - and thus different numbers of unknown parameters to calibrate. Parameter values of this kind of models are obtained with different methods. Some of these parameters can be individually estimated with field observations or experimental data, or are already available in the literature. When this is not possible, groups of parameters defining a relationship between a process variable and environmental variables or other processes are estimated jointly using inverse modelling methods and data on the processes modelled [e.g. @Cailleret2020; @Asse2020].

Calibration (i.e. parameter setting and estimation) is a fundamental step in the modelling process. It allows the model to reproduce the reality with more or less success. The result of the calibration provides insights on the ability of the model to reproduce and explain reality (model predictive power). Calibration of complex models such as process-based SDMs is time-consuming, and modelers are often challenged by the dimension of the parameter space, the complexity of the possible correlations among parameters, and the scarcity of observed experimental data to calibrate them. Therefore, although the philosophy of such models is to measure parameters, statistical inference might be useful when data are not yet available to infer parameter values. Parameter inference can be achieved through several methods which have been developed in the last decades. \textcolor{darkgreen}{Most of them fall into two categories: informal and statistical calibration. On the one hand, statistical calibration assumes a data-generating model, and the likelihood function, which quantifies the probability of the observed data given the model parameters, is expressed mathematically. This likelihood is the foundation of maximum likelihood estimation and Bayesian inference, two commonly used methods for parameter estimation. In practice, deriving the likelihood function can be quite challenging for complex models with many parameters. One approach is to use numerical methods such as simulation-based inference or Approximate Bayesian Computation} [e.g. @Hartig2014]\textcolor{darkgreen}{. However, obtaining a reliable estimate of the likelihood of a complex process-based model can require a large number of simulated samples, which can be computationally expensive. On the other hand, using an informal objective function to measure the discrepancy between the model predictions and the observed data may be a reasonable alternative, even though it may not have the same statistical rigor. By minimizing the objective function, one is able to identify the parameter values that best fit the observed data. This optimization step can be achieved with many methods, either deterministic (e.g. Nelder-Mead method) or stochastic (e.g. simulated annealing, evolutionary algorithms).}     
Recently, an approach belonging to the evolutionary algorithm family, called Covariance Matrix Adaptation Evolution Strategy (CMA-ES), has been proposed [@Hansen2001]. One of the advantages of CMA-ES is its ability to cumulate information over iterations in order to adapt its own parameters (in particular the covariance matrix), which makes it more robust to noise. CMA-ES is especially performant for non-separable problems (i.e. when the model parameters are dependent) and large search space. This method has been successfully applied in various fields such as aerospace [e.g. @Collange2010], optics [e.g. @Gagne2008], and robotics [e.g. @Hill2020]. CMA-ES is acknowledged to be one of the most efficient approaches in continuous black-box optimization [@Hansen2010] but to our knowledge has never been used in ecology.

Here we explored the feasibility and interests of calibrating process-based SDMs with CMA-ES using species occurrence data as correlative SDMs do [fitted process-based models *sensu* @Dormann2012]. We focused on two forest process-based models of varying levels of complexity to evaluate the ability of CMA-ES to calibrate such models. The two models are PHENOFIT [27 to 36 parameters, @Chuine2001] and CASTANEA [77 parameters, @Dufrene2005]. Each model also emphasizes different ecological processes: while PHENOFIT focuses on phenology and how it relates to survival and reproduction, CASTANEA focuses on carbon and water cycles. We focused on three European common tree species, with different range extent and ecological preferences in order to evaluate the algorithms performance in various geographical and climatic conditions. European beech (*Fagus sylvatica L.*) is one of the most widely distributed broadleaved tree in Europe (from southern Sweden to Sicily and from Spain to northwest Turkey), holm oak (*Quercus ilex L.*) is an evergreen broadleaved tree native of the Mediterranean region, and silver fir (*Abies alba Mill.*) is a coniferous tree which mainly occurs in mountain forests of Central Europe and some parts of Southern and Eastern Europe.

# 1. Material and methods {#methods}

## 1.1. Process-based models {#pbmodels}

All versions of the models used for this study are coded in Java and distributed by the [CAPSIS](https://capsis.cirad.fr/capsis/home) platform.

PHENOFIT is a process-based species distribution model for forest tree species which focuses on phenology. It relies on the principle that the distribution of a tree species depends mainly on the synchronization of its timing of development to the local climatic conditions [@Chuine2001]. It is composed of several submodels, including phenology models for leaves, flowers and fruits, and stress resistance models. It simulates the fitness (survival and reproductive success) of an average individual using daily meteorological data, soil water holding capacity and species specific parameters (see [Appendix A](#appendixA) for details). PHENOFIT has been validated for several North American and European species by comparing their known distribution to the modelled fitness [e.g. @Morin2007; @Saltre2013; @Duputie2015; @Gauzere2020].

CASTANEA is an ecophysiological process-based model which simulates carbon and water fluxes in forests [@Dufrene2005]. The model simulates the ecosystem as an average tree with six compartments (leaves, branches, stem, coarse roots, fine roots and reserves). It is much more complex than PHENOFIT, with several processes described and computed, such as photosynthesis, stomatal opening, maintenance and growth respiration, transpiration, and carbon allocation (see [Appendix A](#appendixA) for details). CASTANEA requires daily meteorological variables and soil characteristics. The model has been initially validated at stand scale for beech [@Davi2005], and was then successfully applied to other European species [e.g. @Davi2006; @Delpierre2012; @Davi2017].

Both models, in their standard version (called here after expert calibration), are parameterized using various sources of information. Some parameters are directly measured or found in the literature such as leaf life span, specific leaf area, LT50 (freezing temperature causing 50\% mortality) of leaves, leaf reflectance, and so on. Other parameters cannot be measured at all, or their measurement require an enormous effort that cannot be deployed for a large number of species. These parameters are thus inferred by inverse modelling using either Bayesian methods or optimizing methods and data on the specific process they are describing. This is for example the case for phenology models for which all parameters cannot be measured, especially those describing bud dormancy break regulation, since no method allow to measure dormancy break precisely so far (except for a few fruit tree species). Finally, a few parameters are prescribed based on expert knowledge as no data to estimate them exist. For this study, the standard versions of both models were run for the three species using the same climatic data used to do the inverse calibration (see [1.2.1. Climate data](#climatedata)), in order to compare the correctness of the predictions obtained with the two types of calibration.

## 1.2. Data for the calibration {#data}

### 1.2.1. Climate and soil data {#climatedata}

Raw climatic variables were extracted from ERA5-Land hourly dataset [@Sabater2019; @Sabater2021] from 1970 to 2000, at a spatial resolution of 0.1 degree in latitude and longitude. We calculated the daily mean values of the following variables used by PHENOFIT and CASTANEA: minimum, mean and maximum daily temperatures, mean dewpoint temperature, daily precipitation, daily global radiation and daily mean wind speed. We computed the daily relative humidity with the ratio of vapor pressure and saturation vapor pressure (both calculated with Clausius-Clapeyron equation) using *humidity* R package [@Cai2019]. Daily potential evapotranspiration was calculated with Penman–Monteith equation (FAO standard of hypothetical grass reference surface) using a slightly modified version of the *ET()* function in *Evapotranspiration* R package [@Guo2016].

Water content at field capacity and wilting point data were extracted from EU-SoilHydroGrids [@Toth2017] which is at 1km resolution. Percentage of sand, silt and clay particles, percentage of coarse fragments, bulk density and soil depth were extracted from SoilGrids250m [@Hengl2017] at a 250m resolution. These data (except for soil depth) are provided at seven soil depths, so we summarized them (weighted sum or weighted mean) taking into account each layer width and total soil depth. Finally, all variables were upscaled at the ERA5-Land spatial resolution 0.1° using bilinear interpolation.

### 1.2.2. Tree occurrence data used for the calibration {#occurrencedata}

Sources of occurrence data are known to differ even for common European trees [@Duputie2014] and this makes it quite challenging to gather comprehensive data at a sufficient spatial resolution all over Europe. The occurrence data we used essentially rely on the EU-Forest dataset [@Mauri2017] which benefits from inventory and monitoring programmes implemented in most European countries. 
As EU-Forest is limited to forest ecosystems, we combined it with presence records extracted from the Global Biodiversity Information Facility [@GBIF2022, see [Appendix B](#appendixB) for all download links] but removing observations outside natural species ranges as defined by Atlas Flora Europeae [AFE, @AFE2005] and EuroVegMap [@EVM2003]. By doing so, we also included occurrences of isolated native trees living outside forests, excluding records from arboreta or gardens where the species would have been planted as an exotic. For holm oak, we also added occurrence records in the Mediterranean Basin from the WOODIV database [@Monnet2021], leaving out EU-Forest and GBIF records we had already gathered.
We upscaled all species records at the ERA5-Land resolution (0.1°, see [1.2.1. Climate data](#climatedata)), i.e. the species is considered to be present in the cell if there is at least one record. We finally obtained `r nrow(fsyl_presabs[fsyl_presabs$pres == 1,])` occurrence cells for beech, `r nrow(qile_presabs[qile_presabs$pres == 1,])` for holm oak and `r nrow(aalb_presabs[aalb_presabs$pres == 1,])` for silver fir (see [Appendix B](#appendixB) for details).

All the datasets described above are presence-only data. Therefore, we generated cells where species are supposed to be absent, i.e. pseudo-absence cells. In order to avoid as far as possible creating false absence data, we used EU-Forest cells where the species is not reported present as pseudo-absence cells. We assumed that national forest inventories were exhaustive (which is not true since only specific forest plots in a 0.1° cell are monitored). We obtained `r nrow(fsyl_presabs[fsyl_presabs$pres == 0,])` absence cells for beech, `r nrow(qile_presabs[qile_presabs$pres == 0,])` for holm oak and `r nrow(aalb_presabs[aalb_presabs$pres == 0,])` for silver fir (see [Appendix C](#appendixC)).

We selected subsets of 2000 points (1000 presences and 1000 pseudo-absences) in order to reduce computational costs. For each species, we generated ten presence clusters (k-means algorithm) of similar bioclimatic conditions based on annual climate normals computed with R package *dismo* [@Hijmans2021] and ERA5-Land variables. In each cluster, we randomly sampled a number of cells where the species is present proportional to the total number of  a number of cells where the species is present in the cluster. The aim of this stratified random sampling was to make sure that all species environmental preferences were proportionally represented. We then randomly sampled the same number of pseudo-absence cells (see [Appendix B](#appendixB) for details).

Regarding PHENOFIT model, we calibrated ten times each species parameter set, with 5 repetitions on 2 random subsets of presences/pseudo-absences, except for beech. In the latter case, we ran 10 repetitions on 10 subsets (i.e. 100 calibrations) to investigate both the effect of subsampling and the effect of stochasticity. Since CASTANEA computing time was much higher (see \autoref{tab:modelstable}), we ran only two calibration for each species (on  2 different random subsets).

## 1.3. Model calibration

### 1.3.1. Covariance Matrix Adaptation Evolution Strategy principles

Covariance Matrix Adaptation Evolution Strategy (CMA-ES) is widely accepted as a robust optimization algorithm for non-linear, non-convex, as well as non-separated optimization problems in continuous domain [@Hansen1996; @Hansen2001; @Hansen2006]. 
It is based on the principle of evolutionary biology, via recombination, mutation and selection of the most fit candidate solutions (i.e. parameter sets providing the best predictions). At each iteration:  
- $\lambda$ candidate solutions are evaluated, i.e. model runs $\lambda$ times with $\lambda$ different parameter sets and the objective function is evaluated  
- the best $\mu$ candidate solutionss are selected  
- the weighted mean candidate solution $m$ is computed (mean of the best $\mu$ parameter sets weighted by their objective function value)
- covariance matrix $C$ and step size $\sigma$ are updated (with information accumulated over several consecutive iterations)   
- new $\lambda$ candidate solutions are sampled in a normal distribution $\mathcal{N}(m, \sigma C)$, with both recombination (via the favorite solution $m$) and mutations (via the perturbations $\sigma C$)   
One of the strength of this approach lies in the combination of rank-$\mu$-update, where prior information from previous generations is exploited (mean of the previous covariance matrices, with a higher weight for recent generations), and cumulation, where correlations between generations are retained in an evolution path (sum of consecutive steps), to update the covariance matrix at each step [see @Hansen2016 for a detailed description of the algorithm].

### 1.3.2. CMA-ES in practice

One of the advantages of CMA-ES is that it does not require a complex parameter tuning: as best parameter values at a given time of the optimization process might no longer be efficient later, CMA-ES implements an internal adaptation of its parameters. We only chose the number of candidate solutions $\lambda$, depending on the optimization problem complexity ($\mu$ was set to $\lambda/2$). The default recommended value for $\lambda$ is $4 + 3 ln(N)$, where $N$ is the number of parameters to calibrate (i.e. $\lambda \in [14,17]$ in our case). We set $\lambda = 20$, in order to improve the global search capability [@Hansen2004] and take advantage of the computation power at our disposal. All model parameters were linear scaled into $[0;10]$ so that the same standard deviation can be applied to all parameters: here we chose $\sigma = 2$ (see [Nikolaus Hansen personal website](https://cma-es.github.io/cmaes_sourcecode_page.html#practical) for practical hints on variable encoding). Our stopping criterion for the optimization procedure was the budget, i.e. the number of model runs.

For an easier use and the sake of reproducibility, we chose to use a pure R implementation of CMA-ES available in the R package *cmaes* [@Trautmann2011]. The function *cma_es()* enables us to do $\lambda$ function evaluations in parallel so as to substantially reduce computation time. It also allows us to define lower and upper bound constraints, by penalising the objective function value of the candidate solution if it violates the boundaries. We customized the *cma_es()* function to add an option to define death penalty constraints (rejection of the infeasible candidate solution who is sampled again), in order to define a range of ecologically possible solutions in terms of inequality constraints between parameters (see [Appendix D](#appendixD) for details about boundaries and constraints handling). Death penalty is the easiest way to handle constraints when the feasible region is fairly large, but it is not perfect as there is no use of information from infeasible points (i.e. points which violate the constraints).

The objective function for the calibration was the area under the receiver operating characteristic curve (AUC), evaluated against a subsets of 2000 points (see [1.2.2. Tree occurrence data](#occurrencedata)). Although AUC has been criticized as an imperfect measure of model performance [@Lobo2008; @Leroy2018], we used it as objective function because our goal here was only to calibrate models by maximizing discriminating capacity (i.e. potential to correctly classify presences and absences) with a threshold-independent measure. We used the *AUC* R package [@Ballings2013], and chose the two following model output variables as proxies of classification probabilities (i.e. used to determine if the species can be present or not): fitness index for PHENOFIT and carbon reserves for CASTANEA (see [Appendix A](#appendixA)).

We implemented the CMA-ES calibration on two computing clusters: GenOuest from IRISA-INRIA ([genouest.org](https://www.genouest.org)) and TGCC (*Très Grand Centre de Calcul*) from CEA ([hpc.cea.fr](http://www-hpc.cea.fr/fr/complexe/tgcc.htm)). As the models are coded in Java (see [1.1 Process-based models](#pbmodels)), they need a process of deallocating memory handled by a *garbage collector*. For PHENOFIT, each function evaluation (i.e. each model simulation) was run on a 2-core computing unit in order to have enough computing resources for both simulation and garbage collection. We thus needed twice as many cores as functions evaluated in parallel. CASTANEA model requires a fairly high computation time, so we used a nested parallelism distribution, where each parallel simulation was distributed on 4 computing units. We thus used 4 times as many cores as functions evaluated in parallel, plus some extra cores for garbage collection. We used R package *future* [@Bengtsson2021] for parallel processing.

```{r modelstable, results='asis'}
table <- data.frame(
  model = c('PHENOFIT', 'CASTANEA'),
  output = c('Fitness index', 'Carbon reserves'),
  npar = c('$[27;36]$', '$77$'),
  lambda = c('$20$', '$20$'),
  ncores = c('$40$', '$100$'),
  memory = c('$80$ GB', '$120$ GB'),
  eval = c('6000', '4000'),
  runtime = c("$\\sim 24$ hours", "$\\sim 20$ days")
)
kable(table, booktabs = TRUE, escape = FALSE, 
      col.names = c("Model", "Output variable of interest", "Number of parameters calibrated", "Number of candidate solutions $\\lambda$", "Number of cores", "Total memory", "Number of model evaluations", "Average runtime for one calibration"),
      caption = "Summary of model calibration settings. Average runtime was assessed on the GenOuest cluster.",
      linesep = "\\addlinespace") %>%
  kable_classic(full_width = T, html_font = "Linux Libertine G")

# add_header_above(c("Model" = 2, "Calibration" = 4)) %>%

```

# 2. Results

```{r phenofit_fagus_rep_data}

cal_folder <- 'D:/calibrations/phenofit/fagus_sylvatica/1000pres_1000abs/paper_data'

sim_folder <- 'D:/simulations/phenofit/backward/fagus_sylvatica/paper_data'

# Calibrations
cmaes_files <- mixedsort(list.files(path = file.path(cal_folder,"CMAES"), pattern = "\\.Rdata$", full.names = T, recursive = T))

auc_data_cal <- data.frame()
for(i in 1:length(cmaes_files)){
  auc_data_cal[i, 'subset'] <- paste0(strsplit(cmaes_files[i], "_")[[1]][7])
  auc_data_cal[i, 'rep'] <- strsplit(strsplit(cmaes_files[i], "_")[[1]][8], "[.]")[[1]][1]

  # Load AUC on calibration_points
  load(cmaes_files[i])
  auc_cal <- 1-round(cmaes_fit[[1]][["value"]], 6)

  auc_data_cal[i, 'cal'] <- auc_cal

  # # Load AUC on every pres/abs points
  # sim_dir <- file.path(sim_folder, cal_list[i])
  # fitness_presabs <- read_mean_outputvalue_phenofit(sim_dir, "Fitness", points = fsyl_presabs)
  # auc_presabs <- round(auc(roc(fitness_presabs, as.factor(fsyl_presabs$pres))),3)
  # 
  # auc_data[i, 'presabs'] <- auc_presabs
}

if(recompute_AUC_CMAES){
  simulation_files <- mixedsort(list.files(path = file.path(sim_folder, "CMAES"), pattern = "Fitness", full.names = T, recursive = T))

  auc_data_tot <- data.frame()
  fitness_data <- as.data.frame(cbind(alt$lat, alt$lon))
  for(i in 1:(length(simulation_files))){
    auc_data_tot[i, 'subset'] <- strsplit(strsplit(simulation_files[i], "/")[[1]][8], "_")[[1]][1]
    auc_data_tot[i, 'rep'] <- strsplit(strsplit(simulation_files[i], "/")[[1]][8], "_")[[1]][2]

    # Compute AUC on every pres/abs points
    fitness_presabs <- read_mean_outputvalue_phenofit(simulation_files[i], points = fsyl_presabs)
    auc_tot <- round(auc(roc(fitness_presabs, as.factor(fsyl_presabs$pres))),6)
    auc_data_tot[i, 'tot'] <- auc_tot

    # Best threshold
    youden_index <- sensitivity(fitness_presabs, as.factor(fsyl_presabs$pres), perc.rank = F)$measure +
      specificity(fitness_presabs, as.factor(fsyl_presabs$pres), perc.rank = F)$measure - 1
    thresholds <- sensitivity(fitness_presabs, as.factor(fsyl_presabs$pres), perc.rank = F)$cutoffs
    best_threshold <- thresholds[which(youden_index == max(youden_index))]
    auc_data_tot[i, 'best_threshold'] <- best_threshold

    # Load fitness
    fitness <- read_mean_outputvalue_phenofit(simulation_files[i])
    fitness_data[,2+i] <- t(fitness)

  }

  # Consensus 
  names(fitness_data)[1:2] <- c("lat", "lon")
  fitness_data$mean <- rowMeans(fitness_data[,3:ncol(fitness_data)])
  fitness_data$mad <- rowMads(as.matrix(fitness_data[,3:ncol(fitness_data)]))
  fitness_data$sd <- rowSds(as.matrix(fitness_data[,3:ncol(fitness_data)]))

  saveRDS(auc_data_tot, file.path(wd, "cache", "fagus_auc_data_tot.rds"))
  saveRDS(fitness_data, file.path(wd, "cache", "fagus_fitness_data.rds"))
}else{
  auc_data_tot <- readRDS(file.path(wd, "cache", "fagus_auc_data_tot.rds"))
  fitness_data <- readRDS(file.path(wd, "cache", "fagus_fitness_data.rds"))
}

fagus_auc_data_cal <- auc_data_cal
fagus_auc_data_tot <- auc_data_tot

```

```{r phenofit_quercus_rep_data}

cal_folder <- 'D:/calibrations/phenofit/quercus_ilex/1000pres_1000abs/paper_data'
sim_folder <- 'D:/simulations/phenofit/backward/quercus_ilex/paper_data'

# Calibrations
cmaes_files <- mixedsort(list.files(path = file.path(cal_folder,"CMAES"), pattern = "\\.Rdata$", full.names = T, recursive = T))

auc_data_cal <- data.frame()
for(i in 1:length(cmaes_files)){
  auc_data_cal[i, 'subset'] <- paste0(strsplit(cmaes_files[i], "_")[[1]][7])
  auc_data_cal[i, 'rep'] <- strsplit(strsplit(cmaes_files[i], "_")[[1]][8], "[.]")[[1]][1]

  # Load AUC on calibration_points
  load(cmaes_files[i])
  auc_cal <- 1-round(cmaes_fit[[1]][["value"]], 6)

  auc_data_cal[i, 'cal'] <- auc_cal

  # # Load AUC on every pres/abs points
  # sim_dir <- file.path(sim_folder, cal_list[i])
  # fitness_presabs <- read_mean_outputvalue_phenofit(sim_dir, "Fitness", points = qile_presabs)
  # auc_presabs <- round(auc(roc(fitness_presabs, as.factor(qile_presabs$pres))),3)
  # 
  # auc_data[i, 'presabs'] <- auc_presabs
}

if(recompute_quercus_AUC_CMAES){
  simulation_files <- mixedsort(list.files(path = file.path(sim_folder, "CMAES"), pattern = "Fitness", full.names = T, recursive = T))

  auc_data_tot <- data.frame()

  for(i in 1:(length(simulation_files))){
    auc_data_tot[i, 'subset'] <- strsplit(strsplit(simulation_files[i], "/")[[1]][8], "_")[[1]][1]
    auc_data_tot[i, 'rep'] <- strsplit(strsplit(simulation_files[i], "/")[[1]][8], "_")[[1]][2]

    # Compute AUC on every pres/abs points
    fitness_presabs <- read_mean_outputvalue_phenofit(simulation_files[i], points = qile_presabs)
    auc_tot <- round(auc(roc(fitness_presabs, as.factor(qile_presabs$pres))),6)
    auc_data_tot[i, 'tot'] <- auc_tot

    # Best threshold
    youden_index <- sensitivity(fitness_presabs, as.factor(qile_presabs$pres), perc.rank = F)$measure +
      specificity(fitness_presabs, as.factor(qile_presabs$pres), perc.rank = F)$measure - 1
    thresholds <- sensitivity(fitness_presabs, as.factor(qile_presabs$pres), perc.rank = F)$cutoffs
    best_threshold <- thresholds[which(youden_index == max(youden_index))]
    auc_data_tot[i, 'best_threshold'] <- best_threshold

    # Load fitness
    #fitness <- read_mean_outputvalue_phenofit(simulation_files[i])
    #fitness_data[,2+i] <- t(fitness)

  }

  # Consensus 
  # names(fitness_data)[1:2] <- c("lat", "lon")
  # fitness_data$mean <- rowMeans(fitness_data[,3:ncol(fitness_data)])
  # fitness_data$mad <- rowMads(as.matrix(fitness_data[,3:ncol(fitness_data)]))
  # fitness_data$sd <- rowSds(as.matrix(fitness_data[,3:ncol(fitness_data)]))

  saveRDS(auc_data_tot, file.path(wd, "cache", "quercus_auc_data_tot.rds"))
  # saveRDS(fitness_data, file.path(wd, "cache", "quercus_fitness_data.rds"))
}else{
  auc_data_tot <- readRDS(file.path(wd, "cache", "quercus_auc_data_tot.rds"))
  #fitness_data <- readRDS(file.path(wd, "cache", "fitness_data.rds"))
}

quercus_auc_data_cal <- auc_data_cal
quercus_auc_data_tot <- auc_data_tot

```

```{r phenofit_abies_rep_data}

cal_folder <- 'D:/calibrations/phenofit/abies_alba/1000pres_1000abs/paper_data'
sim_folder <- 'D:/simulations/phenofit/backward/abies_alba/paper_data'

# Calibrations
cmaes_files <- mixedsort(list.files(path = file.path(cal_folder,"CMAES"), pattern = "\\.Rdata$", full.names = T, recursive = T))

auc_data_cal <- data.frame()
for(i in 1:length(cmaes_files)){
  auc_data_cal[i, 'subset'] <- paste0(strsplit(cmaes_files[i], "_")[[1]][7])
  auc_data_cal[i, 'rep'] <- strsplit(strsplit(cmaes_files[i], "_")[[1]][8], "[.]")[[1]][1]

  # Load AUC on calibration_points
  load(cmaes_files[i])
  auc_cal <- 1-round(cmaes_fit[[1]][["value"]], 6)

  auc_data_cal[i, 'cal'] <- auc_cal

  # # Load AUC on every pres/abs points
  # sim_dir <- file.path(sim_folder, cal_list[i])
  # fitness_presabs <- read_mean_outputvalue_phenofit(sim_dir, "Fitness", points = qile_presabs)
  # auc_presabs <- round(auc(roc(fitness_presabs, as.factor(qile_presabs$pres))),3)
  # 
  # auc_data[i, 'presabs'] <- auc_presabs
}

if(recompute_abies_AUC_CMAES){
  simulation_files <- mixedsort(list.files(path = file.path(sim_folder, "CMAES"), pattern = "Fitness", full.names = T, recursive = T))

  auc_data_tot <- data.frame()
  for(i in 1:(length(simulation_files))){
    auc_data_tot[i, 'subset'] <- strsplit(strsplit(simulation_files[i], "/")[[1]][8], "_")[[1]][1]
    auc_data_tot[i, 'rep'] <- strsplit(strsplit(simulation_files[i], "/")[[1]][8], "_")[[1]][2]

    # Compute AUC on every pres/abs points
    fitness_presabs <- read_mean_outputvalue_phenofit(simulation_files[i], points = aalb_presabs)
    auc_tot <- round(auc(roc(fitness_presabs, as.factor(aalb_presabs$pres))),6)
    auc_data_tot[i, 'tot'] <- auc_tot

    # Best threshold
    youden_index <- sensitivity(fitness_presabs, as.factor(aalb_presabs$pres), perc.rank = F)$measure +
      specificity(fitness_presabs, as.factor(aalb_presabs$pres), perc.rank = F)$measure - 1
    thresholds <- sensitivity(fitness_presabs, as.factor(aalb_presabs$pres), perc.rank = F)$cutoffs
    best_threshold <- thresholds[which(youden_index == max(youden_index))]
    auc_data_tot[i, 'best_threshold'] <- best_threshold

    # Load fitness
    #fitness <- read_mean_outputvalue_phenofit(simulation_files[i])
    #fitness_data[,2+i] <- t(fitness)

  }

  # Consensus 
  # names(fitness_data)[1:2] <- c("lat", "lon")
  # fitness_data$mean <- rowMeans(fitness_data[,3:ncol(fitness_data)])
  # fitness_data$mad <- rowMads(as.matrix(fitness_data[,3:ncol(fitness_data)]))
  # fitness_data$sd <- rowSds(as.matrix(fitness_data[,3:ncol(fitness_data)]))

  saveRDS(auc_data_tot, file.path(wd, "cache", "abies_auc_data_tot.rds"))
  # saveRDS(fitness_data, file.path(wd, "cache", "quercus_fitness_data.rds"))
}else{
  auc_data_tot <- readRDS(file.path(wd, "cache", "abies_auc_data_tot.rds"))
  #fitness_data <- readRDS(file.path(wd, "cache", "fitness_data.rds"))
}

abies_auc_data_cal <- auc_data_cal
abies_auc_data_tot <- auc_data_tot

```

## 2.1. Calibration results

```{r phenofit_fitness_maps}
lcol <- "#EBEBD3"
hcol <- "#488B49"

# lcol <- "#ffeaa1"
# hcol <- "#ec572f"

sim_folder <- 'D:/simulations/phenofit/backward/fagus_sylvatica/paper_data'
phenofit_fagus_best_rep <- paste0(fagus_auc_data_tot[fagus_auc_data_tot$tot == max(fagus_auc_data_tot$tot), "subset"],"_", fagus_auc_data_tot[fagus_auc_data_tot$tot == max(fagus_auc_data_tot$tot), "rep"])
best_fit <- read_mean_outputvalue_phenofit(file.path(file.path(sim_folder, "CMAES"), phenofit_fagus_best_rep, "Fitness.txt"), points = ERA5_points)
phenofit_fagus_best_fit <- as.data.frame(cbind(ERA5_points$lat, ERA5_points$lon,t(best_fit)))
colnames(phenofit_fagus_best_fit) <- c("lat", "lon", "fitness")
# best threshold
best_threshold <- fagus_auc_data_tot[fagus_auc_data_tot$tot == max(fagus_auc_data_tot$tot), "best_threshold"]
phenofit_fagus_best_fit[phenofit_fagus_best_fit$fitness < best_threshold, 'presence'] <- 0
phenofit_fagus_best_fit[phenofit_fagus_best_fit$fitness >= best_threshold, 'presence'] <- 1
phenofit_fagus_map <- ggplot(data=phenofit_fagus_best_fit, aes(x = lon, y = lat)) + 
  geom_raster(aes(fill = presence), color = NA) +
  theme_void() +
  labs(fill = "Fitness") +
  ylab("") +
  xlab("") +
  scale_fill_gradient(low = lcol, high = hcol, limits = c(0,1), breaks = c(0,0.5,1), na.value = "#B5BEC6") +
  guides(fill = guide_colorbar(title.position = "top", direction = "horizontal", frame.colour = "black", frame.linewidth = 0.5,
                               ticks = FALSE)) +
  theme(legend.title.align = 0.5, legend.position = "bottom", 
        legend.key.height = unit(0.2, 'cm'), legend.key.width = unit(0.5, 'cm'),
        legend.title = element_text(family = "Linux Libertine G", size = 9),
        legend.text = element_text(family = "Linux Libertine G")) +
  annotate("text", x = -4, y = 65, label = paste("AUC =", round(max(fagus_auc_data_tot$tot), 2)), colour = 'darkgrey', size = 3.6, 
           family = "Linux Libertine G", fontface = "bold")

sim_folder <- 'D:/simulations/phenofit/backward/quercus_ilex/paper_data'
phenofit_quercus_best_rep <- paste0(quercus_auc_data_tot[quercus_auc_data_tot$tot == max(quercus_auc_data_tot$tot), "subset"],"_", quercus_auc_data_tot[quercus_auc_data_tot$tot == max(quercus_auc_data_tot$tot), "rep"])
best_fit <- read_mean_outputvalue_phenofit(file.path(file.path(sim_folder, "CMAES"), phenofit_quercus_best_rep, "Fitness.txt"), points = ERA5_points)
phenofit_quercus_best_fit <- as.data.frame(cbind(ERA5_points$lat, ERA5_points$lon,t(best_fit)))
colnames(phenofit_quercus_best_fit) <- c("lat", "lon", "fitness")
# best threshold
best_threshold <- quercus_auc_data_tot[quercus_auc_data_tot$tot == max(quercus_auc_data_tot$tot), "best_threshold"]
phenofit_quercus_best_fit[phenofit_quercus_best_fit$fitness < best_threshold, 'presence'] <- 0
phenofit_quercus_best_fit[phenofit_quercus_best_fit$fitness >= best_threshold, 'presence'] <- 1
phenofit_quercus_map <- ggplot(data=phenofit_quercus_best_fit, aes(x = lon, y = lat)) + 
  geom_raster(aes(fill = presence), color = NA) +
  theme_void() +
  labs(fill = "Fitness") +
  ylab("") +
  xlab("") +
  scale_fill_gradient(low = lcol, high = hcol, limits = c(0,1), breaks = c(0,0.5,1), na.value = "#B5BEC6") +
  guides(fill = guide_colorbar(title.position = "top", direction = "horizontal", frame.colour = "black", frame.linewidth = 0.5,
                               ticks = FALSE)) +
  theme(legend.title.align = 0.5, legend.position = "bottom", 
        legend.key.height = unit(0.2, 'cm'), legend.key.width = unit(0.5, 'cm'),
        legend.title = element_text(family = "Linux Libertine G", size = 9),
        legend.text = element_text(family = "Linux Libertine G")) +
  annotate("text", x = -4, y = 65, label = paste("AUC =", round(max(quercus_auc_data_tot$tot), 2)), colour = 'darkgrey', size = 3.6, 
           family = "Linux Libertine G", fontface = "bold")

sim_folder <- 'D:/simulations/phenofit/backward/abies_alba/paper_data'
phenofit_abies_best_rep <- paste0(abies_auc_data_tot[abies_auc_data_tot$tot == max(abies_auc_data_tot$tot), "subset"],"_", abies_auc_data_tot[abies_auc_data_tot$tot == max(abies_auc_data_tot$tot), "rep"])
best_fit <- read_mean_outputvalue_phenofit(file.path(file.path(sim_folder, "CMAES"), phenofit_abies_best_rep, "Fitness.txt"), points = ERA5_points)
phenofit_abies_best_fit <- as.data.frame(cbind(ERA5_points$lat, ERA5_points$lon,t(best_fit)))
colnames(phenofit_abies_best_fit) <- c("lat", "lon", "fitness")
# best threshold
best_threshold <- abies_auc_data_tot[abies_auc_data_tot$tot == max(abies_auc_data_tot$tot), "best_threshold"]
phenofit_abies_best_fit[phenofit_abies_best_fit$fitness < best_threshold, 'presence'] <- 0
phenofit_abies_best_fit[phenofit_abies_best_fit$fitness >= best_threshold, 'presence'] <- 1
phenofit_abies_map <- ggplot(data=phenofit_abies_best_fit, aes(x = lon, y = lat)) + 
  geom_raster(aes(fill = presence), color = NA) +
  theme_void() +
  labs(fill = "Fitness") +
  ylab("") +
  xlab("") +
  scale_fill_gradient(low = lcol, high = hcol, limits = c(0,1), breaks = c(0,0.5,1), na.value = "#B5BEC6") +
  guides(fill = guide_colorbar(title.position = "top", direction = "horizontal", frame.colour = "black", frame.linewidth = 0.5,
                               ticks = FALSE)) +
  theme(legend.title.align = 0.5, legend.position = "bottom", 
        legend.key.height = unit(0.2, 'cm'), legend.key.width = unit(0.5, 'cm'),
        legend.title = element_text(family = "Linux Libertine G", size = 9),
        legend.text = element_text(family = "Linux Libertine G")) +
  annotate("text", x = -4, y = 65, label = paste("AUC =", round(max(abies_auc_data_tot$tot), 2)), colour = 'darkgrey', size = 3.6, 
           family = "Linux Libertine G", fontface = "bold")

```

```{r castanea_fitness_maps}

sim_folder <- "D:/simulations/castanea/backward"

if(load_castanea_backward_values){
  backward_fit <- read_mean_outputvalue_castanea(ERA5_grid_points$points, lambda =1, file.path(sim_folder, "fagus_sylvatica/paper_data/CMAES/subset1_rep2"), var = "BiomassOfReserves")
  backward_fit <- t(backward_fit)
  backward_fit[backward_fit < 0 ] <- 0 # negative values in reserve
  castanea_fagus_best_fit <- as.data.frame(cbind(ERA5_grid_points$lat, ERA5_grid_points$lon,backward_fit))
  colnames(castanea_fagus_best_fit) <- c("lat", "lon", "biomass_reserves")
  saveRDS(castanea_fagus_best_fit, file = file.path(wd, "cache", "castanea_fagus_biomassreserves_bestfit.rds"))
}else{
  castanea_fagus_best_fit <- readRDS(file = file.path(wd, "cache", "castanea_fagus_biomassreserves_bestfit.rds"))
}
# auc
presabs_fit <- left_join(fsyl_presabs, castanea_fagus_best_fit)
castanea_fagus_best_auc <- auc(roc(presabs_fit$biomass_reserves, as.factor(fsyl_presabs$pres)))
# best threshold
youden_index <- sensitivity(presabs_fit$biomass_reserves, as.factor(fsyl_presabs$pres), perc.rank = F)$measure +
  specificity(presabs_fit$biomass_reserves, as.factor(fsyl_presabs$pres), perc.rank = F)$measure - 1
thresholds <- sensitivity(presabs_fit$biomass_reserves, as.factor(fsyl_presabs$pres), perc.rank = F)$cutoffs
castanea_fagus_backward_best_threshold <- thresholds[which(youden_index == max(youden_index))]
castanea_fagus_best_fit[is.na(castanea_fagus_best_fit$biomass_reserves), "biomass_reserves"] <- -9999 # NA values impossible for threshold
castanea_fagus_best_fit[castanea_fagus_best_fit$biomass_reserves < castanea_fagus_backward_best_threshold, 'presence'] <- 0
castanea_fagus_best_fit[castanea_fagus_best_fit$biomass_reserves >= castanea_fagus_backward_best_threshold, 'presence'] <- 1
castanea_fagus_best_fit[castanea_fagus_best_fit$biomass_reserves == -9999, "biomass_reserves"] <- NA
castanea_fagus_best_fit[is.na(castanea_fagus_best_fit$biomass_reserves), 'presence'] <- NA
castanea_backward_fagus_map <- ggplot(data=castanea_fagus_best_fit, aes(x = lon, y = lat)) + 
  geom_raster(aes(fill = presence), color = NA) +
  theme_void() +
  labs(fill = "Presence") +
  ylab("") +
  xlab("") +
  scale_fill_gradient(low = lcol, high = hcol, limits = c(0,1), breaks = c(0,0.5,1), na.value = "#EAEAEA") +
  guides(fill = guide_colorbar(title.position = "top", direction = "horizontal", frame.colour = "black", frame.linewidth = 0.5,
                               ticks = FALSE)) +
  theme(legend.title.align = 0.5, legend.position = "bottom", 
        legend.key.height = unit(0.2, 'cm'), legend.key.width = unit(0.5, 'cm'),
        legend.title = element_text(family = "Linux Libertine G", size = 9),
        legend.text = element_text(family = "Linux Libertine G")) +
  annotate("text", x = -4, y = 65, label = paste("AUC =", round(castanea_fagus_best_auc, 2)), colour = 'darkgrey', size = 3.6, 
           family = "Linux Libertine G", fontface = "bold")

if(load_castanea_backward_values){
  backward_fit <- read_mean_outputvalue_castanea(ERA5_grid_points$points, lambda =1, file.path(sim_folder, "abies_alba/paper_data/CMAES/subset1_rep1"), var = "BiomassOfReserves")
  backward_fit <- t(backward_fit)
  backward_fit[backward_fit < 0 ] <- 0 # negative values in reserve
  castanea_abies_best_fit <- as.data.frame(cbind(ERA5_grid_points$lat, ERA5_grid_points$lon,backward_fit))
  colnames(castanea_abies_best_fit) <- c("lat", "lon", "biomass_reserves")
  saveRDS(castanea_abies_best_fit, file = file.path(wd, "cache", "castanea_abies_biomassreserves_bestfit.rds"))
}else{
  castanea_abies_best_fit <- readRDS(file = file.path(wd, "cache", "castanea_abies_biomassreserves_bestfit.rds"))
}
# auc
presabs_fit <- left_join(aalb_presabs, castanea_abies_best_fit)
castanea_abies_best_auc <- auc(roc(presabs_fit$biomass_reserves, as.factor(aalb_presabs$pres)))
# best threshold
youden_index <- sensitivity(presabs_fit$biomass_reserves, as.factor(aalb_presabs$pres), perc.rank = F)$measure +
  specificity(presabs_fit$biomass_reserves, as.factor(aalb_presabs$pres), perc.rank = F)$measure - 1
thresholds <- sensitivity(presabs_fit$biomass_reserves, as.factor(aalb_presabs$pres), perc.rank = F)$cutoffs
castanea_abies_backward_best_threshold <- thresholds[which(youden_index == max(youden_index))]
castanea_abies_best_fit[is.na(castanea_abies_best_fit$biomass_reserves), "biomass_reserves"] <- -9999 # NA values impossible for threshold
castanea_abies_best_fit[castanea_abies_best_fit$biomass_reserves < castanea_abies_backward_best_threshold, 'presence'] <- 0
castanea_abies_best_fit[castanea_abies_best_fit$biomass_reserves >= castanea_abies_backward_best_threshold, 'presence'] <- 1
castanea_abies_best_fit[castanea_abies_best_fit$biomass_reserves == -9999, "biomass_reserves"] <- NA
castanea_abies_best_fit[is.na(castanea_abies_best_fit$biomass_reserves), 'presence'] <- NA
castanea_backward_abies_map <- ggplot(data=castanea_abies_best_fit, aes(x = lon, y = lat)) + 
  geom_raster(aes(fill = presence), color = NA) +
  theme_void() +
  labs(fill = "Presence") +
  ylab("") +
  xlab("") +
  scale_fill_gradient(low = lcol, high = hcol, limits = c(0,1), breaks = c(0,0.5,1), na.value = "#EAEAEA") +
  guides(fill = guide_colorbar(title.position = "top", direction = "horizontal", frame.colour = "black", frame.linewidth = 0.5,
                               ticks = FALSE)) +
  theme(legend.title.align = 0.5, legend.position = "bottom", 
        legend.key.height = unit(0.2, 'cm'), legend.key.width = unit(0.5, 'cm'),
        legend.title = element_text(family = "Linux Libertine G", size = 9),
        legend.text = element_text(family = "Linux Libertine G")) +
  annotate("text", x = -4, y = 65, label = paste("AUC =", round(castanea_abies_best_auc, 2)), colour = 'darkgrey', size = 3.6, 
           family = "Linux Libertine G", fontface = "bold")

if(load_castanea_backward_values){
  backward_fit <- read_mean_outputvalue_castanea(ERA5_grid_points$points, lambda =1, file.path(sim_folder, "quercus_ilex/paper_data/CMAES/subset1_rep1"), var = "BiomassOfReserves")
  backward_fit <- t(backward_fit)
  backward_fit[backward_fit < 0 ] <- 0 # negative values in reserve
  castanea_quercus_best_fit <- as.data.frame(cbind(ERA5_grid_points$lat, ERA5_grid_points$lon,backward_fit))
  colnames(castanea_quercus_best_fit) <- c("lat", "lon", "biomass_reserves")
  saveRDS(castanea_quercus_best_fit, file = file.path(wd, "cache", "castanea_quercus_biomassreserves_bestfit.rds"))
}else{
  castanea_quercus_best_fit <- readRDS(file = file.path(wd, "cache", "castanea_quercus_biomassreserves_bestfit.rds"))
}
# auc
presabs_fit <- left_join(qile_presabs, castanea_quercus_best_fit)
castanea_quercus_best_auc <- auc(roc(presabs_fit$biomass_reserves, as.factor(qile_presabs$pres)))
# best threshold
youden_index <- sensitivity(presabs_fit$biomass_reserves, as.factor(qile_presabs$pres), perc.rank = F)$measure +
  specificity(presabs_fit$biomass_reserves, as.factor(qile_presabs$pres), perc.rank = F)$measure - 1
thresholds <- sensitivity(presabs_fit$biomass_reserves, as.factor(qile_presabs$pres), perc.rank = F)$cutoffs
castanea_quercus_backward_best_threshold <- thresholds[which(youden_index == max(youden_index))]
castanea_quercus_best_fit[is.na(castanea_quercus_best_fit$biomass_reserves), "biomass_reserves"] <- -9999 # NA values impossible for threshold
castanea_quercus_best_fit[castanea_quercus_best_fit$biomass_reserves < castanea_quercus_backward_best_threshold, 'presence'] <- 0
castanea_quercus_best_fit[castanea_quercus_best_fit$biomass_reserves >= castanea_quercus_backward_best_threshold, 'presence'] <- 1
castanea_quercus_best_fit[castanea_quercus_best_fit$biomass_reserves == -9999, "biomass_reserves"] <- NA
castanea_quercus_best_fit[is.na(castanea_quercus_best_fit$biomass_reserves), 'presence'] <- NA
castanea_backward_quercus_map <- ggplot(data=castanea_quercus_best_fit, aes(x = lon, y = lat)) + 
  geom_raster(aes(fill = presence), color = NA) +
  theme_void() +
  labs(fill = "Presence") +
  ylab("") +
  xlab("") +
  scale_fill_gradient(low = lcol, high = hcol, limits = c(0,1), breaks = c(0,0.5,1), na.value = "#EAEAEA") +
  guides(fill = guide_colorbar(title.position = "top", direction = "horizontal", frame.colour = "black", frame.linewidth = 0.5,
                               ticks = FALSE)) +
  theme(legend.title.align = 0.5, legend.position = "bottom", 
        legend.key.height = unit(0.2, 'cm'), legend.key.width = unit(0.5, 'cm'),
        legend.title = element_text(family = "Linux Libertine G", size = 9),
        legend.text = element_text(family = "Linux Libertine G")) +
  annotate("text", x = -4, y = 65, label = paste("AUC =", round(castanea_quercus_best_auc, 2)), colour = 'darkgrey', size = 3.6, 
           family = "Linux Libertine G", fontface = "bold")

false_fit <- castanea_fagus_best_fit
false_fit$biomass_reserves <- 0
castanea_false_map <- ggplot(data=false_fit, aes(x = lon, y = lat)) + 
  geom_raster(aes(fill = biomass_reserves), color = NA) +
  theme_void() +
  labs(fill = "Carbon reserves") +
  ylab("") +
  xlab("") +
  scale_fill_gradient(low = lcol, high = hcol, limits = c(0,1), breaks = c(0,0.5,1)) +
  guides(fill = guide_colorbar(title.position = "top", direction = "horizontal", frame.colour = "black", frame.linewidth = 0.5,
                               ticks = FALSE)) +
  theme(legend.title.align = 0.5, legend.position = "bottom", 
        legend.key.height = unit(0.2, 'cm'), legend.key.width = unit(0.5, 'cm'),
        legend.title = element_text(family = "Linux Libertine G", size = 9),
        legend.text = element_text(family = "Linux Libertine G")) +
  annotate("text", x = -4, y = 65, label = paste("AUC =", round(0, 2)), colour = 'darkgrey', size = 3.6, 
           family = "Linux Libertine G", fontface = "bold")

```

```{r species_distribution_maps}

fsyl_presabs_europe <- right_join(fsyl_presabs, castanea_fagus_best_fit[, c("lon", "lat")])
fsyl_presabs_europe[is.na(fsyl_presabs_europe)] <- 0

fagus_dist_map <- ggplot() +
  geom_raster(data = fsyl_presabs_europe, aes(x = lon, y = lat, fill = pres), color = NA) +
  scale_fill_gradient(low = lcol, high = hcol, limits = c(0,1), breaks = c(0,0.5,1)) +
  theme_void() +
  ylab("") +
  xlab("") +
  guides(fill = guide_colorbar(title.position = "top", direction = "horizontal", frame.colour = "black", frame.linewidth = 0.5,
                               ticks = FALSE)) +
  theme(legend.title.align = 0.5, legend.position = "bottom", 
        legend.key.height = unit(0.2, 'cm'), legend.key.width = unit(0.5, 'cm'),
        legend.title = element_text(family = "Linux Libertine G", size = 9),
        legend.text = element_text(family = "Linux Libertine G"))

qile_presabs_europe <- right_join(qile_presabs, castanea_fagus_best_fit[, c("lon", "lat")])
qile_presabs_europe[is.na(qile_presabs_europe)] <- 0

quercus_dist_map <- ggplot() +
  geom_raster(data = qile_presabs_europe, aes(x = lon, y = lat, fill = pres), color = NA) +
  scale_fill_gradient(low = lcol, high = hcol, limits = c(0,1), breaks = c(0,0.5,1)) +
  theme_void() +
  ylab("") +
  xlab("") +
  guides(fill = guide_colorbar(title.position = "top", direction = "horizontal", frame.colour = "black", frame.linewidth = 0.5,
                               ticks = FALSE)) +
  theme(legend.title.align = 0.5, legend.position = "bottom", 
        legend.key.height = unit(0.2, 'cm'), legend.key.width = unit(0.5, 'cm'),
        legend.title = element_text(family = "Linux Libertine G", size = 9),
        legend.text = element_text(family = "Linux Libertine G"))

aalb_presabs_europe <- right_join(aalb_presabs, castanea_fagus_best_fit[, c("lon", "lat")])
aalb_presabs_europe[is.na(aalb_presabs_europe)] <- 0

abies_dist_map <- ggplot() +
  geom_raster(data = aalb_presabs_europe, aes(x = lon, y = lat, fill = pres), color = NA) +
  scale_fill_gradient(low = lcol, high = hcol, limits = c(0,1), breaks = c(0,0.5,1)) +
  theme_void() +
  ylab("") +
  xlab("") +
  guides(fill = guide_colorbar(title.position = "top", direction = "horizontal", frame.colour = "black", frame.linewidth = 0.5,
                               ticks = FALSE)) +
  theme(legend.title.align = 0.5, legend.position = "bottom", 
        legend.key.height = unit(0.2, 'cm'), legend.key.width = unit(0.5, 'cm'),
        legend.title = element_text(family = "Linux Libertine G", size = 9),
        legend.text = element_text(family = "Linux Libertine G"))

```

```{r phenofit_forward_fitness_maps}

forward_sim_folder <- "D:/simulations/phenofit/forward"

forward_fit <- read_mean_outputvalue_phenofit(file.path(forward_sim_folder, "fagus_sylvatica/VVanderMeersch", "Fitness.txt"), points = ERA5_points)
#auc
fitness_presabs <- read_mean_outputvalue_phenofit(file.path(forward_sim_folder, "fagus_sylvatica/VVanderMeersch", "Fitness.txt"), 
                                          points = fsyl_presabs) # only on pres/abs points
phenofit_fagus_forward_auc <- auc(roc(fitness_presabs, as.factor(fsyl_presabs$pres)))
# best threshold
youden_index <- sensitivity(fitness_presabs, as.factor(fsyl_presabs$pres), perc.rank = F)$measure +
  specificity(fitness_presabs, as.factor(fsyl_presabs$pres), perc.rank = F)$measure - 1
thresholds <- sensitivity(fitness_presabs, as.factor(fsyl_presabs$pres), perc.rank = F)$cutoffs
phenofit_fagus_forward_best_threshold <- thresholds[which(youden_index == max(youden_index))]

phenofit_fagus_forward_fit <- as.data.frame(cbind(ERA5_points$lat, ERA5_points$lon,t(forward_fit)))
colnames(phenofit_fagus_forward_fit) <- c("lat", "lon", "fitness")
phenofit_fagus_forward_fit[phenofit_fagus_forward_fit$fitness < phenofit_fagus_forward_best_threshold, 'presence'] <- 0
phenofit_fagus_forward_fit[phenofit_fagus_forward_fit$fitness >= phenofit_fagus_forward_best_threshold, 'presence'] <- 1

phenofit_forward_fagus_map <- ggplot(data=phenofit_fagus_forward_fit, aes(x = lon, y = lat)) + 
  geom_raster(aes(fill = presence), color = NA) +
  theme_void() +
  labs(fill = "Presence") +
  ylab("") +
  xlab("") +
  scale_fill_gradient(low = lcol, high = hcol, limits = c(0,1), breaks = c(0,0.5,1)) +
  guides(fill = guide_colorbar(title.position = "top", direction = "horizontal", frame.colour = "black", frame.linewidth = 0.5,
                               ticks = FALSE)) +
  theme(legend.title.align = 0.5, legend.position = "bottom", 
        legend.key.height = unit(0.2, 'cm'), legend.key.width = unit(0.5, 'cm'),
        legend.title = element_text(family = "Linux Libertine G", size = 9),
        legend.text = element_text(family = "Linux Libertine G")) +
  annotate("text", x = -4, y = 65, label = paste("AUC =", round(phenofit_fagus_forward_auc, 2)), colour = 'darkgrey', size = 3.6, 
           family = "Linux Libertine G", fontface = "bold")

forward_fit <- read_mean_outputvalue_phenofit(file.path(forward_sim_folder, "abies_alba/VVanderMeersch2", "Fitness.txt"), points = ERA5_points)
# auc
fitness_presabs <- read_mean_outputvalue_phenofit(file.path(forward_sim_folder, "abies_alba/VVanderMeersch2", "Fitness.txt"), 
                                          points = aalb_presabs) # only on pres/abs points
phenofit_abies_forward_auc <- auc(roc(fitness_presabs, as.factor(aalb_presabs$pres)))
# best threshold
youden_index <- sensitivity(fitness_presabs, as.factor(aalb_presabs$pres), perc.rank = F)$measure +
  specificity(fitness_presabs, as.factor(aalb_presabs$pres), perc.rank = F)$measure - 1
thresholds <- sensitivity(fitness_presabs, as.factor(aalb_presabs$pres), perc.rank = F)$cutoffs
phenofit_abies_forward_best_threshold <- thresholds[which(youden_index == max(youden_index))]

phenofit_abies_forward_fit <- as.data.frame(cbind(ERA5_points$lat, ERA5_points$lon,t(forward_fit)))
colnames(phenofit_abies_forward_fit) <- c("lat", "lon", "fitness")
phenofit_abies_forward_fit[phenofit_abies_forward_fit$fitness < phenofit_abies_forward_best_threshold, 'presence'] <- 0
phenofit_abies_forward_fit[phenofit_abies_forward_fit$fitness >= phenofit_abies_forward_best_threshold, 'presence'] <- 1

phenofit_forward_abies_map <- ggplot(data=phenofit_abies_forward_fit, aes(x = lon, y = lat)) + 
  geom_raster(aes(fill = presence), color = NA) +
  theme_void() +
  labs(fill = "Presence") +
  ylab("") +
  xlab("") +
  scale_fill_gradient(low = lcol, high = hcol, limits = c(0,1), breaks = c(0,0.5,1)) +
  guides(fill = guide_colorbar(title.position = "top", direction = "horizontal", frame.colour = "black", frame.linewidth = 0.5,
                               ticks = FALSE)) +
  theme(legend.title.align = 0.5, legend.position = "bottom", 
        legend.key.height = unit(0.2, 'cm'), legend.key.width = unit(0.5, 'cm'),
        legend.title = element_text(family = "Linux Libertine G", size = 9),
        legend.text = element_text(family = "Linux Libertine G")) +
  annotate("text", x = -4, y = 65, label = paste("AUC =", round(phenofit_abies_forward_auc, 2)), colour = 'darkgrey', size = 3.6, 
           family = "Linux Libertine G", fontface = "bold")

forward_fit <- read_mean_outputvalue_phenofit(file.path(forward_sim_folder, "quercus_ilex/FTauc/modcode", "Fitness.txt"), points = ERA5_points)
#auc
fitness_presabs <- read_mean_outputvalue_phenofit(file.path(forward_sim_folder, "quercus_ilex/FTauc/modcode", "Fitness.txt"), 
                                          points = qile_presabs) # only on pres/abs points
phenofit_quercus_forward_auc <- auc(roc(fitness_presabs, as.factor(qile_presabs$pres)))
# best threshold
youden_index <- sensitivity(fitness_presabs, as.factor(qile_presabs$pres), perc.rank = F)$measure +
  specificity(fitness_presabs, as.factor(qile_presabs$pres), perc.rank = F)$measure - 1
thresholds <- sensitivity(fitness_presabs, as.factor(qile_presabs$pres), perc.rank = F)$cutoffs
phenofit_quercus_forward_best_threshold <- thresholds[which(youden_index == max(youden_index))]

phenofit_quercus_forward_fit <- as.data.frame(cbind(ERA5_points$lat, ERA5_points$lon,t(forward_fit)))
colnames(phenofit_quercus_forward_fit) <- c("lat", "lon", "fitness")
phenofit_quercus_forward_fit[phenofit_quercus_forward_fit$fitness < phenofit_quercus_forward_best_threshold, 'presence'] <- 0
phenofit_quercus_forward_fit[phenofit_quercus_forward_fit$fitness >= phenofit_quercus_forward_best_threshold, 'presence'] <- 1

phenofit_forward_quercus_map <- ggplot(data=phenofit_quercus_forward_fit, aes(x = lon, y = lat)) + 
  geom_raster(aes(fill = presence), color = NA) +
  theme_void() +
  labs(fill = "Presence") +
  ylab("") +
  xlab("") +
  scale_fill_gradient(low = lcol, high = hcol, limits = c(0,1), breaks = c(0,0.5,1)) +
  guides(fill = guide_colorbar(title.position = "top", direction = "horizontal", frame.colour = "black", frame.linewidth = 0.5,
                               ticks = FALSE)) +
  theme(legend.title.align = 0.5, legend.position = "bottom", 
        legend.key.height = unit(0.2, 'cm'), legend.key.width = unit(0.5, 'cm'),
        legend.title = element_text(family = "Linux Libertine G", size = 9),
        legend.text = element_text(family = "Linux Libertine G")) +
  annotate("text", x = -4, y = 65, label = paste("AUC =", round(phenofit_quercus_forward_auc, 2)), colour = 'darkgrey', size = 3.6, 
           family = "Linux Libertine G", fontface = "bold")

```

```{r castanea_forward_fitness_maps, eval = T}

forward_sim_folder <- "D:/simulations/castanea/forward"

if(load_castanea_forward_values){
  forward_fit <- read_mean_outputvalue_castanea(ERA5_grid_points$points, lambda =1, file.path(forward_sim_folder, "fagus_sylvatica_capsisrevision18131"), var = "BiomassOfReserves")
  forward_fit <- t(forward_fit)
  forward_fit[forward_fit < 0 ] <- 0 # negative values in reserve
  castanea_fagus_forward_fit <- as.data.frame(cbind(ERA5_grid_points$lat, ERA5_grid_points$lon,forward_fit))
  colnames(castanea_fagus_forward_fit) <- c("lat", "lon", "biomass_reserves")
  saveRDS(castanea_fagus_forward_fit, file = file.path(wd, "cache", "castanea_fagus_biomassreserves_fit.rds"))
}else{
  castanea_fagus_forward_fit <- readRDS(file = file.path(wd, "cache", "castanea_fagus_biomassreserves_fit.rds"))
}
# auc
presabs_fit <- left_join(fsyl_presabs, castanea_fagus_forward_fit)
castanea_fagus_forward_auc <- auc(roc(presabs_fit$biomass_reserves, as.factor(fsyl_presabs$pres)))
# best threshold
youden_index <- sensitivity(presabs_fit$biomass_reserves, as.factor(fsyl_presabs$pres), perc.rank = F)$measure +
  specificity(presabs_fit$biomass_reserves, as.factor(fsyl_presabs$pres), perc.rank = F)$measure - 1
thresholds <- sensitivity(presabs_fit$biomass_reserves, as.factor(fsyl_presabs$pres), perc.rank = F)$cutoffs
castanea_fagus_forward_best_threshold <- thresholds[which(youden_index == max(youden_index))]
castanea_fagus_forward_fit[is.na(castanea_fagus_forward_fit$biomass_reserves), "biomass_reserves"] <- -9999 # NA values impossible for threshold
castanea_fagus_forward_fit[castanea_fagus_forward_fit$biomass_reserves < castanea_fagus_forward_best_threshold, 'presence'] <- 0
castanea_fagus_forward_fit[castanea_fagus_forward_fit$biomass_reserves >= castanea_fagus_forward_best_threshold, 'presence'] <- 1
castanea_fagus_forward_fit[castanea_fagus_forward_fit$biomass_reserves == -9999, "biomass_reserves"] <- NA
castanea_fagus_forward_fit[is.na(castanea_fagus_forward_fit$biomass_reserves), 'presence'] <- NA
castanea_forward_fagus_map <- ggplot(data=castanea_fagus_forward_fit, aes(x = lon, y = lat)) + 
  geom_raster(aes(fill = presence), color = NA) +
  theme_void() +
  labs(fill = "Presence") +
  ylab("") +
  xlab("") +
  scale_fill_gradient(low = lcol, high = hcol, limits = c(0,1), breaks = c(0,0.5,1), na.value = "#EAEAEA") +
  guides(fill = guide_colorbar(title.position = "top", direction = "horizontal", frame.colour = "black", frame.linewidth = 0.5,
                               ticks = FALSE)) +
  theme(legend.title.align = 0.5, legend.position = "bottom", 
        legend.key.height = unit(0.2, 'cm'), legend.key.width = unit(0.5, 'cm'),
        legend.title = element_text(family = "Linux Libertine G", size = 9),
        legend.text = element_text(family = "Linux Libertine G")) +
  annotate("text", x = -4, y = 65, label = paste("AUC =", round(castanea_fagus_forward_auc, 2)), colour = 'darkgrey', size = 3.6, 
           family = "Linux Libertine G", fontface = "bold")

if(load_castanea_forward_values){
  forward_fit <- read_mean_outputvalue_castanea(ERA5_grid_points$points, lambda =1, file.path(forward_sim_folder, "quercus_ilex_capsisrevision17568"), var = "BiomassOfReserves")
  forward_fit <- t(forward_fit)
  forward_fit[forward_fit < 0 ] <- 0 # negative values in reserve
  castanea_quercus_forward_fit <- as.data.frame(cbind(ERA5_grid_points$lat, ERA5_grid_points$lon,forward_fit))
  colnames(castanea_quercus_forward_fit) <- c("lat", "lon", "biomass_reserves")
  saveRDS(castanea_quercus_forward_fit, file = file.path(wd, "cache", "castanea_quercus_biomassreserves_fit.rds"))
}else{
  castanea_quercus_forward_fit <- readRDS(file = file.path(wd, "cache", "castanea_quercus_biomassreserves_fit.rds"))
}
# auc
presabs_fit <- left_join(qile_presabs, castanea_quercus_forward_fit)
castanea_quercus_forward_auc <- auc(roc(presabs_fit$biomass_reserves, as.factor(qile_presabs$pres)))
# best threshold
youden_index <- sensitivity(presabs_fit$biomass_reserves, as.factor(qile_presabs$pres), perc.rank = F)$measure +
  specificity(presabs_fit$biomass_reserves, as.factor(qile_presabs$pres), perc.rank = F)$measure - 1
thresholds <- sensitivity(presabs_fit$biomass_reserves, as.factor(qile_presabs$pres), perc.rank = F)$cutoffs
castanea_quercus_forward_best_threshold <- thresholds[which(youden_index == max(youden_index))]
castanea_quercus_forward_fit[is.na(castanea_quercus_forward_fit$biomass_reserves), "biomass_reserves"] <- -9999 # NA values impossible for threshold
castanea_quercus_forward_fit[castanea_quercus_forward_fit$biomass_reserves < castanea_quercus_forward_best_threshold, 'presence'] <- 0
castanea_quercus_forward_fit[castanea_quercus_forward_fit$biomass_reserves >= castanea_quercus_forward_best_threshold, 'presence'] <- 1
castanea_quercus_forward_fit[castanea_quercus_forward_fit$biomass_reserves == -9999, "biomass_reserves"] <- NA
castanea_quercus_forward_fit[is.na(castanea_quercus_forward_fit$biomass_reserves), 'presence'] <- NA
castanea_forward_quercus_map <- ggplot(data=castanea_quercus_forward_fit, aes(x = lon, y = lat)) + 
  geom_raster(aes(fill = presence)) +
  theme_void() +
  labs(fill = "Presence") +
  ylab("") +
  xlab("") +
  scale_fill_gradient(low = lcol, high = hcol, limits = c(0,1), breaks = c(0,0.5,1), na.value = "#EAEAEA") +
  guides(fill = guide_colorbar(title.position = "top", direction = "horizontal", frame.colour = "black", frame.linewidth = 0.5,
                               ticks = FALSE)) +
  theme(legend.title.align = 0.5, legend.position = "bottom", 
        legend.key.height = unit(0.2, 'cm'), legend.key.width = unit(0.5, 'cm'),
        legend.title = element_text(family = "Linux Libertine G", size = 9),
        legend.text = element_text(family = "Linux Libertine G")) +
  annotate("text", x = -4, y = 65, label = paste("AUC =", round(castanea_quercus_forward_auc, 2)), color = 'darkgrey', size = 3.6, 
           family = "Linux Libertine G", fontface = "bold")

if(load_castanea_forward_values){
  forward_fit <- read_mean_outputvalue_castanea(ERA5_grid_points$points, lambda =1, file.path(forward_sim_folder, "abies_alba_capsisrevision17568"), var = "BiomassOfReserves")
  forward_fit <- t(forward_fit)
  forward_fit[forward_fit < 0 ] <- 0 # negative values in reserve
  castanea_abies_forward_fit <- as.data.frame(cbind(ERA5_grid_points$lat, ERA5_grid_points$lon,forward_fit))
  colnames(castanea_abies_forward_fit) <- c("lat", "lon", "biomass_reserves")
  saveRDS(castanea_abies_forward_fit, file = file.path(wd, "cache", "castanea_abies_biomassreserves_fit.rds"))
}else{
  castanea_abies_forward_fit <- readRDS(file = file.path(wd, "cache", "castanea_abies_biomassreserves_fit.rds"))
}
# auc
presabs_fit <- left_join(aalb_presabs, castanea_abies_forward_fit)
castanea_abies_forward_auc <- auc(roc(presabs_fit$biomass_reserves, as.factor(aalb_presabs$pres)))
# best threshold
youden_index <- sensitivity(presabs_fit$biomass_reserves, as.factor(aalb_presabs$pres), perc.rank = F)$measure +
  specificity(presabs_fit$biomass_reserves, as.factor(aalb_presabs$pres), perc.rank = F)$measure - 1
thresholds <- sensitivity(presabs_fit$biomass_reserves, as.factor(aalb_presabs$pres), perc.rank = F)$cutoffs
castanea_abies_forward_best_threshold <- thresholds[which(youden_index == max(youden_index))]
castanea_abies_forward_fit[is.na(castanea_abies_forward_fit$biomass_reserves), "biomass_reserves"] <- -9999 # NA values impossible for threshold
castanea_abies_forward_fit[castanea_abies_forward_fit$biomass_reserves < castanea_abies_forward_best_threshold, 'presence'] <- 0
castanea_abies_forward_fit[castanea_abies_forward_fit$biomass_reserves >= castanea_abies_forward_best_threshold, 'presence'] <- 1
castanea_abies_forward_fit[castanea_abies_forward_fit$biomass_reserves == -9999, "biomass_reserves"] <- NA
castanea_abies_forward_fit[is.na(castanea_abies_forward_fit$biomass_reserves), 'presence'] <- NA
castanea_forward_abies_map <- ggplot(data=castanea_abies_forward_fit, aes(x = lon, y = lat)) + 
  geom_raster(aes(fill = presence), color = NA) +
  theme_void() +
  labs(fill = "Presence") +
  ylab("") +
  xlab("") +
  scale_fill_gradient(low = lcol, high = hcol, limits = c(0,1), breaks = c(0,0.5,1), na.value = "#EAEAEA") +
  guides(fill = guide_colorbar(title.position = "top", direction = "horizontal", frame.colour = "black", frame.linewidth = 0.5,
                               ticks = FALSE)) +
  theme(legend.title.align = 0.5, legend.position = "bottom", 
        legend.key.height = unit(0.2, 'cm'), legend.key.width = unit(0.5, 'cm'),
        legend.title = element_text(family = "Linux Libertine G", size = 9),
        legend.text = element_text(family = "Linux Libertine G")) +
  annotate("text", x = -4, y = 65, label = paste("AUC =", round(castanea_abies_forward_auc, 2)), colour = 'darkgrey', size = 3.6, 
           family = "Linux Libertine G", fontface = "bold")

```

Calibrations using species distribution data are hereafter called inverse calibrations, and calibrations based on expert knowledge, observations and measurements of the processes modelled are called expert calibrations.

CMA-ES calibration of PHENOFIT model allows an average `r round(((max(fagus_auc_data_tot$tot)-phenofit_fagus_forward_auc)/phenofit_fagus_forward_auc + (max(quercus_auc_data_tot$tot)-phenofit_quercus_forward_auc)/phenofit_quercus_forward_auc + (max(abies_auc_data_tot$tot)-phenofit_abies_forward_auc)/phenofit_abies_forward_auc)/3*100,1)`% increase of AUC across the three species compared to expert calibration (\hyperref[fig:phenofitmaps]{Figure 1}). The maximum increase is obtained for silver fir, from `r round(phenofit_abies_forward_auc,2)` to `r round(max(abies_auc_data_tot$tot),2)` (`r round((max(abies_auc_data_tot$tot)-phenofit_abies_forward_auc)/phenofit_abies_forward_auc*100,1)`%).  
CMA-ES calibration of CASTANEA allows an average `r round(((castanea_fagus_best_auc-castanea_fagus_forward_auc)/castanea_fagus_forward_auc + (castanea_quercus_best_auc-castanea_quercus_forward_auc)/castanea_quercus_forward_auc + (castanea_abies_best_auc-castanea_abies_forward_auc)/castanea_abies_forward_auc)/3*100,1)` % increase of AUC compared to expert calibration (\hyperref[fig:castaneamaps]{Figure 2}), and a maximum increase obtained for holm oak (`r round((castanea_quercus_best_auc-castanea_quercus_forward_auc)/castanea_quercus_forward_auc*100,1)`%).

```{r phenofitmaps, include=TRUE, fig.height = 7, fig.width = 7, fig.align = "center", fig.cap = "Species distribution maps obtained with PHENOFIT expert and inverse calibrations, compared with observed species occurrences. Optimal threshold to dichotomize model predicted fitness index in presence/absence is the Youden index-based cut-off point. Note that models predict species climatic niche which is larger than the realized niche that corresponds to species presence map.", fig.pos="H", fig.showtext=TRUE, eval = T}

p <- plot_grid(
  "",
  plot_grid(
    "",
    phenofit_forward_fagus_map + theme(legend.position = "none"),
    fagus_dist_map + theme(legend.position = "none"),
    phenofit_fagus_map + theme(legend.position = "none"), 
    "",
    phenofit_forward_quercus_map + theme(legend.position = "none"),
    quercus_dist_map + theme(legend.position = "none"),
    phenofit_quercus_map + theme(legend.position = "none"),
    "",
    phenofit_forward_abies_map + theme(legend.position = "none"),
    abies_dist_map + theme(legend.position = "none"),
    phenofit_abies_map + theme(legend.position = "none"), 
    "",
    ncol= 4, rel_widths = c(1, 5, 5, 5),
    nrow= 3, rel_heights = c(5, 5, 5)),
  ncol = 1,
  rel_heights = c(2,20))

ggdraw(p) +
  draw_label("Expert calibration", color = "#495057", size = 10, angle = 0, y = 0.92, x = 0.21,
             fontfamily = "Linux Libertine G", fontface = "italic") +
  draw_label("Species occurrence", color = "#495057", size = 10, angle = 0, y = 0.92, x = 0.53,
             fontfamily = "Linux Libertine G", fontface = "italic") +
  draw_label("Inverse calibration", color = "#495057", size = 10, angle = 0, y = 0.92, x = 0.84,
             fontfamily = "Linux Libertine G", fontface = "italic") +
  draw_label("Fagus sylvatica", color = "#495057", size = 10, angle = 90, y = 0.75, x = 0.04,
             fontfamily = "Linux Libertine G", fontface = "italic") +
  draw_label("Quercus ilex", color = "#495057", size = 10, angle = 90, y = 0.44, x = 0.04,
             fontfamily = "Linux Libertine G", fontface = "italic") +
  draw_label("Abies alba", color = "#495057", size = 10, angle = 90, y = 0.14, x = 0.04,
             fontfamily = "Linux Libertine G", fontface = "italic")

```

```{r castaneamaps, include=TRUE, fig.height = 7, fig.width = 7, fig.align = "center", fig.cap = "Species distribution maps obtained with CASTANEA expert and inverse calibrations, compared with observed species occurrences. Optimal threshold to dichotomize model predicted carbon reserves in presence/absence is the Youden index-based cut-off point. Note that models predict species climatic niche which is larger than the realized niche that corresponds to species presence map. Note also that CASTANEA cannot be used in high-latitude regions (grey area).", fig.pos="H", fig.showtext=TRUE, eval = T}

p <- plot_grid(
  "",
  plot_grid(
    "",
    castanea_forward_fagus_map + theme(legend.position = "none"),
    fagus_dist_map + theme(legend.position = "none"),
    castanea_backward_fagus_map + theme(legend.position = "none"), 
    "",
    castanea_forward_quercus_map + theme(legend.position = "none"),
    quercus_dist_map + theme(legend.position = "none"),
    castanea_backward_quercus_map + theme(legend.position = "none"),
    "",
    castanea_forward_abies_map + theme(legend.position = "none"),
    abies_dist_map + theme(legend.position = "none"),
    castanea_backward_abies_map + theme(legend.position = "none"), 
    "",
    ncol= 4, rel_widths = c(1, 5, 5, 5),
    nrow= 3, rel_heights = c(5, 5, 5)),
  ncol = 1,
  rel_heights = c(2,20))

ggdraw(p) +
  draw_label("Expert calibration", color = "#495057", size = 10, angle = 0, y = 0.92, x = 0.21,
             fontfamily = "Linux Libertine G", fontface = "italic") +
  draw_label("Species occurrence", color = "#495057", size = 10, angle = 0, y = 0.92, x = 0.53,
             fontfamily = "Linux Libertine G", fontface = "italic") +
  draw_label("Inverse calibration", color = "#495057", size = 10, angle = 0, y = 0.92, x = 0.84,
             fontfamily = "Linux Libertine G", fontface = "italic") +
  draw_label("Fagus sylvatica", color = "#495057", size = 10, angle = 90, y = 0.75, x = 0.04,
             fontfamily = "Linux Libertine G", fontface = "italic") +
  draw_label("Quercus ilex", color = "#495057", size = 10, angle = 90, y = 0.44, x = 0.04,
             fontfamily = "Linux Libertine G", fontface = "italic") +
  draw_label("Abies alba", color = "#495057", size = 10, angle = 90, y = 0.14, x = 0.04,
             fontfamily = "Linux Libertine G", fontface = "italic")

```

## 2.2. Impacts of subsampling and calibration stochasticity

### 2.2.1. Variability of calibration performance

```{r cmaesrepAUCcal, include=TRUE, fig.height = 3, fig.width = 6, fig.align = "center", fig.cap = "\\label{fig:cmaesrepAUCcal} Effects of data sub-sampling and stochasticity on CMA-ES calibration using the PHENOFIT model for beech: \\textbf{(A)} calibration AUC (calculated only with calibration cells) and \\textbf{(B)} total AUC (calculated with all presence/absence cells). Each color is a different sub-sampling of occurrence data, each point is a calibration run. Diamonds (with black border) are mean AUC values. On \\textbf{(A)}, the grouping letters represent the multiple comparisons with pairwise Dunn’s tests.", fig.pos="H", fig.showtext=TRUE}

custom_colors = c("#9e0142", "#d53e4f", "#f46d43", "#fdae61", "#fee08b",
                               "#e6f598", "#abdda4", "#66c2a5", "#3288bd", "#5e4fa2")

positions <- c(rep(1,10), rep(1.1,10), rep(1.2,10), rep(1.3, 10), rep(1.4, 10), 
                   rep(1.5, 10), rep(1.6, 10), rep(1.7, 10), rep(1.8, 10), rep(1.9, 10))
fagus_auc_data_cal$positions <- positions

# Wilcoxon
# wilcoxon_test_cal <- compare_means(cal ~ positions,  data = fagus_auc_data_cal) 
# wilcoxon_test_s7 <- wilcoxon_test_cal %>% filter(p.signif != "ns") %>% filter(group1 == "1.7" | group2 == "1.7")
# list_comparisons <- lapply(1:nrow(wilcoxon_test_s7), function(i){
#   return(c(wilcoxon_test_s7$group1[i], wilcoxon_test_s7$group2[i]))})
# list_positions <- c(0.875, 0.9225, 0.9245, 0.917, 0.9145, 0.8785, 0.882, 0.894, 0.888)
# 
symnum.args <- list(cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 1), symbols = c("****", "***", "**", "*", "ns"))
# 
# p_auc_cal <-fagus_auc_data_cal %>% ggplot(aes(y = cal, x = as.character(positions), group = subset, 
#                     color = factor(subset, levels = paste0("subset", 1:10)), 
#                     fill = factor(subset, levels = paste0("subset", 1:10)))) +
#   geom_boxplot(width = 0.5, alpha = 0.3, outlier.shape = NA, fatten = 1, lwd=0.3) +
#   geom_beeswarm(alpha = 0.8, cex = 2.5, size = 0.5) +
#   stat_compare_means(comparisons = list_comparisons, symnum.args = symnum.args, 
#                      tip.length = c(-0.005, rep(0.005,4), rep(-0.005,4)),
#                      label.y = list_positions, size=2, vjust = 0.6, bracket.size = 0.2) + 
#   annotate("text", label = "Kruskal-Wallis", x = 1.65, y = 0.917, family = "Linux Libertine G", size = 2.5) +
#   annotate("text", label = paste0("p = ", compare_means(cal ~ positions,  
#                                                         data = fagus_auc_data_cal, method = "kruskal.test")$p.format), 
#            x = 1.65, y = 0.915, family = "Linux Libertine G", size = 3, fontface = "bold") +
#   scale_fill_manual(values = hcl.colors(10, palette = "Dynamic")) +
#   scale_color_manual(values = hcl.colors(10, palette = "Dynamic")) +
#   theme_minimal() +
#   theme(axis.text.x = element_blank(),
#         axis.title.x = element_blank(),
#         axis.text.y = element_text(family = "Linux Libertine G"),
#         axis.title.y = element_text(family = "Linux Libertine G", margin = margin(t = 0, r = 10, b = 0, l = 0)),
#         axis.ticks.x=element_blank(),
#         panel.grid.major = element_line(size = 0.2),
#         panel.grid.major.x = element_blank(),
#         panel.grid.minor.x = element_blank(),
#         panel.grid.minor.y = element_blank(),
#         legend.position="none") +
#   coord_cartesian(ylim = c(0.875, 0.925)) +
#   labs(y = "AUC")

# Dunn better after Kruskal
library(rstatix)
fagus_auc_data_cal$positions <- as.character(positions)
dunn.stat.test.all <- fagus_auc_data_cal %>% dunn_test(cal ~ positions, p.adjust.method = "holm")
dunn.stat.test <- fagus_auc_data_cal %>% dunn_test(cal ~ positions, p.adjust.method = "holm") %>% 
  dplyr::filter((group1 == "1.7" | group2 == "1.7") & p.adj.signif !="ns")

p_auc_cal <-  ggplot(data = fagus_auc_data_cal) +
  geom_boxplot(aes(y = cal, x = as.character(positions), group = subset, 
                   color = factor(subset, levels = paste0("subset", 1:10)), 
                   fill = factor(subset, levels = paste0("subset", 1:10))), 
               width = 0.5, alpha = 0.3, outlier.shape = NA, fatten = 1, lwd=0.3) +
  geom_beeswarm(aes(y = cal, x = as.character(positions), group = subset, 
                    color = factor(subset, levels = paste0("subset", 1:10)), 
                    fill = factor(subset, levels = paste0("subset", 1:10))),
                alpha = 0.8, cex = 2.5, size = 0.5) +
  stat_pvalue_manual(data = dunn.stat.test, hide.ns = FALSE, y.position = c(0.9205, 0.9225, 0.9245, 0.9165, 0.9185),
                     tip.length = 0.005, size=2, vjust = 0.6, bracket.size = 0.2) +
  annotate("text", label = "Kruskal-Wallis", x = 1.67, y = 0.917, family = "Linux Libertine G", size = 2.5) +
  annotate("text", label = paste0("p = ", compare_means(cal ~ positions,  
                                                        data = fagus_auc_data_cal, method = "kruskal.test")$p.format), 
           x = 1.67, y = 0.915, family = "Linux Libertine G", size = 3, fontface = "bold") +
  geom_point(data = fagus_auc_data_cal %>% group_by(subset) %>% summarise(mean_cal = mean(cal), position = mean(as.numeric(positions))),
             aes(x = as.character(position), y = mean_cal, fill = factor(subset, levels = paste0("subset", 1:10))), inherit.aes = F,
             size = 0.8, alpha = 0.8, shape = 23, colour="white") +
  scale_fill_manual(values = hcl.colors(10, palette = "Dynamic")) +
  scale_color_manual(values = hcl.colors(10, palette = "Dynamic")) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        axis.text.y = element_text(family = "Linux Libertine G"),
        axis.title.y = element_text(family = "Linux Libertine G", margin = margin(t = 0, r = 10, b = 0, l = 0)),
        axis.ticks.x=element_blank(),
        panel.grid.major = element_line(size = 0.2),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position="none") +
  coord_cartesian(ylim = c(0.875, 0.925)) +
  labs(y = "AUC")

# Test lettres
dunnTest_out <- FSA::dunnTest(cal ~ positions, method = "bh", data = fagus_auc_data_cal)
letters <- rcompanion::cldList(P.adj ~ Comparison, data = dunnTest_out$res, threshold = 0.05)

data_letters <- data.frame(letters)
data_letters$y <- unlist(fagus_auc_data_cal %>% 
  group_by(positions) %>%
  summarise(
    min = min(cal, na.rm = T)
  ) %>%
  select(min) - 0.004)

p_auc_cal <- fagus_auc_data_cal %>% ggplot() +
  geom_boxplot(aes(y = cal, x = as.character(positions), group = subset, 
                   color = factor(subset, levels = paste0("subset", 1:10)), 
                   fill = factor(subset, levels = paste0("subset", 1:10))), 
               width = 0.5, alpha = 0.3, outlier.shape = NA, fatten = 1, lwd=0.3) +
  geom_beeswarm(aes(y = cal, x = as.character(positions), group = subset, 
                    color = factor(subset, levels = paste0("subset", 1:10)), 
                    fill = factor(subset, levels = paste0("subset", 1:10))),
                alpha = 0.8, cex = 2.5, size = 0.5) +
  geom_text(data = data_letters, aes(x = as.character(Group), label = Letter, y = y), vjust = -0.5, inherit.aes = F,
            family = "Linux Libertine G", size = 2.5) +
  annotate("text", label = "Kruskal-Wallis", x = 1.68, y = 0.917, family = "Linux Libertine G", size = 2.5) +
  annotate("text", label = paste0("p = ", compare_means(cal ~ positions,  
                                                        data = fagus_auc_data_cal, method = "kruskal.test")$p.format), 
           x = 1.68, y = 0.915, family = "Linux Libertine G", size = 3, fontface = "bold") +
  geom_point(data = fagus_auc_data_cal %>% group_by(subset) %>% summarise(mean_cal = mean(cal), position = mean(as.numeric(positions))),
             aes(x = as.character(position), y = mean_cal, fill = factor(subset, levels = paste0("subset", 1:10))), inherit.aes = F,
             size = 0.8, alpha = 1, shape = 23, colour="black", stroke = 0.2) +
  scale_fill_manual(values = hcl.colors(10, palette = "Dynamic")) +
  scale_color_manual(values = hcl.colors(10, palette = "Dynamic")) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        axis.text.y = element_text(family = "Linux Libertine G"),
        axis.title.y = element_text(family = "Linux Libertine G", margin = margin(t = 0, r = 10, b = 0, l = 0)),
        axis.ticks.x=element_blank(),
        panel.grid.major = element_line(size = 0.2),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position="none") +
  coord_cartesian(ylim = c(0.875, 0.925)) +
  labs(y = "AUC")

fagus_auc_data_tot$positions <- positions

wilcoxon_test_tot <- compare_means(tot ~ positions,  data = fagus_auc_data_tot) 
wilcoxon_test_s7 <- wilcoxon_test_tot %>% filter(p.signif != "ns") %>% filter(group1 == "1.7" | group2 == "1.7")
list_comparisons <- lapply(1:nrow(wilcoxon_test_s7), function(i){
  return(c(wilcoxon_test_s7$group1[i], wilcoxon_test_s7$group2[i]))})

p_auc_tot <- na.omit(fagus_auc_data_tot) %>% ggplot(aes(y = tot, as.character(positions), group = subset, 
                    color = factor(subset, levels = paste0("subset", 1:10)), fill = factor(subset, levels = paste0("subset", 1:10)))) +
  geom_boxplot(width = 0.5, alpha = 0.2, outlier.shape = NA, fatten = 1, lwd=0.3) +
  geom_beeswarm(alpha = 0.8, cex = 2.5, size = 0.5) +
  annotate("text", label = "Kruskal-Wallis", x = 1.68, y = 0.917, family = "Linux Libertine G", size = 2.5) +
  annotate("text", label = paste0("p = ", compare_means(tot ~ positions,  
                                                        data = fagus_auc_data_tot, method = "kruskal.test")$p.format), 
           x = 1.68, y = 0.915, family = "Linux Libertine G", size = 3, fontface = "bold") +
  geom_point(data = fagus_auc_data_tot %>% group_by(subset) %>% summarise(mean_tot = mean(tot), position = mean(as.numeric(positions))),
             aes(x = as.character(position), y = mean_tot, fill = factor(subset, levels = paste0("subset", 1:10))), inherit.aes = F,
             size = 0.8, alpha = 1, shape = 23, colour="black", stroke = 0.2) +
  scale_fill_manual(values = hcl.colors(10, palette = "Dynamic")) +
  scale_color_manual(values = hcl.colors(10, palette = "Dynamic")) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        axis.text.y = element_text(family = "Linux Libertine G"),
        axis.title.y = element_text(family = "Linux Libertine G"),
        axis.ticks.x=element_blank(),
        panel.grid.major = element_line(size = 0.2),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position="none") +
  coord_cartesian(ylim = c(0.875, 0.925)) +
  labs(y = "")

plot_grid(p_auc_cal, p_auc_tot, labels=c("(A)", "(B)"), ncol = 2, nrow = 1,
          hjust = 0, vjust = 1.1, align = "h", label_size = 12, label_colour = "#495057")

```

```{r cvs}

library(sjstats)

cv_cal <- fagus_auc_data_cal %>%
    group_by(subset) %>%
    dplyr::summarize(mean = mean(cal), cv = cv(cal))

cv_tot <- fagus_auc_data_tot %>%
    group_by(subset) %>%
    dplyr::summarize(mean = mean(tot), cv = cv(tot))

```

The 100 calibrations of the PHENOFIT model realized for beech showed that random data subsampling had an effect on the final objective function value (i.e. the AUC computed on the 2000 calibration points). Kruskal-Wallis test was significant (p = `r compare_means(cal ~ positions,  data = fagus_auc_data_cal, method = "kruskal.test")$p.format`), meaning that at least one subset provided better AUC during calibration. According to Dunn's tests, `r nrow(dunn.stat.test.all %>% filter(p.adj.signif != "ns"))` pairwise comparisons out of `r nrow(dunn.stat.test.all)` were significant (\hyperref[fig:cmaesrepAUCcal]{Figure 3.A.}). The calibration AUC ranged from `r round(min(fagus_auc_data_cal$cal),3)` to `r round(max(fagus_auc_data_cal$cal),3)` over all subsets, with a mean value of `r round(mean(fagus_auc_data_cal$cal),3)`.

However, more importantly, the repetition of calibrations on different subsets had no significant impact on the total AUC computed on all presence/absence points (Kruskal-Wallis test, p = `r compare_means(tot ~ positions,  data = fagus_auc_data_tot, method = "kruskal.test")$p.format`). Thus, no subset led to an overall better prediction of the species distribution (see \hyperref[fig:cmaesrepAUCcal]{Figure 3.B.}). The total AUC ranged from `r round(min(fagus_auc_data_tot$tot),3)` to `r round(max(fagus_auc_data_tot$tot),3)`, with a mean value of `r round(mean(fagus_auc_data_tot$tot),3)`. 
\newline

### 2.2.2. Non-identifiability of parameters

```{r parameters_values}
cal_folder <- 'D:/calibrations/phenofit/fagus_sylvatica/1000pres_1000abs/paper_data'
sim_folder <- 'D:/simulations/phenofit/backward/fagus_sylvatica/paper_data'

forward_species_file <- "C:/Users/vandermeersch/Dropbox/These_Victor/Phenofit4/species/forward/Fagus_sylvatica_VVanderMeersch.species"
forward_values <- as.numeric(read_species_file(forward_species_file))

# Bounds
bd_folder <- 'C:/Users/vandermeersch/Dropbox/These_Victor/Phenofit4/species/backward/fagus_sylvatica'
species_lb  <- as.numeric(read_species_file(file.path(bd_folder, 'Fagus_sylvatica_lb.species')))
species_ub <- as.numeric(read_species_file(file.path(bd_folder, 'Fagus_sylvatica_ub.species')))
species_init <- as.numeric(read_species_file(file.path(bd_folder, 'Fagus_sylvatica_init.species')))

# Calibrations
species_files <- mixedsort(list.files(path = file.path(cal_folder,"CMAES"), pattern = "\\.species$", full.names = T, recursive = T))

# Load values
best_values <- as.numeric(read_species_file("D:/calibrations/phenofit/fagus_sylvatica/1000pres_1000abs/paper_data/CMAES/subset_4/cmaes_fit_subset4_rep1.species"))
species_parameters <- lapply(species_files, function(x){
  species_values <- data.frame(value = read_species_file(x)) %>% 
    mutate_all(as.numeric)
  subset <- strsplit(x, "/")[[1]][8]
  rep <- strsplit(strsplit(strsplit(x, "/")[[1]][9], "_")[[1]][4], "[.]")[[1]][1]
  species_values$rep <- paste0(subset, "_", rep)
  species_values$var <- rownames(species_values)
  species_values$lb <- species_lb
  species_values$ub <- species_ub
  species_values$init <- species_init
  species_values$forward <- forward_values
  species_values$best <- best_values
  species_values
  })
species_parameters <- do.call(rbind, species_parameters)
species_parameters <- na.omit(species_parameters) #remove fixed parameters (bounds = NA)

# Calibrations
species_files <- mixedsort(list.files(path = file.path(cal_folder,"ABC"), pattern = "\\.species$", full.names = T, recursive = T))

# Load values
best_values <- as.numeric(read_species_file("D:/calibrations/phenofit/fagus_sylvatica/1000pres_1000abs/paper_data/ABC/subset_1/abc_rej_subset1_rep9.species" ))
species_parameters_ABC <- lapply(species_files, function(x){
  species_values <- data.frame(value = read_species_file(x)) %>% 
    mutate_all(as.numeric)
  subset <- strsplit(x, "/")[[1]][8]
  rep <- strsplit(strsplit(strsplit(x, "/")[[1]][9], "_")[[1]][4], "[.]")[[1]][1]
  species_values$rep <- paste0(subset, "_", rep)
  species_values$var <- rownames(species_values)
  species_values$lb <- species_lb
  species_values$ub <- species_ub
  species_values$init <- species_init
  species_values$forward <- forward_values
  species_values$best <- best_values
  species_values
  })
species_parameters_ABC <- do.call(rbind, species_parameters_ABC)
species_parameters_ABC <- na.omit(species_parameters_ABC) #remove fixed parameters (bounds = NA)

```

```{r unfolding_data}

if(recompute_unfold_date){
  simulation_files <- mixedsort(list.files(path = file.path(sim_folder, "CMAES"), pattern = "LeafUnfoldingDate", full.names = T, recursive = T))

  unfold_date_data <- as.data.frame(cbind(alt$lat, alt$lon))
  for(i in 1:(length(simulation_files))){
    # Load date
    date <- read_mean_outputvalue_phenofit(simulation_files[i])
    unfold_date_data[,2+i] <- t(date)
  }

  # Summary 
  names(unfold_date_data)[1:2] <- c("lat", "lon")
  unfold_date_data$mean <- rowMeans(unfold_date_data[,3:ncol(unfold_date_data)])
  unfold_date_data$median <- rowMedians(as.matrix(unfold_date_data[,3:ncol(unfold_date_data)]))
  unfold_date_data$sd <- rowSds(as.matrix(unfold_date_data[,3:ncol(unfold_date_data)]))
  unfold_date_data$mad <- rowMads(as.matrix(unfold_date_data[,3:ncol(unfold_date_data)]))

  saveRDS(unfold_date_data, file.path(wd, "cache", "unfold_date_data.rds"))
}else{
  unfold_date_data <- readRDS(file.path(wd, "cache", "unfold_date_data.rds"))
}

data_leaf_par <- species_parameters %>%
  dplyr::filter(grepl("leaf", species_parameters$var))
data_leaf_par$var_name <- factor(data_leaf_par$var, labels = c('leaf1' = parse(text=TeX('$t_0$')),
                                                               'leaf2' = parse(text=TeX('$T_b$')),
                                                               'leaf3' = parse(text=TeX('$d_T$')),
                                                               'leaf4' = parse(text=TeX('$T_{50}$')),
                                                               'leaf5' = parse(text=TeX('$C_{crit}$')),
                                                               'leaf6' = parse(text=TeX('$F_{crit}$'))))

data_leaf_par_ABC <- species_parameters_ABC %>%
  dplyr::filter(grepl("leaf", species_parameters_ABC$var))
data_leaf_par_ABC$var_name <- factor(data_leaf_par_ABC$var, labels = c('leaf1' = parse(text=TeX('$t_0$')),
                                                               'leaf2' = parse(text=TeX('$T_b$')),
                                                               'leaf3' = parse(text=TeX('$d_T$')),
                                                               'leaf4' = parse(text=TeX('$T_{50}$')),
                                                               'leaf5' = parse(text=TeX('$C_{crit}$')),
                                                               'leaf6' = parse(text=TeX('$F_{crit}$'))))

```

To illustrate the variability in the parameter estimates that can be obtained with the inverse calibration, we focused on the leaf unfolding date submodel of PHENOFIT (see [Appendix A](#appendixA) and [Appendix G](#appendixG)). The parameter values found by CMA-ES varied greatly across the 100 calibrations (\hyperref[fig:unfoldingplots]{Figure 4}). For example, the critical amount of chilling $C_{crit}$ required to break bud dormancy and the critical amount of forcing $F_{crit}$ required to break bud ranged from `r round(min(data_leaf_par[data_leaf_par$var == "leaf5", "value"]),2)` to `r round(max(data_leaf_par[data_leaf_par$var == "leaf5", "value"]),2)` and from `r round(min(data_leaf_par[data_leaf_par$var == "leaf6", "value"]),2)` to `r round(max(data_leaf_par[data_leaf_par$var == "leaf6", "value"]),2)` respectively, with a mean value of `r round(mean(data_leaf_par[data_leaf_par$var == "leaf5", "value"]),2)` and `r round(mean(data_leaf_par[data_leaf_par$var == "leaf6", "value"]),2)`. Their coefficient of variations were `r round(cv(data_leaf_par[data_leaf_par$var == "leaf5", "value"])*100,1)`% and `r round(cv(data_leaf_par[data_leaf_par$var == "leaf6", "value"])*100,1)`% respectively. Kendall correlation coefficient between $C_{crit}$ and the threshold temperature of the response function to temperature during dormancy $T_b$ is `r round(cor.test(data_leaf_par[data_leaf_par$var == "leaf2", "value"], data_leaf_par[data_leaf_par$var == "leaf5", "value"], method="kendall")$estimate,2)` ($p <$ `r max(0.001, cor.test(data_leaf_par[data_leaf_par$var == "leaf2", "value"], data_leaf_par[data_leaf_par$var == "leaf5", "value"], method="kendall")$p.value)`). Kendall correlation coefficient between $F_{crit}$ and the mid-response temperature $T_{50}$ is `r round(cor.test(data_leaf_par[data_leaf_par$var == "leaf4", "value"], data_leaf_par[data_leaf_par$var == "leaf6", "value"], method="kendall")$estimate,2)` ($p <$ `r max(0.001, cor.test(data_leaf_par[data_leaf_par$var == "leaf4", "value"], data_leaf_par[data_leaf_par$var == "leaf6", "value"], method="kendall")$p.value)`).

```{r unfoldingplots, include=TRUE, fig.height = 4, fig.width = 4, fig.align = "center", fig.cap = "Effects of stochasticity of CMA-ES calibrationd on PHENOFIT leaf unfolding model parameter values for beech. Each panel is a parameter. Y-axis limits are lower and upper bounds used during calibration. Each point is a calibrated parameter value, color gradient is based on $F_{crit}$ values. Red diamonds are parameter values obtained with expert calibration, blue ones are parameter values obtained with the best inverse calibration.", fig.pos="H", fig.showtext=TRUE}
count1 <- 0
breaks_fun1 <- function(x) {
  count1 <<- count1 + 1
  switch(
    count1,
    c(-122, -100, -75, -50, -25, 0),
    c(-122, -100, -75, -50, -25, 0),
    c(-5, 0, 5, 12),
    c(-5, 0, 5, 12),
    c(-2, -1.5, -1, -0.5, -0.05),
    c(-2, -1.5, -1, -0.5, -0.05),
    c(5, 10, 15, 20),
    c(5, 10, 15, 20),
    c(1, 50, 100, 150),
    c(1, 50, 100, 150),
    c(1, 25, 50, 75, 100, 120),
    c(1, 25, 50, 75, 100, 120)
  )
}

count2 <- 0
breaks_fun2 <- function(x) {
  count2 <<- count2 + 1
  switch(
    count2,
    c(-122, -100, -75, -50, -25, 0),
    c(-122, -100, -75, -50, -25, 0),
    c(-5, 0, 5, 12),
    c(-5, 0, 5, 12),
    c(-2, -1.5, -1, -0.5, -0.05),
    c(-2, -1.5, -1, -0.5, -0.05),
    c(5, 10, 15, 20),
    c(5, 10, 15, 20),
    c(1, 50, 100, 150),
    c(1, 50, 100, 150),
    c(1, 25, 50, 75, 100, 120),
    c(1, 25, 50, 75, 100, 120)
  )
}

col_Fcrit_temp <- data_leaf_par[data_leaf_par$var == "leaf6",]
col_Fcrit_temp <- col_Fcrit_temp[order(col_Fcrit_temp$value),]
col_Fcrit_temp <- split(col_Fcrit_temp, ceiling(seq_along(col_Fcrit_temp$value)/10))
for(i in 1:10){
  col_Fcrit_temp[[i]]['col_group'] <- i
}
col_Fcrit_temp<- do.call("rbind", col_Fcrit_temp) %>% dplyr::select(c(rep, col_group))

data_leaf_par <- left_join(data_leaf_par, col_Fcrit_temp)

parameter_plot <- data_leaf_par %>%
  ggplot(aes(y = value, x = 1)) +
  geom_violin(fill = 'lightgrey', alpha= 0.2, size = 0.4, colour = 'darkgrey') +
  geom_beeswarm(aes(fill = factor(col_group), col = factor(col_group)), 
                alpha = 0.5, cex = 2.5, size = 0.4) +
  facet_wrap("var_name", scales="free_y",
             labeller= label_parsed) +
  scale_y_continuous(breaks = breaks_fun1) +
  geom_point(aes(y = forward, x= 0.55), fill = 'darkred', size = 1.5, alpha = 0.9, shape = 23, colour="white", stat = "unique") +
  geom_point(data = data.frame(best = unique(data_leaf_par$best), var_name = unique(data_leaf_par$var_name)),
             aes(y = best, x= c(0.55, 0.65, 0.55, 0.55, 0.55, 0.55)), fill = 'darkblue', size = 1.5, 
             alpha = 0.9, shape = 23, colour="white", inherit.aes = F) +
  scale_fill_manual(values = c("#002f61", "#004a72", "#00647e", "#007c84", "#009382", "#00aa76", "#3bbd60", "#83c846", "#bdce29", "#f4d004")) +
  scale_color_manual(values = c("#002f61", "#004a72", "#00647e", "#007c84", "#009382", "#00aa76", "#3bbd60", "#83c846", "#bdce29", "#f4d004")) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        axis.text.y = element_text(family = "Linux Libertine G"),
        axis.title.y = element_text(family = "Linux Libertine G"),
        strip.text.x = element_text(family = "Linux Libertine G", face = "bold", size = 11),
        axis.ticks.x=element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position="none") +
  geom_blank(aes(y = lb)) +
  geom_blank(aes(y = ub)) +
  labs(y = "")

col_Fcrit_temp <- data_leaf_par_ABC[data_leaf_par_ABC$var == "leaf6",]
col_Fcrit_temp <- col_Fcrit_temp[order(col_Fcrit_temp$value),]
col_Fcrit_temp <- split(col_Fcrit_temp, ceiling(seq_along(col_Fcrit_temp$value)/10))
for(i in 1:3){
  col_Fcrit_temp[[i]]['col_group'] <- i
}
col_Fcrit_temp<- do.call("rbind", col_Fcrit_temp) %>% dplyr::select(c(rep, col_group))

data_leaf_par_ABC <- left_join(data_leaf_par_ABC, col_Fcrit_temp)

parameter_plot_ABC <- data_leaf_par_ABC %>%
  ggplot(aes(y = value, x = 1)) +
  geom_violin(fill = 'lightgrey', alpha= 0.2, size = 0.4, colour = 'darkgrey') +
  geom_beeswarm(aes(fill = factor(col_group), col = factor(col_group)), 
                alpha = 0.5, cex = 2.5, size = 0.4) +
  facet_wrap("var_name", scales="free_y",
             labeller= label_parsed) +
  scale_y_continuous(breaks = breaks_fun2) +
  geom_point(aes(y = forward, x= 0.55), fill = 'darkred', size = 1.5, alpha = 0.9, shape = 23, colour="white", stat = "unique") +
  geom_point(data = data.frame(best = unique(data_leaf_par_ABC$best), var_name = unique(data_leaf_par_ABC$var_name)),
             aes(y = best, x= 0.55), fill = 'darkblue', size = 1.5, 
             alpha = 0.9, shape = 23, colour="white", inherit.aes = F) +
  scale_fill_manual(values = c("#002f61", "#00aa76", "#f4d004")) +
  scale_color_manual(values = c("#002f61", "#00aa76", "#f4d004")) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        axis.text.y = element_text(family = "Linux Libertine G"),
        axis.title.y = element_text(family = "Linux Libertine G"),
        strip.text.x = element_text(family = "Linux Libertine G", face = "bold", size = 11),
        axis.ticks.x=element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position="none") +
  geom_blank(aes(y = lb)) +
  geom_blank(aes(y = ub)) +
  labs(y = "")

# with ABC:
# plot_grid(parameter_plot, parameter_plot_ABC, labels=c("(A)", "(B)"), ncol = 2, nrow = 1,
#           hjust = 0, vjust = 1.1, align = "h", label_size = 12, rel_widths = c(1, 1),
#           label_colour = "#495057")

parameter_plot

# geom_dotplot(binaxis = 'y', stackdir='center', alpha=0.5, col = 'darkred') can also be an option ?

unfold_date_data_fsyl <- left_join(fsyl_presabs, unfold_date_data)

```

# 3. Discussion

\textcolor{darkgreen}{Our results showed that CMA-ES is a feasible and efficient optimizer for inverse calibration of complex ecological models. The algorithm was able to find parameter sets that significantly improved the predictions of the calibrated models compared to the expert parametrization.  However, our study also highlighted the issue of non-identifiability of parameter values due to the data limitation and the dependence between process-based model parameters, which may result in diverging parameter values even though the calibrated models describe the observed species distribution well. Nonetheless, our study provides a promising methodological framework for calibrating process-based models using CMA-ES, and for spreading the use of such models in predictive ecology.}

## 3.1. Performance and advantages of CMA-ES to calibrate complex models in ecology

Here we demonstrated that inverse calibration with CMA-ES is feasible and provide good results for complex and runtime-expensive ecological models.  
With a subsampling of species occurrence data, the algorithm succeeds in finding parameter sets which provide \textcolor{darkgreen}{higher} AUC values. The predictions of the calibrated models are sharply improved compared to expert parametrization (\hyperref[fig:phenofitmaps]{Figure 1} and \hyperref[fig:castaneamaps]{Figure 2}). Two striking examples are the increase in the performance of PHENOFIT model for silver fir, from 0.72 to 0.9, and of CASTANEA model for holm oak, from 0.7 to 0.95. Moreover, CMA-ES performed equally well regardless of the species occurrence subset used during calibration (\hyperref[fig:cmaesrepAUCcal]{Figure 3.B.}), and thus permitted to find a good compromise between computational cost and calibration efficiency.

CMA-ES is a "generic" optimizer which can be applied to various problems. It is easy to use as it does not require an extensive tuning to  efficiently explore the parameter space. We only had to choose the number of candidate solutions $\lambda$, and the initial search region (initial starting point and step size $\sigma$). As well as being quasi parameter-free, CMA-ES has several structural advantages particularly useful for complex optimization problems. \textcolor{darkgreen}{Firstly, the algorithm's covariance matrix enables the learning of second-order information, which provides insights into pairwise dependencies between parameters. Secondly, the covariance matrix adaptation of CMA-ES is highly efficient in handling ill-conditioned and non-separable problems }[@Hansen2011]\textcolor{darkgreen}{. Lastly, CMA-ES's update mechanism of step size $\sigma$ (i.e. the mutation force) helps prevent premature convergence }[@Hansen2001]\textcolor{darkgreen}{, allowing the algorithm to explore more of the search space.}
CMA-ES has been shown to outperform several other optimization algorithms [@Hansen2010], and is usually the most efficient method when the target cost (i.e. the number of objective function evaluations) is about $100*N$ [$N$ being the dimension of the parameter search space, @Baeck2013].

## 3.2. Non-identifiability of parameter values

\textcolor{darkgreen}{Equifinality is a common problem encountered during model calibration, where multiple parameter sets can produce equally good fits to the observed data. This issue can arise due to several factors, including limitations in the quantity or quality of available data, competing processes within the model, and parameter interactions that affect the model output in complex ways.}

In both models used in this study, biological mechanisms are explicitly calculated in several submodels (e.g. a leaf unfolding submodel or a stomatal opening submodel). A submodel output has inevitably a significant influence on the other submodels as biological processes can be highly dependent with feedbacks: in CASTANEA, for example, the stomatal opening affects the photosynthesis, and *vice versa*. Within each submodel, parameters are also strongly dependent because of structural correlations. To illustrate this problem, we focused on the beech leaf unfolding submodel of PHENOFIT (see [Appendix A](#appendixA)). This model has 6 parameters [@Chuine2000]: a starting date of the processes ($t_0$), one parameter describing the response function to temperature during the dormancy phase ($T_b$), two parameters describing the response function to temperature during the phase of bud growth ($d_T$, $T_{50}$), and two parameters representing the sums of the daily responses to temperature during bud dormancy ($C_{crit}$) and during bud growth ($F_{crit}$) that respectively determine the date of bud dormancy break and the date of leaf unfolding (see [Appendix G](#appendixG) for details). Since no information on the date of bud dormancy break is available for the calibration, a first structural negative correlation exists between $C_{crit}$ and $F_{crit}$: the same leaf unfolding date can be obtained with either a long dormancy phase and short bud growth phase or a short dormancy phase and a long bud growth phase. Other structural correlations exist between $T_b$ and $C_{crit}$ on the one hand and $d_T$/$T_{50}$ and $F_{crit}$ on the other hand: for example, a rapid accumulation of chilling units with a high critical chilling requirement could yield identical results as a slow accumulation with a low critical chilling requirement (i.e. the threshold temperature $T_{b}$ and the critical chilling requirement $C_{crit}$ are dependent, see Figure G.1.A. in [Appendix G](#appendixG)).

Consequently, several parameter sets may be statistically equivalent and parameters non-identifiable. In fact, calibration repetitions gave diverging parameter values (\hyperref[fig:unfoldingplots]{Figure 4}) while being efficient in distinguishing between species presence and absence (i.e. AUC ~ 0.9, \hyperref[fig:cmaesrepAUCcal]{Figure 3}). Thus, even if the calibrated model describes the observed species distribution very well, it does not necessarily mean that parameter values are ecologically relevant. This concern is similar to the criticisms against correlative SDMs, in which parameter values and correlations that well reproduce species ranges do not necessarily describe a complex biological reality. In our case, the constraints imposed by the explicit mathematical equations embedded in the models we used were not sufficient to ensure calibration convergence towards similar solutions that would have suggested a high biological realism. However, it is worth noting that we deliberately chose large parameter ranges (although biologically realistic, i.e. corresponding to the observations made on the different processes modelled across different species) in order to give free rein to the optimization algorithm. As our goal was to assess the performance of CMA-ES objectively, we did not attempt to minimize this non-identifiability issue by restricting the parameter space. To deal with equifinality issues, an avenue to explore could be the use of multiple objective functions during model calibration to assess different aspects of the model performance. \textcolor{darkgreen}{Additionally, if a closed-form likelihood can be derived, one could use a Bayesian framework to combine prior knowledge and inverse estimation of parameters to constrain the parameter space and study the nature of trade-offs between parameters} [@Hartig2012; @Cailleret2020].

## 3.3. Methodological issues and perspectives

Our goal here was to investigate the performance of CMA-ES to calibrate quite complex process-based species distribution models using species occurrence data. We did not attempt to validate our parametrizations using temporally or spatially independent data, and AUC was only used to determine if model outputs were consistent with species distributions. However, AUC is scale-invariant (it measures how well predictions are ranked rather than their absolute values), and prevents us from obtaining calibrated outputs with more ecological meaning. For example, in PHENOFIT, one species with a fitness of $0.8$ could be considered as absent whereas another one as present. Further work could thus be conducted to examine the effects of choosing a different objective function.

It would also be valuable to use a significantly higher computing power, with an adapted version of CMA-ES. To improve the global search performance of CMA-ES, we slightly increased the number of candidate solutions $\lambda$ [@Hansen2004] and used a computing cluster to evaluate $\lambda$ functions in parallel. We were able to use between 40 and 120 cores, which is far from the computing power of some GPUs (> 2000 cores). In this case, choosing a very large number of candidate solutions might not be the best choice. To use efficiently this large parallel computing power, one could rather use a CMA-ES restart strategy [e.g. IPOP-CMA-ES, @Auger2005], where the number of candidate solutions is successively increased (by a factor of 2), and run these calibrations in parallel. Moreover, when a model  requires a high computation time and thus only a small budget can be afforded, the original fitness function could be approximated with a surrogate model in order to reduce the number of original function evaluations required [e.g. @Auger2004 ; @Loshchilov2013].

There are several issues regarding the process-based models we used which can impact their calibration and bias their parameter estimates. First, like any model, although they have a certain level of complexity, they are not a perfect representation of the reality, and their inherent structural errors increase the probability of finding parameter values that deviate from the true values of the underlying processes [see @Oberpriller2021]. Secondly, they do not necessarily include all the environmental factors at stake. For example, pedologic variable in PHENOFIT only involve the water holding capacity. In this model, other variables such as pH or soil texture are not considered. Thirdly, and more importantly, they are used here to represent the fundamental niche, and to estimate the potential distribution of the species using pedoclimatic variables. When calibrated against observed distributions, which represent the realized niche, they face the same issues as correlative models, and their parameter estimated can be distorted because compensating for processes not represented in the model (e.g. biotic interactions). In addition, in our case here, land use management probably also play an important role in shaping tree realized distribution, while not being addressed in the models. We included GBIF occurrence data, and especially as much as possible isolated native tree records outside forests, to help correct this problem, but it is impossible to be exhaustive at the spatial resolution of 0.1°. At this scale, local variations of soil characteristics and of competitive interactions among trees (e.g. along an altitudinal gradient) can also not be considered.

Finally, several authors advocate for process-based modeling approaches relying upon species response functions that are a priori defined [e.g. @Higgins2020]. However, the main limitation of such models is the data availability to infer their parameters [@Urban2016]. Expert parameterization is often long and arduous. One possible way to facilitate parameter value estimation would be to use inverse calibration, and we demonstrated here that CMA-ES can be a powerful optimizer to this end. For example, CMA-ES driven by species occurrence data could be used to calibrate submodels whose parameter values cannot be measured or are too hard to measure experimentally. However, when a structural correlation exists (i.e. trade-off between processes that are inherently present and interconnected in the model, as in the leaf unfolding submodel), inverse calibration might not provide the right parameter estimates. In such a case, expert knowledge, observations and measurements are necessary to determine *a posteriori* which estimates are the most realistic. This is possible in the case of process-based SDMs, and usually not feasable in the case of correlative SDMs. A combination of both expert and inverse calibrations might offer a new perspective for spreading the use of process-based models in predictive ecology, especially for climate change impact studies.

<!-- # Conclusion (optionnal) -->

<!-- **[Commentaire : ce n’est pas une conclusion, il faut la refaire. Ca ressemble plus à une partie perspectives et ouverture d’une discussions]** -->
<!-- \newline -->

<!-- ~~We do recognize that the calibration approach proposed here is not a perfect method, the main weakness being the questionable biological plausibility of parameter values we found. Nevertheless, CMA-ES is particularly useful to generate inverse optimized versions of the process-based models. These fitted model versions integrate explicit processes but are calibrated on species distributions like cSDMs. They thus fall within the continuum between correlative and forward models [see @Dormann2012]. Both types of models have been compared in several studies, which found strong discrepancies between projections [@Morin2009; @Cheaib2012; @Gritti2013]. No studies, though, have examined the structural differences between the two approaches in order to determine the origin of their robustness. Comparison of correlative models to optimized process-based model projections will allow testing whether model robustness come from the cause-to-effect relationships or not, while comparison of optimized process-based model to standard process-based models will allow testing whether model robustness come from a calibration independent species distribution or not. This study is, to our knowledge, the first one to assess the feasibility of CMA-ES to calibrate ecological process-based models, and will permit to carry out a thorough comparison of different versions of the models.~~ -->
<!-- \newline -->

# Acknowledgements

The authors would like to thank Hendrik Davi for helping us in using the CASTANEA model. We are also deeply grateful for many helpful comments from Florence Tauc. We would also like to thank François de Coligny, manager of the CAPSIS platform, Gilles Le Moguedec, and the GenOuest and TGCC teams for their support. Finally, we would like to thank Florian Hartig and another anonymous reviewer whose comments and suggestions helped us improve and clarify this manuscript.   
V.V. was supported by a GAIA doctoral school PhD Fellowship.

# Conflict of interest

The authors have no conflicts of interest to declare.

# Author contributions

I.C. devised the main conceptual ideas. V.V. worked out the technical details, performed the numerical calculations and wrote the first draft of the manuscript. The two authors discussed the analyses and the results, and contributed to the final manuscript.

# Data availability

ERA5-Land dataset is available on the [Copernicus Climate Change Service website](https://cds.climate.copernicus.eu/cdsapp#!/dataset/reanalysis-era5-land?tab=overview). EU-SoilHydroGrids is available on the 
[European Soil Data Centre website](https://esdac.jrc.ec.europa.eu/content/3d-soil-hydraulic-database-europe-1-km-and-250-m-resolution). SoilGrids250m is available on the [International Soil Reference and Information Centre website](https://www.isric.org/explore/soilgrids). EU-Forest database is available on [FigShare](https://figshare.com/collections/A_high-resolution_pan-European_tree_occurrence_dataset/3288407). The R code associated with this work will be made available on GitLab upon acceptance of the manuscript.

# References
\setlength{\parindent}{-0.2in}
\setlength{\leftskip}{0.2in}
\setlength{\parskip}{8pt}
\vspace*{-0.2in}
\noindent
<div id="refs"></div>

\newpage

# Supplementary Appendix A: Insights on the models {#appendixA}

\renewcommand*\thetable{A.\arabic{table}}
\renewcommand*\thefigure{A.\arabic{figure}}

\setcounter{figure}{0}
\setcounter{table}{0}

\pagenumbering{arabic}
\renewcommand*{\thepage}{A--\arabic{page}}

```{r appendixA, child = c('appendix/appendixA.Rmd'), eval=T}
```

# Supplementary Appendix B: Processing of occurrence data {#appendixB}

\renewcommand*\thetable{B.\arabic{table}}
\renewcommand*\thefigure{B.\arabic{figure}}

\setcounter{figure}{0}
\setcounter{table}{0}

\pagenumbering{arabic}
\renewcommand*{\thepage}{B--\arabic{page}}

```{r appendixB, child = c('appendix/appendixB.Rmd'), eval=T}
```

\newpage

# Supplementary Appendix C: Species distributions {#appendixC}

\renewcommand*\thetable{C.\arabic{table}}
\renewcommand*\thefigure{C.\arabic{figure}}

\setcounter{figure}{0}
\setcounter{table}{0}

\pagenumbering{arabic}
\renewcommand*{\thepage}{C--\arabic{page}}

```{r appendixC, child = c('appendix/appendixC.Rmd'), eval=T}
```

\newpage

# Supplementary Appendix D: Objective function and constraint handling in CMA-ES {#appendixD}

\renewcommand*\thetable{D.\arabic{table}}
\renewcommand*\thefigure{D.\arabic{figure}}

\setcounter{figure}{0}
\setcounter{table}{0}

\pagenumbering{arabic}
\renewcommand*{\thepage}{D--\arabic{page}}

```{r appendixD, child = c('appendix/appendixD.Rmd'), eval=T}
```

\newpage

# Supplementary Appendix E: holm oak and silver fir calibrations {#appendixE}

\renewcommand*\thetable{E.\arabic{table}}
\renewcommand*\thefigure{E.\arabic{figure}}

\setcounter{figure}{0}
\setcounter{table}{0}

\pagenumbering{arabic}
\renewcommand*{\thepage}{E--\arabic{page}}

```{r appendixE, child = c('appendix/appendixE.Rmd'), eval=T}
```

\newpage

# Supplementary Appendix F: Raw model outputs  {#appendixF}

\renewcommand*\thetable{F.\arabic{table}}
\renewcommand*\thefigure{F.\arabic{figure}}

\setcounter{figure}{0}
\setcounter{table}{0}

\pagenumbering{arabic}
\renewcommand*{\thepage}{F--\arabic{page}}

```{r appendixF, child = c('appendix/appendixF.Rmd'), eval=T}
```

\newpage

# Supplementary Appendix G: Leaf unfolding submodel  {#appendixG}

\renewcommand*\thetable{G.\arabic{table}}
\renewcommand*\thefigure{G.\arabic{figure}}

\setcounter{figure}{0}
\setcounter{table}{0}

\pagenumbering{arabic}
\renewcommand*{\thepage}{G--\arabic{page}}

```{r appendixG, child = c('appendix/appendixG.Rmd'), eval=T}
```

### 2023-03-21 manuscript/paper_revision.tex
add manuscript Rmarkdown/Latex files

% ARTICLE ----
% This is just here so I know exactly what I'm looking at in Rstudio when messing with stuff.
\documentclass[11pt,]{article}
\usepackage[left=1in,top=1in,right=1in,bottom=1in]{geometry}
\newcommand*{\authorfont}{\fontfamily{phv}\selectfont}
\usepackage[]{libertine}

  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}

\usepackage{abstract}
\renewcommand{\abstractname}{}    % clear the title
\renewcommand{\absnamepos}{empty} % originally center

\renewenvironment{abstract}
 {{%
    \setlength{\leftmargin}{0mm}
    \setlength{\rightmargin}{\leftmargin}%
  }%
  \relax}
 {\endlist}

\makeatletter
\def\@maketitle{%
  \newpage
%  \null
%  \vskip 2em%
%  \begin{center}%
  \let \footnote \thanks
    {\fontsize{18}{20}\selectfont\raggedright  \setlength{\parindent}{0pt} \@title \par}%
}
%\fi
\makeatother

\setcounter{secnumdepth}{0}

\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}

\title{A novel approach to inverse calibration of process-based models
using evolutionary algorithm and species distribution data  }

\author{\Large Victor Van der Meersch
\footnote{Corresponding author - victor.vandermeersch@cefe.cnrs.fr},
Isabelle Chuine\vspace{0.05in} \newline\newline\normalsize\emph{CEFE,
Université de Montpellier, CNRS, EPHE, IRD, Montpellier, France}  }

\date{}

\usepackage{titlesec}

\titleformat*{\section}{\normalsize\bfseries}
\titleformat*{\subsection}{\normalsize\itshape}
\titleformat*{\subsubsection}{\normalsize\itshape}
\titleformat*{\paragraph}{\normalsize\itshape}
\titleformat*{\subparagraph}{\normalsize\itshape}

\usepackage{biblatex}

\addbibresource{bib/paper.bib}

\newtheorem{hypothesis}{Hypothesis}
\usepackage{setspace}

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usepackage{graphicx}
\usepackage{float}
\floatplacement{figure}{H}
\usepackage[width=.9\textwidth, font=small]{caption}
\usepackage{libertine}
\usepackage[left]{lineno}
\linenumbers
\renewcommand*{\thefootnote}{\fnsymbol{footnote}}
\usepackage{xcolor}
\definecolor{darkgreen}{RGB}{112,173,71}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

% move the hyperref stuff down here, after header-includes, to allow for - \usepackage{hyperref}

\makeatletter

\makeatother

% Add an option for endnotes. -----

% add tightlist ----------
\providecommand{\tightlist}{%
\setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% add some other packages ----------

% \usepackage{multicol}
% This should regulate where figures float
% See: https://tex.stackexchange.com/questions/2275/keeping-tables-figures-close-to-where-they-are-mentioned
\usepackage[section]{placeins}

% CSL environment change -----

\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

% Last minute stuff -----
% \usepackage{amssymb,amsmath} % HT @ashenkin

% Add by VV
% \usepackage[dvipsnames]{xcolor} % clash with kableExtra ?
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage[unicode=true]{hyperref}
  \PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
\definecolor{um-red}{RGB}{233,78,82}
\definecolor{um-gray}{RGB}{114,146,162}
\definecolor{blue-sapphire}{RGB}{0, 110, 144}
\hypersetup{
    colorlinks=true,
    linkcolor=um-red,
    citecolor=um-red,
    urlcolor=um-gray,
    breaklinks=true}
\urlstyle{same} % don't use monospace font for urls
\usepackage[capitalise, nameinlink]{cleveref}

\begin{document}

% \pagenumbering{arabic}% resets `page` counter to 1
%
% \maketitle

{% \usefont{T1}{pnc}{m}{n}
\setlength{\parindent}{0pt}
\thispagestyle{plain}
{\fontsize{18}{20}\selectfont\raggedright
\maketitle  % title \par

}

{
   \vskip 13.5pt\relax \normalsize\fontsize{11}{12}
\textbf{\authorfont Victor Van der Meersch
\footnote{Corresponding author - victor.vandermeersch@cefe.cnrs.fr},
Isabelle Chuine} \hskip 15pt \emph{\small CEFE, Université de
Montpellier, CNRS, EPHE, IRD, Montpellier, France}

}

}

\begin{abstract}

    \hbox{\vrule height .2pt width 39.14pc}

    \vskip 8.5pt % \small

\noindent \begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Two main types of species distribution models are used to project
  species range shifts in future climatic conditions: correlative and
  process-based models. Although there is some continuity between these
  two types of models, they are fundamentally different in their
  hypotheses (statistical relationships \emph{vs} mechanistic
  relationships) and their calibration methods (dependent \emph{vs}
  generally independent of the species observed distributions).
\item
  One of the limitations to the use of process-based models is the
  difficulty to parameterize them for a large number of species, while
  one of the advantages of correlative SDMs is their easier calibration
  using solely species occurrences\}. We investigated the feasibility of
  using an evolutionary algorithm (called covariance matrix adaptation
  evolution strategy, CMA-ES) to calibrate process-based models using
  species distribution data. This method is well established in some
  fields (robotics, aerospace research, \ldots), but has never been
  used, to our knowledge, in ecology, despite its ability to deal with
  very large space dimensions. Using tree species occurrence data across
  Europe, we adapted the CMA-ES algorithm to find appropriate values of
  model parameters. We estimated simultaneously 27 to 77 parameters of
  two process-based models simulating forest tree's ecophysiology for
  three species with varying range sizes and geographical distributions.
\item
  CMA-ES provided parameter estimates leading to better prediction of
  species distribution than parameter estimates based on experts
  knowledge. Our results also revealed that some model parameters and
  processes were strongly dependent, and different parameter
  combinations could therefore lead to high model accuracy.
\item
  \textcolor{darkgreen}{We conclude that} CMA-ES is an efficient
  state-of-the-art method to calibrate process-based models with a large
  number of parameters using species occurrence data. Inverse modelling
  using CMA-ES is a powerful method to calibrate process-based
  parameters which can hardly be measured. However, the method does not
  warranty that parameter estimates are correct because of several
  sources of bias, similarly to correlative models, and expert knowledge
  is required to validate results.
\end{enumerate}

\vskip 8.5pt \noindent \emph{Keywords}: calibration, evolutionary
algorithm, cma-es, species distribution model, process-based model,
trees \par

    \hbox{\vrule height .2pt width 39.14pc}

\end{abstract}

\vskip -8.5pt

 % removetitleabstract

\noindent \doublespacing

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

The speed and magnitude of projected climate changes are profoundly
affecting species distributions, ecological communities and ecosystem
processes, and numerous ecological systems are now approaching tipping
points (\protect\hyperlink{ref-Lenton2008}{Lenton \emph{et al.} 2008};
\protect\hyperlink{ref-Barnosky2012}{Barnosky \emph{et al.} 2012};
\protect\hyperlink{ref-Steffen2018}{Steffen \emph{et al.} 2018}; but see
\protect\hyperlink{ref-Brook2013}{Brook \emph{et al.} 2013}). Large
uncertainties on the persistence and the resilience of ecosystems exist.
Ecological forecasting has now become a critical tool for managers and
decision-makers (\protect\hyperlink{ref-Urban2015}{Urban 2015}), and
robust predictive approaches are necessary to provide reliable
projections of species geographic range shifts and ecosystem functioning
(\protect\hyperlink{ref-Mouquet2015}{Mouquet \emph{et al.} 2015}).
Forecasting the dynamics of ecological systems for the upcoming decades
and centuries is very difficult, because ecological systems are
extremely complex, influenced by a lot of factors and processes, and
climatic conditions with no analogues in the recent past are forecasted
to become common (\protect\hyperlink{ref-Williams2007}{Williams \emph{et
al.} 2007}; \protect\hyperlink{ref-Radeloff2015}{Radeloff \emph{et al.}
2015}; \protect\hyperlink{ref-Fitzpatrick2018}{Fitzpatrick \emph{et al.}
2018}). Ecological models have thus increased in complexity over the
last 50 years, incorporating more and more processes described with
various degrees of complexity depending on their objectives.

Nowadays, two main types of species distribution models (SDM) are used
to project species range shifts in future climatic conditions:
correlative and process-based models
(\protect\hyperlink{ref-Dormann2012}{Dormann \emph{et al.} 2012}). The
vast majority of currently used SDMs are correlative: they seek to find
statistical relationships between various environmental descriptors and
species presence and absence. They assume there is an equilibrium
between species distribution and environment (equilibrium postulate,
\protect\hyperlink{ref-Guisan2005}{Guisan \& Thuiller 2005}), that there
is no adaptive responses within a generation (no trait plasticity,
\protect\hyperlink{ref-Berzaghi2020}{Berzaghi \emph{et al.} 2020}), and
that species niche is stable over macroevolutionary time (niche
conservatism, \protect\hyperlink{ref-Pearman2008}{Pearman \emph{et al.}
2008}). Most of them include a fairly large number of predictors
(particularly in machine-learning approaches), and consider flexible
transformations (linear, quadratic\ldots) and interactions between them
(\protect\hyperlink{ref-Merow2014}{Merow \emph{et al.} 2014}). Even
though some authors advocate for ``putting more biology into SDMs''
(\protect\hyperlink{ref-Higgins2012}{Higgins \emph{et al.} 2012}),
parameters have no a priori defined ecological meaning
(\protect\hyperlink{ref-Dormann2012}{Dormann \emph{et al.} 2012}) and
shape of response curves to environmental variables is generally not
constrained based on biological considerations. Although these models
are not always used correctly (\protect\hyperlink{ref-Araujo2019}{Araújo
\emph{et al.} 2019}; \protect\hyperlink{ref-Santini2021}{Santini
\emph{et al.} 2021}), their flexibility makes them an important tool in
predictive ecology (\protect\hyperlink{ref-Mouquet2015}{Mouquet \emph{et
al.} 2015}). They have been widely used especially to generate species
range projections under current and future climates (e.g.
\protect\hyperlink{ref-Guisan2005}{Guisan \& Thuiller 2005}).
Nevertheless, their ability to accurately describe the effects of
climate on species distributions has recently been questioned (e.g.
\protect\hyperlink{ref-Fourcade2018}{Fourcade \emph{et al.} 2018};
\protect\hyperlink{ref-Journe2020}{Journé \emph{et al.} 2020};
\protect\hyperlink{ref-Warren2021}{Warren \emph{et al.} 2021}).\\
For all these reasons, another kind of models has been developed.
Process-based models aim to translate into mathematical equations our
knowledge about the physiological and ecological processes involved in
an organism's life, such as growth, reproduction, survival, movement,
and interactions with other livebeings. Process-based models take more
time to develop and are more challenging to use, but they might provide
a greater comprehension of the complexity of ecosystem dynamics and more
robust projections in novel conditions
(\protect\hyperlink{ref-Evans2012}{Evans 2012};
\protect\hyperlink{ref-Zurell2016}{Zurell \emph{et al.} 2016};
\protect\hyperlink{ref-Singer2016}{Singer \emph{et al.} 2016};
\protect\hyperlink{ref-Urban2016}{Urban \emph{et al.} 2016}). A wide
variety of process-based models exists, from quite simple models (e.g.
\protect\hyperlink{ref-Kleidon2000}{Kleidon \& Mooney 2000}) to much
more complex ones (e.g. \protect\hyperlink{ref-Dufrene2005}{Dufrêne
\emph{et al.} 2005}). They all rely on an explicit representation of
processes at stake, with a direct biological interpretation - at least
in principle (\protect\hyperlink{ref-Connolly2017}{Connolly \emph{et
al.} 2017}). Process-based models may also include some phenomenological
relationships on lower-level processes, but never on the pattern itself.
The choices about the specific processes to include into the model are
made based on theory, empirical observations and the objectives of the
research, and modeler subjectivity may play an important role. One of
the challenges is to build a model with the appropriate amount of
complexity: a too simplistic model might be unrealistic whereas a very
complex model could be far beyond our ability to understand it (because
of interconnected mechanisms) and calibrate it. Each model relies on
different hypotheses with its own balance of complexity, accuracy and
parsimony - and thus different numbers of unknown parameters to
calibrate. Parameter values of this kind of models are obtained with
different methods. Some of these parameters can be individually
estimated with field observations or experimental data, or are already
available in the literature. When this is not possible, groups of
parameters defining a relationship between a process variable and
environmental variables or other processes are estimated jointly using
inverse modelling methods and data on the processes modelled (e.g.
\protect\hyperlink{ref-Cailleret2020}{Cailleret \emph{et al.} 2020};
\protect\hyperlink{ref-Asse2020}{Asse \emph{et al.} 2020}).

Calibration (i.e.~parameter setting and estimation) is a fundamental
step in the modelling process. It allows the model to reproduce the
reality with more or less success. The result of the calibration
provides insights on the ability of the model to reproduce and explain
reality (model predictive power). Calibration of complex models such as
process-based SDMs is time-consuming, and modelers are often challenged
by the dimension of the parameter space, the complexity of the possible
correlations among parameters, and the scarcity of observed experimental
data to calibrate them. Therefore, although the philosophy of such
models is to measure parameters, statistical inference might be useful
when data are not yet available to infer parameter values. Parameter
inference can be achieved through several methods which have been
developed in the last decades.
\textcolor{darkgreen}{Most of them fall into two categories: informal and statistical calibration. On the one hand, statistical calibration assumes a data-generating model, and the likelihood function, which quantifies the probability of the observed data given the model parameters, is expressed mathematically. This likelihood is the foundation of maximum likelihood estimation and Bayesian inference, two commonly used methods for parameter estimation. In practice, deriving the likelihood function can be quite challenging for complex models with many parameters. One approach is to use numerical methods such as simulation-based inference or Approximate Bayesian Computation}
(e.g. \protect\hyperlink{ref-Hartig2014}{Hartig \emph{et al.}
2014})\textcolor{darkgreen}{. However, obtaining a reliable estimate of the likelihood of a complex process-based model can require a large number of simulated samples, which can be computationally expensive. On the other hand, using an informal objective function to measure the discrepancy between the model predictions and the observed data may be a reasonable alternative, even though it may not have the same statistical rigor. By minimizing the objective function, one is able to identify the parameter values that best fit the observed data. This optimization step can be achieved with many methods, either deterministic (e.g. Nelder-Mead method) or stochastic (e.g. simulated annealing, evolutionary algorithms).}\\
Recently, an approach belonging to the evolutionary algorithm family,
called Covariance Matrix Adaptation Evolution Strategy (CMA-ES), has
been proposed (\protect\hyperlink{ref-Hansen2001}{Hansen \& Ostermeier
2001}). One of the advantages of CMA-ES is its ability to cumulate
information over iterations in order to adapt its own parameters (in
particular the covariance matrix), which makes it more robust to noise.
CMA-ES is especially performant for non-separable problems (i.e.~when
the model parameters are dependent) and large search space. This method
has been successfully applied in various fields such as aerospace (e.g.
\protect\hyperlink{ref-Collange2010}{Collange \emph{et al.} 2010}),
optics (e.g. \protect\hyperlink{ref-Gagne2008}{Gagné \emph{et al.}
2008}), and robotics (e.g. \protect\hyperlink{ref-Hill2020}{Hill
\emph{et al.} 2020}). CMA-ES is acknowledged to be one of the most
efficient approaches in continuous black-box optimization
(\protect\hyperlink{ref-Hansen2010}{Hansen \emph{et al.} 2010}) but to
our knowledge has never been used in ecology.

Here we explored the feasibility and interests of calibrating
process-based SDMs with CMA-ES using species occurrence data as
correlative SDMs do (fitted process-based models \emph{sensu}
\protect\hyperlink{ref-Dormann2012}{Dormann \emph{et al.} 2012}). We
focused on two forest process-based models of varying levels of
complexity to evaluate the ability of CMA-ES to calibrate such models.
The two models are PHENOFIT (27 to 36 parameters,
\protect\hyperlink{ref-Chuine2001}{Chuine \& Beaubien 2001}) and
CASTANEA (77 parameters, \protect\hyperlink{ref-Dufrene2005}{Dufrêne
\emph{et al.} 2005}). Each model also emphasizes different ecological
processes: while PHENOFIT focuses on phenology and how it relates to
survival and reproduction, CASTANEA focuses on carbon and water cycles.
We focused on three European common tree species, with different range
extent and ecological preferences in order to evaluate the algorithms
performance in various geographical and climatic conditions. European
beech (\emph{Fagus sylvatica L.}) is one of the most widely distributed
broadleaved tree in Europe (from southern Sweden to Sicily and from
Spain to northwest Turkey), holm oak (\emph{Quercus ilex L.}) is an
evergreen broadleaved tree native of the Mediterranean region, and
silver fir (\emph{Abies alba Mill.}) is a coniferous tree which mainly
occurs in mountain forests of Central Europe and some parts of Southern
and Eastern Europe.

\hypertarget{methods}{%
\section{1. Material and methods}\label{methods}}

\hypertarget{pbmodels}{%
\subsection{1.1. Process-based models}\label{pbmodels}}

All versions of the models used for this study are coded in Java and
distributed by the \href{https://capsis.cirad.fr/capsis/home}{CAPSIS}
platform.

PHENOFIT is a process-based species distribution model for forest tree
species which focuses on phenology. It relies on the principle that the
distribution of a tree species depends mainly on the synchronization of
its timing of development to the local climatic conditions
(\protect\hyperlink{ref-Chuine2001}{Chuine \& Beaubien 2001}). It is
composed of several submodels, including phenology models for leaves,
flowers and fruits, and stress resistance models. It simulates the
fitness (survival and reproductive success) of an average individual
using daily meteorological data, soil water holding capacity and species
specific parameters (see \protect\hyperlink{appendixA}{Appendix A} for
details). PHENOFIT has been validated for several North American and
European species by comparing their known distribution to the modelled
fitness (e.g. \protect\hyperlink{ref-Morin2007}{Morin \emph{et al.}
2007}; \protect\hyperlink{ref-Saltre2013}{Saltré \emph{et al.} 2013};
\protect\hyperlink{ref-Duputie2015}{Duputié \emph{et al.} 2015};
\protect\hyperlink{ref-Gauzere2020}{Gauzere \emph{et al.} 2020}).

CASTANEA is an ecophysiological process-based model which simulates
carbon and water fluxes in forests
(\protect\hyperlink{ref-Dufrene2005}{Dufrêne \emph{et al.} 2005}). The
model simulates the ecosystem as an average tree with six compartments
(leaves, branches, stem, coarse roots, fine roots and reserves). It is
much more complex than PHENOFIT, with several processes described and
computed, such as photosynthesis, stomatal opening, maintenance and
growth respiration, transpiration, and carbon allocation (see
\protect\hyperlink{appendixA}{Appendix A} for details). CASTANEA
requires daily meteorological variables and soil characteristics. The
model has been initially validated at stand scale for beech
(\protect\hyperlink{ref-Davi2005}{Davi \emph{et al.} 2005}), and was
then successfully applied to other European species (e.g.
\protect\hyperlink{ref-Davi2006}{Davi \emph{et al.} 2006};
\protect\hyperlink{ref-Delpierre2012}{Delpierre \emph{et al.} 2012};
\protect\hyperlink{ref-Davi2017}{Davi \& Cailleret 2017}).

Both models, in their standard version (called here after expert
calibration), are parameterized using various sources of information.
Some parameters are directly measured or found in the literature such as
leaf life span, specific leaf area, LT50 (freezing temperature causing
50\% mortality) of leaves, leaf reflectance, and so on. Other parameters
cannot be measured at all, or their measurement require an enormous
effort that cannot be deployed for a large number of species. These
parameters are thus inferred by inverse modelling using either Bayesian
methods or optimizing methods and data on the specific process they are
describing. This is for example the case for phenology models for which
all parameters cannot be measured, especially those describing bud
dormancy break regulation, since no method allow to measure dormancy
break precisely so far (except for a few fruit tree species). Finally, a
few parameters are prescribed based on expert knowledge as no data to
estimate them exist. For this study, the standard versions of both
models were run for the three species using the same climatic data used
to do the inverse calibration (see
\protect\hyperlink{climatedata}{1.2.1. Climate data}), in order to
compare the correctness of the predictions obtained with the two types
of calibration.

\hypertarget{data}{%
\subsection{1.2. Data for the calibration}\label{data}}

\hypertarget{climatedata}{%
\subsubsection{1.2.1. Climate and soil data}\label{climatedata}}

Raw climatic variables were extracted from ERA5-Land hourly dataset
(\protect\hyperlink{ref-Sabater2019}{Muñoz Sabater 2019},
\protect\hyperlink{ref-Sabater2021}{2021}) from 1970 to 2000, at a
spatial resolution of 0.1 degree in latitude and longitude. We
calculated the daily mean values of the following variables used by
PHENOFIT and CASTANEA: minimum, mean and maximum daily temperatures,
mean dewpoint temperature, daily precipitation, daily global radiation
and daily mean wind speed. We computed the daily relative humidity with
the ratio of vapor pressure and saturation vapor pressure (both
calculated with Clausius-Clapeyron equation) using \emph{humidity} R
package (\protect\hyperlink{ref-Cai2019}{Cai 2019}). Daily potential
evapotranspiration was calculated with Penman--Monteith equation (FAO
standard of hypothetical grass reference surface) using a slightly
modified version of the \emph{ET()} function in
\emph{Evapotranspiration} R package (\protect\hyperlink{ref-Guo2016}{Guo
\emph{et al.} 2016}).

Water content at field capacity and wilting point data were extracted
from EU-SoilHydroGrids (\protect\hyperlink{ref-Toth2017}{Tóth \emph{et
al.} 2017}) which is at 1km resolution. Percentage of sand, silt and
clay particles, percentage of coarse fragments, bulk density and soil
depth were extracted from SoilGrids250m
(\protect\hyperlink{ref-Hengl2017}{Hengl \emph{et al.} 2017}) at a 250m
resolution. These data (except for soil depth) are provided at seven
soil depths, so we summarized them (weighted sum or weighted mean)
taking into account each layer width and total soil depth. Finally, all
variables were upscaled at the ERA5-Land spatial resolution 0.1° using
bilinear interpolation.

\hypertarget{occurrencedata}{%
\subsubsection{1.2.2. Tree occurrence data used for the
calibration}\label{occurrencedata}}

Sources of occurrence data are known to differ even for common European
trees (\protect\hyperlink{ref-Duputie2014}{Duputié \emph{et al.} 2014})
and this makes it quite challenging to gather comprehensive data at a
sufficient spatial resolution all over Europe. The occurrence data we
used essentially rely on the EU-Forest dataset
(\protect\hyperlink{ref-Mauri2017}{Mauri \emph{et al.} 2017}) which
benefits from inventory and monitoring programmes implemented in most
European countries. As EU-Forest is limited to forest ecosystems, we
combined it with presence records extracted from the Global Biodiversity
Information Facility (\protect\hyperlink{ref-GBIF2022}{GBIF 2022}, see
\protect\hyperlink{appendixB}{Appendix B} for all download links) but
removing observations outside natural species ranges as defined by Atlas
Flora Europeae (AFE, \protect\hyperlink{ref-AFE2005}{Jalas \& Suominen
1972--2005}) and EuroVegMap (\protect\hyperlink{ref-EVM2003}{Bohn
\emph{et al.} 2003}). By doing so, we also included occurrences of
isolated native trees living outside forests, excluding records from
arboreta or gardens where the species would have been planted as an
exotic. For holm oak, we also added occurrence records in the
Mediterranean Basin from the WOODIV database
(\protect\hyperlink{ref-Monnet2021}{Monnet \emph{et al.} 2021}), leaving
out EU-Forest and GBIF records we had already gathered. We upscaled all
species records at the ERA5-Land resolution (0.1°, see
\protect\hyperlink{climatedata}{1.2.1. Climate data}), i.e.~the species
is considered to be present in the cell if there is at least one record.
We finally obtained 21458 occurrence cells for beech, 6653 for holm oak
and 5385 for silver fir (see \protect\hyperlink{appendixB}{Appendix B}
for details).

All the datasets described above are presence-only data. Therefore, we
generated cells where species are supposed to be absent,
i.e.~pseudo-absence cells. In order to avoid as far as possible creating
false absence data, we used EU-Forest cells where the species is not
reported present as pseudo-absence cells. We assumed that national
forest inventories were exhaustive (which is not true since only
specific forest plots in a 0.1° cell are monitored). We obtained 25423
absence cells for beech, 37931 for holm oak and 38365 for silver fir
(see \protect\hyperlink{appendixC}{Appendix C}).

We selected subsets of 2000 points (1000 presences and 1000
pseudo-absences) in order to reduce computational costs. For each
species, we generated ten presence clusters (k-means algorithm) of
similar bioclimatic conditions based on annual climate normals computed
with R package \emph{dismo} (\protect\hyperlink{ref-Hijmans2021}{Hijmans
\emph{et al.} 2021}) and ERA5-Land variables. In each cluster, we
randomly sampled a number of cells where the species is present
proportional to the total number of a number of cells where the species
is present in the cluster. The aim of this stratified random sampling
was to make sure that all species environmental preferences were
proportionally represented. We then randomly sampled the same number of
pseudo-absence cells (see \protect\hyperlink{appendixB}{Appendix B} for
details).

Regarding PHENOFIT model, we calibrated ten times each species parameter
set, with 5 repetitions on 2 random subsets of
presences/pseudo-absences, except for beech. In the latter case, we ran
10 repetitions on 10 subsets (i.e.~100 calibrations) to investigate both
the effect of subsampling and the effect of stochasticity. Since
CASTANEA computing time was much higher (see \autoref{tab:modelstable}),
we ran only two calibration for each species (on 2 different random
subsets).

\hypertarget{model-calibration}{%
\subsection{1.3. Model calibration}\label{model-calibration}}

\hypertarget{covariance-matrix-adaptation-evolution-strategy-principles}{%
\subsubsection{1.3.1. Covariance Matrix Adaptation Evolution Strategy
principles}\label{covariance-matrix-adaptation-evolution-strategy-principles}}

Covariance Matrix Adaptation Evolution Strategy (CMA-ES) is widely
accepted as a robust optimization algorithm for non-linear, non-convex,
as well as non-separated optimization problems in continuous domain
(\protect\hyperlink{ref-Hansen1996}{Hansen \& Ostermeier 1996};
\protect\hyperlink{ref-Hansen2001}{Hansen \& Ostermeier 2001};
\protect\hyperlink{ref-Hansen2006}{Hansen 2006}). It is based on the
principle of evolutionary biology, via recombination, mutation and
selection of the most fit candidate solutions (i.e.~parameter sets
providing the best predictions). At each iteration:\\
- \(\lambda\) candidate solutions are evaluated, i.e.~model runs
\(\lambda\) times with \(\lambda\) different parameter sets and the
objective function is evaluated\\
- the best \(\mu\) candidate solutionss are selected\\
- the weighted mean candidate solution \(m\) is computed (mean of the
best \(\mu\) parameter sets weighted by their objective function value)
- covariance matrix \(C\) and step size \(\sigma\) are updated (with
information accumulated over several consecutive iterations)\\
- new \(\lambda\) candidate solutions are sampled in a normal
distribution \(\mathcal{N}(m, \sigma C)\), with both recombination (via
the favorite solution \(m\)) and mutations (via the perturbations
\(\sigma C\))\\
One of the strength of this approach lies in the combination of
rank-\(\mu\)-update, where prior information from previous generations
is exploited (mean of the previous covariance matrices, with a higher
weight for recent generations), and cumulation, where correlations
between generations are retained in an evolution path (sum of
consecutive steps), to update the covariance matrix at each step (see
\protect\hyperlink{ref-Hansen2016}{Hansen 2016} for a detailed
description of the algorithm).

\hypertarget{cma-es-in-practice}{%
\subsubsection{1.3.2. CMA-ES in practice}\label{cma-es-in-practice}}

One of the advantages of CMA-ES is that it does not require a complex
parameter tuning: as best parameter values at a given time of the
optimization process might no longer be efficient later, CMA-ES
implements an internal adaptation of its parameters. We only chose the
number of candidate solutions \(\lambda\), depending on the optimization
problem complexity (\(\mu\) was set to \(\lambda/2\)). The default
recommended value for \(\lambda\) is \(4 + 3 ln(N)\), where \(N\) is the
number of parameters to calibrate (i.e.~\(\lambda \in [14,17]\) in our
case). We set \(\lambda = 20\), in order to improve the global search
capability (\protect\hyperlink{ref-Hansen2004}{Hansen \& Kern 2004}) and
take advantage of the computation power at our disposal. All model
parameters were linear scaled into \([0;10]\) so that the same standard
deviation can be applied to all parameters: here we chose \(\sigma = 2\)
(see
\href{https://cma-es.github.io/cmaes_sourcecode_page.html\#practical}{Nikolaus
Hansen personal website} for practical hints on variable encoding). Our
stopping criterion for the optimization procedure was the budget,
i.e.~the number of model runs.

For an easier use and the sake of reproducibility, we chose to use a
pure R implementation of CMA-ES available in the R package \emph{cmaes}
(\protect\hyperlink{ref-Trautmann2011}{Trautmann \emph{et al.} 2011}).
The function \emph{cma\_es()} enables us to do \(\lambda\) function
evaluations in parallel so as to substantially reduce computation time.
It also allows us to define lower and upper bound constraints, by
penalising the objective function value of the candidate solution if it
violates the boundaries. We customized the \emph{cma\_es()} function to
add an option to define death penalty constraints (rejection of the
infeasible candidate solution who is sampled again), in order to define
a range of ecologically possible solutions in terms of inequality
constraints between parameters (see
\protect\hyperlink{appendixD}{Appendix D} for details about boundaries
and constraints handling). Death penalty is the easiest way to handle
constraints when the feasible region is fairly large, but it is not
perfect as there is no use of information from infeasible points
(i.e.~points which violate the constraints).

The objective function for the calibration was the area under the
receiver operating characteristic curve (AUC), evaluated against a
subsets of 2000 points (see \protect\hyperlink{occurrencedata}{1.2.2.
Tree occurrence data}). Although AUC has been criticized as an imperfect
measure of model performance (\protect\hyperlink{ref-Lobo2008}{Lobo
\emph{et al.} 2008}; \protect\hyperlink{ref-Leroy2018}{Leroy \emph{et
al.} 2018}), we used it as objective function because our goal here was
only to calibrate models by maximizing discriminating capacity
(i.e.~potential to correctly classify presences and absences) with a
threshold-independent measure. We used the \emph{AUC} R package
(\protect\hyperlink{ref-Ballings2013}{Ballings \& Van den Poel 2013}),
and chose the two following model output variables as proxies of
classification probabilities (i.e.~used to determine if the species can
be present or not): fitness index for PHENOFIT and carbon reserves for
CASTANEA (see \protect\hyperlink{appendixA}{Appendix A}).

We implemented the CMA-ES calibration on two computing clusters:
GenOuest from IRISA-INRIA
(\href{https://www.genouest.org}{genouest.org}) and TGCC (\emph{Très
Grand Centre de Calcul}) from CEA
(\href{http://www-hpc.cea.fr/fr/complexe/tgcc.htm}{hpc.cea.fr}). As the
models are coded in Java (see \protect\hyperlink{pbmodels}{1.1
Process-based models}), they need a process of deallocating memory
handled by a \emph{garbage collector}. For PHENOFIT, each function
evaluation (i.e.~each model simulation) was run on a 2-core computing
unit in order to have enough computing resources for both simulation and
garbage collection. We thus needed twice as many cores as functions
evaluated in parallel. CASTANEA model requires a fairly high computation
time, so we used a nested parallelism distribution, where each parallel
simulation was distributed on 4 computing units. We thus used 4 times as
many cores as functions evaluated in parallel, plus some extra cores for
garbage collection. We used R package \emph{future}
(\protect\hyperlink{ref-Bengtsson2021}{Bengtsson 2021}) for parallel
processing.

\begin{table}

\caption{\label{tab:modelstable}Summary of model calibration settings. Average runtime was assessed on the GenOuest cluster.}
\centering
\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
\toprule
Model & Output variable of interest & Number of parameters calibrated & Number of candidate solutions $\lambda$ & Number of cores & Total memory & Number of model evaluations & Average runtime for one calibration\\
\midrule
PHENOFIT & Fitness index & $[27;36]$ & $20$ & $40$ & $80$ GB & 6000 & $\sim 24$ hours\\
\addlinespace
CASTANEA & Carbon reserves & $77$ & $20$ & $100$ & $120$ GB & 4000 & $\sim 20$ days\\
\bottomrule
\end{tabu}
\end{table}

\hypertarget{results}{%
\section{2. Results}\label{results}}

\hypertarget{calibration-results}{%
\subsection{2.1. Calibration results}\label{calibration-results}}

Calibrations using species distribution data are hereafter called
inverse calibrations, and calibrations based on expert knowledge,
observations and measurements of the processes modelled are called
expert calibrations.

CMA-ES calibration of PHENOFIT model allows an average 17.2\% increase
of AUC across the three species compared to expert calibration
(\hyperref[fig:phenofitmaps]{Figure 1}). The maximum increase is
obtained for silver fir, from 0.72 to 0.9 (25\%).\\
CMA-ES calibration of CASTANEA allows an average 23.7 \% increase of AUC
compared to expert calibration (\hyperref[fig:castaneamaps]{Figure 2}),
and a maximum increase obtained for holm oak (34.7\%).

\begin{figure}[H]

{\centering \includegraphics{figs/phenofitmaps}

}

\caption{Species distribution maps obtained with PHENOFIT expert and inverse calibrations, compared with observed species occurrences. Optimal threshold to dichotomize model predicted fitness index in presence/absence is the Youden index-based cut-off point. Note that models predict species climatic niche which is larger than the realized niche that corresponds to species presence map.}\label{fig:phenofitmaps}
\end{figure}

\begin{figure}[H]

{\centering \includegraphics{figs/castaneamaps}

}

\caption{Species distribution maps obtained with CASTANEA expert and inverse calibrations, compared with observed species occurrences. Optimal threshold to dichotomize model predicted carbon reserves in presence/absence is the Youden index-based cut-off point. Note that models predict species climatic niche which is larger than the realized niche that corresponds to species presence map. Note also that CASTANEA cannot be used in high-latitude regions (grey area).}\label{fig:castaneamaps}
\end{figure}

\hypertarget{impacts-of-subsampling-and-calibration-stochasticity}{%
\subsection{2.2. Impacts of subsampling and calibration
stochasticity}\label{impacts-of-subsampling-and-calibration-stochasticity}}

\hypertarget{variability-of-calibration-performance}{%
\subsubsection{2.2.1. Variability of calibration
performance}\label{variability-of-calibration-performance}}

\begin{figure}[H]

{\centering \includegraphics{figs/cmaesrepAUCcal}

}

\caption{\label{fig:cmaesrepAUCcal} Effects of data sub-sampling and stochasticity on CMA-ES calibration using the PHENOFIT model for beech: \textbf{(A)} calibration AUC (calculated only with calibration cells) and \textbf{(B)} total AUC (calculated with all presence/absence cells). Each color is a different sub-sampling of occurrence data, each point is a calibration run. Diamonds (with black border) are mean AUC values. On \textbf{(A)}, the grouping letters represent the multiple comparisons with pairwise Dunn’s tests.}\label{fig:cmaesrepAUCcal}
\end{figure}

The 100 calibrations of the PHENOFIT model realized for beech showed
that random data subsampling had an effect on the final objective
function value (i.e.~the AUC computed on the 2000 calibration points).
Kruskal-Wallis test was significant (p = 2.1e-08), meaning that at least
one subset provided better AUC during calibration. According to Dunn's
tests, 11 pairwise comparisons out of 45 were significant
(\hyperref[fig:cmaesrepAUCcal]{Figure 3.A.}). The calibration AUC ranged
from 0.879 to 0.923 over all subsets, with a mean value of 0.9.

However, more importantly, the repetition of calibrations on different
subsets had no significant impact on the total AUC computed on all
presence/absence points (Kruskal-Wallis test, p = 0.96). Thus, no subset
led to an overall better prediction of the species distribution (see
\hyperref[fig:cmaesrepAUCcal]{Figure 3.B.}). The total AUC ranged from
0.881 to 0.914, with a mean value of 0.896. \newline

\hypertarget{non-identifiability-of-parameters}{%
\subsubsection{2.2.2. Non-identifiability of
parameters}\label{non-identifiability-of-parameters}}

To illustrate the variability in the parameter estimates that can be
obtained with the inverse calibration, we focused on the leaf unfolding
date submodel of PHENOFIT (see \protect\hyperlink{appendixA}{Appendix A}
and \protect\hyperlink{appendixG}{Appendix G}). The parameter values
found by CMA-ES varied greatly across the 100 calibrations
(\hyperref[fig:unfoldingplots]{Figure 4}). For example, the critical
amount of chilling \(C_{crit}\) required to break bud dormancy and the
critical amount of forcing \(F_{crit}\) required to break bud ranged
from 1.02 to 149.96 and from 1.5 to 79.26 respectively, with a mean
value of 51.52 and 38.78. Their coefficient of variations were 126.7\%
and 51.9\% respectively. Kendall correlation coefficient between
\(C_{crit}\) and the threshold temperature of the response function to
temperature during dormancy \(T_b\) is 0.64 (\(p <\) 0.001). Kendall
correlation coefficient between \(F_{crit}\) and the mid-response
temperature \(T_{50}\) is -0.55 (\(p <\) 0.001).

\begin{figure}[H]

{\centering \includegraphics{figs/unfoldingplots}

}

\caption{Effects of stochasticity of CMA-ES calibrationd on PHENOFIT leaf unfolding model parameter values for beech. Each panel is a parameter. Y-axis limits are lower and upper bounds used during calibration. Each point is a calibrated parameter value, color gradient is based on $F_{crit}$ values. Red diamonds are parameter values obtained with expert calibration, blue ones are parameter values obtained with the best inverse calibration.}\label{fig:unfoldingplots}
\end{figure}

\hypertarget{discussion}{%
\section{3. Discussion}\label{discussion}}

\textcolor{darkgreen}{Our results showed that CMA-ES is a feasible and efficient optimizer for inverse calibration of complex ecological models. The algorithm was able to find parameter sets that significantly improved the predictions of the calibrated models compared to the expert parametrization.  However, our study also highlighted the issue of non-identifiability of parameter values due to the data limitation and the dependence between process-based model parameters, which may result in diverging parameter values even though the calibrated models describe the observed species distribution well. Nonetheless, our study provides a promising methodological framework for calibrating process-based models using CMA-ES, and for spreading the use of such models in predictive ecology.}

\hypertarget{performance-and-advantages-of-cma-es-to-calibrate-complex-models-in-ecology}{%
\subsection{3.1. Performance and advantages of CMA-ES to calibrate
complex models in
ecology}\label{performance-and-advantages-of-cma-es-to-calibrate-complex-models-in-ecology}}

Here we demonstrated that inverse calibration with CMA-ES is feasible
and provide good results for complex and runtime-expensive ecological
models.\\
With a subsampling of species occurrence data, the algorithm succeeds in
finding parameter sets which provide \textcolor{darkgreen}{higher} AUC
values. The predictions of the calibrated models are sharply improved
compared to expert parametrization
(\hyperref[fig:phenofitmaps]{Figure 1} and
\hyperref[fig:castaneamaps]{Figure 2}). Two striking examples are the
increase in the performance of PHENOFIT model for silver fir, from 0.72
to 0.9, and of CASTANEA model for holm oak, from 0.7 to 0.95. Moreover,
CMA-ES performed equally well regardless of the species occurrence
subset used during calibration
(\hyperref[fig:cmaesrepAUCcal]{Figure 3.B.}), and thus permitted to find
a good compromise between computational cost and calibration efficiency.

CMA-ES is a ``generic'' optimizer which can be applied to various
problems. It is easy to use as it does not require an extensive tuning
to efficiently explore the parameter space. We only had to choose the
number of candidate solutions \(\lambda\), and the initial search region
(initial starting point and step size \(\sigma\)). As well as being
quasi parameter-free, CMA-ES has several structural advantages
particularly useful for complex optimization problems.
\textcolor{darkgreen}{Firstly, the algorithm's covariance matrix enables the learning of second-order information, which provides insights into pairwise dependencies between parameters. Secondly, the covariance matrix adaptation of CMA-ES is highly efficient in handling ill-conditioned and non-separable problems }(\protect\hyperlink{ref-Hansen2011}{Hansen
\emph{et al.}
2011})\textcolor{darkgreen}{. Lastly, CMA-ES's update mechanism of step size $\sigma$ (i.e. the mutation force) helps prevent premature convergence }(\protect\hyperlink{ref-Hansen2001}{Hansen
\& Ostermeier
2001})\textcolor{darkgreen}{, allowing the algorithm to explore more of the search space.}
CMA-ES has been shown to outperform several other optimization
algorithms (\protect\hyperlink{ref-Hansen2010}{Hansen \emph{et al.}
2010}), and is usually the most efficient method when the target cost
(i.e.~the number of objective function evaluations) is about \(100*N\)
(\(N\) being the dimension of the parameter search space,
\protect\hyperlink{ref-Baeck2013}{Bäck \emph{et al.} 2013}).

\hypertarget{non-identifiability-of-parameter-values}{%
\subsection{3.2. Non-identifiability of parameter
values}\label{non-identifiability-of-parameter-values}}

\textcolor{darkgreen}{Equifinality is a common problem encountered during model calibration, where multiple parameter sets can produce equally good fits to the observed data. This issue can arise due to several factors, including limitations in the quantity or quality of available data, competing processes within the model, and parameter interactions that affect the model output in complex ways.}

In both models used in this study, biological mechanisms are explicitly
calculated in several submodels (e.g.~a leaf unfolding submodel or a
stomatal opening submodel). A submodel output has inevitably a
significant influence on the other submodels as biological processes can
be highly dependent with feedbacks: in CASTANEA, for example, the
stomatal opening affects the photosynthesis, and \emph{vice versa}.
Within each submodel, parameters are also strongly dependent because of
structural correlations. To illustrate this problem, we focused on the
beech leaf unfolding submodel of PHENOFIT (see
\protect\hyperlink{appendixA}{Appendix A}). This model has 6 parameters
(\protect\hyperlink{ref-Chuine2000}{Chuine 2000}): a starting date of
the processes (\(t_0\)), one parameter describing the response function
to temperature during the dormancy phase (\(T_b\)), two parameters
describing the response function to temperature during the phase of bud
growth (\(d_T\), \(T_{50}\)), and two parameters representing the sums
of the daily responses to temperature during bud dormancy (\(C_{crit}\))
and during bud growth (\(F_{crit}\)) that respectively determine the
date of bud dormancy break and the date of leaf unfolding (see
\protect\hyperlink{appendixG}{Appendix G} for details). Since no
information on the date of bud dormancy break is available for the
calibration, a first structural negative correlation exists between
\(C_{crit}\) and \(F_{crit}\): the same leaf unfolding date can be
obtained with either a long dormancy phase and short bud growth phase or
a short dormancy phase and a long bud growth phase. Other structural
correlations exist between \(T_b\) and \(C_{crit}\) on the one hand and
\(d_T\)/\(T_{50}\) and \(F_{crit}\) on the other hand: for example, a
rapid accumulation of chilling units with a high critical chilling
requirement could yield identical results as a slow accumulation with a
low critical chilling requirement (i.e.~the threshold temperature
\(T_{b}\) and the critical chilling requirement \(C_{crit}\) are
dependent, see Figure G.1.A. in \protect\hyperlink{appendixG}{Appendix
G}).

Consequently, several parameter sets may be statistically equivalent and
parameters non-identifiable. In fact, calibration repetitions gave
diverging parameter values (\hyperref[fig:unfoldingplots]{Figure 4})
while being efficient in distinguishing between species presence and
absence (i.e.~AUC \textasciitilde{} 0.9,
\hyperref[fig:cmaesrepAUCcal]{Figure 3}). Thus, even if the calibrated
model describes the observed species distribution very well, it does not
necessarily mean that parameter values are ecologically relevant. This
concern is similar to the criticisms against correlative SDMs, in which
parameter values and correlations that well reproduce species ranges do
not necessarily describe a complex biological reality. In our case, the
constraints imposed by the explicit mathematical equations embedded in
the models we used were not sufficient to ensure calibration convergence
towards similar solutions that would have suggested a high biological
realism. However, it is worth noting that we deliberately chose large
parameter ranges (although biologically realistic, i.e.~corresponding to
the observations made on the different processes modelled across
different species) in order to give free rein to the optimization
algorithm. As our goal was to assess the performance of CMA-ES
objectively, we did not attempt to minimize this non-identifiability
issue by restricting the parameter space. To deal with equifinality
issues, an avenue to explore could be the use of multiple objective
functions during model calibration to assess different aspects of the
model performance.
\textcolor{darkgreen}{Additionally, if a closed-form likelihood can be derived, one could use a Bayesian framework to combine prior knowledge and inverse estimation of parameters to constrain the parameter space and study the nature of trade-offs between parameters}
(\protect\hyperlink{ref-Hartig2012}{Hartig \emph{et al.} 2012};
\protect\hyperlink{ref-Cailleret2020}{Cailleret \emph{et al.} 2020}).

\hypertarget{methodological-issues-and-perspectives}{%
\subsection{3.3. Methodological issues and
perspectives}\label{methodological-issues-and-perspectives}}

Our goal here was to investigate the performance of CMA-ES to calibrate
quite complex process-based species distribution models using species
occurrence data. We did not attempt to validate our parametrizations
using temporally or spatially independent data, and AUC was only used to
determine if model outputs were consistent with species distributions.
However, AUC is scale-invariant (it measures how well predictions are
ranked rather than their absolute values), and prevents us from
obtaining calibrated outputs with more ecological meaning. For example,
in PHENOFIT, one species with a fitness of \(0.8\) could be considered
as absent whereas another one as present. Further work could thus be
conducted to examine the effects of choosing a different objective
function.

It would also be valuable to use a significantly higher computing power,
with an adapted version of CMA-ES. To improve the global search
performance of CMA-ES, we slightly increased the number of candidate
solutions \(\lambda\) (\protect\hyperlink{ref-Hansen2004}{Hansen \& Kern
2004}) and used a computing cluster to evaluate \(\lambda\) functions in
parallel. We were able to use between 40 and 120 cores, which is far
from the computing power of some GPUs (\textgreater{} 2000 cores). In
this case, choosing a very large number of candidate solutions might not
be the best choice. To use efficiently this large parallel computing
power, one could rather use a CMA-ES restart strategy (e.g.~IPOP-CMA-ES,
\protect\hyperlink{ref-Auger2005}{Auger \& Hansen 2005}), where the
number of candidate solutions is successively increased (by a factor of
2), and run these calibrations in parallel. Moreover, when a model
requires a high computation time and thus only a small budget can be
afforded, the original fitness function could be approximated with a
surrogate model in order to reduce the number of original function
evaluations required (e.g. \protect\hyperlink{ref-Auger2004}{Auger
\emph{et al.} 2004} ; \protect\hyperlink{ref-Loshchilov2013}{Loshchilov
\emph{et al.} 2013}).

There are several issues regarding the process-based models we used
which can impact their calibration and bias their parameter estimates.
First, like any model, although they have a certain level of complexity,
they are not a perfect representation of the reality, and their inherent
structural errors increase the probability of finding parameter values
that deviate from the true values of the underlying processes (see
\protect\hyperlink{ref-Oberpriller2021}{Oberpriller \emph{et al.}
2021}). Secondly, they do not necessarily include all the environmental
factors at stake. For example, pedologic variable in PHENOFIT only
involve the water holding capacity. In this model, other variables such
as pH or soil texture are not considered. Thirdly, and more importantly,
they are used here to represent the fundamental niche, and to estimate
the potential distribution of the species using pedoclimatic variables.
When calibrated against observed distributions, which represent the
realized niche, they face the same issues as correlative models, and
their parameter estimated can be distorted because compensating for
processes not represented in the model (e.g.~biotic interactions). In
addition, in our case here, land use management probably also play an
important role in shaping tree realized distribution, while not being
addressed in the models. We included GBIF occurrence data, and
especially as much as possible isolated native tree records outside
forests, to help correct this problem, but it is impossible to be
exhaustive at the spatial resolution of 0.1°. At this scale, local
variations of soil characteristics and of competitive interactions among
trees (e.g.~along an altitudinal gradient) can also not be considered.

Finally, several authors advocate for process-based modeling approaches
relying upon species response functions that are a priori defined (e.g.
\protect\hyperlink{ref-Higgins2020}{Higgins \emph{et al.} 2020}).
However, the main limitation of such models is the data availability to
infer their parameters (\protect\hyperlink{ref-Urban2016}{Urban \emph{et
al.} 2016}). Expert parameterization is often long and arduous. One
possible way to facilitate parameter value estimation would be to use
inverse calibration, and we demonstrated here that CMA-ES can be a
powerful optimizer to this end. For example, CMA-ES driven by species
occurrence data could be used to calibrate submodels whose parameter
values cannot be measured or are too hard to measure experimentally.
However, when a structural correlation exists (i.e.~trade-off between
processes that are inherently present and interconnected in the model,
as in the leaf unfolding submodel), inverse calibration might not
provide the right parameter estimates. In such a case, expert knowledge,
observations and measurements are necessary to determine \emph{a
posteriori} which estimates are the most realistic. This is possible in
the case of process-based SDMs, and usually not feasable in the case of
correlative SDMs. A combination of both expert and inverse calibrations
might offer a new perspective for spreading the use of process-based
models in predictive ecology, especially for climate change impact
studies.

\hypertarget{acknowledgements}{%
\section{Acknowledgements}\label{acknowledgements}}

The authors would like to thank Hendrik Davi for helping us in using the
CASTANEA model. We are also deeply grateful for many helpful comments
from Florence Tauc. We would also like to thank François de Coligny,
manager of the CAPSIS platform, Gilles Le Moguedec, and the GenOuest and
TGCC teams for their support. Finally, we would like to thank Florian
Hartig and another anonymous reviewer whose comments and suggestions
helped us improve and clarify this manuscript.\\
V.V. was supported by a GAIA doctoral school PhD Fellowship.

\hypertarget{conflict-of-interest}{%
\section{Conflict of interest}\label{conflict-of-interest}}

The authors have no conflicts of interest to declare.

\hypertarget{author-contributions}{%
\section{Author contributions}\label{author-contributions}}

I.C. devised the main conceptual ideas. V.V. worked out the technical
details, performed the numerical calculations and wrote the first draft
of the manuscript. The two authors discussed the analyses and the
results, and contributed to the final manuscript.

\hypertarget{data-availability}{%
\section{Data availability}\label{data-availability}}

ERA5-Land dataset is available on the
\href{https://cds.climate.copernicus.eu/cdsapp\#!/dataset/reanalysis-era5-land?tab=overview}{Copernicus
Climate Change Service website}. EU-SoilHydroGrids is available on the
\href{https://esdac.jrc.ec.europa.eu/content/3d-soil-hydraulic-database-europe-1-km-and-250-m-resolution}{European
Soil Data Centre website}. SoilGrids250m is available on the
\href{https://www.isric.org/explore/soilgrids}{International Soil
Reference and Information Centre website}. EU-Forest database is
available on
\href{https://figshare.com/collections/A_high-resolution_pan-European_tree_occurrence_dataset/3288407}{FigShare}.
The R code associated with this work will be made available on GitLab
upon acceptance of the manuscript.

\hypertarget{references}{%
\section{References}\label{references}}

\setlength{\parindent}{-0.2in}
\setlength{\leftskip}{0.2in}
\setlength{\parskip}{8pt}
\vspace*{-0.2in}

\noindent

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-Araujo2019}{}}%
Araújo, M.B., Anderson, R.P., Márcia Barbosa, A., Beale, C.M., Dormann,
C.F., Early, R., Garcia, R.A., Guisan, A., Maiorano, L., Naimi, B.,
O'Hara, R.B., Zimmermann, N.E. \& Rahbek, C. (2019). Standards for
distribution models in biodiversity assessments. \emph{Science
Advances}, \textbf{5}.
\href{https://doi.org/10.1126/sciadv.aat4858}{10.1126/sciadv.aat4858}

\leavevmode\vadjust pre{\hypertarget{ref-Asse2020}{}}%
Asse, D., Randin, C.F., Bonhomme, M., Delestrade, A. \& Chuine, I.
(2020). Process-based models outcompete correlative models in projecting
spring phenology of trees in a future warmer climate. \emph{Agricultural
and Forest Meteorology}, \textbf{285-286}, 107931.
\href{https://doi.org/10.1016/j.agrformet.2020.107931}{10.1016/j.agrformet.2020.107931}

\leavevmode\vadjust pre{\hypertarget{ref-Auger2005}{}}%
Auger, A. \& Hansen, N. (2005). A restart {CMA} evolution strategy with
increasing population size. \emph{2005 {IEEE} {Congress} on
{Evolutionary} {Computation}}, pp. 1769--1776 Vol. 2.

\leavevmode\vadjust pre{\hypertarget{ref-Auger2004}{}}%
Auger, A., Schoenauer, M. \& Vanhaecke, N. (2004). {LS}-{CMA}-{ES}: {A}
{Second}-{Order} {Algorithm} for {Covariance} {Matrix} {Adaptation}.
\emph{Parallel {Problem} {Solving} from {Nature} - {PPSN} {VIII}} (eds
X. Yao, E.K. Burke, J.A. Lozano, J. Smith, J.J. Merelo-Guervós, J.A.
Bullinaria, J.E. Rowe, P. Tino, A. Kabán \& H.-P. Schwefel), pp.
182--191. Lecture {Notes} in {Computer} {Science}. Springer, Berlin,
Heidelberg.

\leavevmode\vadjust pre{\hypertarget{ref-Baeck2013}{}}%
Bäck, T., Foussette, C. \& Krause, P. (2013). Empirical {Analysis}. (eds
T. Bäck, C. Foussette \& P. Krause), pp. 55--83. Natural {Computing}
{Series}. Springer, Berlin, Heidelberg.

\leavevmode\vadjust pre{\hypertarget{ref-Ballings2013}{}}%
Ballings, M. \& Van den Poel, D. (2013). \emph{AUC: Threshold
independent performance measures for probabilistic classifiers.}

\leavevmode\vadjust pre{\hypertarget{ref-Barnosky2012}{}}%
Barnosky, A.D., Hadly, E.A., Bascompte, J., Berlow, E.L., Brown, J.H.,
Fortelius, M., Getz, W.M., Harte, J., Hastings, A., Marquet, P.A.,
Martinez, N.D., Mooers, A., Roopnarine, P., Vermeij, G., Williams, J.W.,
Gillespie, R., Kitzes, J., Marshall, C., Matzke, N., Mindell, D.P.,
Revilla, E. \& Smith, A.B. (2012). Approaching a state shift in
{Earth}'s biosphere. \emph{Nature}, \textbf{486}, 52--58.
\href{https://doi.org/10.1038/nature11018}{10.1038/nature11018}

\leavevmode\vadjust pre{\hypertarget{ref-Bengtsson2021}{}}%
Bengtsson, H. (2021). A unifying framework for parallel and distributed
processing in r using futures.

\leavevmode\vadjust pre{\hypertarget{ref-Berzaghi2020}{}}%
Berzaghi, F., Wright, I.J., Kramer, K., Oddou-Muratorio, S., Bohn, F.J.,
Reyer, C.P.O., Sabaté, S., Sanders, T.G.M. \& Hartig, F. (2020). Towards
a {New} {Generation} of {Trait}-{Flexible} {Vegetation} {Models}.
\emph{Trends in Ecology \& Evolution}, \textbf{35}, 191--205.
\href{https://doi.org/10.1016/j.tree.2019.11.006}{10.1016/j.tree.2019.11.006}

\leavevmode\vadjust pre{\hypertarget{ref-EVM2003}{}}%
Bohn, U., Neuhäusl, R., Gisela Gollub, Hettwer, C., Neuhäuslová, Z.,
Raus, T., Schlüter, H. \& Weber, H. (2003). \emph{Map of the natural
vegetation of europe - scale 1:2500000}.

\leavevmode\vadjust pre{\hypertarget{ref-Brook2013}{}}%
Brook, B.W., Ellis, E.C., Perring, M.P., Mackay, A.W. \& Blomqvist, L.
(2013). Does the terrestrial biosphere have planetary tipping points?
\emph{Trends in Ecology \& Evolution}, \textbf{28}, 396--401.
\href{https://doi.org/10.1016/j.tree.2013.01.016}{10.1016/j.tree.2013.01.016}

\leavevmode\vadjust pre{\hypertarget{ref-Cai2019}{}}%
Cai, J. (2019). \emph{Humidity: Calculate water vapor measures from
temperature and dew point}.

\leavevmode\vadjust pre{\hypertarget{ref-Cailleret2020}{}}%
Cailleret, M., Bircher, N., Hartig, F., Hülsmann, L. \& Bugmann, H.
(2020). Bayesian calibration of a growth-dependent tree mortality model
to simulate the dynamics of {European} temperate forests.
\emph{Ecological Applications}, \textbf{30}, e02021.
\href{https://doi.org/10.1002/eap.2021}{10.1002/eap.2021}

\leavevmode\vadjust pre{\hypertarget{ref-Chuine2000}{}}%
Chuine, I. (2000). A {Unified} {Model} for {Budburst} of {Trees}.
\emph{Journal of Theoretical Biology}, \textbf{207}, 337--347.
\href{https://doi.org/10.1006/jtbi.2000.2178}{10.1006/jtbi.2000.2178}

\leavevmode\vadjust pre{\hypertarget{ref-Chuine2001}{}}%
Chuine, I. \& Beaubien, E.G. (2001). Phenology is a major determinant of
tree species range. \emph{Ecology Letters}, \textbf{4}, 500--510.
\href{https://doi.org/10.1046/j.1461-0248.2001.00261.x}{10.1046/j.1461-0248.2001.00261.x}

\leavevmode\vadjust pre{\hypertarget{ref-Collange2010}{}}%
Collange, G., Reynaud, S. \& Hansen, N. (2010). Covariance matrix
adaptation evolution strategy for multidisciplinary optimization of
expendable launcher family. \emph{13th AIAA/ISSMO multidisciplinary
analysis optimization conference}.

\leavevmode\vadjust pre{\hypertarget{ref-Connolly2017}{}}%
Connolly, S.R., Keith, S.A., Colwell, R.K. \& Rahbek, C. (2017).
Process, {Mechanism}, and {Modeling} in {Macroecology}. \emph{Trends in
Ecology \& Evolution}, \textbf{32}, 835--844.
\href{https://doi.org/10.1016/j.tree.2017.08.011}{10.1016/j.tree.2017.08.011}

\leavevmode\vadjust pre{\hypertarget{ref-Davi2017}{}}%
Davi, H. \& Cailleret, M. (2017). Assessing drought-driven mortality
trees with physiological process-based models. \emph{Agricultural and
Forest Meteorology}, \textbf{232}, 279--290.
\href{https://doi.org/10.1016/j.agrformet.2016.08.019}{10.1016/j.agrformet.2016.08.019}

\leavevmode\vadjust pre{\hypertarget{ref-Davi2006}{}}%
Davi, H., Dufrêne, E., Francois, C., Le Maire, G., Loustau, D., Bosc,
A., Rambal, S., Granier, A. \& Moors, E. (2006). Sensitivity of water
and carbon fluxes to climate changes from 1960 to 2100 in {European}
forest ecosystems. \emph{Agricultural and Forest Meteorology},
\textbf{141}, 35--56.
\href{https://doi.org/10.1016/j.agrformet.2006.09.003}{10.1016/j.agrformet.2006.09.003}

\leavevmode\vadjust pre{\hypertarget{ref-Davi2005}{}}%
Davi, H., Dufrêne, E., Granier, A., Le Dantec, V., Barbaroux, C.,
François, C. \& Bréda, N. (2005). Modelling carbon and water cycles in a
beech forest: {Part} {II}.: {Validation} of the main processes from
organ to stand scale. \emph{Ecological Modelling}, \textbf{185},
387--405.
\href{https://doi.org/10.1016/j.ecolmodel.2005.01.003}{10.1016/j.ecolmodel.2005.01.003}

\leavevmode\vadjust pre{\hypertarget{ref-Delpierre2012}{}}%
Delpierre, N., Soudani, K., François, C., Le Maire, G., Bernhofer, C.,
Kutsch, W., Misson, L., Rambal, S., Vesala, T. \& Dufrêne, E. (2012).
Quantifying the influence of climate and biological drivers on the
interannual variability of carbon exchanges in {European} forests
through process-based modelling. \emph{Agricultural and Forest
Meteorology}, \textbf{154-155}, 99--112.
\href{https://doi.org/10.1016/j.agrformet.2011.10.010}{10.1016/j.agrformet.2011.10.010}

\leavevmode\vadjust pre{\hypertarget{ref-Dormann2012}{}}%
Dormann, C.F., Schymanski, S.J., Cabral, J., Chuine, I., Graham, C.,
Hartig, F., Kearney, M., Morin, X., Römermann, C., Schröder, B. \&
Singer, A. (2012). Correlation and process in species distribution
models: Bridging a dichotomy. \emph{Journal of Biogeography},
\textbf{39}, 2119--2131.
\href{https://doi.org/10.1111/j.1365-2699.2011.02659.x}{10.1111/j.1365-2699.2011.02659.x}

\leavevmode\vadjust pre{\hypertarget{ref-Dufrene2005}{}}%
Dufrêne, E., Davi, H., François, C., Maire, G. le, Dantec, V.L. \&
Granier, A. (2005). Modelling carbon and water cycles in a beech forest:
{Part} {I}: {Model} description and uncertainty analysis on modelled
{NEE}. \emph{Ecological Modelling}, \textbf{185}, 407--436.
\href{https://doi.org/10.1016/j.ecolmodel.2005.01.004}{10.1016/j.ecolmodel.2005.01.004}

\leavevmode\vadjust pre{\hypertarget{ref-Duputie2015}{}}%
Duputié, A., Rutschmann, A., Ronce, O. \& Chuine, I. (2015).
Phenological plasticity will not help all species adapt to climate
change. \emph{Global Change Biology}, \textbf{21}, 3062--3073.
\href{https://doi.org/10.1111/gcb.12914}{10.1111/gcb.12914}

\leavevmode\vadjust pre{\hypertarget{ref-Duputie2014}{}}%
Duputié, A., Zimmermann, N.E. \& Chuine, I. (2014). Where are the wild
things? Why we need better data on species distribution. \emph{Global
Ecology and Biogeography}, \textbf{23}, 457--467.
\href{https://doi.org/10.1111/geb.12118}{10.1111/geb.12118}

\leavevmode\vadjust pre{\hypertarget{ref-Evans2012}{}}%
Evans, M.R. (2012). Modelling ecological systems in a changing world.
\emph{Philosophical Transactions of the Royal Society B: Biological
Sciences}, \textbf{367}, 181--190.
\href{https://doi.org/10.1098/rstb.2011.0172}{10.1098/rstb.2011.0172}

\leavevmode\vadjust pre{\hypertarget{ref-Fitzpatrick2018}{}}%
Fitzpatrick, M.C., Blois, J.L., Williams, J.W., Nieto-Lugilde, D.,
Maguire, K.C. \& Lorenz, D.J. (2018). How will climate novelty influence
ecological forecasts? {Using} the {Quaternary} to assess future
reliability. \emph{Global Change Biology}, \textbf{24}, 3575--3586.
\href{https://doi.org/10.1111/gcb.14138}{10.1111/gcb.14138}

\leavevmode\vadjust pre{\hypertarget{ref-Fourcade2018}{}}%
Fourcade, Y., Besnard, A.G. \& Secondi, J. (2018). Paintings predict the
distribution of species, or the challenge of selecting environmental
predictors and evaluation statistics. \emph{Global Ecology and
Biogeography}, \textbf{27}, 245--256.
\href{https://doi.org/10.1111/geb.12684}{10.1111/geb.12684}

\leavevmode\vadjust pre{\hypertarget{ref-Gagne2008}{}}%
Gagné, C., Beaulieu, J., Parizeau, M. \& Thibault, S. (2008).
Human-competitive lens system design with evolution strategies.
\emph{Applied Soft Computing}, \textbf{8}, 1439--1452.
\href{https://doi.org/10.1016/j.asoc.2007.10.018}{10.1016/j.asoc.2007.10.018}

\leavevmode\vadjust pre{\hypertarget{ref-Gauzere2020}{}}%
Gauzere, J., Teuf, B., Davi, H., Chevin, L.-M., Caignard, T., Leys, B.,
Delzon, S., Ronce, O. \& Chuine, I. (2020). Where is the optimum?
{Predicting} the variation of selection along climatic gradients and the
adaptive value of plasticity. {A} case study on tree phenology.
\emph{Evolution Letters}, \textbf{4}, 109--123.
\href{https://doi.org/10.1002/evl3.160}{10.1002/evl3.160}

\leavevmode\vadjust pre{\hypertarget{ref-GBIF2022}{}}%
GBIF. (2022). The global biodiversity information facility.
\url{https://www.gbif.org} {[}accessed 24 January 2022{]}

\leavevmode\vadjust pre{\hypertarget{ref-Guisan2005}{}}%
Guisan, A. \& Thuiller, W. (2005). Predicting species distribution:
Offering more than simple habitat models. \emph{Ecology Letters},
\textbf{8}, 993--1009.
\href{https://doi.org/10.1111/j.1461-0248.2005.00792.x}{10.1111/j.1461-0248.2005.00792.x}

\leavevmode\vadjust pre{\hypertarget{ref-Guo2016}{}}%
Guo, D., Westra, S. \& Maier, H.R. (2016). An {R} package for modelling
actual, potential and reference evapotranspiration. \emph{Environmental
Modelling \& Software}, \textbf{78}, 216--224.
\href{https://doi.org/10.1016/j.envsoft.2015.12.019}{10.1016/j.envsoft.2015.12.019}

\leavevmode\vadjust pre{\hypertarget{ref-Hansen2016}{}}%
Hansen, N. (2016). The CMA evolution strategy: A tutorial.

\leavevmode\vadjust pre{\hypertarget{ref-Hansen2006}{}}%
Hansen, N. (2006). The {CMA} {Evolution} {Strategy}: {A} {Comparing}
{Review}. (eds J.A. Lozano, P. Larrañaga, I. Inza \& E. Bengoetxea), pp.
75--102. Studies in {Fuzziness} and {Soft} {Computing}. Springer,
Berlin, Heidelberg.

\leavevmode\vadjust pre{\hypertarget{ref-Hansen2010}{}}%
Hansen, N., Auger, A., Ros, R., Finck, S. \& Posik, P. (2010).
{Comparing Results of 31 Algorithms from the Black-Box Optimization
Benchmarking BBOB-2009}. \emph{{ACM-GECCO Genetic and Evolutionary
Computation Conference}}. Portland, United States.

\leavevmode\vadjust pre{\hypertarget{ref-Hansen2004}{}}%
Hansen, N. \& Kern, S. (2004). Evaluating the {CMA} {Evolution}
{Strategy} on {Multimodal} {Test} {Functions}. \emph{Parallel {Problem}
{Solving} from {Nature} - {PPSN} {VIII}} (eds X. Yao, E.K. Burke, J.A.
Lozano, J. Smith, J.J. Merelo-Guervós, J.A. Bullinaria, J.E. Rowe, P.
Tino, A. Kabán \& H.-P. Schwefel), pp. 282--291. Lecture {Notes} in
{Computer} {Science}. Springer, Berlin, Heidelberg.

\leavevmode\vadjust pre{\hypertarget{ref-Hansen2009}{}}%
Hansen, N., Niederberger, A.S.P., Guzzella, L. \& Koumoutsakos, P.
(2009). A {Method} for {Handling} {Uncertainty} in {Evolutionary}
{Optimization} {With} an {Application} to {Feedback} {Control} of
{Combustion}. \emph{IEEE Transactions on Evolutionary Computation},
\textbf{13}, 180--197.
\href{https://doi.org/10.1109/TEVC.2008.924423}{10.1109/TEVC.2008.924423}

\leavevmode\vadjust pre{\hypertarget{ref-Hansen1996}{}}%
Hansen, N. \& Ostermeier, A. (1996). Adapting arbitrary normal mutation
distributions in evolution strategies: The covariance matrix adaptation.
\emph{Proceedings of {IEEE} {International} {Conference} on
{Evolutionary} {Computation}}, pp. 312--317.

\leavevmode\vadjust pre{\hypertarget{ref-Hansen2001}{}}%
Hansen, N. \& Ostermeier, A. (2001). Completely {Derandomized}
{Self}-{Adaptation} in {Evolution} {Strategies}. \emph{Evolutionary
Computation}, \textbf{9}, 159--195.
\href{https://doi.org/10.1162/106365601750190398}{10.1162/106365601750190398}

\leavevmode\vadjust pre{\hypertarget{ref-Hansen2011}{}}%
Hansen, N., Ros, R., Mauny, N., Schoenauer, M. \& Auger, A. (2011).
Impacts of invariance in search: {When} {CMA}-{ES} and {PSO} face
ill-conditioned and non-separable problems. \emph{Applied Soft
Computing}, \textbf{11}, 5755--5769.
\href{https://doi.org/10.1016/j.asoc.2011.03.001}{10.1016/j.asoc.2011.03.001}

\leavevmode\vadjust pre{\hypertarget{ref-Hartig2014}{}}%
Hartig, F., Dislich, C., Wiegand, T. \& Huth, A. (2014). Technical
{Note}: {Approximate} {Bayesian} parameterization of a process-based
tropical forest model. \emph{Biogeosciences}, \textbf{11}, 1261--1272.
\href{https://doi.org/10.5194/bg-11-1261-2014}{10.5194/bg-11-1261-2014}

\leavevmode\vadjust pre{\hypertarget{ref-Hartig2012}{}}%
Hartig, F., Dyke, J., Hickler, T., Higgins, S.I., O'Hara, R.B.,
Scheiter, S. \& Huth, A. (2012). Connecting dynamic vegetation models to
data -- an inverse perspective. \emph{Journal of Biogeography},
\textbf{39}, 2240--2252.
\href{https://doi.org/10.1111/j.1365-2699.2012.02745.x}{10.1111/j.1365-2699.2012.02745.x}

\leavevmode\vadjust pre{\hypertarget{ref-Hengl2017}{}}%
Hengl, T., Jesus, J.M. de, Heuvelink, G.B.M., Gonzalez, M.R., Kilibarda,
M., Blagotic, A., Shangguan, W., Wright, M.N., Geng, X.,
Bauer-Marschallinger, B., Guevara, M.A., Vargas, R., MacMillan, R.A.,
Batjes, N.H., Leenaars, J.G.B., Ribeiro, E., Wheeler, I., Mantel, S. \&
Kempen, B. (2017). {SoilGrids250m}: {Global} gridded soil information
based on machine learning. \emph{PLOS ONE}, \textbf{12}, e0169748.
\href{https://doi.org/10.1371/journal.pone.0169748}{10.1371/journal.pone.0169748}

\leavevmode\vadjust pre{\hypertarget{ref-Higgins2020}{}}%
Higgins, S.I., Larcombe, M.J., Beeton, N.J., Conradi, T. \& Nottebrock,
H. (2020). Predictive ability of a process-based versus a correlative
species distribution model. \emph{Ecology and Evolution}, \textbf{10},
11043--11054.
\href{https://doi.org/10.1002/ece3.6712}{10.1002/ece3.6712}

\leavevmode\vadjust pre{\hypertarget{ref-Higgins2012}{}}%
Higgins, S.I., O'Hara, R.B. \& Römermann, C. (2012). A niche for biology
in species distribution models. \emph{Journal of Biogeography},
\textbf{39}, 2091--2095.
\href{https://doi.org/10.1111/jbi.12029}{10.1111/jbi.12029}

\leavevmode\vadjust pre{\hypertarget{ref-Hijmans2021}{}}%
Hijmans, R.J., Phillips, S., Leathwick, J. \& Elith, J. (2021).
\emph{Dismo: Species distribution modeling}.

\leavevmode\vadjust pre{\hypertarget{ref-Hill2020}{}}%
Hill, A., Laneurit, J., Lenain, R. \& Lucet, E. (2020). {Online gain
setting method for path tracking using CMA-ES: Application to off-road
mobile robot control}. \emph{{IROS 2020, International Conference on
Intelligent Robots and Systems}}. Las Vegas, United States.

\leavevmode\vadjust pre{\hypertarget{ref-AFE2005}{}}%
Jalas, J. \& Suominen, J. (1972--2005). \emph{Atlas florae europaeae}.
Committee for Mapping the Flora of Europe; Societas Biologica Fennica
Vanamo, Helsinki, Finland.

\leavevmode\vadjust pre{\hypertarget{ref-Journe2020}{}}%
Journé, V., Barnagaud, J., Bernard, C., Crochet, P. \& Morin, X. (2020).
Correlative climatic niche models predict real and virtual species
distributions equally well. \emph{Ecology}, \textbf{101}.
\href{https://doi.org/10.1002/ecy.2912}{10.1002/ecy.2912}

\leavevmode\vadjust pre{\hypertarget{ref-Kleidon2000}{}}%
Kleidon, A. \& Mooney, H.A. (2000). A global distribution of
biodiversity inferred from climatic constraints: Results from a
process-based modelling study. \emph{Global Change Biology}, \textbf{6},
507--523.
\href{https://doi.org/10.1046/j.1365-2486.2000.00332.x}{10.1046/j.1365-2486.2000.00332.x}

\leavevmode\vadjust pre{\hypertarget{ref-Lenton2008}{}}%
Lenton, T.M., Held, H., Kriegler, E., Hall, J.W., Lucht, W., Rahmstorf,
S. \& Schellnhuber, H.J. (2008). Tipping elements in the {Earth}'s
climate system. \emph{Proceedings of the National Academy of Sciences},
\textbf{105}, 1786--1793.
\href{https://doi.org/10.1073/pnas.0705414105}{10.1073/pnas.0705414105}

\leavevmode\vadjust pre{\hypertarget{ref-Leroy2018}{}}%
Leroy, B., Delsol, R., Hugueny, B., Meynard, C.N., Barhoumi, C.,
Barbet-Massin, M. \& Bellard, C. (2018). Without quality
presence--absence data, discrimination metrics such as {TSS} can be
misleading measures of model performance. \emph{Journal of
Biogeography}, \textbf{45}, 1994--2002.
\href{https://doi.org/10.1111/jbi.13402}{10.1111/jbi.13402}

\leavevmode\vadjust pre{\hypertarget{ref-Lobo2008}{}}%
Lobo, J.M., Jiménez-Valverde, A. \& Real, R. (2008). {AUC}: A misleading
measure of the performance of predictive distribution models.
\emph{Global Ecology and Biogeography}, \textbf{17}, 145--151.
\href{https://doi.org/10.1111/j.1466-8238.2007.00358.x}{10.1111/j.1466-8238.2007.00358.x}

\leavevmode\vadjust pre{\hypertarget{ref-Loshchilov2013}{}}%
Loshchilov, I., Schoenauer, M. \& Sèbag, M. (2013). Bi-population
{CMA}-{ES} agorithms with surrogate models and line searches.
\emph{Proceedings of the 15th annual conference companion on {Genetic}
and evolutionary computation}, pp. 1177--1184. {GECCO} '13 {Companion}.
Association for Computing Machinery, New York, NY, USA.

\leavevmode\vadjust pre{\hypertarget{ref-Mauri2017}{}}%
Mauri, A., Strona, G. \& San-Miguel-Ayanz, J. (2017). {EU}-{Forest}, a
high-resolution tree occurrence dataset for {Europe}. \emph{Scientific
Data}, \textbf{4}, 160123.
\href{https://doi.org/10.1038/sdata.2016.123}{10.1038/sdata.2016.123}

\leavevmode\vadjust pre{\hypertarget{ref-Merow2014}{}}%
Merow, C., Smith, M.J., Edwards, T.C., Guisan, A., McMahon, S.M.,
Normand, S., Thuiller, W., Wüest, R.O., Zimmermann, N.E. \& Elith, J.
(2014). What do we gain from simplicity versus complexity in species
distribution models? \emph{Ecography}, \textbf{37}, 1267--1281.
\href{https://doi.org/10.1111/ecog.00845}{10.1111/ecog.00845}

\leavevmode\vadjust pre{\hypertarget{ref-Monnet2021}{}}%
Monnet, A.-C., Cilleros, K., Médail, F., Albassatneh, M.C., Arroyo, J.,
Bacchetta, G., Bagnoli, F., Barina, Z., Cartereau, M., Casajus, N.,
Dimopoulos, P., Domina, G., Doxa, A., Escudero, M., Fady, B., Hampe, A.,
Matevski, V., Misfud, S., Nikolic, T., Pavon, D., Roig, A., Barea, E.S.,
Spanu, I., Strid, A., Vendramin, G.G. \& Leriche, A. (2021). {WOODIV}, a
database of occurrences, functional traits, and phylogenetic data for
all {Euro}-{Mediterranean} trees. \emph{Scientific Data}, \textbf{8},
89.
\href{https://doi.org/10.1038/s41597-021-00873-3}{10.1038/s41597-021-00873-3}

\leavevmode\vadjust pre{\hypertarget{ref-Morin2007}{}}%
Morin, X., Augspurger, C. \& Chuine, I. (2007). Process-{Based}
{Modeling} of {Species}' {Distributions}: {What} {Limits} {Temperate}
{Tree} {Species}' {Range} {Boundaries}? \emph{Ecology}, \textbf{88},
2280--2291. \href{https://doi.org/10.1890/06-1591.1}{10.1890/06-1591.1}

\leavevmode\vadjust pre{\hypertarget{ref-Mouquet2015}{}}%
Mouquet, N., Lagadeuc, Y., Devictor, V., Doyen, L., Duputié, A.,
Eveillard, D., Faure, D., Garnier, E., Gimenez, O., Huneman, P., Jabot,
F., Jarne, P., Joly, D., Julliard, R., Kéfi, S., Kergoat, G.J., Lavorel,
S., Le Gall, L., Meslin, L., Morand, S., Morin, X., Morlon, H., Pinay,
G., Pradel, R., Schurr, F.M., Thuiller, W. \& Loreau, M. (2015).
{REVIEW}: {Predictive} ecology in a changing world. \emph{Journal of
Applied Ecology}, \textbf{52}, 1293--1310.
\href{https://doi.org/10.1111/1365-2664.12482}{10.1111/1365-2664.12482}

\leavevmode\vadjust pre{\hypertarget{ref-Sabater2021}{}}%
Muñoz Sabater, J. (2021). ERA5-land hourly data from 1950 to 1980.
\url{https://cds.climate.copernicus.eu/cdsapp\#!/dataset/reanalysis-era5-land}

\leavevmode\vadjust pre{\hypertarget{ref-Sabater2019}{}}%
Muñoz Sabater, J. (2019). ERA5-land hourly data from 1981 to present.
\url{https://cds.climate.copernicus.eu/cdsapp\#!/dataset/reanalysis-era5-land}

\leavevmode\vadjust pre{\hypertarget{ref-Oberpriller2021}{}}%
Oberpriller, J., Cameron, D.R., Dietze, M.C. \& Hartig, F. (2021).
Towards robust statistical inference for complex computer models.
\emph{Ecology Letters}, \textbf{24}, 1251--1261.
\href{https://doi.org/10.1111/ele.13728}{10.1111/ele.13728}

\leavevmode\vadjust pre{\hypertarget{ref-Pearman2008}{}}%
Pearman, P.B., Guisan, A., Broennimann, O. \& Randin, C.F. (2008). Niche
dynamics in space and time. \emph{Trends in Ecology \& Evolution},
\textbf{23}, 149--158.
\href{https://doi.org/10.1016/j.tree.2007.11.005}{10.1016/j.tree.2007.11.005}

\leavevmode\vadjust pre{\hypertarget{ref-Radeloff2015}{}}%
Radeloff, V.C., Williams, J.W., Bateman, B.L., Burke, K.D., Carter,
S.K., Childress, E.S., Cromwell, K.J., Gratton, C., Hasley, A.O.,
Kraemer, B.M., Latzka, A.W., Marin-Spiotta, E., Meine, C.D., Munoz,
S.E., Neeson, T.M., Pidgeon, A.M., Rissman, A.R., Rivera, R.J.,
Szymanski, L.M. \& Usinowicz, J. (2015). The rise of novelty in
ecosystems. \emph{Ecological Applications}, \textbf{25}, 2051--2068.
\href{https://doi.org/10.1890/14-1781.1}{10.1890/14-1781.1}

\leavevmode\vadjust pre{\hypertarget{ref-Saltre2013}{}}%
Saltré, F., Saint-Amant, R., Gritti, E.S., Brewer, S., Gaucherel, C.,
Davis, B.A.S. \& Chuine, I. (2013). Climate or migration: What limited
{European} beech post-glacial colonization? \emph{Global Ecology and
Biogeography}, \textbf{22}, 1217--1227.
\href{https://doi.org/10.1111/geb.12085}{10.1111/geb.12085}

\leavevmode\vadjust pre{\hypertarget{ref-Santini2021}{}}%
Santini, L., Benítez-López, A., Maiorano, L., Cengic, M. \& Huijbregts,
M.A.J. (2021). Assessing the reliability of species distribution
projections in climate change research. \emph{Diversity and
Distributions}, \textbf{27}, 1035--1050.
\href{https://doi.org/10.1111/ddi.13252}{10.1111/ddi.13252}

\leavevmode\vadjust pre{\hypertarget{ref-Singer2016}{}}%
Singer, A., Johst, K., Banitz, T., Fowler, M.S., Groeneveld, J.,
Gutiérrez, A.G., Hartig, F., Krug, R.M., Liess, M., Matlack, G., Meyer,
K.M., Pe'er, G., Radchuk, V., Voinopol-Sassu, A.-J. \& Travis, J.M.J.
(2016). Community dynamics under environmental change: {How} can next
generation mechanistic models improve projections of species
distributions? \emph{Ecological Modelling}, \textbf{326}, 63--74.
\href{https://doi.org/10.1016/j.ecolmodel.2015.11.007}{10.1016/j.ecolmodel.2015.11.007}

\leavevmode\vadjust pre{\hypertarget{ref-Steffen2018}{}}%
Steffen, W., Rockström, J., Richardson, K., Lenton, T.M., Folke, C.,
Liverman, D., Summerhayes, C.P., Barnosky, A.D., Cornell, S.E.,
Crucifix, M., Donges, J.F., Fetzer, I., Lade, S.J., Scheffer, M.,
Winkelmann, R. \& Schellnhuber, H.J. (2018). Trajectories of the {Earth}
{System} in the {Anthropocene}. \emph{Proceedings of the National
Academy of Sciences}, \textbf{115}, 8252--8259.
\href{https://doi.org/10.1073/pnas.1810141115}{10.1073/pnas.1810141115}

\leavevmode\vadjust pre{\hypertarget{ref-Toth2017}{}}%
Tóth, B., Weynants, M., Pásztor, L. \& Hengl, T. (2017). {3D} soil
hydraulic database of {Europe} at 250 m resolution. \emph{Hydrological
Processes}, \textbf{31}, 2662--2666.
\href{https://doi.org/10.1002/hyp.11203}{10.1002/hyp.11203}

\leavevmode\vadjust pre{\hypertarget{ref-Trautmann2011}{}}%
Trautmann, H., Mersmann, O. \& Arnu, D. (2011). \emph{Cmaes: Covariance
matrix adapting evolutionary strategy}.

\leavevmode\vadjust pre{\hypertarget{ref-Urban2015}{}}%
Urban, M.C. (2015). Accelerating extinction risk from climate change.
\emph{Science}, \textbf{348}, 571--573.
\href{https://doi.org/10.1126/science.aaa4984}{10.1126/science.aaa4984}

\leavevmode\vadjust pre{\hypertarget{ref-Urban2016}{}}%
Urban, M.C., Bocedi, G., Hendry, A.P., Mihoub, J.-B., Pe'er, G., Singer,
A., Bridle, J.R., Crozier, L.G., De Meester, L., Godsoe, W., Gonzalez,
A., Hellmann, J.J., Holt, R.D., Huth, A., Johst, K., Krug, C.B.,
Leadley, P.W., Palmer, S.C.F., Pantel, J.H., Schmitz, A., Zollner, P.A.
\& Travis, J.M.J. (2016). Improving the forecast for biodiversity under
climate change. \emph{Science}, \textbf{353}, aad8466.
\href{https://doi.org/10.1126/science.aad8466}{10.1126/science.aad8466}

\leavevmode\vadjust pre{\hypertarget{ref-Warren2021}{}}%
Warren, D.L., Dornburg, A., Zapfe, K. \& Iglesias, T.L. (2021). The
effects of climate change on {Australia}'s only endemic {Pokémon}:
{Measuring} bias in species distribution models. \emph{Methods in
Ecology and Evolution}, \textbf{12}, 985--995.
\href{https://doi.org/10.1111/2041-210X.13591}{10.1111/2041-210X.13591}

\leavevmode\vadjust pre{\hypertarget{ref-Williams2007}{}}%
Williams, J.W., Jackson, S.T. \& Kutzbach, J.E. (2007). Projected
distributions of novel and disappearing climates by 2100 {AD}.
\emph{Proceedings of the National Academy of Sciences}, \textbf{104},
5738--5742.
\href{https://doi.org/10.1073/pnas.0606292104}{10.1073/pnas.0606292104}

\leavevmode\vadjust pre{\hypertarget{ref-Zurell2016}{}}%
Zurell, D., Thuiller, W., Pagel, J., Cabral, J.S., Münkemüller, T.,
Gravel, D., Dullinger, S., Normand, S., Schiffers, K.H., Moore, K.A. \&
Zimmermann, N.E. (2016). Benchmarking novel approaches for modelling
species range dynamics. \emph{Global Change Biology}, \textbf{22},
2651--2664. \href{https://doi.org/10.1111/gcb.13251}{10.1111/gcb.13251}

\end{CSLReferences}

\newpage

\hypertarget{appendixA}{%
\section{Supplementary Appendix A: Insights on the
models}\label{appendixA}}

\renewcommand*\thetable{A.\arabic{table}}
\renewcommand*\thefigure{A.\arabic{figure}}

\setcounter{figure}{0}
\setcounter{table}{0}

\pagenumbering{arabic}
\renewcommand*{\thepage}{A--\arabic{page}}

\begin{figure}[htbp]

{\centering \includegraphics{figs/files/phenofit}

}

\caption{PHENOFIT model in a nutshell.}\label{fig:phenofit_model}
\end{figure}

\begin{figure}[htbp]

{\centering \includegraphics{figs/files/castanea_schema}

}

\caption{CASTANEA model in a nutshell.}\label{fig:phenofit_model}
\end{figure}

\newpage

\hypertarget{appendixB}{%
\section{Supplementary Appendix B: Processing of occurrence
data}\label{appendixB}}

\renewcommand*\thetable{B.\arabic{table}}
\renewcommand*\thefigure{B.\arabic{figure}}

\setcounter{figure}{0}
\setcounter{table}{0}

\pagenumbering{arabic}
\renewcommand*{\thepage}{B--\arabic{page}}

\hfill \break

\begin{table}[!h]

\caption{\label{tab:unnamed-chunk-1}GBIF download links}
\centering
\begin{tabular}[t]{ccc}
\toprule
Species & Number of occurrences & Download link\\
\midrule
Fagus sylvatica & 718.898 & https://doi.org/10.15468/dl.e9wasa\\
Quercus ilex & 78.979 & https://doi.org/10.15468/dl.2a4haw\\
Abies alba & 119.891 & https://doi.org/10.15468/dl.my6c9t\\
\bottomrule
\end{tabular}
\end{table}

\hfill \break

\begin{figure}[htbp]

{\centering \includegraphics{figs/files/presence_clustering}

}

\caption{Stratified random sampling of beech presence records based on climate clusters.}\label{fig:pres_clustering}
\end{figure}

\begin{landscape}
\begin{figure}[htbp]

{\centering \includegraphics{figs/files/occurrence_processing}

}

\caption{Processing of holm oak occurrence records. GBIF: Global Biodiversity Information Facility, AFE: Atlas Flora Europeae, EVM: EuroVegMap.}\label{fig:occ_processing}
\end{figure}
\end{landscape}

\newpage

\newpage

\hypertarget{appendixC}{%
\section{Supplementary Appendix C: Species
distributions}\label{appendixC}}

\renewcommand*\thetable{C.\arabic{table}}
\renewcommand*\thefigure{C.\arabic{figure}}

\setcounter{figure}{0}
\setcounter{table}{0}

\pagenumbering{arabic}
\renewcommand*{\thepage}{C--\arabic{page}}

\begin{figure}
\centering
\includegraphics{figs/distributionmaps.pdf}
\caption{Species distributions of \textbf{(A)} beech, \textbf{(B)} holm
oak and \textbf{(C)} silver fir. Green cells are \(0.1^\circ\) cells
where species is present, orange cells where species is supposed to be
absent.}
\end{figure}

\newpage

\hypertarget{appendixD}{%
\section{Supplementary Appendix D: Objective function and constraint
handling in CMA-ES}\label{appendixD}}

\renewcommand*\thetable{D.\arabic{table}}
\renewcommand*\thefigure{D.\arabic{figure}}

\setcounter{figure}{0}
\setcounter{table}{0}

\pagenumbering{arabic}
\renewcommand*{\thepage}{D--\arabic{page}}

\hfill \break

\hypertarget{the-auc-as-an-objective-function}{%
\subsection{The AUC as an objective
function}\label{the-auc-as-an-objective-function}}

The AUC is the area under the receiver operating characteristic curve,
which plots sensitivity (correctly predicted positive fraction) as a
function of commission error (falsely predicted positive fraction), as
the probability threshold discriminating presence/absence varies. It is
a discrimination metric, which has been widely used in the species
distribution modelling literature.

\hypertarget{box-constraint-handling}{%
\subsection{Box constraint handling}\label{box-constraint-handling}}

With this constraint handling - implemented by default in the R package
\emph{cmaes} (\protect\hyperlink{ref-Trautmann2011}{Trautmann \emph{et
al.} 2011}) - each evaluated solution is guaranteed to lie within the
feasible space. Let's say we have a parameter vector \(x\). For each
parameter \(x_i\), we have a lower bound \(lb_i\) and an upper bound
\(ub_i\). If a parameter \(x_i\) violates one of this bound, we set
\(x_i\) to a new value \(x_i^{repaired}\) equal to the closest boundary
value (\(lb_i\) or \(ub_i\)). We thus obtained a new parameter set
\(x^{repaired}\), with a minimal \(\|x-x^{repaired}\|\) value. This new
feasible solution \(x^{repaired}\) is used for the evaluation of the
objective function \(AUC_{model}(x^{repaired})\), and to compute a
penalty term
\(pen=\sum\limits_{i}(x_i-x_i^{repaired})^2=\|x^{repaired}-x\|^2\). Then
\(x^{repaired}\) is discarded, and the algorithm computes the penalized
objective function of \(x^{repaired}\) as follows:
\(AUC_{model}(x)=AUC_{model}(x^{repaired})+pen\). This boundary handling
could be improved with adaptive weights (see
\protect\hyperlink{ref-Hansen2009}{Hansen \emph{et al.} 2009}).

\hypertarget{ecological-infeasibility-constraint}{%
\subsection{Ecological infeasibility
constraint}\label{ecological-infeasibility-constraint}}

We added a simple way to handle ecological constraint (e.g.~unfolding
before flowering in beech mixed bud) with a death penalty. When a
parameter vector \(x\) violates a constraint, it is rejected and
generated again. The main drawback of this approach is that CMA-ES does
not use information from unfeasible points. An other approach could be
to set \(AUC_{model}(x)=0\). However, as our feasible space was large,
the death penalty constraint worked well in our case.

We applied an inequality constraint on both \textit{Fagus sylvatica} and
\textit{Quercus ilex}, which have mixed buds (leaves and flowers within
the same bud): unfolding must happen before flowering. On the contrary,
we did not apply any inequality constraint on \textit{Abies alba} simple
bud phenology parameters.

\newpage

\hypertarget{appendixE}{%
\section{Supplementary Appendix E: holm oak and silver fir
calibrations}\label{appendixE}}

\renewcommand*\thetable{E.\arabic{table}}
\renewcommand*\thefigure{E.\arabic{figure}}

\setcounter{figure}{0}
\setcounter{table}{0}

\pagenumbering{arabic}
\renewcommand*{\thepage}{E--\arabic{page}}

\hfill \break

\begin{figure}[H]

{\centering \includegraphics{figs/calibCMAESquercus}

}

\caption{CMA-ES calibration using the PHENOFIT model and holm oak: \textbf{(A)} calibration AUC (only calibration cells) and \textbf{(B)} total AUC (every presence/absence cells). Each color is a different sub-sampling of occurrence data, each point is a calibration run. The black horizontal bars represent the pairwise Mann–Whitney tests between the two subsets.}\label{fig:calibCMAESquercus}
\end{figure}

\begin{figure}[H]

{\centering \includegraphics{figs/calibCMAESabies}

}

\caption{CMA-ES calibration using the PHENOFIT model and silver fir: \textbf{(A)} calibration AUC (only calibration cells) and \textbf{(B)} total AUC (every presence/absence cells). Each color is a different sub-sampling of occurrence data, each point is a calibration run. The black horizontal bars represent the pairwise Mann–Whitney tests between the two subsets.}\label{fig:calibCMAESabies}
\end{figure}

\newpage

\hypertarget{appendixF}{%
\section{Supplementary Appendix F: Raw model outputs}\label{appendixF}}

\renewcommand*\thetable{F.\arabic{table}}
\renewcommand*\thefigure{F.\arabic{figure}}

\setcounter{figure}{0}
\setcounter{table}{0}

\pagenumbering{arabic}
\renewcommand*{\thepage}{F--\arabic{page}}

\hfill \break

\begin{figure}[H]

{\centering \includegraphics{figs/phenofit_output_maps}

}

\caption{Fitness index predicted by PHENOFIT with the expert and the inverse calibrations.}\label{fig:phenofit_output_maps}
\end{figure}

\begin{figure}[H]

{\centering \includegraphics{figs/best_forward_maps}

}

\caption{Carbon reserves predicted by CASTANEA with the expert and the inverse calibrations. Note that CASTANEA cannot be used in high-latitude regions (grey area).}\label{fig:best_forward_maps}
\end{figure}

\newpage

\hypertarget{appendixG}{%
\section{Supplementary Appendix G: Leaf unfolding
submodel}\label{appendixG}}

\renewcommand*\thetable{G.\arabic{table}}
\renewcommand*\thefigure{G.\arabic{figure}}

\setcounter{figure}{0}
\setcounter{table}{0}

\pagenumbering{arabic}
\renewcommand*{\thepage}{G--\arabic{page}}

\hfill \break

\hypertarget{fagus-sylvatica-leaf-unfolding-submodel}{%
\subsection{Fagus sylvatica leaf unfolding
submodel}\label{fagus-sylvatica-leaf-unfolding-submodel}}

This model, called UniChill (\protect\hyperlink{ref-Chuine2000}{Chuine
2000}), is a sequential two-phase model (endodormancy and ecodormancy
phases).\\
The endodormancy phase begins at day \(t_0\). The daily rate of chilling
\(R_c\) is defined as a threshold function of the daily mean temperature
\(T_d\): \[ R_c(T_d) = \left\{
\begin{array}{ll}
      0 & T_d \geq T_b \\
      1 & T_d < T_b \\
\end{array} 
\right. \] where \(T_b\) is the threshold temperature below which the
bud accumulates chilling units.\\
The endodormancy releases at day \(t_c\) when the accumulated rate of
chilling has reached the level \(C_{crit}\):
\[\sum\limits_{t_0}^{t_c} R_c(T_d) \geq C_{crit} \] Then, the
ecodormancy phase begins. The daily rate of forcing \(R_f\) is defined
as a sigmoid function of the daily mean temperature \(T_d\):
\[R_f(T_d) = \frac{1}{1 + e^{-d_T(T_d-T_{50})}} \] where \(d_T\) is the
slope and \(T_{50}\) the mid-response temperature. Bud break occurs at
day \(t_f\) when the accumulated rate of forcing has reached the level
\(F_{crit}\): \[\sum\limits_{t_c}^{t_f} R_f(T_d) \geq F_{crit} \] Thus,
the UniChill model has 6 parameters: \(t_0\), \(T_b\) and \(C_{crit}\)
for the first phase, \(d_T\), \(T_{50}\) and \(F_{crit}\) for the second
phase.

\begin{figure}[H]

{\centering \includegraphics{figs/leafpardensity}

}

\caption{Beech leaf unfolding model parameter density. Y-axis and X-axis limits are lower and upper bounds used during calibration.}\label{fig:leafpardensity}
\end{figure}

\begin{figure}[H]

{\centering \includegraphics{figs/consensusfitnessmap}

}

\caption{Median absolute deviation of beech \textbf{(A)} leaf unfolding date and \textbf{(B)} fitness, predicted with 100 calibrated parameter sets of PHENOFIT.}\label{fig:consensusfitnessmap}
\end{figure}

The median standard deviation of unfolding date across Europe was about
15.4 days. On beech presence points, it was about 16.2 days. Nearly
90.3\% of cells had a median absolute deviation lower than 30 days
(\hyperref[fig:consensusfitnessmap]{Figure F.2.A.}). The median standard
deviation of fitness across Europe was about 0.148. On beech presence
points, it was about 0.153. Nearly 46.4\% and 91.5\% of total cells had
a median absolute deviation lower than 0.1 and 0.2 respectively
(\hyperref[fig:consensusfitnessmap]{Figure F.2.B.}).

\begin{figure}[H]

{\centering \includegraphics{figs/unfoldingdatesbackforw}

}

\caption{Mean leaf unfolding day of beech with \textbf{(A)} best CMA-ES calibrated parameters and \textbf{(B)} expert parameters. Values above June solstice day (167) are in grey. Note that PHENOFIT model assign a value of 365 when unfolding has not happened at all due to climate conditions.}\label{fig:unfoldingdatesbackforw}
\end{figure}

\newpage
\singlespacing
\printbibliography

\end{document}

### 2023-03-21 manuscript/functions/command_file_setup.R

@@ -1,0 +1,21 @@
+command_file_setup <- function(command_file, species_files, output_folder, climate_folder, 
+                       climate_scenario, starting_year, ending_year, quiet_mode){
+  
+  command_lines <- c()
+  command_lines[1] <- c("# File for Phenofit4 CommandScript")
+  command_lines[2] <- c(paste0("# Generated with Rstudio (time: ", Sys.time(),", user: ",Sys.info()[["user"]],")"))
+  command_lines[3] <- c("\t\t\t\t\t")
+  command_lines[4] <- c(paste("speciesInputDir ="))
+  command_lines[5] <- c(paste("climateInputDir ="))
+  command_lines[6] <- c(paste("outputDir =", output_folder))
+  command_lines[7] <- c(paste("quietMode =", quiet_mode))
+  command_lines[8] <- c("\t\t\t\t\t")
+  command_lines[9] <- c("# speciesFileName\tclimateFolderName\tclimateScenario\tstartingYear\tendingYear")
+  for(i in 1:length(species_files)){
+    command_lines[9+i] <- paste(species_files[i], climate_folder, 
+                              climate_scenario, starting_year, ending_year, sep="\t")
+  }
+  
+  writeLines(command_lines, command_file)
+  
+}

### 2023-03-21 manuscript/functions/read_mean_outputvalue_castanea.R

@@ -1,0 +1,29 @@
+library(vroom)
+
+read_mean_outputvalue_castanea <- function(grid, lambda, output_folder = 'D:/applications/capsis4/var', var = "BiomassOfReserves"){
+  
+  output <- matrix(ncol=length(grid), nrow=lambda, byrow=TRUE)
+  
+  for(l in 1:lambda){
+    
+    filenames <- sapply(grid, function(i){
+      ind <- paste0("ind", l, "_")
+      file.path(output_folder, paste0(format(i, scientific=F), "_", ind, "yearlyResults.log"))})
+    
+    yearly_results <- vroom(filenames, col_select = c(var), show_col_types = FALSE, progress=FALSE)
+    yearly_results[is.na(yearly_results)] <- 0 # NA values mean tree is dead (thus biomass = 0) 
+    
+    # if(nrow(yearly_results) == 0){
+    #   yearly_results[1:(length(grid)*30), var] <- NA
+    # }
+    
+    nyears <- nrow(yearly_results)/length(grid)
+    
+    mean_results <- aggregate(yearly_results, list(rep(1:(nrow(yearly_results) %/% nyears + 1), each = nyears, len = nrow(yearly_results))), mean)[-1]
+    
+    output[l,] <- t(mean_results)
+    
+  }
+  
+  return(output)
+}

### 2023-03-21 manuscript/functions/read_mean_outputvalue_forceeps.R

@@ -1,0 +1,33 @@
+library(vroom)
+
+# v2 with meanstandexport
+
+read_mean_outputvalue_forceeps <- function(grid, ind_paths, var = "basalAreaMean"){
+  
+  
+  
+  cnames <- c("date", "nTrees", "totalBiomass", "totalBasalArea", "totalHeight", "totalVolume", "LAI", "droughtIndexAnnual", 
+              "droughtIndexSeasonal", "winterTMIN", "GDDannual", "GDDseasonal", "nTreesMean", "nTreesSD", "biomassMean", "biomassSD",
+              "basalAreaMean", "basalAreaSD", "heightMean", "heightSD", "volumeMean", "volumeSD")
+  
+  temp <- future_lapply(1:length(ind_paths), function(l){
+    
+    filenames <- sapply(grid, function(i){
+      ind <- paste0(format(i, scientific=F), ".sitemeanstand.txt")
+      file.path(ind_paths[l], ind)})
+    
+    yearly_results <- vroom(filenames, show_col_types = FALSE, progress=FALSE, skip = 3, col_names=cnames) %>%
+      dplyr::select(all_of(var))
+    
+    nyears <- nrow(yearly_results)/length(grid)
+    
+    mean_results <- aggregate(yearly_results, list(rep(1:(nrow(yearly_results) %/% nyears + 1), each = nyears, len = nrow(yearly_results))), mean)[-1]
+    
+    return(mean_results)
+    
+  })
+  
+  output <- matrix(unlist(temp), ncol=length(grid), nrow=length(ind_paths), byrow=TRUE)
+  
+  return(output)
+}

### 2023-03-21 manuscript/functions/read_mean_outputvalue_phenofit.R

@@ -1,0 +1,33 @@
+library(data.table) #fastest
+setDTthreads(1) #only one thread (in case of parallel cmaes)
+
+read_mean_outputvalue_phenofit <- function(output_folder, points = NULL){
+  if(is.null(points)){
+    #read all file
+    fitness <- fread(output_folder, header=T, sep="\t", fill=T)
+    fitness_mean <- matrix(ncol=ncol(fitness)-1, nrow=1, byrow=TRUE)
+    
+    fitness <- as.data.frame(fitness)
+    fitness[fitness == -999] <- 366
+    fitness_mean[1,] <- apply(fitness[c(-1,-2),-1], 2, mean)
+  }
+  else{
+    #select some points
+    fitness <- fread(output_folder, header=F, sep="\t", fill=T)
+    col_names <- t(fitness[-1,1])
+    fitness <- fitness[-1,-1]
+    fitness <- as.data.frame(t(fitness))
+    names(fitness) <- col_names
+    names(fitness)[1:2] <- c("lat", "lon")
+    fitness$lat <- round(as.numeric(fitness$lat), 1)
+    fitness$lon <- round(as.numeric(fitness$lon), 1)
+    fitness <- inner_join(points, fitness, by = c("lat", "lon")) %>% dplyr::select(-c('pres'))
+    fitness <- as.data.frame(t(fitness))
+    fitness[fitness == -999] <- 366
+    fitness_mean <- matrix(ncol=ncol(fitness), nrow=1, byrow=TRUE)
+    fitness_mean[1,] <- apply(fitness[c(-1,-2),], 2, mean)
+  }
+  
+  return(fitness_mean)
+  
+}

### 2023-03-21 manuscript/functions/read_species_file.R

@@ -1,0 +1,22 @@
+library("xml2")
+
+read_species_file <- function(file){
+  species_xml <- read_xml(file)
+  species <- xml_children(xml_children(species_xml)[[1]])
+  
+  
+  Leaf <- xml_children(xml_find_all(species, "//Leaf"))
+  Flower <- xml_children(xml_find_all(species, "//Flower"))
+  FruitMat <- xml_children(xml_find_all(species, "//FruitMaturation"))
+  Senes <- xml_children(xml_find_all(species, "//Senescence"))
+  CpdBud <- (xml_find_all(species, "//CompoundBud"))
+  LeafHabit <- (xml_find_all(species, "//LeafHabit"))
+  Tdeath <- (xml_find_all(species, "//Tdeath"))
+  Frost <- xml_children(xml_find_all(species, "//Frost"))
+  Drought <- xml_children(xml_find_all(species, "//Drought"))
+  
+  params <- c(leaf = xml_text(Leaf), flower = xml_text(Flower), fruitmat = xml_text(FruitMat), senes = xml_text(Senes), 
+              cpdbud = xml_text(CpdBud), leafhabit = xml_text(LeafHabit), tdeath= xml_text(Tdeath), frost = xml_text(Frost), drought = xml_text(Drought))
+  
+  return(params)
+}

### 2023-03-21 manuscript/scripts/generate_density_plot.R

@@ -1,0 +1,51 @@
+lb1 <- unique(data_leaf_par[data_leaf_par$var == par_names[parameter_name1], "lb"])
+ub1 <- unique(data_leaf_par[data_leaf_par$var == par_names[parameter_name1], "ub"])
+lb2 <- unique(data_leaf_par[data_leaf_par$var == par_names[parameter_name2], "lb"])
+ub2 <- unique(data_leaf_par[data_leaf_par$var == par_names[parameter_name2], "ub"] )
+
+data_leaf_par_density <- data.frame(x = data_leaf_par[data_leaf_par$var == par_names[parameter_name1], "value"], 
+                                    y = data_leaf_par[data_leaf_par$var == par_names[parameter_name2], "value"] )
+
+
+
+
+
+plot <- ggplot(data = data_leaf_par_density, aes(x = x, y = y)) +
+  geom_density_2d_filled(
+    aes(fill = ..level..),
+    contour_var = "ndensity",
+    alpha = 1,
+    colour="white") +
+  scale_fill_grey(start = 0.4, end = 0.9) +
+  # scale_fill_manual(values = c("#FFEAA1","#FDDA94","#FBC988","#F9B97B","#F7A96E",
+  #                              "#F49862","#F28855","#F07848","#EE673C","#EC572F")) +
+  geom_segment(aes(x=lb1,xend=ub1,y=lb2,yend=lb2), col = NA, size = 1) +
+  geom_segment(aes(x=lb1,xend=ub1,y=ub2,yend=ub2), col = NA, size = 1) +
+  geom_segment(aes(x=lb1,xend=lb1,y=lb2,yend=ub2), col = NA, size = 1) +
+  geom_segment(aes(x=ub1,xend=ub1,y=lb2,yend=ub2), col = NA, size = 1) +
+  scale_y_continuous(limits = c(lb2, ub2), expand = c(0,0), 
+                     breaks = c(lb2, breaks2[1], breaks2[2], ub2)) +
+  scale_x_continuous(limits = c(lb1, ub1), expand = c(0,0),
+                     breaks = c(lb1, breaks1[1], breaks1[2], ub1)) +
+  theme_minimal() +
+  theme(axis.ticks.y = element_line(colour = "black", size = 0.5),
+        axis.ticks.x = element_line(colour = "black", size = 0.5),
+        panel.border = element_rect(colour = "black", fill=NA, size=1)) +
+  theme(axis.text.y = element_text(family = "Linux Libertine G"),
[…]

### 2023-03-21 manuscript/scripts/run_phenofit.R

@@ -1,0 +1,37 @@
+library(gtools)
+
+# Run Phenofit for every calibrated parameter sets
+wd <- 'C:/Users/vandermeersch/Documents/CEFE/phd/present_calibration/calibration_paper'
+source(file.path(wd, "functions", "command_file_setup.R"))
+
+cal_folder <- 'D:/calibrations/phenofit/fagus_sylvatica/1000pres_1000abs/paper_data/ABC'
+sim_folder <- 'D:/simulations/phenofit/backward/fagus_sylvatica/paper_data/ABC'
+
+species_files <- mixedsort(list.files(path = cal_folder, pattern = "\\.species$", full.names = F, recursive = T))
+
+climate_folder <- "D:/climate/ERA5-Land/phenofit_format/transformed"
+capsis_settings=list(java8="java8.cmd", cd_java8="cd C:/Program Files/Java/scripts", cd_capsis="cd/d D:/applications/capsis4 && setmem 30000")
+
+for(cal in species_files){
+  subset <- strsplit(cal, "_")[[1]][4]
+  rep <- substr(strsplit(cal, "_")[[1]][5], 1, nchar(strsplit(cal, "_")[[1]][5])-8)
+  output_folder <- file.path(sim_folder, paste0(subset, "_", rep))
+  print(paste0("Repetition ", rep, " of subset ", subset))
+  
+  species_file <- file.path(cal_folder, cal)
+  command_file <- file.path(sim_folder, "command_file.txt")
+  
+  command_file_setup(command_file, species_file, output_folder, climate_folder,
+                     climate_scenario="ERA5LAND", starting_year="1970", ending_year="2000", quiet_mode="true")
+  
+  run_capsis <- paste("capsis -p script phenofit4.myscripts.ScriptVictor", command_file)
+  run <- paste(capsis_settings$cd_java8, capsis_settings$java8, capsis_settings$cd_capsis, run_capsis, sep=' && ')
+  
+  system.time(shell(run, intern=F))
+}
+
+cal_folder <- 'D:/calibrations/phenofit/fagus_sylvatica/1000pres_1000abs/paper_data/ABC'
+sim_folder <- 'D:/simulations/phenofit/backward/fagus_sylvatica/paper_data/ABC'
+
+species_files <- mixedsort(list.files(path = cal_folder, pattern = "\\.species$", full.names = F, recursive = T))
+

### 2023-03-21 manuscript/scripts/test_color_by_auc.R

@@ -1,0 +1,59 @@
+
+auc_tot_temp <- data.frame(auc_tot = fagus_auc_data_tot$tot, 
+                           rep =paste0("subset_", parse_number(fagus_auc_data_tot$subset), "_", fagus_auc_data_tot$rep))
+auc_tot_temp  <- auc_tot_temp[order(auc_tot_temp$auc_tot),]
+auc_tot_temp <- split(auc_tot_temp, ceiling(seq_along(auc_tot_temp$auc_tot)/10))
+for(i in 1:10){
+  auc_tot_temp[[i]]['col_group_auc'] <- i
+}
+auc_tot_temp<- do.call("rbind", auc_tot_temp) %>% dplyr::select(c(rep, col_group_auc))
+
+data_leaf_par <- left_join(data_leaf_par, auc_tot_temp)
+
+data_leaf_par %>%
+  ggplot(aes(y = value, x = 1)) +
+  geom_violin(fill = 'lightgrey', alpha= 0.2, size = 0.4, colour = 'darkgrey') +
+  geom_beeswarm(aes(fill = factor(col_group_auc), col = factor(col_group_auc)), 
+                alpha = 0.5, cex = 2.5, size = 2) +
+  facet_wrap("var_name", scales="free_y",
+             labeller= label_parsed) +
+  scale_y_continuous(breaks = breaks_fun1) +
+  geom_point(aes(y = forward, x= 0.55), fill = 'darkred', size = 1.5, alpha = 0.9, shape = 23, colour="white", stat = "unique") +
+  geom_point(data = data.frame(best = unique(data_leaf_par$best), var_name = unique(data_leaf_par$var_name)),
+             aes(y = best, x= c(0.55, 0.65, 0.55, 0.55, 0.55, 0.55)), fill = 'darkblue', size = 1.5, 
+             alpha = 0.9, shape = 23, colour="white", inherit.aes = F) +
+  scale_fill_manual(values = c("#ef7e56", "#e28143", "#d38531", "#c18820", "#ae8c0f", "#988e00", "#819100", "#67920c", "#49921d", "#16922e")) +
+  scale_color_manual(values = c("#ef7e56", "#e28143", "#d38531", "#c18820", "#ae8c0f", "#988e00", "#819100", "#67920c", "#49921d", "#16922e")) +
+  theme_minimal() +
+  theme(axis.text.x = element_blank(),
+        axis.title.x = element_blank(),
+        axis.text.y = element_text(family = "Linux Libertine G"),
[…]

### 2023-03-21 code/calibration.R
example of R code used to run calibration on GenOuest cluster

@@ -1,0 +1,95 @@
+# Genouest cluster adaptation
+
+# Working directory
+wd <- "/scratch/vvandermeersch/phenofit_calibration"
+
+#wd <- 'C:/Users/vandermeersch/Documents/CEFE/thesis/phenofit/Phenofit_from_R/inverse_modelling_cluster'
+
+#library(progressr)
+
+source(paste0(wd, "/functions/cmaes_calibration.R"))
+source(paste0(wd, "/functions/auc_test.R"))
+source(paste0(wd, "/functions/read_species_file.R"))
+source("functions/linear_scaling.R")
+
+
+
+################
+#### SET-UP ####
+################
+
+# Parameter inital values and bounds
+param_init <- read_species_file(paste0(wd, "/input_files/pinus_sylvestris/Pinus_sylvestris_init.species"))
+param_fixed <- read_species_file(paste0(wd, "/input_files/pinus_sylvestris/Pinus_sylvestris_lb.species"))
+param_lb <- read_species_file(paste0(wd, "/input_files/pinus_sylvestris/Pinus_sylvestris_lb.species"))
+param_ub <- read_species_file(paste0(wd, "/input_files/pinus_sylvestris/Pinus_sylvestris_ub.species"))
+scale_factor <- 10
+parameters=parameters=list(param_init=param_init, param_fixed=param_fixed, 
+                           param_lb=param_lb, param_ub=param_ub, scale_factor=scale_factor)
+
+# Constraints
+constraint_function <- function(x){
+  x[8] <= x[9] # Leaf Fcrit <= Fflower Fcrit for P. sylvestris
+}
+
+# CMA-ES settings
+cmaes_controls=list(sigma=2, mu=10, lambda=20, maxit=300)
+cmaes_settings=list(rand_init=FALSE, ipop=FALSE, ipop_factor=NULL)
+
+# Parallelization settings
+parallel_settings=list(ncores_runs=1, ncores_eval=20)
+
+# Capsis settings
+structure_file <- paste0(wd, "/input_files/Pinus_sylvestris_ADuputie.species")
+species=list(structure=structure_file, file="pinus_sylvestris")
+folders=list(wd=wd, 
+             climate="/scratch/vvandermeersch/data/climate/phenofit_format/ERA5Land_pinsyl/1000pres_1000abs/subset_1")
+simulation=list(climate_scenario="ERA5LAND", starting_year="1970", ending_year="2000", quiet_mode="true")
[…]

### 2023-03-21 code/calibration.sh

@@ -1,0 +1,14 @@
+#!/bin/sh
+#SBATCH --mail-type=ALL
+#SBATCH --mail-user=victor.vandermeersch@cefe.cnrs.fr
+#SBATCH --job-name=CMAES1
+#SBATCH --output=/scratch/vvandermeersch/phenofit_calibration/logs/CMAES_log.txt
+#SBATCH --cpus-per-task=30
+#SBATCH --mem=80G
+#SBATCH --time 1-12:00:00
+
+. /local/env/envconda.sh
+
+conda activate /home/genouest/mnhn_cesco/vvandermeersch/env/env_capsis_cmaes
+
+Rscript calibration.R

### 2023-03-21 code/functions/auc.R

@@ -1,0 +1,78 @@
+suppressPackageStartupMessages(library(AUC))
+source(paste0(wd,"/functions/linear_scaling.R"))
+source(paste0(wd, "/functions/read_mean_fitness.R"))
+
+auc_test <- function(x, Yobs, run, run_all, wd, species_files, lambda, param){
+  
+  # Inverse transformation
+  # param_invs <- matrix(ncol=ncol(param), nrow=nrow(param), byrow=TRUE)
+  # for(i in 1:ncol(param)){
+  #   param_invs[,i] <- inv_linearscaling(param[,i], domain_lb, domain_ub)
+  # }
+  
+  
+  param_f <- which(param$param_fixed!="FIXED")
+  
+  # If mixt bud
+  if(parameters$param_init["cpdbud"]=="2"){
+    flower_param <- grepl("flower", names(param_f)) # get indices
+    flower_param[max(which(flower_param))] <- FALSE # exclude last parameter of flower model (Fcrit)
+    
+    param_f <- param_f[!flower_param] # new fixed parameter indice list
+  }
+  
+  param_lb <- as.numeric(param$param_lb[param_f])
+  param_ub <- as.numeric(param$param_ub[param_f])
+  
+  param_f <- which(param$param_fixed!="FIXED") # inital fixed parameter indice list
+  
+  for(i in 1:lambda){
+    x_inv <- inv_linearscaling(x[,i], param_lb, param_ub, 
+                               param$scale_factor)
+    
+    
+    # If mixt bud
+    if(parameters$param_init["cpdbud"]=="2"){
+      len <- length(x_inv) #number of parameters found by cmaes algorithm 
+      nb <- length(which(flower_param)) #number of parameters to copy from leaf model to flower model
+      x_inv <- c(x_inv[1:(nb+1)], x_inv[1:nb], x_inv[(nb+2):len]) # new parameter list to modify
+      
+    }
+    
+    
+    modify_species_file(as.character(x_inv), param$param_fixed, species_files[i])
+  }
+  
+  
+  # Run Capsis simulations
+  temp <- future_lapply(run_all, function(x){
+    out <- system(x, ignore.stdout = TRUE, ignore.stderr = TRUE)
+  })
+  #shell(run_all)
+  
+  # eval(parse(text=paste0("it <- it_run", run)))
+  # line <- paste("Iteration", it, "of run", run, "completed\n")
[…]

### 2023-03-21 code/functions/cmaes_calibration.R

@@ -1,0 +1,257 @@
+library(future)
+library(future.apply)
+library(cmaes)
+
+
+
+source(paste0(wd, "/functions/linear_scaling.R"))
+source(paste0(wd, "/functions/create_init_species_file.R"))
+source(paste0(wd, "/functions/command_file_setup.R"))
+source(paste0(wd,"/functions/cmaes_mod.R"))
+
+#cmaes_settings=list(rand_init=FALSE, ipop=FALSE, ipop_factor=NULL)
+#parallel=list(ncores_runs=1, ncores_eval=1)
+#parameters=list(param_init, param_lb, param_ub, scale_factor)
+#controls=list(sigma=NULL, mu=NULL, lambda=NULL, maxit=NULL)
+#species=list(structure="/input_files/Fagus_sylvatica_structure.species", file="/process_files/params/Fagus_sylvatica")
+#folders=list(output, climate)
+#simulation=list(command_file, climate_scenario, starting_year, ending_year, tmp_means)
+#capsis_settings=list(java8=TRUE, cd_java8=NA, cd_capsis)
+
+
+# Function to run a CMA-ES calibration
+cmaes_calibration <- function(nruns, obj_function, ..., 
+                         parameters=list(param_init, param_fixed, param_lb, param_ub, scale_factor),
+                         is_feasible = function(x){TRUE},
+                         controls=list(sigma=NULL, mu=NULL, lambda=NULL, maxit=NULL),
+                         parallel=list(ncores_runs=1, ncores_eval=1),
+                         cmaes_settings=list(rand_init=FALSE, ipop=FALSE, ipop_factor=NULL),
+                         species=list(structure, file),
+                         folders=list(wd, climate),
+                         simulation=list(climate_scenario, starting_year, ending_year, tmp_means),
+                         capsis_settings=list(java8=NA, cd_java8=NA, cd_capsis)){
+  
+  
+  
+  start_time_total <- Sys.time()
+  
+  # Print messages
+  cat(paste0(Sys.time(),"\n"))
+  cat("Starting CMA-ES calibration\n")
+  cat("-------------------------------------\n")
+  cat(paste0(nruns, " run(s) with ", parallel$ncores_eval, " parallel simulations\n"))
+  cat("-------------------------------------\n")
+  
+  # Parameter initialisation
+  structure_file <- species$structure
+  sp_file <- paste0("/process_files/params/",species$file)
[…]

### 2023-03-21 code/functions/cmaes_mod.R

@@ -1,0 +1,254 @@
+# Adapted from cmaes::cma_es function by Olaf Mersmann
+
+# Add message to a log
+# Add a is_feasible option for constraints (adapted from rCMA package, by Wolfgang Konen)
+
+
+cmaes_mod <- function (par, fn, ..., lower, upper, control = list(), is_feasible, log_file, trace_log = NA, trace_iter = c())
+{
+  norm <- function(x) drop(sqrt(crossprod(x)))
+  
+  controlParam <- function(name, default) {
+    v <- control[[name]]
+    if (is.null(v)) 
+      return(default)
+    else return(v)
+  }
+  
+  ## Inital solution:
+  xmean <- par
+  N <- length(xmean)
+  
+  ## Box constraints:
+  if (missing(lower)) 
+    lower <- rep(-Inf, N)
+  else if (length(lower) == 1) 
+    lower <- rep(lower, N)
+  if (missing(upper)) 
+    upper <- rep(Inf, N)
+  else if (length(upper) == 1) 
+    upper <- rep(upper, N)
+  
+  ## Parameters
+  trace <- controlParam("trace", FALSE)
+  fnscale <- controlParam("fnscale", 1)
+  stopfitness <- controlParam("stopfitness", -Inf)
+  maxiter <- controlParam("maxit", 100 * N^2)
+  sigma <- controlParam("sigma", 0.5)
+  sc_tolx <- controlParam("stop.tolx", 1e-12 * sigma)
+  keep.best <- controlParam("keep.best", TRUE)
+  vectorized <- controlParam("vectorized", FALSE)
+  log.all <- controlParam("diag", FALSE)
+  log.sigma <- controlParam("diag.sigma", log.all)
+  log.eigen <- controlParam("diag.eigen", log.all)
+  log.value <- controlParam("diag.value", log.all)
+  log.pop <- controlParam("diag.pop", log.all)
+  
+  ## Strategy parameter setting (defaults as recommended by Nicolas Hansen):
+  lambda <- controlParam("lambda", 4 + floor(3 * log(N)))
+  mu <- controlParam("mu", floor(lambda/2))
+  weights <- controlParam("weights", log(mu + 1) - log(1:mu))
+  weights <- weights/sum(weights)
+  mueff <- controlParam("mueff", sum(weights)^2/sum(weights^2))
+  cc <- controlParam("ccum", 4/(N + 4))
[…]

### 2023-03-21 code/functions/create_init_species_file.R

@@ -1,0 +1,13 @@
+source(paste0(wd, "/functions/create_species_file.R"))
+source(paste0(wd, "/functions/modify_species_file.R"))
+
+create_init_species_file <- function(name, params_init, structure_file, lambda){
+  species_files <- c()
+  for(i in 1:lambda){
+    species_files[i] <- paste0(wd, name, "_", as.character(i), ".species")
+    create_species_file(species_files[i], structure_file)
+    modify_species_file(params_init, param_fixed=NULL, species_files[i])
+  }
+  return(species_files)
+}
+

### 2023-03-21 code/functions/create_species_file.R

@@ -1,0 +1,6 @@
+library("xml2")
+
+create_species_file <- function(file, structure_file){
+  species_structure_xml <- read_xml(structure_file)
+  write_xml(species_structure_xml, file)
+}

### 2023-03-21 code/functions/linear_scaling.R

@@ -1,0 +1,15 @@
+linearscaling <- function(x, min, max, scale_factor){
+  y <- 1:length(x)
+  for(i in 1:length(x)){
+    y[i] <- (x[i]-min[i])/(max[i]-min[i])*10
+  }
+  return(y)
+}
+
+inv_linearscaling <- function(x, min, max, scale_factor){
+  y <- 1:length(x)
+  for(i in 1:length(x)){
+    y[i] <- x[i]*(max[i]-min[i])/10+min[i]
+  }
+  return(y)
+}

### 2023-03-21 code/functions/modify_species_file.R

@@ -1,0 +1,30 @@
+modify_species_file <- function(new_params, param_fixed=NULL, file){
+  species_xml <- read_xml(file)
+  species <- xml_children(xml_children(species_xml)[[1]])
+  
+  Leaf <- xml_children(xml_find_all(species, "//Leaf"))
+  Flower <- xml_children(xml_find_all(species, "//Flower"))
+  FruitMat <- xml_children(xml_find_all(species, "//FruitMaturation"))
+  Senes <- xml_children(xml_find_all(species, "//Senescence"))
+  CpdBud <- (xml_find_all(species, "//CompoundBud"))
+  LeafHabit <- (xml_find_all(species, "//LeafHabit"))
+  Tdeath <- (xml_find_all(species, "//Tdeath"))
+  Frost <- xml_children(xml_find_all(species, "//Frost"))
+  Drought <- xml_children(xml_find_all(species, "//Drought"))
+  
+  params <- c(Leaf, Flower, FruitMat, Senes, CpdBud, LeafHabit, Tdeath, Frost, Drought)
+  
+  if(is.null(param_fixed)){
+    for(i in 1:length(params)){
+      xml_text(params[[i]]) <- new_params[i]
+    }
+  }else{
+    j <- 1
+    for(i in which(param_fixed!="FIXED")){
+      xml_text(params[[i]]) <- new_params[j]
+      j <- j + 1
+    }
+  }
+  
+  write_xml(species_xml, file)
+}

### 2023-03-21 code/functions/read_mean_fitness.R

@@ -1,0 +1,32 @@
+
+# Old version, phenofit without continuous export
+
+# read_mean_fitness <- function(lambda, output_folder, years_mean){
+#   fitness <- read.table(paste0(output_folder,"-",as.character(1),"/FitnessYear.txt"), header=T)
+#   fitness_mean <- matrix(ncol=lambda, nrow=nrow(fitness), byrow=TRUE)
+#   for(i in 1:lambda){
+#     fitness <- read.table(paste0(output_folder,"-",as.character(i),"/FitnessYear.txt"), header=T)
+#     fitness_mean[,i] <- apply(fitness[, grepl(paste(years_mean, collapse = "|"), names(fitness))], 1, mean)
+#   }
+#   return(fitness_mean)
+# }
+
+library(data.table) #fastest
+setDTthreads(1) #only one thread (in case of parallel cmaes)
+
+read_mean_fitness <- function(lambda, output_folder){
+  fitness <- fread(paste0(output_folder,as.character(1),"/Fitness.txt"), skip=3, sep="\t", fill = TRUE)
+  fitness_mean <- matrix(ncol=ncol(fitness)-1, nrow=lambda, byrow=TRUE)
+  
+  fitness <- as.data.frame(fitness)
+  fitness_mean[1,] <- apply(fitness[,-1], 2, mean)
+  if(lambda>1){
+    for(i in 2:lambda){
+      fitness <- fread(paste0(output_folder,as.character(i),"/Fitness.txt"), skip=3, sep="\t", fill = TRUE)
+      fitness <- as.data.frame(fitness)
+      fitness_mean[i,] <- apply(fitness[,-1], 2, mean)
+    }
+  }
+  
+  return(fitness_mean)
+}

### 2023-03-21 manuscript/paper.Rmd
did some cleaning

@@
 
-<!-- # Conclusion (optionnal) -->
-
-<!-- **[Commentaire : ce n’est pas une conclusion, il faut la refaire. Ca ressemble plus à une partie perspectives et ouverture d’une discussions]** -->
-<!-- \newline -->
-
-<!-- ~~We do recognize that the calibration approach proposed here is not a perfect method, the main weakness being the questionable biological plausibility of parameter values we found.
-Nevertheless, CMA-ES is particularly useful to generate inverse optimized versions of the process-based models.
-These fitted model versions integrate explicit processes but are calibrated on species distributions like cSDMs.
-They thus fall within the continuum between correlative and forward models [see @Dormann2012].
-Both types of models have been compared in several studies, which found strong discrepancies between projections [@Morin2009; @Cheaib2012; @Gritti2013].
-No studies, though, have examined the structural differences between the two approaches in order to determine the origin of their robustness.
-Comparison of correlative models to optimized process-based model projections will allow testing whether model robustness come from the cause-to-effect relationships or not, while comparison of optimized process-based model to standard process-based models will allow testing whether model robustness come from a calibration independent species distribution or not.
-This study is, to our knowledge, the first one to assess the feasibility of CMA-ES to calibrate ecological process-based models, and will permit to carry out a thorough comparison of different versions of the models.~~ -->
-<!-- \newline -->
-
 # Acknowledgements

### 2023-03-28 manuscript/paper.Rmd
small changes before resubmission (minor revision)

@@
 csl: bib/mee.csl
-title: "A novel approach to inverse calibration of process-based models using evolutionary algorithm and species distribution data"
+title: "Estimating process-based model parameters from species distribution data using the evolutionary algorithm CMA-ES"
 author:
@@
 Two main types of species distribution models are used to project species range shifts in future climatic conditions: correlative and process-based models.
-Although there is some continuity between these two types of models, they are fundamentally different in their hypotheses (statistical relationships *vs* mechanistic relationships) and their calibration methods (dependent *vs* generally independent of the species observed distributions).
+Although there is some continuity between these two types of models, they are fundamentally different in their hypotheses (statistical relationships *vs* mechanistic relationships) and and their calibration methods \\textcolor{darkgreen}{(SDMs tend to be occurrence data-driven while PBMs tend to be prior-driven)}.
 \n\n 2.
-One of the limitations to the use of process-based models is the difficulty to parameterize them for a large number of species, while one of the advantages of correlative SDMs is their easier calibration using solely species occurrences}.
+One of the limitations to the use of process-based models is the difficulty to parameterize them for a large number of species \\textcolor{darkgreen}{compared to correlative SDMs}.
 We investigated the feasibility of using an evolutionary algorithm (called covariance matrix adaptation evolution strategy, CMA-ES) to calibrate process-based models using species distribution data.
@@
 \textcolor{darkgreen}{Most of them fall into two categories: informal and statistical calibration.
-On the one hand, statistical calibration assumes a data-generating model, and the likelihood function, which quantifies the probability of the observed data given the model parameters, is expressed mathematically.
-This likelihood is the foundation of maximum likelihood estimation and Bayesian inference, two commonly used methods for parameter estimation.
-In practice, deriving the likelihood function can be quite challenging for complex models with many parameters.
-One approach is to use numerical methods such as simulation-based inference or Approximate Bayesian Computation} [e.g.
+On the one hand, statistical calibration assumes a data-generating model, and a likelihood function, which quantifies the probability of the observed data given the model parameters, is expressed mathematically.
+This likelihood is the foundation of *maximum likelihood estimation* and *Bayesian inference*, two commonly used methods for parameter estimation.
+In practice, deriving the likelihood function can be quite challenging for complex models with many parameters and thresholding.
+One approach is to use numerical methods such as *simulation-based inference* or *approximate Bayesian computation*} [e.g.
 @Hartig2014]\textcolor{darkgreen}{.
 However, obtaining a reliable estimate of the likelihood of a complex process-based model can require a large number of simulated samples, which can be computationally expensive.
-On the other hand, using an informal objective function to measure the discrepancy between the model predictions and the observed data may be a reasonable alternative, even though it may not have the same statistical rigor.
-By minimizing the objective function, one is able to identify the parameter values that best fit the observed data.
-This optimization step can be achieved with many methods, either deterministic (e.g.
+On the other hand, informal calibration uses an informal objective function to measure the discrepancy between the model predictions and the observed data may be a reasonable alternative.
+By minimizing the objective function, one is able to identify the parameter values that best fit the observed data, even though it may not have the same statistical rigor as statistical calibration.
+This optimization step can be achieved with several methods, either deterministic (e.g.
 Nelder-Mead method) or stochastic (e.g.
@@
 
-\textcolor{darkgreen}{Our results showed that CMA-ES is a feasible and efficient optimizer for inverse calibration of complex ecological models.
+\textcolor{darkgreen}{Our results showed that CMA-ES is an efficient optimizer for inverse calibration of complex ecological models.
 The algorithm was able to find parameter sets that significantly improved the predictions of the calibrated models compared to the expert parametrization.
-However, our study also highlighted the issue of non-identifiability of parameter values due to the data limitation and the dependence between process-based model parameters, which may result in diverging parameter values even though the calibrated models describe the observed species distribution well.
-Nonetheless, our study provides a promising methodological framework for calibrating process-based models using CMA-ES, and for spreading the use of such models in predictive ecology.}
+However, our study also highlighted the issue of non-identifiability of parameter values due to the data limitation and the dependence between process-based model parameters, which may result in diverging parameter values even though the calibrated models describe the observed species distribution well.}
 
@@
 As well as being quasi parameter-free, CMA-ES has several structural advantages particularly useful for complex optimization problems.
-\textcolor{darkgreen}{Firstly, the algorithm's covariance matrix enables the learning of second-order information, which provides insights into pairwise dependencies between parameters.
-Secondly, the covariance matrix adaptation of CMA-ES is highly efficient in handling ill-conditioned and non-separable problems }[@Hansen2011]\textcolor{darkgreen}{.
-Lastly, CMA-ES's update mechanism of step size $\sigma$ (i.e.
+\textcolor{darkgreen}{First, the algorithm's covariance matrix enables the learning of second-order information, which provides insights into pairwise dependencies between parameters.
+Second, the covariance matrix adaptation of CMA-ES is highly efficient in handling ill-conditioned and non-separable problems }[@Hansen2011]\textcolor{darkgreen}{.
+Last, CMA-ES's update mechanism of step size $\sigma$ (i.e.
 the mutation force) helps prevent premature convergence }[@Hansen2001]\textcolor{darkgreen}{, allowing the algorithm to explore more of the search space.}
@@
 We did not attempt to validate our parametrizations using temporally or spatially independent data, and AUC was only used to determine if model outputs were consistent with species distributions.
-However, AUC is scale-invariant (it measures how well predictions are ranked rather than their absolute values), and prevents us from obtaining calibrated outputs with more ecological meaning.
-For example, in PHENOFIT, one species with a fitness of $0.8$ could be considered as absent whereas another one as present.
-Further work could thus be conducted to examine the effects of choosing a different objective function.
+\textcolor{darkgreen}{However, AUC is scale-invariant: it measures how well predictions are ranked rather than their absolute values.
+For example, with PHENOFIT, a species with a fitness of $0.8$ could be considered as absent while another one with the same fitness could be considered as present.
+Therefore, when it is used as an objective function for model calibration, we probably lack some precision and consistency among species’ parameters estimation.} Further work could thus be conducted to examine the effects of choosing a different objective function.
 
@@
 EU-Forest database is available on [FigShare](https://figshare.com/collections/A_high-resolution_pan-European_tree_occurrence_dataset/3288407).
-The R code associated with this work will be made available on GitLab upon acceptance of the manuscript.
+The R code associated with this work is available on [this GitHub repository](https://github.com/vvandermeersch/inverse_calibration), as well as on Zenodo ([10.5281/zenodo.7774981](https://doi.org/10.5281/zenodo.7774981)).
 

### 2023-03-28 manuscript/paper.tex
small changes before resubmission (minor revision)

--- 2023-03-21 manuscript/paper_revision.tex
@@
 
-\title{A novel approach to inverse calibration of process-based models
-using evolutionary algorithm and species distribution data  }
+\title{Estimating process-based model parameters from species
+distribution data using the evolutionary algorithm CMA-ES  }
 
@@
   hypotheses (statistical relationships \emph{vs} mechanistic
-  relationships) and their calibration methods (dependent \emph{vs}
-  generally independent of the species observed distributions).
+  relationships) and and their calibration methods
+  \textcolor{darkgreen}{(SDMs tend to be occurrence data-driven while PBMs tend to be prior-driven)}.
 \item
   One of the limitations to the use of process-based models is the
-  difficulty to parameterize them for a large number of species, while
-  one of the advantages of correlative SDMs is their easier calibration
-  using solely species occurrences\}. We investigated the feasibility of
-  using an evolutionary algorithm (called covariance matrix adaptation
-  evolution strategy, CMA-ES) to calibrate process-based models using
-  species distribution data. This method is well established in some
-  fields (robotics, aerospace research, \ldots), but has never been
-  used, to our knowledge, in ecology, despite its ability to deal with
-  very large space dimensions. Using tree species occurrence data across
-  Europe, we adapted the CMA-ES algorithm to find appropriate values of
-  model parameters. We estimated simultaneously 27 to 77 parameters of
-  two process-based models simulating forest tree's ecophysiology for
-  three species with varying range sizes and geographical distributions.
+  difficulty to parameterize them for a large number of species
+  \textcolor{darkgreen}{compared to correlative SDMs}. We investigated
+  the feasibility of using an evolutionary algorithm (called covariance
+  matrix adaptation evolution strategy, CMA-ES) to calibrate
+  process-based models using species distribution data. This method is
+  well established in some fields (robotics, aerospace research,
+  \ldots), but has never been used, to our knowledge, in ecology,
+  despite its ability to deal with very large space dimensions. Using
+  tree species occurrence data across Europe, we adapted the CMA-ES
+  algorithm to find appropriate values of model parameters. We estimated
+  simultaneously 27 to 77 parameters of two process-based models
+  simulating forest tree's ecophysiology for three species with varying
+  range sizes and geographical distributions.
 \item
@@
 \textcolor{darkgreen}{Most of them fall into two categories: informal and statistical calibration.
-On the one hand, statistical calibration assumes a data-generating model, and the likelihood function, which quantifies the probability of the observed data given the model parameters, is expressed mathematically.
-This likelihood is the foundation of maximum likelihood estimation and Bayesian inference, two commonly used methods for parameter estimation.
-In practice, deriving the likelihood function can be quite challenging for complex models with many parameters.
-One approach is to use numerical methods such as simulation-based inference or Approximate Bayesian Computation}
+On the one hand, statistical calibration assumes a data-generating model, and a likelihood function, which quantifies the probability of the observed data given the model parameters, is expressed mathematically.
+This likelihood is the foundation of *maximum likelihood estimation* and *Bayesian inference*, two commonly used methods for parameter estimation.
+In practice, deriving the likelihood function can be quite challenging for complex models with many parameters and thresholding.
+One approach is to use numerical methods such as *simulation-based inference* or *approximate Bayesian computation*}
 (e.g. \protect\hyperlink{ref-Hartig2014}{Hartig \emph{et al.}
@@
 However, obtaining a reliable estimate of the likelihood of a complex process-based model can require a large number of simulated samples, which can be computationally expensive.
-On the other hand, using an informal objective function to measure the discrepancy between the model predictions and the observed data may be a reasonable alternative, even though it may not have the same statistical rigor.
-By minimizing the objective function, one is able to identify the parameter values that best fit the observed data.
-This optimization step can be achieved with many methods, either deterministic (e.g.
+On the other hand, informal calibration uses an informal objective function to measure the discrepancy between the model predictions and the observed data may be a reasonable alternative.
+By minimizing the objective function, one is able to identify the parameter values that best fit the observed data, even though it may not have the same statistical rigor as statistical calibration.
+This optimization step can be achieved with several methods, either deterministic (e.g.
 Nelder-Mead method) or stochastic (e.g.
@@
 particularly useful for complex optimization problems.
-\textcolor{darkgreen}{Firstly, the algorithm's covariance matrix enables the learning of second-order information, which provides insights into pairwise dependencies between parameters.
-Secondly, the covariance matrix adaptation of CMA-ES is highly efficient in handling ill-conditioned and non-separable problems }(\protect\hyperlink{ref-Hansen2011}{Hansen
+\textcolor{darkgreen}{First, the algorithm's covariance matrix enables the learning of second-order information, which provides insights into pairwise dependencies between parameters.
+Second, the covariance matrix adaptation of CMA-ES is highly efficient in handling ill-conditioned and non-separable problems }(\protect\hyperlink{ref-Hansen2011}{Hansen
 \emph{et al.}
-2011})\textcolor{darkgreen}{. Lastly, CMA-ES's update mechanism of step size $\sigma$ (i.e. the mutation force) helps prevent premature convergence }(\protect\hyperlink{ref-Hansen2001}{Hansen
+2011})\textcolor{darkgreen}{. Last, CMA-ES's update mechanism of step size $\sigma$ (i.e. the mutation force) helps prevent premature convergence }(\protect\hyperlink{ref-Hansen2001}{Hansen
 \& Ostermeier
@@
 determine if model outputs were consistent with species distributions.
-However, AUC is scale-invariant (it measures how well predictions are
-ranked rather than their absolute values), and prevents us from
-obtaining calibrated outputs with more ecological meaning. For example,
-in PHENOFIT, one species with a fitness of \(0.8\) could be considered
-as absent whereas another one as present. Further work could thus be
-conducted to examine the effects of choosing a different objective
-function.
+\textcolor{darkgreen}{However, AUC is scale-invariant: it measures how well predictions are ranked rather than their absolute values.
+For example, with PHENOFIT, a species with a fitness of $0.8$ could be considered as absent while another one with the same fitness could be considered as present.
+Therefore, when it is used as an objective function for model calibration, we probably lack some precision and consistency among species’ parameters estimation.}
+Further work could thus be conducted to examine the effects of choosing
+a different objective function.
 
@@
 \href{https://figshare.com/collections/A_high-resolution_pan-European_tree_occurrence_dataset/3288407}{FigShare}.
-The R code associated with this work will be made available on GitLab
-upon acceptance of the manuscript.
+The R code associated with this work is available on
+\href{https://github.com/vvandermeersch/inverse_calibration}{this GitHub
+repository}, as well as on Zenodo
+(\href{https://doi.org/10.5281/zenodo.7774981}{10.5281/zenodo.7774981}).
 

### 2023-03-30 manuscript/paper.Rmd
change title and some little stuff

@@
 Two main types of species distribution models are used to project species range shifts in future climatic conditions: correlative and process-based models.
-Although there is some continuity between these two types of models, they are fundamentally different in their hypotheses (statistical relationships *vs* mechanistic relationships) and and their calibration methods \\textcolor{darkgreen}{(SDMs tend to be occurrence data-driven while PBMs tend to be prior-driven)}.
+Although there is some continuity between these two types of models, they are fundamentally different in their hypotheses (statistical relationships *vs* mechanistic relationships) and their calibration methods \\textcolor{darkgreen}{(SDMs tend to be occurrence data-driven while PBMs tend to be prior-driven)}.
 \n\n 2.
@@
 On the one hand, statistical calibration assumes a data-generating model, and a likelihood function, which quantifies the probability of the observed data given the model parameters, is expressed mathematically.
-This likelihood is the foundation of *maximum likelihood estimation* and *Bayesian inference*, two commonly used methods for parameter estimation.
+This likelihood is the foundation of \emph{maximum likelihood estimation} and \emph{Bayesian inference}, two commonly used methods for parameter estimation.
 In practice, deriving the likelihood function can be quite challenging for complex models with many parameters and thresholding.
-One approach is to use numerical methods such as *simulation-based inference* or *approximate Bayesian computation*} [e.g.
+One approach is to use numerical methods such as \emph{simulation-based inference} or \emph{approximate Bayesian computation} [e.g.
 @Hartig2014]\textcolor{darkgreen}{.
 However, obtaining a reliable estimate of the likelihood of a complex process-based model can require a large number of simulated samples, which can be computationally expensive.
-On the other hand, informal calibration uses an informal objective function to measure the discrepancy between the model predictions and the observed data may be a reasonable alternative.
+On the other hand, informal calibration uses an informal objective function to measure the discrepancy between the model predictions and the observed data.
 By minimizing the objective function, one is able to identify the parameter values that best fit the observed data, even though it may not have the same statistical rigor as statistical calibration.
@@
 - $\lambda$ candidate solutions are evaluated, i.e. model runs $\lambda$ times with $\lambda$ different parameter sets and the objective function is evaluated  
-- the best $\mu$ candidate solutionss are selected  
+- the best $\mu$ candidate solutions are selected  
 - the weighted mean candidate solution $m$ is computed (mean of the best $\mu$ parameter sets weighted by their objective function value)  
@@
 - new $\lambda$ candidate solutions are sampled in a normal distribution $\mathcal{N}(m, \sigma C)$, with both recombination (via the favorite solution $m$) and mutations (via the perturbations $\sigma C$)
-One of the strength of this approach lies in the combination of rank-$\mu$-update, where prior information from previous generations is exploited (mean of the previous covariance matrices, with a higher weight for recent generations), and cumulation, where correlations between generations are retained in an evolution path (sum of consecutive steps), to update the covariance matrix at each step [see @Hansen2016 for a detailed description of the algorithm].
+One of the strengths of this approach lies in the combination of rank-$\mu$-update, where prior information from previous generations is exploited (mean of the previous covariance matrices, with a higher weight for recent generations), and cumulation, where correlations between generations are retained in an evolution path (sum of consecutive steps), to update the covariance matrix at each step [see @Hansen2016 for a detailed description of the algorithm].
 
@@
 First, like any model, although they have a certain level of complexity, they are not a perfect representation of the reality, and their inherent structural errors increase the probability of finding parameter values that deviate from the true values of the underlying processes [see @Oberpriller2021].
-Secondly, they do not necessarily include all the environmental factors at stake.
+Second, they do not necessarily include all the environmental factors at stake.
 For example, pedologic variable in PHENOFIT only involve the water holding capacity.
 In this model, other variables such as pH or soil texture are not considered.
-Thirdly, and more importantly, they are used here to represent the fundamental niche, and to estimate the potential distribution of the species using pedoclimatic variables.
+Third, and more importantly, they are used here to represent the fundamental niche, and to estimate the potential distribution of the species using pedoclimatic variables.
 When calibrated against observed distributions, which represent the realized niche, they face the same issues as correlative models, and their parameter estimated can be distorted because compensating for processes not represented in the model (e.g.
@@
 In such a case, expert knowledge, observations and measurements are necessary to determine *a posteriori* which estimates are the most realistic.
-This is possible in the case of process-based SDMs, and usually not feasable in the case of correlative SDMs.
+This is possible in the case of process-based SDMs, and usually not feasible in the case of correlative SDMs.
 A combination of both expert and inverse calibrations might offer a new perspective for spreading the use of process-based models in predictive ecology, especially for climate change impact studies.

### 2023-03-30 manuscript/paper.tex
change title and some little stuff

@@
   hypotheses (statistical relationships \emph{vs} mechanistic
-  relationships) and and their calibration methods
+  relationships) and their calibration methods
   \textcolor{darkgreen}{(SDMs tend to be occurrence data-driven while PBMs tend to be prior-driven)}.
@@
 developed in the last decades.
-\textcolor{darkgreen}{Most of them fall into two categories: informal and statistical calibration.
-On the one hand, statistical calibration assumes a data-generating model, and a likelihood function, which quantifies the probability of the observed data given the model parameters, is expressed mathematically.
-This likelihood is the foundation of *maximum likelihood estimation* and *Bayesian inference*, two commonly used methods for parameter estimation.
-In practice, deriving the likelihood function can be quite challenging for complex models with many parameters and thresholding.
-One approach is to use numerical methods such as *simulation-based inference* or *approximate Bayesian computation*}
-(e.g. \protect\hyperlink{ref-Hartig2014}{Hartig \emph{et al.}
+\textbackslash textcolor\{darkgreen\}\{Most of them fall into two
+categories: informal and statistical calibration. On the one hand,
+statistical calibration assumes a data-generating model, and a
+likelihood function, which quantifies the probability of the observed
+data given the model parameters, is expressed mathematically. This
+likelihood is the foundation of \emph{maximum likelihood estimation} and
+\emph{Bayesian inference}, two commonly used methods for parameter
+estimation. In practice, deriving the likelihood function can be quite
+challenging for complex models with many parameters and thresholding.
+One approach is to use numerical methods such as
+\emph{simulation-based inference} or
+\emph{approximate Bayesian computation} (e.g.
+\protect\hyperlink{ref-Hartig2014}{Hartig \emph{et al.}
 2014})\textcolor{darkgreen}{.
 However, obtaining a reliable estimate of the likelihood of a complex process-based model can require a large number of simulated samples, which can be computationally expensive.
-On the other hand, informal calibration uses an informal objective function to measure the discrepancy between the model predictions and the observed data may be a reasonable alternative.
+On the other hand, informal calibration uses an informal objective function to measure the discrepancy between the model predictions and the observed data.
 By minimizing the objective function, one is able to identify the parameter values that best fit the observed data, even though it may not have the same statistical rigor as statistical calibration.
@@
 objective function is evaluated\\
-- the best \(\mu\) candidate solutionss are selected\\
+- the best \(\mu\) candidate solutions are selected\\
 - the weighted mean candidate solution \(m\) is computed (mean of the
-best \(\mu\) parameter sets weighted by their objective function value)
+best \(\mu\) parameter sets weighted by their objective function
+value)\\
 - covariance matrix \(C\) and step size \(\sigma\) are updated (with
@@
 \(\sigma C\))\\
-One of the strength of this approach lies in the combination of
+One of the strengths of this approach lies in the combination of
 rank-\(\mu\)-update, where prior information from previous generations
@@
 \protect\hyperlink{ref-Oberpriller2021}{Oberpriller \emph{et al.}
-2021}). Secondly, they do not necessarily include all the environmental
+2021}). Second, they do not necessarily include all the environmental
 factors at stake. For example, pedologic variable in PHENOFIT only
 involve the water holding capacity. In this model, other variables such
-as pH or soil texture are not considered. Thirdly, and more importantly,
+as pH or soil texture are not considered. Third, and more importantly,
 they are used here to represent the fundamental niche, and to estimate
@@
 posteriori} which estimates are the most realistic. This is possible in
-the case of process-based SDMs, and usually not feasable in the case of
+the case of process-based SDMs, and usually not feasible in the case of
 correlative SDMs. A combination of both expert and inverse calibrations

### 2023-04-17 climate_compilation/format/formatting_data.R
add script and functions to compile daily climate required by the models

@@ -1,0 +1,79 @@
+#########################################################################
+# Script to format ERA5-Land climate data as required by Phenofit model #
+#########################################################################
+
+# author : V. Van der Meersch
+# date : 16/02/2022
+
+wd <- "C:/Users/vandermeersch/Documents/CEFE/phd/present_calibration/climate_data/format/phenofit/"
+
+
+## Packages needed
+
+library(ncdf4)
+library(future)
+library(future.apply)
+library(dplyr)
+library(raster)
+library(data.table)
+
+
+## Functions needed
+
+source(paste0(wd, 'functions/phenofit_formatting.R')) # just to format variables for Phenofit, v2 with raster format (faster)
+source(paste0(wd, 'functions/phenofit_processing_and_formatting.R')) # to calculate supplementary variables and format them for Phenofit, v2 with raster format (faster)
+source(paste0(wd, 'functions/phenofit_compute_PET2.R')) # to calculate evapotranspiration, v2 with different parallel architecture
+source(paste0(wd, 'functions/phenofit_get_altitude.R')) # to get altitude
+source(paste0(wd, 'functions/phenofit_get_WHC.R')) # to get WHC
+
+
+## Settings
+
+rawdata_folder <- "D:/climate/ERA5-Land/raw/"
+processeddata_folder <- "D:/climate/ERA5-Land/phenofit_format/"
+
+
+## Compile daily data from hourly data
+phenofit_formatting(1969:2000, "2m_dewpoint_temperature", 'mean', rawdata_folder, processeddata_folder, ncores=5)
+gc()
+
+phenofit_formatting(1969:2000, "2m_temperature", "min", rawdata_folder, processeddata_folder, ncores=5)
+gc()
+
+phenofit_formatting(1969:2000, "2m_temperature", "max", rawdata_folder, processeddata_folder, ncores=5)
+gc()
+
+phenofit_formatting(1969:2000, "total_precipitation", "sum", rawdata_folder, processeddata_folder, ncores=4)
+gc()
+
+phenofit_formatting(1969:2000, "surface_solar_radiation_downwards", "sum", rawdata_folder, processeddata_folder, ncores=5) #used only for ETP calculation
+gc()
+
+phenofit_processing_and_formatting(1969:2000, "wind", "mean", rawdata_folder, processeddata_folder, ncores=5) #used only for ETP calculation
[…]

### 2023-04-17 climate_compilation/format/functions/phenofit_formatting.R

@@ -1,0 +1,202 @@
+source(paste0(wd, 'functions/get_var_name.R'))
+source(paste0(wd, 'functions/get_comments.R'))
+source(paste0(wd, 'functions/convert_units.R'))
+
+# author : V. Van der Meersch
+
+# Package for reporting progress updates
+library(progressr)
+handlers(global = TRUE)
+handlers("txtprogressbar")
+
+phenofit_formatting2 <- function(years, var, stat, rd_folder, pd_folder, ncores=1, landsea_mask=FALSE){
+  
+  if(landsea_mask){
+    landsea_file <- paste0(rd_folder, 'lsm_1279l4_0.1x0.1.grb_v4_unpack.nc')
+    landsea_mask <- brick(landsea_file, varname='lsm')
+    landsea_mask <- rotate(landsea_mask) # longitude : go from 0 - 360 to -180 - 180
+    e <- extent(-14, 40, 34,72)
+    landsea_mask <- crop(landsea_mask, e)
+    landsea_mask[landsea_mask<0.1] <- NA
+    landsea_mask_df <- na.omit(as.data.frame(landsea_mask, xy=TRUE))
+    names(landsea_mask_df) <- c("lon", "lat", "mask")
+    landsea_mask_df <- round(landsea_mask_df, 2)
+  }
+  
+  if(var == "wind" | var == "relative_humidity"){
+    stop("For wind or relative humidity, you need to use phenofit_processing_and_formatting function !")
+  }
+  else if( !(var %in% c("2m_temperature", "total_precipitation", "surface_solar_radiation_downwards", 
+                        "potential_evaporation", "2m_dewpoint_temperature")) ){
+    stop("Cannot recognize the variable you asked for !")
+  }
+  
+  # To follow the progress
+  p <- progressor(length(years)*12)
+  
+  # Get variable name as coded in NC file
+  var_name <- get_var_name(var, type = "ERA5")
+  
+  months <- c("01", "02", "03", "04", "05", "06",
+               "07", "08", "09", "10", "11", "12")
+  
+  plan(multisession, workers = ncores)
+  
+  # Parallel loop (in case of parallel computing)
+  out <- future_apply(array(1:ncores), 1, function(i){
+    
+    # To determine number of years processed by core (in case of parallel computing)
+    quot <- length(years)%/%ncores
+    rem <- length(years)%%ncores
[…]

### 2023-04-17 climate_compilation/format/functions/phenofit_get_altitude.R

@@ -1,0 +1,51 @@
+# author : V. Van der Meersch
+
+source(paste0(wd, 'functions/get_comments.R'))
+source(paste0(wd, 'functions/geopotential_to_geometric.R'))
+
+# loc : data.frame containing longitude and latitude for which altitude is needed
+# if NULL, compute for every point !
+
+phenofit_get_altitude <- function(loc = NULL, rd_folder, pd_folder){
+
+  # Open geopotential file
+  geopotential_file <- paste0(rd_folder, "geo_1279l4_0.1x0.1.grib2_v4_unpack.nc")
+  
+  geopotential_raster <- brick(geopotential_file, varname="z")
+  
+  geopotential_df <- as.data.frame(geopotential_raster, xy=TRUE)
+  names(geopotential_df) <- c("lon", "lat", "z")
+  
+  # Need to convert longitude from (0-360) to (-180, 180)
+  geopotential_df$lon <- ((geopotential_df$lon+180)%%360)-180.0
+  
+  # Convert to geometric height
+  geopotential_df$z <- geopotential_to_geometric(geopotential_df$z)
+  
+  if(is.null(loc)){
+    alt_needed <- geopotential_df
+  }else{
+    alt_needed <- left_join(round(loc,2),round(geopotential_df,2),by=c("lat","lon"))  # round is important, of left_join won't work
+  }
+  
+  alt_needed$lon <- loc$lon
+  alt_needed$lat <- loc$lat
+  
+  # Fill name
+  processed_file <- paste0(pd_folder, "ERA5LAND_Altitude.fit")
+  
+  # Fill file
+  con <- file(processed_file, open="wt")
+  writeLines("Elevation datafile for Phenofit model", con)
+  writeLines(paste("Created on RStudio, by user", Sys.getenv("USERNAME") ,", on the", Sys.Date()), con)
+  comments <- get_comments(var = "Altitude")
+  writeLines(comments, con)
+  writeLines(" ", con)
+  write.table(alt_needed, file = con, sep = "\t",
+              row.names = FALSE, col.names= FALSE)
+  close(con)
+  
+  # Free memory
+  gc(verbose = FALSE)
+
+}

### 2023-04-17 climate_compilation/format/functions/phenofit_processing_and_formatting.R

@@ -1,0 +1,184 @@
+source(paste0(wd, 'functions/get_var_name.R'))
+source(paste0(wd, 'functions/get_comments.R'))
+source(paste0(wd, 'functions/convert_units.R'))
+
+# author : V. Van der Meersch
+
+# Package to calculate relative humidity with the ratio of saturation vapor pressure
+library(humidity)
+
+# Package for reporting progress updates
+library(progressr)
+handlers(global = TRUE)
+handlers("txtprogressbar")
+
+phenofit_processing_and_formatting <- function(years, var, stat, rd_folder, pd_folder, ncores=1){
+  
+  if(var %in% c("2m_temperature", "total_precipitation", "surface_solar_radiation_downwards")){
+    stop(paste0("For ", var, ", you need to use phenofit_formatting function !"))
+  }
+  else if( var != "wind" & var != "relative_humidity" ){
+    stop("Cannot recognize the variable you asked for !")
+  }
+  
+  # To follow the progress
+  p <- progressor(length(years)*12)
+  
+  months <- c("01", "02", "03", "04", "05", "06",
+              "07", "08", "09", "10", "11", "12")
+  
+  # Get variable names as coded in NC file
+  if(var == "wind"){
+    var_1 <- "10m_u_component_of_wind"
+    var_name_1 <- get_var_name(var_1, type = "ERA5")
+    var_2 <- "10m_v_component_of_wind"
+    var_name_2 <- get_var_name(var_2, type = "ERA5")
+  }
+  if(var == "relative_humidity"){
+    var_1 <- "2m_temperature"
+    var_name_1 <- get_var_name(var_1, type = "ERA5")
+    var_2 <- "2m_dewpoint_temperature"
+    var_name_2 <- get_var_name(var_2, type = "ERA5")
+  }
+  
+  
+  plan(multisession, workers = ncores)
+  
+  # Parallel loop in case of parallel computing
+  out <- future_apply(array(1:ncores), 1, function(i){
+    
+    # To determine number of years processed by core in case of parallel computing
+    quot <- length(years)%/%ncores
+    rem <- length(years)%%ncores
+    if(i<ncores){ years_i <- years[(1+(i-1)*quot):(quot*i)] }
+    else{ years_i <- years[(1+(i-1)*quot):((quot*i)+rem)] }
+    
+    
+    # Loop on years
+    for(yr in years_i){
+      
+      p(message = paste("Processing year", yr, "..."), class = "sticky", amount = 0)
+      
[…]

### 2023-04-18 data_compilation/format/formatting_soil_data.R
script to format WHC as requried by PHENOFIT

@@ -1,0 +1,82 @@
+
+#----------------------------------------------------------#
+# Script to format soil data as required by PHENOFIT model #
+#----------------------------------------------------------#
+
+### Settings ###
+
+rawdata_folder <- "D:/soil"
+
+# ERA5_Land grid (0.1deg)
+ERA5land_grid <- raster("D:/climate/ERA5-Land/raw/2m_dewpoint_temperature_1969_01.nc")
+
+# 7 soil depths : 0, 5, 15, 30, 60, 100, and 200 cm depth
+dep <- c(0, 5, 15, 30, 60, 100, 200)
+# 6 layers ?
+thick <- c(5, 10, 15, 30, 40, 100)
+
+# Folders
+EU_SHD_folder <- file.path(rawdata_folder, "EU_SoilHydroGrids_1km/raw") # folder where EU-SoilHydroGrids data are stored
+SoilGrids_folder <- file.path(rawdata_folder, "SoilGrids250m") # folder where SoilGrids data are stored
+
+
+
+### Load data ###
+
+# Load field capacity data
+filenames <- list.files(EU_SHD_folder, pattern="^FC", full.names=TRUE)
+fc_stack <- raster::stack(filenames) %>% crop(ERA5land_grid)
+
+# Load wilting point data
+filenames <- list.files(EU_SHD_folder, pattern="^WP", full.names=TRUE)
+wp_stack <- raster::stack(filenames) %>% crop(ERA5land_grid)
+
+# Load volumetric percentage of coarse fragments (>2 mm)
+filenames <- list.files(SoilGrids_folder, pattern="^CRFVOL", full.names=TRUE)
+crf_stack <- raster::stack(filenames) %>% crop(ERA5land_grid)
+
+
+
+### Compile data ###
+
+# Change resolution, from 250m/1km to 0.1deg
+fc_stack <- resample(fc_stack, ERA5land_grid, method="bilinear")
+wp_stack <- resample(wp_stack, ERA5land_grid, method="bilinear")
+crf_stack <- resample(crf_stack, ERA5land_grid, method="bilinear")
+depth <- resample(depth, ERA5land_grid, method="bilinear")
+
+# From one depth layer to layered weights
+depth_weights <- depth
[…]

### 2023-04-18 data_compilation/download/ERA5_land_download.py
moving scripts to new folder

@@ -1,0 +1,58 @@
+import cdsapi
+import requests
+
+c = cdsapi.Client(full_stack=True)
+
+# Accumulations (ie variables accumulated from the beginning of the forecast to the end of the forecast step, see ERA5-Land documentation)
+acc_vars = ['total_precipitation', 'surface_solar_radiation_downwards', 'potential_evaporation']
+
+# Retrieve all months for a given year
+months = ['01', '02', '03',
+            '04', '05', '06',
+            '07', '08', '09',
+            '10', '11', '12']
+            
+# Retrieve all days for a given month
+days = ['01','02','03', '04','05','06',
+		'07','08','09', '10','11','12',
+		'13','14','15', '16','17','18',
+		'19','20','21', '22','23','24',
+		'25','26','27', '28','29','30',
+		'31']
+        
+
+
+
+
+
+def era5_land(years, var):
+    print(var)
+    # check if the variable is accumulated    
+    if var in acc_vars:     
+        print("Accumulated variable, script will download only the end of the forecast step (00:00)")
+        hours = ['00:00']
+    else:
+        hours = ['00:00','01:00','02:00', '03:00','04:00','05:00',
+	     '06:00','07:00','08:00', '09:00','10:00','11:00',
+	     '12:00','13:00','14:00', '15:00','16:00','17:00',
+	     '18:00','19:00','20:00', '21:00','22:00','23:00']
+
+    # Loop over years and months
+    for yr in years:
+        for mn in months:
+            r = c.service("catalogue.retrieve",
+            'reanalysis-era5-land',
+            {"product_type": "reanalysis", 
+            "variable": var,
+            "year": yr,
+            "month": mn,
+            "day": days,
+            'time': hours,
+            "area" : str(72)+'/'+str(-14)+'/'+str(34)+'/'+str(40)})
+            
+            file_name = var + "_" + yr + "_" + mn + ".nc"
+            
+            r.download(file_name)
+        
+        
+            

### 2023-04-18 data_compilation/download/downloading_era5_land.R

@@ -1,0 +1,31 @@
+library(reticulate)
+
+setwd("D:/climate/ERA5_Land/raw") # folder where files will be downloaded
+
+# Python config
+use_python("C:/Users/vandermeersch/AppData/Local/Programs/Python/Python310/python.exe", required=T)
+cdsapi <- import("cdsapi")
+source_python("C:/Users/vandermeersch/Documents/CEFE/projects/inverse_calibration/climate_compilation/download/ERA5_land_download.py") # custom script
+
+
+# select the variable; name must be a valid ERA5 CDS API name
+var <- '10m_u_component_of_wind'
+
+# For valid keywords, see Table 2 of:
+# https://confluence.ecmwf.int/display/CKB/ERA5-Land%3A+data+documentation#ERA5Land:datadocumentation-parameterlistingParameterlistings
+
+# Select years
+years <- as.character(1969:2000)
+
+# Maximum number of per-user requests that access the online CDS data is 2
+# plan(multisession, workers = 2) ?!
+# But Python from reticulate cannot run in parallel because of external pointers !
+# We can launch manually 2 R sessions, with these two subsets:
+# years_1 <- years[1:(length(years)%/%2)]
+# years_2 <- years[(length(years)%/%2+1):length(years)]
+
+#Run script
+era5_land(years = years, var = var)
+
+
+

### 2023-04-18 data_compilation/format/formatting_climate_data.R

@@ -1,0 +1,79 @@
+#-----------------------------------------------------------------------#
+# Script to format ERA5-Land climate data as required by Phenofit model #
+#-----------------------------------------------------------------------#
+
+# author : V. Van der Meersch
+# date : 16/02/2022
+
+wd <- "~/projects/inverse_calibration/data_compilation/format"
+
+
+## Packages needed
+
+library(ncdf4)
+library(future)
+library(future.apply)
+library(dplyr)
+library(raster)
+library(data.table)
+
+
+## Functions needed
+
+source(paste0(wd, 'functions/phenofit_formatting.R')) # just to format variables for Phenofit, v2 with raster format (faster)
+source(paste0(wd, 'functions/phenofit_processing_and_formatting.R')) # to calculate supplementary variables and format them for Phenofit, v2 with raster format (faster)
+source(paste0(wd, 'functions/phenofit_compute_PET2.R')) # to calculate evapotranspiration, v2 with different parallel architecture
+source(paste0(wd, 'functions/phenofit_get_altitude.R')) # to get altitude
+source(paste0(wd, 'functions/phenofit_get_WHC.R')) # to get WHC
+
+
+## Settings
+
+rawdata_folder <- "D:/climate/ERA5-Land/raw/"
+processeddata_folder <- "D:/climate/ERA5-Land/phenofit_format/"
+
+
+## Compile daily data from hourly data
+phenofit_formatting(1969:2000, "2m_dewpoint_temperature", 'mean', rawdata_folder, processeddata_folder, ncores=5)
+gc()
+
+phenofit_formatting(1969:2000, "2m_temperature", "min", rawdata_folder, processeddata_folder, ncores=5)
+gc()
+
+phenofit_formatting(1969:2000, "2m_temperature", "max", rawdata_folder, processeddata_folder, ncores=5)
+gc()
+
+phenofit_formatting(1969:2000, "total_precipitation", "sum", rawdata_folder, processeddata_folder, ncores=4)
+gc()
+
+phenofit_formatting(1969:2000, "surface_solar_radiation_downwards", "sum", rawdata_folder, processeddata_folder, ncores=5) #used only for ETP calculation
+gc()
+
+phenofit_processing_and_formatting(1969:2000, "wind", "mean", rawdata_folder, processeddata_folder, ncores=5) #used only for ETP calculation
+gc()
+
[…]

### 2023-04-18 data_compilation/format/functions/compute_PET.R

@@ -1,0 +1,120 @@
+
+# Function adapted by V. Van der Meersch from Evapotranspiration package (created by Danlu Guo)
+# Added a modification to correct polar night effect...
+
+
+# method = PenmanMonteith or PriestleyTaylor
+# physically based PenmanMonteith (PM) approach is presently considered as the state-of-the art 
+# PriestleyTaylor (PT) data requirement does not include some hardly available meteorological variables as wind speed or relative humidity
+# PT could fail in mimicking PM values under dry conditions or in zones where wind speed is relatively high
+
+
+# elev : ground elevation above mean sea level in m
+# latitude in radians
+
+
+compute_PET <- function(data, elev, lat_rad, constants, alpha = 0.23, method){
+  
+  if(method == "PriestleyTaylor"){
+    
+    Ta <- (data$Tmax + data$Tmin)/2
+    
+    # Saturated vapour pressure (August–Roche–Magnus approximation)
+    vs_Tmax <- 0.6108 * exp(17.27 * data$Tmax/(data$Tmax + 237.3))
+    vs_Tmin <- 0.6108 * exp(17.27 * data$Tmin/(data$Tmin + 237.3))
+    vas <- (vs_Tmax + vs_Tmin)/2
+    
+    # Vapour pressure
+    vabar <- (vs_Tmin * data$RHmax/100 + vs_Tmax * data$RHmin/100)/2 
+    
+    P <- 101.3 * ((293 - 0.0065 * elev)/293)^5.26 # atmospheric pressure
+    # effect of atmospheric pressure (i.e. Eevaporation at high altitudes promoted due to low atmospheric pressure) is small
+    # thus, the average value of atmospheric pressure for a location is sufficient
+    
+    delta <- 4098 * (0.6108 * exp((17.27 * Ta)/(Ta + 237.3)))/((Ta + 237.3)^2) # slope of vapour pressure curve
+    gamma <- 0.00163 * P/constants$lambda # psychrometric constant
+    d_r2 <- 1 + 0.033 * cos(2 * pi/365 * data$J) # dr is the inverse relative distance Earth-Sun
+    delta2 <- 0.409 * sin(2 * pi/365 * data$J - 1.39) # solar dedication
+    
+    # Added by Victor
+    aux <- -tan(lat_rad) * tan(delta2) 
[…]

### 2023-04-18 data_compilation/format/functions/convert_units.R

@@ -1,0 +1,29 @@
+# Function to convert units as needed by Phenofit
+
+# author : V. Van der Meersch
+
+convert_units <- function(value, var){
+  if(var == "2m_temperature" | var == "2m_dewpoint_temperature"){
+    # Kelvin to Celsius
+    value_c <- sapply(value, function(i){i-273.15})
+  }
+  if(var == "total_precipitation"){
+    # m to mm
+    value_c <- sapply(value, function(i){i*1000})
+  }
+  if(var == "surface_solar_radiation_downwards"){
+    # J/m2 to MJ/m2
+    value_c <- sapply(value, function(i){i/1000000})
+  }
+  if(var == "surface_pressure"){
+    # Pa to kPa
+    value_c <- sapply(value, function(i){i/1000})
+  }
+  if(var == "potential_evaporation"){
+    # m to mm, and negative to positive
+    value_c <- sapply(value, function(i){- i*1000})
+  }
+  
+  return(value_c)
+}
+

### 2023-04-18 data_compilation/format/functions/geopotential_to_geometric.R

@@ -1,0 +1,22 @@
+# Function to convert geopotential height to geometric height
+# see https://confluence.ecmwf.int/display/CKB/ERA5%3A+compute+pressure+and+geopotential+on+model+levels%2C+geopotential+height+and+geometric+height
+
+# author : V. Van der Meersch
+
+geopotential_to_geometric <- function(h){
+  
+  # Earth's gravitational acceleration
+  g0 <-  9.80665
+  
+  # Mean radius of the earth
+  re <- 6371000
+  
+  # Get geopotential height in meters
+  h_m <- h/g0
+  
+  # Convert to geometric height 
+  z <- re*h_m/(re-h_m)
+  
+  return(z)
+  
+}

### 2023-04-18 data_compilation/format/functions/get_comments.R

@@ -1,0 +1,53 @@
+# Function to get Phenofit input variable definitions
+
+# author : V. Van der Meersch
+
+get_comments <- function(var){
+  if(var=='glo'){
+    return("Variable : daily global radiation (MJ/m²)")
+  }
+  else if(var=='pre'){
+    return("Variable : daily precipitation, comprising rain and snow (mm)")
+  }
+  else if(var=='prs'){
+    return("Variable : daily mean surface pressure (kPa)")
+  }
+  else if(var=='RH'){
+    return("Variable : daily mean 2m relative humidity (%) calculated with vapor pressure ratio (Clausius-Clapeyron relation)")
+  }
+  else if(var=='tmn'){
+    return("Variable : daily mimimal 2m temperature (°C)")
+  }
+  else if(var=='tmp'){
+    return("Variable : daily mean 2m temperature (°C)")
+  }
+  else if(var=='tmx'){
+    return("Variable : daily maximal 2m temperature (°C)")
+  }
+  else if(var=='dtm'){
+    return("Variable : daily mean 2m dewpoint temperature (°C)")
+  }
+  else if(var=='wnd'){
+    return("Variable : daily mean 10m wind speed (m/s) - used only to compute evapotranspiration")
+  }
+  else if(var=='RHmin'){
+    return("Variable : daily miminal 2m relative humidity (%) - used only to compute evapotranspiration")
+  }
+  else if(var=='RHmax'){
+    return("Variable : daily maximal 2m relative humidity (%) - used only to compute evapotranspiration")
+  }
+  else if(var=='petPenmanMonteith'){
+    return("Variable : potential evapotranspiration (mm) calculated with Penman-Monteith formulation (FAO-56 hypothetical short grass method) ")
+  }
+  else if(var=='petPriestleyTaylor'){
+    return("Variable : potential evapotranspiration (mm) calculated with Priestley-Taylor formulation")
+  }
+  else if(var=='Altitude'){
+    return("Variable : altitude (m) calculated from geopotential height")
+  }
+  else if(var=='WHC'){
+    return("Variable : water holding capacity (mm)")
+  }
+  
+  
+}

### 2023-04-18 data_compilation/format/functions/get_var_name.R

@@ -1,0 +1,33 @@
+# Function to get variable name as coded in NC files of ERA5-Land database
+# Or as needed as input variables for Phenofit
+
+# author : V. Van der Meersch
+
+get_var_name <- function(var, stat=NULL, type){
+  long_names <- c("2m_temperature", "total_precipitation", "surface_solar_radiation_downwards", "surface_pressure", 
+                  "wind", "relative_humidity", "2m_dewpoint_temperature", "10m_u_component_of_wind", "10m_v_component_of_wind",
+                  "potential_evaporation")
+  era5_names <- c("t2m", "tp", "ssrd", "ps",  NA, NA, "dpt", "u10", "v10", "pev")
+  
+  if(type == "ERA5"){ return(era5_names[which(long_names == var)]) }
+  
+  if(type == "PHENOFIT" | type=="Phenofit"){
+    if(stat=="mean"){
+      phenofit_names <- c("tmp", NA, NA, "prs", "wnd", "RH", "dtm", NA, NA, NA)
+      return(phenofit_names[which(long_names == var)])
+    }
+    if(stat=="max"){
+      phenofit_names <- c("tmx", NA, NA, NA, NA, "RHmax", NA, NA, NA, NA)
+      return(phenofit_names[which(long_names == var)])
+    }
+    if(stat=="min"){
+      phenofit_names <- c("tmn", NA, NA, NA, NA, "RHmin", NA, NA, NA, NA)
+      return(phenofit_names[which(long_names == var)])
+    }
+    if(stat=="sum"){
+      phenofit_names <- c(NA, "pre", "glo", NA, NA, NA, NA, NA, NA, "pet")
+      return(phenofit_names[which(long_names == var)])
+    }
+  }
+  
+}

### 2023-04-18 data_compilation/format/functions/phenofit_compute_PET.R

@@ -1,0 +1,155 @@
+# author : V. Van der Meersch
+
+# Loading custom functions
+source(paste0(wd, 'functions/compute_PET.R')) 
+source(paste0(wd, 'functions/get_comments.R'))
+
+# Package for reporting progress updates
+library(progressr)
+handlers(global = TRUE)
+handlers("txtprogressbar")
+
+
+# Needed constants
+PET_constants <- data.frame(lambda = 2.45, Gsc = 0.082, sigma = 4.903e-09, alphaPT = 1.26, G = 0, z = 10)
+  # lambda : latent heat of vaporisation - Gsc : solar constant - sigma : Stefan-Boltzmann constan
+  # alphaPT : Priestley-Taylor coefficient - G : soil heat flux (= 0 with daily time step)
+  # z : height of wind instrument in m (= 10 in ERA5-Land data)
+
+
+# Function parameters :
+  # - pd_folder : folder where climate data (Phenofit format) are located
+  # this pd_folder needs two subfolders : "/pet_PenmanMonteith" and "/pet_PriestleyTaylor"
+  #
+  # - method = PenmanMonteith or PriestleyTaylor (see compute_PET function)
+  #
+  # - years : e.g. 1990:2000
+  #
+  # - ncores : for parallelisation over years
+ 
+
+phenofit_compute_PET <- function(years, pd_folder, method, ncores, transform_NA_and_zero=FALSE){
+  
+  pd_folder_method <- paste0(pd_folder, "pet_", method, "/")
+  if(transform_NA_and_zero){
+    pd_folder_method <- paste0(pd_folder, "pet_", method, "_trans/")
+  }
+  
+  plan(multisession, workers = ncores)
+  
+  # Get altitude
+  alt_file <- paste0(pd_folder, "ERA5LAND_Altitude.fit")
+  alt <- fread(alt_file, showProgress=F)
+  nrows <- nrow(alt)
+  
+  p <- progressor(length(years))
+  
+  # Parallel loop in case of parallel computing
+  out <- future_apply(array(1:ncores), 1, function(i){
+    
+    # To determine number of years processed by core in case of parallel computing
+    quot <- length(years)%/%ncores
+    rem <- length(years)%%ncores
+    if(i<ncores){ years_i <- years[(1+(i-1)*quot):(quot*i)] 
[…]

### 2023-04-18 climate_compilation/download/ERA5_land_download.py
do some cleaning

@@ -1,58 +1,0 @@
-import cdsapi
-import requests
-
-c = cdsapi.Client(full_stack=True)
-
-# Accumulations (ie variables accumulated from the beginning of the forecast to the end of the forecast step, see ERA5-Land documentation)
-acc_vars = ['total_precipitation', 'surface_solar_radiation_downwards', 'potential_evaporation']
-
-# Retrieve all months for a given year
-months = ['01', '02', '03',
-            '04', '05', '06',
-            '07', '08', '09',
-            '10', '11', '12']
-            
-# Retrieve all days for a given month
-days = ['01','02','03', '04','05','06',
-		'07','08','09', '10','11','12',
-		'13','14','15', '16','17','18',
-		'19','20','21', '22','23','24',
-		'25','26','27', '28','29','30',
-		'31']
-        
-
-
-
-
-
-def era5_land(years, var):
-    print(var)
-    # check if the variable is accumulated    
-    if var in acc_vars:     
-        print("Accumulated variable, script will download only the end of the forecast step (00:00)")
-        hours = ['00:00']
-    else:
-        hours = ['00:00','01:00','02:00', '03:00','04:00','05:00',
-	     '06:00','07:00','08:00', '09:00','10:00','11:00',
-	     '12:00','13:00','14:00', '15:00','16:00','17:00',
-	     '18:00','19:00','20:00', '21:00','22:00','23:00']
-
-    # Loop over years and months
-    for yr in years:
-        for mn in months:
-            r = c.service("catalogue.retrieve",
-            'reanalysis-era5-land',
-            {"product_type": "reanalysis", 
-            "variable": var,
-            "year": yr,
-            "month": mn,
-            "day": days,
-            'time': hours,
-            "area" : str(72)+'/'+str(-14)+'/'+str(34)+'/'+str(40)})
-            
-            file_name = var + "_" + yr + "_" + mn + ".nc"
-            
-            r.download(file_name)
-        
-        
-            

### 2023-04-18 climate_compilation/download/downloading_era5_land.R

@@ -1,31 +1,0 @@
-library(reticulate)
-
-setwd("D:/climate/ERA5_Land/raw") # folder where files will be downloaded
-
-# Python config
-use_python("C:/Users/vandermeersch/AppData/Local/Programs/Python/Python310/python.exe", required=T)
-cdsapi <- import("cdsapi")
-source_python("C:/Users/vandermeersch/Documents/CEFE/projects/inverse_calibration/climate_compilation/download/ERA5_land_download.py") # custom script
-
-
-# select the variable; name must be a valid ERA5 CDS API name
-var <- '10m_u_component_of_wind'
-
-# For valid keywords, see Table 2 of:
-# https://confluence.ecmwf.int/display/CKB/ERA5-Land%3A+data+documentation#ERA5Land:datadocumentation-parameterlistingParameterlistings
-
-# Select years
-years <- as.character(1969:2000)
-
-# Maximum number of per-user requests that access the online CDS data is 2
-# plan(multisession, workers = 2) ?!
-# But Python from reticulate cannot run in parallel because of external pointers !
-# We can launch manually 2 R sessions, with these two subsets:
-# years_1 <- years[1:(length(years)%/%2)]
-# years_2 <- years[(length(years)%/%2+1):length(years)]
-
-#Run script
-era5_land(years = years, var = var)
-
-
-

### 2023-04-18 climate_compilation/format/formatting_data.R

@@ -1,79 +1,0 @@
-#########################################################################
-# Script to format ERA5-Land climate data as required by Phenofit model #
-#########################################################################
-
-# author : V. Van der Meersch
-# date : 16/02/2022
-
-wd <- "C:/Users/vandermeersch/Documents/CEFE/phd/present_calibration/climate_data/format/phenofit/"
-
-
-## Packages needed
-
-library(ncdf4)
-library(future)
-library(future.apply)
-library(dplyr)
-library(raster)
-library(data.table)
-
-
-## Functions needed
-
-source(paste0(wd, 'functions/phenofit_formatting.R')) # just to format variables for Phenofit, v2 with raster format (faster)
-source(paste0(wd, 'functions/phenofit_processing_and_formatting.R')) # to calculate supplementary variables and format them for Phenofit, v2 with raster format (faster)
-source(paste0(wd, 'functions/phenofit_compute_PET2.R')) # to calculate evapotranspiration, v2 with different parallel architecture
-source(paste0(wd, 'functions/phenofit_get_altitude.R')) # to get altitude
-source(paste0(wd, 'functions/phenofit_get_WHC.R')) # to get WHC
-
-
-## Settings
-
-rawdata_folder <- "D:/climate/ERA5-Land/raw/"
-processeddata_folder <- "D:/climate/ERA5-Land/phenofit_format/"
-
-
-## Compile daily data from hourly data
-phenofit_formatting(1969:2000, "2m_dewpoint_temperature", 'mean', rawdata_folder, processeddata_folder, ncores=5)
-gc()
-
-phenofit_formatting(1969:2000, "2m_temperature", "min", rawdata_folder, processeddata_folder, ncores=5)
-gc()
-
-phenofit_formatting(1969:2000, "2m_temperature", "max", rawdata_folder, processeddata_folder, ncores=5)
-gc()
-
-phenofit_formatting(1969:2000, "total_precipitation", "sum", rawdata_folder, processeddata_folder, ncores=4)
-gc()
-
-phenofit_formatting(1969:2000, "surface_solar_radiation_downwards", "sum", rawdata_folder, processeddata_folder, ncores=5) #used only for ETP calculation
-gc()
-
-phenofit_processing_and_formatting(1969:2000, "wind", "mean", rawdata_folder, processeddata_folder, ncores=5) #used only for ETP calculation
[…]

### 2023-04-18 climate_compilation/format/functions/compute_PET.R

@@ -1,120 +1,0 @@
-
-# Function adapted by V. Van der Meersch from Evapotranspiration package (created by Danlu Guo)
-# Added a modification to correct polar night effect...
-
-
-# method = PenmanMonteith or PriestleyTaylor
-# physically based PenmanMonteith (PM) approach is presently considered as the state-of-the art 
-# PriestleyTaylor (PT) data requirement does not include some hardly available meteorological variables as wind speed or relative humidity
-# PT could fail in mimicking PM values under dry conditions or in zones where wind speed is relatively high
-
-
-# elev : ground elevation above mean sea level in m
-# latitude in radians
-
-
-compute_PET <- function(data, elev, lat_rad, constants, alpha = 0.23, method){
-  
-  if(method == "PriestleyTaylor"){
-    
-    Ta <- (data$Tmax + data$Tmin)/2
-    
-    # Saturated vapour pressure (August–Roche–Magnus approximation)
-    vs_Tmax <- 0.6108 * exp(17.27 * data$Tmax/(data$Tmax + 237.3))
-    vs_Tmin <- 0.6108 * exp(17.27 * data$Tmin/(data$Tmin + 237.3))
-    vas <- (vs_Tmax + vs_Tmin)/2
-    
-    # Vapour pressure
-    vabar <- (vs_Tmin * data$RHmax/100 + vs_Tmax * data$RHmin/100)/2 
-    
-    P <- 101.3 * ((293 - 0.0065 * elev)/293)^5.26 # atmospheric pressure
-    # effect of atmospheric pressure (i.e. Eevaporation at high altitudes promoted due to low atmospheric pressure) is small
-    # thus, the average value of atmospheric pressure for a location is sufficient
-    
-    delta <- 4098 * (0.6108 * exp((17.27 * Ta)/(Ta + 237.3)))/((Ta + 237.3)^2) # slope of vapour pressure curve
-    gamma <- 0.00163 * P/constants$lambda # psychrometric constant
-    d_r2 <- 1 + 0.033 * cos(2 * pi/365 * data$J) # dr is the inverse relative distance Earth-Sun
-    delta2 <- 0.409 * sin(2 * pi/365 * data$J - 1.39) # solar dedication
-    
-    # Added by Victor
-    aux <- -tan(lat_rad) * tan(delta2) 
[…]

### 2023-04-18 climate_compilation/format/functions/convert_units.R

@@ -1,29 +1,0 @@
-# Function to convert units as needed by Phenofit
-
-# author : V. Van der Meersch
-
-convert_units <- function(value, var){
-  if(var == "2m_temperature" | var == "2m_dewpoint_temperature"){
-    # Kelvin to Celsius
-    value_c <- sapply(value, function(i){i-273.15})
-  }
-  if(var == "total_precipitation"){
-    # m to mm
-    value_c <- sapply(value, function(i){i*1000})
-  }
-  if(var == "surface_solar_radiation_downwards"){
-    # J/m2 to MJ/m2
-    value_c <- sapply(value, function(i){i/1000000})
-  }
-  if(var == "surface_pressure"){
-    # Pa to kPa
-    value_c <- sapply(value, function(i){i/1000})
-  }
-  if(var == "potential_evaporation"){
-    # m to mm, and negative to positive
-    value_c <- sapply(value, function(i){- i*1000})
-  }
-  
-  return(value_c)
-}
-

### 2023-04-18 climate_compilation/format/functions/geopotential_to_geometric.R

@@ -1,22 +1,0 @@
-# Function to convert geopotential height to geometric height
-# see https://confluence.ecmwf.int/display/CKB/ERA5%3A+compute+pressure+and+geopotential+on+model+levels%2C+geopotential+height+and+geometric+height
-
-# author : V. Van der Meersch
-
-geopotential_to_geometric <- function(h){
-  
-  # Earth's gravitational acceleration
-  g0 <-  9.80665
-  
-  # Mean radius of the earth
-  re <- 6371000
-  
-  # Get geopotential height in meters
-  h_m <- h/g0
-  
-  # Convert to geometric height 
-  z <- re*h_m/(re-h_m)
-  
-  return(z)
-  
-}

### 2023-04-18 climate_compilation/format/functions/get_comments.R

@@ -1,53 +1,0 @@
-# Function to get Phenofit input variable definitions
-
-# author : V. Van der Meersch
-
-get_comments <- function(var){
-  if(var=='glo'){
-    return("Variable : daily global radiation (MJ/m²)")
-  }
-  else if(var=='pre'){
-    return("Variable : daily precipitation, comprising rain and snow (mm)")
-  }
-  else if(var=='prs'){
-    return("Variable : daily mean surface pressure (kPa)")
-  }
-  else if(var=='RH'){
-    return("Variable : daily mean 2m relative humidity (%) calculated with vapor pressure ratio (Clausius-Clapeyron relation)")
-  }
-  else if(var=='tmn'){
-    return("Variable : daily mimimal 2m temperature (°C)")
-  }
-  else if(var=='tmp'){
-    return("Variable : daily mean 2m temperature (°C)")
-  }
-  else if(var=='tmx'){
-    return("Variable : daily maximal 2m temperature (°C)")
-  }
-  else if(var=='dtm'){
-    return("Variable : daily mean 2m dewpoint temperature (°C)")
-  }
-  else if(var=='wnd'){
-    return("Variable : daily mean 10m wind speed (m/s) - used only to compute evapotranspiration")
-  }
-  else if(var=='RHmin'){
-    return("Variable : daily miminal 2m relative humidity (%) - used only to compute evapotranspiration")
-  }
-  else if(var=='RHmax'){
-    return("Variable : daily maximal 2m relative humidity (%) - used only to compute evapotranspiration")
-  }
-  else if(var=='petPenmanMonteith'){
-    return("Variable : potential evapotranspiration (mm) calculated with Penman-Monteith formulation (FAO-56 hypothetical short grass method) ")
-  }
-  else if(var=='petPriestleyTaylor'){
-    return("Variable : potential evapotranspiration (mm) calculated with Priestley-Taylor formulation")
-  }
-  else if(var=='Altitude'){
-    return("Variable : altitude (m) calculated from geopotential height")
-  }
-  else if(var=='WHC'){
-    return("Variable : water holding capacity (mm)")
-  }
-  
-  
-}

### 2023-04-18 climate_compilation/format/functions/get_var_name.R

@@ -1,33 +1,0 @@
-# Function to get variable name as coded in NC files of ERA5-Land database
-# Or as needed as input variables for Phenofit
-
-# author : V. Van der Meersch
-
-get_var_name <- function(var, stat=NULL, type){
-  long_names <- c("2m_temperature", "total_precipitation", "surface_solar_radiation_downwards", "surface_pressure", 
-                  "wind", "relative_humidity", "2m_dewpoint_temperature", "10m_u_component_of_wind", "10m_v_component_of_wind",
-                  "potential_evaporation")
-  era5_names <- c("t2m", "tp", "ssrd", "ps",  NA, NA, "dpt", "u10", "v10", "pev")
-  
-  if(type == "ERA5"){ return(era5_names[which(long_names == var)]) }
-  
-  if(type == "PHENOFIT" | type=="Phenofit"){
-    if(stat=="mean"){
-      phenofit_names <- c("tmp", NA, NA, "prs", "wnd", "RH", "dtm", NA, NA, NA)
-      return(phenofit_names[which(long_names == var)])
-    }
-    if(stat=="max"){
-      phenofit_names <- c("tmx", NA, NA, NA, NA, "RHmax", NA, NA, NA, NA)
-      return(phenofit_names[which(long_names == var)])
-    }
-    if(stat=="min"){
-      phenofit_names <- c("tmn", NA, NA, NA, NA, "RHmin", NA, NA, NA, NA)
-      return(phenofit_names[which(long_names == var)])
-    }
-    if(stat=="sum"){
-      phenofit_names <- c(NA, "pre", "glo", NA, NA, NA, NA, NA, NA, "pet")
-      return(phenofit_names[which(long_names == var)])
-    }
-  }
-  
-}

### 2023-04-18 climate_compilation/format/functions/phenofit_compute_PET.R

@@ -1,155 +1,0 @@
-# author : V. Van der Meersch
-
-# Loading custom functions
-source(paste0(wd, 'functions/compute_PET.R')) 
-source(paste0(wd, 'functions/get_comments.R'))
-
-# Package for reporting progress updates
-library(progressr)
-handlers(global = TRUE)
-handlers("txtprogressbar")
-
-
-# Needed constants
-PET_constants <- data.frame(lambda = 2.45, Gsc = 0.082, sigma = 4.903e-09, alphaPT = 1.26, G = 0, z = 10)
-  # lambda : latent heat of vaporisation - Gsc : solar constant - sigma : Stefan-Boltzmann constan
-  # alphaPT : Priestley-Taylor coefficient - G : soil heat flux (= 0 with daily time step)
-  # z : height of wind instrument in m (= 10 in ERA5-Land data)
-
-
-# Function parameters :
-  # - pd_folder : folder where climate data (Phenofit format) are located
-  # this pd_folder needs two subfolders : "/pet_PenmanMonteith" and "/pet_PriestleyTaylor"
-  #
-  # - method = PenmanMonteith or PriestleyTaylor (see compute_PET function)
-  #
-  # - years : e.g. 1990:2000
-  #
-  # - ncores : for parallelisation over years
- 
-
-phenofit_compute_PET <- function(years, pd_folder, method, ncores, transform_NA_and_zero=FALSE){
-  
-  pd_folder_method <- paste0(pd_folder, "pet_", method, "/")
-  if(transform_NA_and_zero){
-    pd_folder_method <- paste0(pd_folder, "pet_", method, "_trans/")
-  }
-  
-  plan(multisession, workers = ncores)
-  
-  # Get altitude
-  alt_file <- paste0(pd_folder, "ERA5LAND_Altitude.fit")
-  alt <- fread(alt_file, showProgress=F)
-  nrows <- nrow(alt)
-  
-  p <- progressor(length(years))
-  
-  # Parallel loop in case of parallel computing
-  out <- future_apply(array(1:ncores), 1, function(i){
-    
-    # To determine number of years processed by core in case of parallel computing
-    quot <- length(years)%/%ncores
-    rem <- length(years)%%ncores
-    if(i<ncores){ years_i <- years[(1+(i-1)*quot):(quot*i)] 
[…]

### 2023-04-18 climate_compilation/format/functions/phenofit_formatting.R

@@ -1,202 +1,0 @@
-source(paste0(wd, 'functions/get_var_name.R'))
-source(paste0(wd, 'functions/get_comments.R'))
-source(paste0(wd, 'functions/convert_units.R'))
-
-# author : V. Van der Meersch
-
-# Package for reporting progress updates
-library(progressr)
-handlers(global = TRUE)
-handlers("txtprogressbar")
-
-phenofit_formatting2 <- function(years, var, stat, rd_folder, pd_folder, ncores=1, landsea_mask=FALSE){
-  
-  if(landsea_mask){
-    landsea_file <- paste0(rd_folder, 'lsm_1279l4_0.1x0.1.grb_v4_unpack.nc')
-    landsea_mask <- brick(landsea_file, varname='lsm')
-    landsea_mask <- rotate(landsea_mask) # longitude : go from 0 - 360 to -180 - 180
-    e <- extent(-14, 40, 34,72)
-    landsea_mask <- crop(landsea_mask, e)
-    landsea_mask[landsea_mask<0.1] <- NA
-    landsea_mask_df <- na.omit(as.data.frame(landsea_mask, xy=TRUE))
-    names(landsea_mask_df) <- c("lon", "lat", "mask")
-    landsea_mask_df <- round(landsea_mask_df, 2)
-  }
-  
-  if(var == "wind" | var == "relative_humidity"){
-    stop("For wind or relative humidity, you need to use phenofit_processing_and_formatting function !")
-  }
-  else if( !(var %in% c("2m_temperature", "total_precipitation", "surface_solar_radiation_downwards", 
-                        "potential_evaporation", "2m_dewpoint_temperature")) ){
-    stop("Cannot recognize the variable you asked for !")
-  }
-  
-  # To follow the progress
-  p <- progressor(length(years)*12)
-  
-  # Get variable name as coded in NC file
-  var_name <- get_var_name(var, type = "ERA5")
-  
-  months <- c("01", "02", "03", "04", "05", "06",
-               "07", "08", "09", "10", "11", "12")
-  
-  plan(multisession, workers = ncores)
-  
-  # Parallel loop (in case of parallel computing)
-  out <- future_apply(array(1:ncores), 1, function(i){
-    
-    # To determine number of years processed by core (in case of parallel computing)
-    quot <- length(years)%/%ncores
-    rem <- length(years)%%ncores
[…]

### 2023-04-18 climate_compilation/format/functions/phenofit_get_altitude.R

@@ -1,51 +1,0 @@
-# author : V. Van der Meersch
-
-source(paste0(wd, 'functions/get_comments.R'))
-source(paste0(wd, 'functions/geopotential_to_geometric.R'))
-
-# loc : data.frame containing longitude and latitude for which altitude is needed
-# if NULL, compute for every point !
-
-phenofit_get_altitude <- function(loc = NULL, rd_folder, pd_folder){
-
-  # Open geopotential file
-  geopotential_file <- paste0(rd_folder, "geo_1279l4_0.1x0.1.grib2_v4_unpack.nc")
-  
-  geopotential_raster <- brick(geopotential_file, varname="z")
-  
-  geopotential_df <- as.data.frame(geopotential_raster, xy=TRUE)
-  names(geopotential_df) <- c("lon", "lat", "z")
-  
-  # Need to convert longitude from (0-360) to (-180, 180)
-  geopotential_df$lon <- ((geopotential_df$lon+180)%%360)-180.0
-  
-  # Convert to geometric height
-  geopotential_df$z <- geopotential_to_geometric(geopotential_df$z)
-  
-  if(is.null(loc)){
-    alt_needed <- geopotential_df
-  }else{
-    alt_needed <- left_join(round(loc,2),round(geopotential_df,2),by=c("lat","lon"))  # round is important, of left_join won't work
-  }
-  
-  alt_needed$lon <- loc$lon
-  alt_needed$lat <- loc$lat
-  
-  # Fill name
-  processed_file <- paste0(pd_folder, "ERA5LAND_Altitude.fit")
-  
-  # Fill file
-  con <- file(processed_file, open="wt")
-  writeLines("Elevation datafile for Phenofit model", con)
-  writeLines(paste("Created on RStudio, by user", Sys.getenv("USERNAME") ,", on the", Sys.Date()), con)
-  comments <- get_comments(var = "Altitude")
-  writeLines(comments, con)
-  writeLines(" ", con)
-  write.table(alt_needed, file = con, sep = "\t",
-              row.names = FALSE, col.names= FALSE)
-  close(con)
-  
-  # Free memory
-  gc(verbose = FALSE)
-
-}

### 2023-04-18 climate_compilation/format/functions/phenofit_processing_and_formatting.R

@@ -1,184 +1,0 @@
-source(paste0(wd, 'functions/get_var_name.R'))
-source(paste0(wd, 'functions/get_comments.R'))
-source(paste0(wd, 'functions/convert_units.R'))
-
-# author : V. Van der Meersch
-
-# Package to calculate relative humidity with the ratio of saturation vapor pressure
-library(humidity)
-
-# Package for reporting progress updates
-library(progressr)
-handlers(global = TRUE)
-handlers("txtprogressbar")
-
-phenofit_processing_and_formatting <- function(years, var, stat, rd_folder, pd_folder, ncores=1){
-  
-  if(var %in% c("2m_temperature", "total_precipitation", "surface_solar_radiation_downwards")){
-    stop(paste0("For ", var, ", you need to use phenofit_formatting function !"))
-  }
-  else if( var != "wind" & var != "relative_humidity" ){
-    stop("Cannot recognize the variable you asked for !")
-  }
-  
-  # To follow the progress
-  p <- progressor(length(years)*12)
-  
-  months <- c("01", "02", "03", "04", "05", "06",
-              "07", "08", "09", "10", "11", "12")
-  
-  # Get variable names as coded in NC file
-  if(var == "wind"){
-    var_1 <- "10m_u_component_of_wind"
-    var_name_1 <- get_var_name(var_1, type = "ERA5")
-    var_2 <- "10m_v_component_of_wind"
-    var_name_2 <- get_var_name(var_2, type = "ERA5")
-  }
-  if(var == "relative_humidity"){
-    var_1 <- "2m_temperature"
-    var_name_1 <- get_var_name(var_1, type = "ERA5")
-    var_2 <- "2m_dewpoint_temperature"
-    var_name_2 <- get_var_name(var_2, type = "ERA5")
-  }
-  
-  
-  plan(multisession, workers = ncores)
-  
-  # Parallel loop in case of parallel computing
-  out <- future_apply(array(1:ncores), 1, function(i){
-    
-    # To determine number of years processed by core in case of parallel computing
-    quot <- length(years)%/%ncores
-    rem <- length(years)%%ncores
-    if(i<ncores){ years_i <- years[(1+(i-1)*quot):(quot*i)] }
-    else{ years_i <- years[(1+(i-1)*quot):((quot*i)+rem)] }
-    
-    
-    # Loop on years
-    for(yr in years_i){
-      
-      p(message = paste("Processing year", yr, "..."), class = "sticky", amount = 0)
-      
[…]

### 2023-04-18 manuscript/paper.Rmd
removed green annotations for reviewers

@@
 Two main types of species distribution models are used to project species range shifts in future climatic conditions: correlative and process-based models.
-Although there is some continuity between these two types of models, they are fundamentally different in their hypotheses (statistical relationships *vs* mechanistic relationships) and their calibration methods \\textcolor{darkgreen}{(SDMs tend to be occurrence data-driven while PBMs tend to be prior-driven)}.
+Although there is some continuity between these two types of models, they are fundamentally different in their hypotheses (statistical relationships *vs* mechanistic relationships) and their calibration methods (SDMs tend to be occurrence data-driven while PBMs tend to be prior-driven).
 \n\n 2.
-One of the limitations to the use of process-based models is the difficulty to parameterize them for a large number of species \\textcolor{darkgreen}{compared to correlative SDMs}.
+One of the limitations to the use of process-based models is the difficulty to parameterize them for a large number of species compared to correlative SDMs.
 We investigated the feasibility of using an evolutionary algorithm (called covariance matrix adaptation evolution strategy, CMA-ES) to calibrate process-based models using species distribution data.
@@
 \n\n 4.
-\\textcolor{darkgreen}{We conclude that} CMA-ES is an efficient state-of-the-art method to calibrate process-based models with a large number of parameters using species occurrence data.
+We conclude that CMA-ES is an efficient state-of-the-art method to calibrate process-based models with a large number of parameters using species occurrence data.
 Inverse modelling using CMA-ES is a powerful method to calibrate process-based parameters which can hardly be measured.
@@
 Parameter inference can be achieved through several methods which have been developed in the last decades.
-\textcolor{darkgreen}{Most of them fall into two categories: informal and statistical calibration.
+Most of them fall into two categories: informal and statistical calibration.
 On the one hand, statistical calibration assumes a data-generating model, and a likelihood function, which quantifies the probability of the observed data given the model parameters, is expressed mathematically.
@@
 In practice, deriving the likelihood function can be quite challenging for complex models with many parameters and thresholding.
-One approach is to use numerical methods such as \emph{simulation-based inference} or \emph{approximate Bayesian computation} [e.g.
-@Hartig2014]\textcolor{darkgreen}{.
+One approach is to use numerical methods such as \emph{simulation-based inference} or \emph{approximate Bayesian computation [e.g.
+@Hartig2014].
 However, obtaining a reliable estimate of the likelihood of a complex process-based model can require a large number of simulated samples, which can be computationally expensive.
@@
 Nelder-Mead method) or stochastic (e.g.
-simulated annealing, evolutionary algorithms).}
+simulated annealing, evolutionary algorithms).
 Recently, an approach belonging to the evolutionary algorithm family, called Covariance Matrix Adaptation Evolution Strategy (CMA-ES), has been proposed [@Hansen2001].
@@
 
-\textcolor{darkgreen}{Our results showed that CMA-ES is an efficient optimizer for inverse calibration of complex ecological models.
+Our results showed that CMA-ES is an efficient optimizer for inverse calibration of complex ecological models.
 The algorithm was able to find parameter sets that significantly improved the predictions of the calibrated models compared to the expert parametrization.
-However, our study also highlighted the issue of non-identifiability of parameter values due to the data limitation and the dependence between process-based model parameters, which may result in diverging parameter values even though the calibrated models describe the observed species distribution well.}
+However, our study also highlighted the issue of non-identifiability of parameter values due to the data limitation and the dependence between process-based model parameters, which may result in diverging parameter values even though the calibrated models describe the observed species distribution well.
 
@@
 Here we demonstrated that inverse calibration with CMA-ES is feasible and provide good results for complex and runtime-expensive ecological models.  
-With a subsampling of species occurrence data, the algorithm succeeds in finding parameter sets which provide \textcolor{darkgreen}{higher} AUC values.
+With a subsampling of species occurrence data, the algorithm succeeds in finding parameter sets which provide higher AUC values.
 The predictions of the calibrated models are sharply improved compared to expert parametrization (\hyperref[fig:phenofitmaps]{Figure 1} and \hyperref[fig:castaneamaps]{Figure 2}).
@@
 As well as being quasi parameter-free, CMA-ES has several structural advantages particularly useful for complex optimization problems.
-\textcolor{darkgreen}{First, the algorithm's covariance matrix enables the learning of second-order information, which provides insights into pairwise dependencies between parameters.
-Second, the covariance matrix adaptation of CMA-ES is highly efficient in handling ill-conditioned and non-separable problems }[@Hansen2011]\textcolor{darkgreen}{.
+First, the algorithm's covariance matrix enables the learning of second-order information, which provides insights into pairwise dependencies between parameters.
+Second, the covariance matrix adaptation of CMA-ES is highly efficient in handling ill-conditioned and non-separable problems [@Hansen2011].
 Last, CMA-ES's update mechanism of step size $\sigma$ (i.e.
-the mutation force) helps prevent premature convergence }[@Hansen2001]\textcolor{darkgreen}{, allowing the algorithm to explore more of the search space.}
+the mutation force) helps prevent premature convergence [@Hansen2001], allowing the algorithm to explore more of the search space.
 CMA-ES has been shown to outperform several other optimization algorithms [@Hansen2010], and is usually the most efficient method when the target cost (i.e.
@@
 
-\textcolor{darkgreen}{Equifinality is a common problem encountered during model calibration, where multiple parameter sets can produce equally good fits to the observed data.
-This issue can arise due to several factors, including limitations in the quantity or quality of available data, competing processes within the model, and parameter interactions that affect the model output in complex ways.}
+Equifinality is a common problem encountered during model calibration, where multiple parameter sets can produce equally good fits to the observed data.
+This issue can arise due to several factors, including limitations in the quantity or quality of available data, competing processes within the model, and parameter interactions that affect the model output in complex ways.
 
@@
 To deal with equifinality issues, an avenue to explore could be the use of multiple objective functions during model calibration to assess different aspects of the model performance.
-\textcolor{darkgreen}{Additionally, if a closed-form likelihood can be derived, one could use a Bayesian framework to combine prior knowledge and inverse estimation of parameters to constrain the parameter space and study the nature of trade-offs between parameters} [@Hartig2012; @Cailleret2020].
+Additionally, if a closed-form likelihood can be derived, one could use a Bayesian framework to combine prior knowledge and inverse estimation of parameters to constrain the parameter space and study the nature of trade-offs between parameters [@Hartig2012; @Cailleret2020].
 
@@
 We did not attempt to validate our parametrizations using temporally or spatially independent data, and AUC was only used to determine if model outputs were consistent with species distributions.
-\textcolor{darkgreen}{However, AUC is scale-invariant: it measures how well predictions are ranked rather than their absolute values.
+However, AUC is scale-invariant: it measures how well predictions are ranked rather than their absolute values.
 For example, with PHENOFIT, a species with a fitness of $0.8$ could be considered as absent while another one with the same fitness could be considered as present.
-Therefore, when it is used as an objective function for model calibration, we probably lack some precision and consistency among species’ parameters estimation.} Further work could thus be conducted to examine the effects of choosing a different objective function.
+Therefore, when it is used as an objective function for model calibration, we probably lack some precision and consistency among species’ parameters estimation.
+Further work could thus be conducted to examine the effects of choosing a different objective function.
 

### 2023-04-18 manuscript/paper.tex
removed green annotations for reviewers

@@
   hypotheses (statistical relationships \emph{vs} mechanistic
-  relationships) and their calibration methods
-  \textcolor{darkgreen}{(SDMs tend to be occurrence data-driven while PBMs tend to be prior-driven)}.
+  relationships) and their calibration methods (SDMs tend to be
+  occurrence data-driven while PBMs tend to be prior-driven).
 \item
   One of the limitations to the use of process-based models is the
-  difficulty to parameterize them for a large number of species
-  \textcolor{darkgreen}{compared to correlative SDMs}. We investigated
-  the feasibility of using an evolutionary algorithm (called covariance
-  matrix adaptation evolution strategy, CMA-ES) to calibrate
-  process-based models using species distribution data. This method is
-  well established in some fields (robotics, aerospace research,
-  \ldots), but has never been used, to our knowledge, in ecology,
-  despite its ability to deal with very large space dimensions. Using
-  tree species occurrence data across Europe, we adapted the CMA-ES
-  algorithm to find appropriate values of model parameters. We estimated
-  simultaneously 27 to 77 parameters of two process-based models
-  simulating forest tree's ecophysiology for three species with varying
-  range sizes and geographical distributions.
+  difficulty to parameterize them for a large number of species compared
+  to correlative SDMs. We investigated the feasibility of using an
+  evolutionary algorithm (called covariance matrix adaptation evolution
+  strategy, CMA-ES) to calibrate process-based models using species
+  distribution data. This method is well established in some fields
+  (robotics, aerospace research, \ldots), but has never been used, to
+  our knowledge, in ecology, despite its ability to deal with very large
+  space dimensions. Using tree species occurrence data across Europe, we
+  adapted the CMA-ES algorithm to find appropriate values of model
+  parameters. We estimated simultaneously 27 to 77 parameters of two
+  process-based models simulating forest tree's ecophysiology for three
+  species with varying range sizes and geographical distributions.
 \item
@@
 \item
-  \textcolor{darkgreen}{We conclude that} CMA-ES is an efficient
-  state-of-the-art method to calibrate process-based models with a large
-  number of parameters using species occurrence data. Inverse modelling
-  using CMA-ES is a powerful method to calibrate process-based
-  parameters which can hardly be measured. However, the method does not
-  warranty that parameter estimates are correct because of several
-  sources of bias, similarly to correlative models, and expert knowledge
-  is required to validate results.
+  We conclude that CMA-ES is an efficient state-of-the-art method to
+  calibrate process-based models with a large number of parameters using
+  species occurrence data. Inverse modelling using CMA-ES is a powerful
+  method to calibrate process-based parameters which can hardly be
+  measured. However, the method does not warranty that parameter
+  estimates are correct because of several sources of bias, similarly to
+  correlative models, and expert knowledge is required to validate
+  results.
 \end{enumerate}
@@
 inference can be achieved through several methods which have been
-developed in the last decades.
-\textbackslash textcolor\{darkgreen\}\{Most of them fall into two
-categories: informal and statistical calibration. On the one hand,
-statistical calibration assumes a data-generating model, and a
-likelihood function, which quantifies the probability of the observed
-data given the model parameters, is expressed mathematically. This
-likelihood is the foundation of \emph{maximum likelihood estimation} and
+developed in the last decades. Most of them fall into two categories:
+informal and statistical calibration. On the one hand, statistical
+calibration assumes a data-generating model, and a likelihood function,
+which quantifies the probability of the observed data given the model
+parameters, is expressed mathematically. This likelihood is the
+foundation of \emph{maximum likelihood estimation} and
 \emph{Bayesian inference}, two commonly used methods for parameter
@@
 One approach is to use numerical methods such as
-\emph{simulation-based inference} or
-\emph{approximate Bayesian computation} (e.g.
-\protect\hyperlink{ref-Hartig2014}{Hartig \emph{et al.}
-2014})\textcolor{darkgreen}{.
-However, obtaining a reliable estimate of the likelihood of a complex process-based model can require a large number of simulated samples, which can be computationally expensive.
-On the other hand, informal calibration uses an informal objective function to measure the discrepancy between the model predictions and the observed data.
-By minimizing the objective function, one is able to identify the parameter values that best fit the observed data, even though it may not have the same statistical rigor as statistical calibration.
-This optimization step can be achieved with several methods, either deterministic (e.g.
-Nelder-Mead method) or stochastic (e.g.
-simulated annealing, evolutionary algorithms).}\\
+\emph{simulation-based inference} or \textbackslash emph\{approximate
+Bayesian computation (e.g. \protect\hyperlink{ref-Hartig2014}{Hartig
+\emph{et al.} 2014}). However, obtaining a reliable estimate of the
+likelihood of a complex process-based model can require a large number
+of simulated samples, which can be computationally expensive. On the
+other hand, informal calibration uses an informal objective function to
+measure the discrepancy between the model predictions and the observed
+data. By minimizing the objective function, one is able to identify the
+parameter values that best fit the observed data, even though it may not
+have the same statistical rigor as statistical calibration. This
+optimization step can be achieved with several methods, either
+deterministic (e.g.~Nelder-Mead method) or stochastic (e.g.~simulated
+annealing, evolutionary algorithms).\\
 Recently, an approach belonging to the evolutionary algorithm family,
@@
 
-\textcolor{darkgreen}{Our results showed that CMA-ES is an efficient optimizer for inverse calibration of complex ecological models.
-The algorithm was able to find parameter sets that significantly improved the predictions of the calibrated models compared to the expert parametrization.
-However, our study also highlighted the issue of non-identifiability of parameter values due to the data limitation and the dependence between process-based model parameters, which may result in diverging parameter values even though the calibrated models describe the observed species distribution well.}
+Our results showed that CMA-ES is an efficient optimizer for inverse
+calibration of complex ecological models. The algorithm was able to find
+parameter sets that significantly improved the predictions of the
+calibrated models compared to the expert parametrization. However, our
+study also highlighted the issue of non-identifiability of parameter
+values due to the data limitation and the dependence between
+process-based model parameters, which may result in diverging parameter
+values even though the calibrated models describe the observed species
+distribution well.
 
@@
 With a subsampling of species occurrence data, the algorithm succeeds in
-finding parameter sets which provide \textcolor{darkgreen}{higher} AUC
-values. The predictions of the calibrated models are sharply improved
-compared to expert parametrization
-(\hyperref[fig:phenofitmaps]{Figure 1} and
+finding parameter sets which provide higher AUC values. The predictions
+of the calibrated models are sharply improved compared to expert
+parametrization (\hyperref[fig:phenofitmaps]{Figure 1} and
 \hyperref[fig:castaneamaps]{Figure 2}). Two striking examples are the
@@
 quasi parameter-free, CMA-ES has several structural advantages
-particularly useful for complex optimization problems.
-\textcolor{darkgreen}{First, the algorithm's covariance matrix enables the learning of second-order information, which provides insights into pairwise dependencies between parameters.
-Second, the covariance matrix adaptation of CMA-ES is highly efficient in handling ill-conditioned and non-separable problems }(\protect\hyperlink{ref-Hansen2011}{Hansen
-\emph{et al.}
-2011})\textcolor{darkgreen}{. Last, CMA-ES's update mechanism of step size $\sigma$ (i.e. the mutation force) helps prevent premature convergence }(\protect\hyperlink{ref-Hansen2001}{Hansen
-\& Ostermeier
-2001})\textcolor{darkgreen}{, allowing the algorithm to explore more of the search space.}
-CMA-ES has been shown to outperform several other optimization
-algorithms (\protect\hyperlink{ref-Hansen2010}{Hansen \emph{et al.}
-2010}), and is usually the most efficient method when the target cost
-(i.e.~the number of objective function evaluations) is about \(100*N\)
-(\(N\) being the dimension of the parameter search space,
+particularly useful for complex optimization problems. First, the
+algorithm's covariance matrix enables the learning of second-order
+information, which provides insights into pairwise dependencies between
+parameters. Second, the covariance matrix adaptation of CMA-ES is highly
+efficient in handling ill-conditioned and non-separable problems
+(\protect\hyperlink{ref-Hansen2011}{Hansen \emph{et al.} 2011}). Last,
+CMA-ES's update mechanism of step size \(\sigma\) (i.e.~the mutation
+force) helps prevent premature convergence
+(\protect\hyperlink{ref-Hansen2001}{Hansen \& Ostermeier 2001}),
+allowing the algorithm to explore more of the search space. CMA-ES has
+been shown to outperform several other optimization algorithms
+(\protect\hyperlink{ref-Hansen2010}{Hansen \emph{et al.} 2010}), and is
+usually the most efficient method when the target cost (i.e.~the number
+of objective function evaluations) is about \(100*N\) (\(N\) being the
+dimension of the parameter search space,
 \protect\hyperlink{ref-Baeck2013}{Bäck \emph{et al.} 2013}).
@@
 
-\textcolor{darkgreen}{Equifinality is a common problem encountered during model calibration, where multiple parameter sets can produce equally good fits to the observed data.
-This issue can arise due to several factors, including limitations in the quantity or quality of available data, competing processes within the model, and parameter interactions that affect the model output in complex ways.}
+Equifinality is a common problem encountered during model calibration,
+where multiple parameter sets can produce equally good fits to the
+observed data. This issue can arise due to several factors, including
+limitations in the quantity or quality of available data, competing
+processes within the model, and parameter interactions that affect the
+model output in complex ways.
 
@@
 functions during model calibration to assess different aspects of the
-model performance.
-\textcolor{darkgreen}{Additionally, if a closed-form likelihood can be derived, one could use a Bayesian framework to combine prior knowledge and inverse estimation of parameters to constrain the parameter space and study the nature of trade-offs between parameters}
+model performance. Additionally, if a closed-form likelihood can be
+derived, one could use a Bayesian framework to combine prior knowledge
+and inverse estimation of parameters to constrain the parameter space
+and study the nature of trade-offs between parameters
 (\protect\hyperlink{ref-Hartig2012}{Hartig \emph{et al.} 2012};
@@
 determine if model outputs were consistent with species distributions.
-\textcolor{darkgreen}{However, AUC is scale-invariant: it measures how well predictions are ranked rather than their absolute values.
-For example, with PHENOFIT, a species with a fitness of $0.8$ could be considered as absent while another one with the same fitness could be considered as present.
-Therefore, when it is used as an objective function for model calibration, we probably lack some precision and consistency among species’ parameters estimation.}
-Further work could thus be conducted to examine the effects of choosing
-a different objective function.
+However, AUC is scale-invariant: it measures how well predictions are
+ranked rather than their absolute values. For example, with PHENOFIT, a
+species with a fitness of \(0.8\) could be considered as absent while
+another one with the same fitness could be considered as present.
+Therefore, when it is used as an objective function for model
+calibration, we probably lack some precision and consistency among
+species' parameters estimation. Further work could thus be conducted to
+examine the effects of choosing a different objective function.
 

### 2023-04-19 data_compilation/format/formatting_climate_data.R
Update formatting_climate_data.R

@@ -1,9 +1,7 @@
+
 #-----------------------------------------------------------------------#
 # Script to format ERA5-Land climate data as required by Phenofit model #
 #-----------------------------------------------------------------------#
-
-# author : V. Van der Meersch
-# date : 16/02/2022
 
 wd <- "~/projects/inverse_calibration/data_compilation/format"
 

### 2023-04-19 data_compilation/sampling/functions/compute_biovars.R
add script to sample presence/absence data with environ. clustering

@@ -1,0 +1,94 @@
+
+# Function to compute bioclim variables from PHENOFIT daily variables
+
+compute_biovars2 <- function(years, pd_folder, bio_folder, ncores){
+  
+  # initialise progressor
+  p <- progressor(length(years)*5)
+  
+  plan(multisession, workers = ncores)
+  
+  # Parallel loop (in case of parallel computing)
+  out <- future_apply(array(1:ncores), 1, function(i){
+    
+    # To determine number of years processed by core (in case of parallel computing)
+    quot <- length(years)%/%ncores
+    rem <- length(years)%%ncores
+    if(i<ncores){ years_i <- years[(1+(i-1)*quot):(quot*i)] }
+    else{ years_i <- years[(1+(i-1)*quot):((quot*i)+rem)] }
+    
+    # Loop on years
+    for(yr in years_i){
+    
+      p(message = paste("Processing year", yr, "..."), class = "sticky", amount = 0)
+      
+      # Load data 
+      tmin_file <- paste0(pd_folder, "ERA5LAND_", "tmn", "_", yr, "_dly.fit")
+      tmin <- fread(tmin_file, showProgress=F)
+      nbdays <- ncol(tmin) - 2
+      
+      tmax_file <- paste0(pd_folder, "ERA5LAND_", "tmx", "_", yr, "_dly.fit")
+      tmax <- fread(tmax_file, showProgress=F)
+      
+      pre_file <- paste0(pd_folder, "ERA5LAND_", "pre", "_", yr, "_dly.fit")
+      pre <- fread(pre_file, showProgress=F)
+      
+      p(message = paste("Climate files loaded !"), class = "sticky")
+      
+      # Monthly means
+      p(message = paste("Computing monthly values for year", yr, "..."), class = "sticky", amount = 0)
+      tmin_mn <- monthly_calc(tmin, nbdays, stat = "mean")
+      p()
+      tmax_mn <- monthly_calc(tmax, nbdays, stat = "mean")
+      p()
+      pre_mn <- monthly_calc(pre, nbdays, stat = "sum")
+      p()
+      
+      biovars <- matrix(nrow = nrow(tmin_mn), ncol = 21) # lat + lon + 19 biovars
+      biovars[,1:2] <- as.matrix(tmin[,1:2])
[…]

### 2023-04-19 data_compilation/sampling/functions/generate_subsets.R

@@ -1,0 +1,143 @@
+
+# Function to generate random pres/(pseudo)abs combination
+
+generate_subsets <- function(N, npres, nabs, nclusters, occ_data, output_dir){
+  
+  sub_dir <- paste0(npres, "pres_", nabs, "abs")
+  work_dir <- file.path(output_dir, sub_dir)
+  dir.create(work_dir, showWarnings = FALSE)
+  
+  for(i in 1:N){
+    
+    fagussylvatica_pres <- .sample_presence_by_env(occ_data, biovars_30y, k = nclusters, nb_samples = npres)
+    fagussylvatica_abs <- sample_absence(occ_data, EUForest, ERA5land, nb_samples = nabs, env_data = biovars_30y)
+    
+    species_occurrence <- .create_occurrence(fagussylvatica_pres, fagussylvatica_abs)
+    
+    save(species_occurrence, file=paste0(work_dir, "/occurrence_subset_", i, ".Rdata"))
+    gc()
+    
+  }
+  
+}
+
+.create_occurrence <- function(presence, absence){
+  
+  presence <- presence %>%
+    dplyr::select(-c(nb_src, fold))
+  presence$pres <- 1
+  absence$pres <- 0
+  
+  return(rbind(presence,absence))
+  
+}
+
+
+
+
+# Inspired by blockCV package (Valavi et al.)
+
+# species_data : sf object containing species presence
+# env_data : data.frame object of covariates to identify environmental groups
+# k : number of the folds (e.g. k=10)
+# nb_samples : total number of samples (e.g. nb_samples=1000)
+.sample_presence_by_env <- function(species_data, env_data, k, nb_samples, debug=FALSE){
+  
+  # transform species data
+  species_data_df <- as.data.frame(species_data, xy=T) 
+  species_data_df <- na.omit(species_data_df) %>%
+    dplyr::mutate(lon = round(x,1),
+                  lat = round(y,1)) %>% 
+    dplyr::select(-c(x,y))
+  
+  # extract environmental values
+  env_data$lat <- round(env_data$lat,1)
+  env_data$lon <- round(env_data$lon,1)
+  
+  ext <- left_join(species_data_df, env_data, by = c("lon", "lat"))
+  if(anyNA(ext)){
+    message("The input environmental dataframe does not cover all the study area !")
+    if(debug){
+      message("Debug = TRUE, returning missing points...")
+      
[…]

### 2023-04-19 data_compilation/sampling/functions/process_occurrence.R

@@ -1,0 +1,79 @@
+
+# Function to extract and assemble occurrence data from multiple databases
+
+process_occurrence <- function(species, EUForest, GBIF, EVM, AFE, ERA5land, WOODIV=NULL){
+  
+  # GBIF data
+  ## set-up filters
+  EVM_filter <- EVM %>% filter(!is.na(!!as.symbol(substr(paste0(word(species, 1), '_', word(species, 2)), 1, 10)))) %>% 
+    dplyr::select(geometry) %>% 
+    st_transform(crs=st_crs(4326))
+  
+  AFE$taxon_name <- word(AFE$taxon_name, 1,2)
+  AFE_filter <- AFE %>% filter(taxon_name == species) %>% 
+    dplyr::select(geometry = geom) %>% 
+    st_transform(crs=st_crs(4326))
+  
+  ## filtering
+  GBIF_filtered <- .read_process_GBIF(GBIF, filter = rbind(AFE_filter, EVM_filter))
+  rm(GBIF)
+  
+  # EU-Forest data
+  EUForest_sf <- st_as_sf(EUForest, coords = c("X","Y"), crs=st_crs(3035))
+  colnames(EUForest_sf)[2] <- "species"
+  EUForest_sf <- EUForest_sf[EUForest_sf$species==species,] %>% 
+    dplyr::select(species, geometry)
+  EUForest_sf <- st_transform(EUForest_sf, crs = st_crs(4326))
+ 
+  # WOODIV
+  if(!is.null(WOODIV)){
+    WOODIV <- WOODIV %>% dplyr::select(species, geometry) %>% st_transform(crs = st_crs(4326)) %>% st_centroid() %>% unique()
+  }
+  
+  # Transform to grid
+  GBIF_gridded <-  rasterize(GBIF_filtered, ERA5land, field=1, fun='count')
+  GBIF_gridded[GBIF_gridded > 0] <- 1
+  EUForest_gridded <- rasterize(EUForest_sf, ERA5land, field=1, fun='count')
+  EUForest_gridded[EUForest_gridded > 0] <- 1
+  
+  # Find consensus
+  if(is.null(WOODIV)){
+    rs <- stack(GBIF_gridded, EUForest_gridded)
+    cons <- calc(rs, sum, na.rm=T)
+  }else{
+    if(species== "Quercus ilex"){
+      WOODIV <- WOODIV[WOODIV$species=='QILE',]
+    }else if(species=="Quercus pubescens"){
[…]

### 2023-04-19 data_compilation/sampling/sampling_occurrence_data.R

@@ -1,0 +1,85 @@
+
+#----------------------------------------#
+# Script to sample presence/absence data #
+#----------------------------------------#
+
+# Example with F. sylvatica
+
+wd <- "~/projects/inverse_calibration/data_compilation/sampling"
+
+
+# Packages needed
+library(future)
+library(future.apply)
+options(future.globals.maxSize= 1000*1024^2)
+library(data.table)
+library(dplyr)
+library(stringr)
+library(raster)
+library(abind)
+library(dismo, include.only = 'biovars')
+# setDTthreads(4)
+
+# Package for reporting progress updates
+library(progressr)
+handlers(global = TRUE)
+handlers("txtprogressbar")
+
+# Functions needed
+source(paste0(wd, 'functions/compute_biovars.R')) # to compute bioclim variables
+source(paste0(wd, 'functions/process_occurrence.R')) # extract and assemble occurrence data from multiple databases
+source(paste0(wd, 'functions/generate_subsets.R')) # generate pres/abs combination
+
+# Settings 
+processeddata_folder <- "D:/climate/ERA5-Land/phenofit_format/transformed/"
+rawdata_folder <- "D:/climate/ERA5-Land/raw/"
+bioclimdata_folder <- "D:/climate/ERA5-Land/bioclim_format/"
+speciesdata_folder <- "D:/species/"
+
+
+
+# 1. Compute bioclim variables... save in bioclimdata_folder
+compute_biovars(1970:2000, processeddata_folder, bioclimdata_folder, ncores=3)
+gc()
+
+# Mean of 31 years of bioclim data
+biovars_all <- list()
+for(i in as.character(1970:2000)){
+  load(paste0(bioclimdata_folder, "biovars_", i, ".Rdata"))
+  biovars_all <- append(biovars_all, list(biovars))
+}
+rm(biovars)
+biovars_30y <- abind(biovars_all, along=3)
+biovars_30y <- apply(biovars_30y, c(1,2), mean)
+biovars_30y <- as.data.frame(biovars_30y)
+
+
+
+# 2. Process occurrence data
+
+# load data
+EVM <- read_sf(paste0(speciesdata_folder,"EuroVegMap/evm_bonusspecies.shp"))
+AFE <- read_sf(paste0(speciesdata_folder,"AFE/species/all_bonusspecies.gpkg")) %>% dplyr::filter(status == 6)
[…]

### 2023-04-19 data_compilation/sampling/functions/generate_subsets.R
small adjustments

@@ -1,7 +1,7 @@
 
 # Function to generate random pres/(pseudo)abs combination
 
-generate_subsets <- function(N, npres, nabs, nclusters, occ_data, output_dir){
+generate_subsets <- function(N, npres, nabs, nclusters, occ_data, env_data, output_dir){
   
   sub_dir <- paste0(npres, "pres_", nabs, "abs")
   work_dir <- file.path(output_dir, sub_dir)
@@ -9,8 +9,8 @@ generate_subsets <- function(N, npres, nabs, nclusters, occ_data, output_dir){
   
   for(i in 1:N){
     
-    fagussylvatica_pres <- .sample_presence_by_env(occ_data, biovars_30y, k = nclusters, nb_samples = npres)
-    fagussylvatica_abs <- sample_absence(occ_data, EUForest, ERA5land, nb_samples = nabs, env_data = biovars_30y)
+    fagussylvatica_pres <- .sample_presence_by_env(occ_data, env_data, k = nclusters, nb_samples = npres)
+    fagussylvatica_abs <- .sample_absence(occ_data, EUForest, ERA5land, nb_samples = nabs, env_data)
     
     species_occurrence <- .create_occurrence(fagussylvatica_pres, fagussylvatica_abs)
     

### 2023-04-19 data_compilation/sampling/sampling_occurrence_data.R

@@ -67,7 +67,7 @@ ERA5land <- raster(paste0(rawdata_folder,"2m_dewpoint_temperature_1969_01.nc"))
 GBIF <- fread(paste0(speciesdata_folder,"GBIF/Fagus_sylvatica_20012022/0112486-210914110416597.csv"), sep="\t", dec=".", quote = "",
                              header = TRUE, fill = TRUE, stringsAsFactors = FALSE)
 fagussylvatica_occ_rs <- process_occurrence("Fagus sylvatica", EUForest, GBIF, EVM, AFE, ERA5land)
-save(fagussylvatica_occ_rs, file = "D:/species/processed/fagus_sylvatica/fagus_sylvatica_occ_rs.Rdata") # one can save file to prevent future computations
+saveRDS(fagussylvatica_occ_rs, file = "D:/species/processed/fagus_sylvatica/fagus_sylvatica_occ_rs.rds") # one can save file to prevent future computations
 
 
 
@@ -81,5 +81,7 @@ nclusters <- 10 # number of clusters for presence data
 
 output_folder <- "D:/species/processed/fagus_sylvatica"
 
-generate_subsets(N, npres, nabs, nclusters, quercuspubescens_occ_rs, output_folder)
+generate_subsets(N, npres, nabs, nclusters, 
+                 fagussylvatica_occ_rs, env_data = biovars_30y, # k-means is not affected by collinearity, we can use all bioclimatic variables
+                 output_folder)
 
